### 2023-06-14 manuscript/wp4.tex
Commit initial version of manuscript

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small}
\usepackage{amsmath}
\usepackage{fontspec}
\setmainfont{Times New Roman}

\title{How does the writing system get informative? Differentiation or conservation?}

\shorttitle{Informativity in the writing system}

\authorsnames[1,1]{Jon W. Carr, Kathleen Rastle}

\authorsaffiliations{{Department of Psychology, Royal Holloway, University of London, England}}

% \authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023). Informativity in the writing system. \textit{OSF Preprints}. Version~0.}

\abstract{Abstract.}

\keywords{communication; cultural evolution; informativity; iterated learning; language evolution; orthography; spelling; sound change; writing}

\begin{document}

\maketitle

\noindent
Written languages represent spoken languages. However, there are many ways in which written languages diverge. E.g. spacing between words helps readers parse a sentence even if they are not technically required. Written languages can also have there own regularities such as spelling related words similarly (magician), past tense spellings directly conveying meaning, certain spellings conveying word class ous. These features abound in writing, especially in languages such as English that have not been subject to serious regulation.

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings. For example, <meat> and <meet> are both pronounced /miːt/. Furthermore, these spellings can be cognitively useful . For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, allowing the reader to get directly to the meaning. This idiosyncratic spelling of words that are homophonous in speech may confer certain benefits in reading, particularly in terms of comprehension. Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart.

However, this additional informativeness comes at a cost. A system of words that conveys more information will---in general---be more complex than one that conveys less information. There exists a tradeoff between the simplicity of the spelling system and its ability to be expressive. A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology. The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Carr:2020, Kirby:2015} studies. Of particular note here is the idea that the pressure for simplicity in language structure derives principally from learning, while the pressure for informativeness derives from the need for precise communication.

Any explanation for heterographic homophones must, therefore, demonstrate that heterographic spellings convey a benefit in terms of informativeness that is outweighed by the cost of learning the (often arbitrary) spelling distinction. What are the benefits? There seem to be broadly two kinds of benefit. (1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently. (2) permitting the cognitive system to rapidly and accurately attain the meaning of a word. In this paper we limit ourselves to informativeness of the first kind, which is more amenable to experimental testing.

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}). The first model, \textit{the differentiation model}, explains heterography as the result of the creation of a new orthographic form to differentiate one meaning from another. Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing. The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved by the orthography. For example, the words \textit{night} and \textit{knight}, which are heterographic homophones in modern English, were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permitted consonant cluster, as it still is in modern Dutch and German).

%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

\section{Experiment 1}

Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography. In particular we had two main hypotheses:

variation

communication

Preregistered \url{https://aspredicted.org/YLG_CKD}

\subsection{Methods}

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants through repeated learning and production. Participant $i$ in the transmission chain learns the system based on the linguistic output of participant $i-i$ and then produces new linguistic output for participant $i+1$ to learn from. Through several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.

To explore the hypotheses outlines above, we conducted the experiment under our different conditions: high variation vs.\@ no variation and communication vs.\@ no communication.

\subsubsection{Participants}

We recruited 539 participants via the Prolific platform. Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.xxx). Participants were required to have English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied. Participants' most common first languages were: xxx. 44 participants (8\%) were lost to rejections or communication-game matching failures, leaving us with 495 participants in our final dataset.

\subsubsection{Stimuli}

Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}). The alien words had both a spoken and a written form and were composed of a stem and a suffix. The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>. To make the stems easy to learn, they were designed to be graphically and phonetically iconic of the shapes they represent. The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to evolve over time. Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color. The spoken forms were synthesized using the Apple text-to-speech synthesizer (the Tessa voice).

The transmission chains were seeded with an orthographic system that may exhibit high variation or no variation according to condition. In the no-variation conditions, the suffix had only one spelling, although each chain was initialized with a different randomly generated spelling. To generate a spelling we randomly choose a grapheme from each of the sets \{<ee>, <i>, <y>\}, \{<c>, <ck>, <k>\}, and \{<oe>, <oh>, <ow>\} and concatenate them. For example, in one chain the suffix might initially be spelled <icoe>, while in another chain it might initially be spelled <yckow>. In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects. Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors will all have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow). These spellings were not systematic, however; the <yckoh> spelling was not systematically used to represent yellow across different shapes. Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.

\subsubsection{Transmission procedure}

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain. The first participant in a chain was taught the randomly generated orthographic system explained in the previous section, and this system was then free to evolve as it was culturally transmitted. This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). This procedure continued until the chain converged on a particular orthographic system. A chain is said to converge when two consecutive generations produce identical spellings for all 16 items, suggesting that the chain is unlikely to experience significant further change. We adopted this convergence-based approach because we were interested in revealing what kinds of orthographic system are stable under the four different conditions.

\subsubsection{Training procedure}

Participants were trained on the spoken and written forms through a combination of passive exposure trials and mini-test trials. In the passive exposure trials, the alien object appeared first, accompanied by the written and spoken forms after a 500~ms delay. In the mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The participant could freely type in a word and would then be given feedback on any spelling errors (additions shown with green text and deletions shown with red strikethrough text). Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly.

\subsubsection{Test procedure}

\subsubsection{Communication game procedure}

\subsection{Results}

Learnability, measured as transmission error (lower error = higher learnability). Mean Levenshtein edit distance between the words produced at generation i and the corresponding words produced at generation i-1.

Simplicity, measured as complexity (lower complexity = higher simplicity). The length (in bits) of the shortest grammar that losslessly compresses the lexicon, as computed by the Grammarette package: https://github.com/jwcarr/grammarette. Finding the shortest grammar for some lexicon is a non-trivial problem, and it is likely that we will need to adapt the method in response to the collected data.

Communicative cost is given by
\begin{equation}
\mathrm{cost}(L) := \frac{1}{|U|} \sum_{m \in U} -\log \mathrm{Pr}(m|s_m),
\end{equation}
where $\mathrm{Pr}(m|s_m)$ is the probability that a listener would infer meaning $m$ given that a speaker produced signal $s$ for meaning $m$. Here we assume this probability to be $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$.

\subsection{Summary}

%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

\section{Experiment 2}

Our first experiment tests the ability of the conservation model to explain the emergence of informativeness in orthography.

\subsection{Methods}

The methods were mostly identical to Experiment~1 with three major exceptions.

sound change

non convergence

no variation

compositional start

\subsubsection{Participants}

\subsubsection{Stimuli}

The alien objects and word stems were identical to Experiment~1. Unlike Experiment~1, however, the chains were initialized with a fully compositional language that uses four distinct suffixes to systematically express each of the four colors. The suffixes took the form /ɪ\textit{CV}/ where $C$ is a voiceless fricative consonant from the set \{/θ/, /s/, /ʃ/, /h/\} and $V$ is a vowel from the set \{/ə/, /ɛɪ/, /əʊ/, /u/\}. The consonant and vowel sets were randomly permuted for each iterated learning chain, such that the suffix forms were unique to each chain. For example, under the permutations given above, the resulting suffixes would be /ɪθə/, /ɪsɛɪ/, /ɪʃəʊ/, and /ɪhu/, but another iterated learning chain might use the suffixes /ɪsəʊ/, /ɪʃɛɪ/, /ɪθu/, and /ɪhə/. The initial orthographic system used to seed the chain was generated based on the following phoneme–grapheme mapping \{/θ/→<th>, /s/→<s>, /ʃ/→<sh>, /h/→<x>, /ə/→<a>, /ɛɪ/→<e>, /əʊ/→<o>, /u/→<u>\}. The suffixes were designed to be distinctive (and therefore easy to memorize and recall), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /θ/ might be supplanted by /s/ in speech or that /θ/ might be spelled <x> in writing). We attempted to achieve this balance by combining consonants that are very similar (differing only in place of articulation) with vowels that are very dissimilar.

\subsubsection{Sound change}

Each iterated learning chain was run for 12 generations, divided into four epochs of three generations. In the first epoch, all suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color. This was achieved through three sound changes occurring between the four epochs. In the first sound change, two of the suffixes were chosen at random to be merged. To perform the merger, the consonant from one of the suffixes (chosen at random) was paired with the vowel from the other suffix, resulting in a new suffix form that would replace the original two. For example, assuming the language has the suffixes /ɪθə/ (pink), /ɪsɛɪ/ (gray), /ɪʃəʊ/ (blue), and /ɪhu/ (yellow), we might merge /ɪsɛɪ/ and /ɪhu/ into /ɪhɛɪ/, such that the spoken language is no longer able to mark a distinction between gray and yellow. In the second sound change, the same procedure is applied to the two suffixes that were not selected previously (e.g., /ɪθə/ and /ɪʃəʊ/ become /ɪθəʊ/). And in the third and final sound change, the two suffix forms created through the first and second sounds changes---/ɪhɛɪ/ (grey/yellow) and /ɪθəʊ/ (pink/blue)---merge into a single form (e.g., /ɪθɛɪ/). Importantly, the spellings do not change automatically following a sound change event; instead, the orthographic system is free to adapt (or not) in response to sound changes.

%    ORIG.    SC1      SC2      SC3

% P: /ɪθə/           > /ɪθəʊ/ > /ɪθɛɪ/
% G: /ɪsɛɪ/ > /ɪhɛɪ/          > /ɪθɛɪ/
% B: /ɪʃəʊ/          > /ɪθəʊ/ > /ɪθɛɪ/
% Y: /ɪhu/  > /ɪhɛɪ/          > /ɪθɛɪ/

\subsection{Results}

\subsection{Summary}

%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

\section{Discussion}

%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

\section{Acknowledgments}

\noindent This work was funded by a grant from the Leverhulme Trust.

\printbibliography

\end{document}

### 2023-06-14 manuscript/wp4.tex
ms

@@
 \textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}).
-The first model, \textit{the differentiation model}, explains heterography as the result of the creation of a new orthographic form to differentiate one meaning from another.
+The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form to differentiate between homophones.
 Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing.
-The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved by the orthography.
-For example, the words \textit{night} and \textit{knight}, which are heterographic homophones in modern English, were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permitted consonant cluster, as it still is in modern Dutch and German).
+The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography.
+For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).
+
+byte bite
+
+bizarre bazaar
+
+bombe
+
+cache
+
+<cheque>, which the OED describes as a ``a differentiated spelling of \textit{check}.''
+
+Tindr
+
+Innovation
+
+French spellings
+double letters inn
+add -e
+
+Processes
+
+Sound change
+dialect merger (bury is the spelling from one dialect with the pronunciation from another dialect)
+Borrowing (English tends to retain spelling from the source language)
+Printers innovations
+- Etymological spellings (doubt)
+- Deliberate design of heterographs
+- Morphological spellings
 

### 2023-06-15 experiment/server.js
Update experiment code to handle variable spoken forms

@@ -379,6 +379,7 @@ socket.on('connection', function(client) {
 						max_pay: EXP_CONFIG.max_pay,
 						session_time: EXP_CONFIG.session_time,
 						object_array_dims: EXP_CONFIG.object_array_dims,
+						spoken_forms: subject.spoken_forms,
 					});
 				});
 			} else { // If we haven't seen this subject before, create a new subject
@@ -392,6 +393,7 @@ socket.on('connection', function(client) {
 					chain_id: null,
 					generation: null,
 					input_lexicon: null,
+					spoken_forms: null,
 					training_items: null,
 					trial_sequence: generateTrialSequenceStub(),
 					n_reinitializations: 0,
@@ -448,6 +450,7 @@ socket.on('connection', function(client) {
 				// subject's lexicon, training items, and trial sequence, and
 				// write to the database
 				const input_lexicon = chain.lexicon;
+				const spoken_forms = chain.spoken_forms[chain.sound_epoch];
 				const training_items = generateItems(chain.task.n_shapes, chain.task.n_colors, chain.task.bottleneck);
 				const trial_sequence = generateTrialSequence(chain.task, input_lexicon, training_items, lead_communicator);
 				db.subjects.findAndModify({
@@ -458,6 +461,7 @@ socket.on('connection', function(client) {
 							chain_id: chain.chain_id,
 							generation: chain.current_gen + 1,
 							input_lexicon: input_lexicon,
+							spoken_forms: spoken_forms,
 							training_items: training_items,
 							trial_sequence: trial_sequence,
 						},
@@ -471,6 +475,8 @@ socket.on('connection', function(client) {
 						return reportError(client, 128, 'Unrecognized participant ID.');
 					// tell client to begin the next trial
 					const next = prepareNextTrial(subject);
+					next.payload.spoken_forms = spoken_forms;
+					console.log(next);
 					return client.emit(next.event, next.payload);
 				});
 			});

### 2023-06-21 manuscript/wp4.tex
ms

@@
+\PassOptionsToPackage{hidelinks}{hyperref}
 \documentclass[doc,biblatex]{apa7}
@@
 \usepackage{amsmath}
-\usepackage{fontspec}
+\usepackage{mathspec}
+\setallmainfonts{Times New Roman}
 \setmainfont{Times New Roman}
@@
 And if so, what might be the mechanism through which such selection could operate?
-\textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}).
-The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form to differentiate between homophones.
-Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing.
+\textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{berg_aronoff_models}.
+The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form.
+For example, in (British) English, <cheque>, which the OED describes as a ``a differentiated spelling of \textit{check},'' allows the written language to distinguish the bank order from other meanings of the word \textit{check}.
+The word \textit{byte} is a respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
+In English, surnames that are homophonous with ordinary nouns often add <e> (Clarke, Greene, Wilde) or double the final consonant (Carr, Hogg, Mann), and it is common for brand names to adopt creative spellings, such as <flickr> (from <flicker>), <Froot Loops> (from <fruit>), or MVMT (from <movement>).
+
 The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography.
-For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).
-
-byte bite
-
-bizarre bazaar
-
-bombe
-
-cache
-
-<cheque>, which the OED describes as a ``a differentiated spelling of \textit{check}.''
-
-Tindr
+For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).
 
@@
 - Morphological spellings
+
+\subsection{Research questions}
+
+Q1) Does a learning pressure (induced through cultural transmission) result in the emergence of a simple, systematic, transparent orthography that is easier to learn?
+Hypotheses: In the high-variation conditions, where the orthography starts out chaotic (resulting in holistic systems), we expect simpler systems to emerge.
+Simplicity may be realized in a number of ways.
+Transparent systems use a one-to-one mapping to directly represent the spoken forms in writing.
+Redundant systems retain some spelling variation (with idiosyncratic spellings of the suffix for each shape) but become simpler by collapsing across colour.
+Expressive systems continue to express colour, but do so with a non-transparent but easy to learn compositional suffix system.
+The prevalence of these three types of system will be influenced by the presence of the communicative pressure.
+When communication is not present, transparent systems will tend to dominate, alongside some redundant systems and possibly some expressive systems.
+When communication is present, expressive systems will tend to dominate, perhaps with some redundant and holistic forms.
+The gradual emergence of simplicity, whatever form it takes, should be accompanied by increased learnability.
+
+Q2) Can a writing system evolve to become more informative than its uninformative spoken counterpart?
+Hypotheses: Yes, but preferentially when two vital ingredients are present: a communicative pressure for informativeness (which makes an expressive system desirable) and high variation (which provides the building blocks to make that happen).
+When both ingredients are present, expressive systems will tend to dominate.
+If only one of the ingredients is available (either communication or high variation), informative systems may emerge but are less likely.
+In the case of learning-only and high-variation, expressive systems may emerge if the learners – yearning for an explanation for the variation – assume that colour must somehow be involved and therefore impose (perhaps unconsciously) colour-expressive structure on the language (i.e.
+learners may, to a certain extent, have a preconceived notion that languages ought to be informative).
+In the case of communication and no variation (where, again, only one of the ingredients is met), expressive systems may emerge but this is highly dependent on participants being 'innovators' who are willing to develop novel spelling systems that diverge from the spoken system in order to express colour.
+If neither ingredient is present, expressive systems will not emerge.
 
@@
 
-We recruited 539 participants via the Prolific platform.
-Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.xxx).
-Participants were required to have English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied.
-Participants' most common first languages were: xxx.
-44 participants (8\%) were lost to rejections or communication-game matching failures, leaving us with 495 participants in our final dataset.
+We recruited 538 participants via the Prolific platform.
+Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.64).
+The median completion time was 19:36, resulting in an hourly rate of £9.18 (£11.03 including bonus).
+We limited recruitment to participants who had English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied.
+The most common first languages were: English (59\%), Polish (9.7\%), Portuguese (6.1\%), Italian (5.4\%), Spanish (5\%), and Greek (3.7\%).
+Five participants were rejected because they did not pass the random audio checks, and a further three participants were removed because they had been partnered with one of the five rejected participants.
+In addition, 35 participants were lost to communication-game pairing failures, resulting in a final dataset of 495 participants.
 
@@
 Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}).
-The alien words had both a spoken and a written form and were composed of a stem and a suffix.
+The alien words had a spoken and written form composed of a stem and suffix.
 The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>.
-To make the stems easy to learn, they were designed to be graphically and phonetically iconic of the shapes they represent.
-The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to evolve over time.
+Since we are not interested in participants' learning of the stems, they were designed to be quick and easy to learn by being graphically and phonetically iconic of the shapes they represent, and they were fixed and unchanging throughout the experiment.
+The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to change over time.
 Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.
@@
 In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects.
-Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors will all have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow).
-These spellings were not systematic, however; the <yckoh> spelling was not systematically used to represent yellow across different shapes.
+Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow).
+These spellings were not systematic, however; for example, the <yckoh> spelling was not systematically used to represent yellow across different shapes.
 Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.
@@
 As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain.
-The first participant in a chain was taught the randomly generated orthographic system explained in the previous section, and this system was then free to evolve as it was culturally transmitted.
+The first participant in a chain was taught the initial, randomly generated orthographic system, and this system was then free to evolve as it was culturally transmitted.
 This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color).
-This procedure continued until the chain converged on a particular orthographic system.
-A chain is said to converge when two consecutive generations produce identical spellings for all 16 items, suggesting that the chain is unlikely to experience significant further change.
-We adopted this convergence-based approach because we were interested in revealing what kinds of orthographic system are stable under the four different conditions.
+Transmission continued until the chain converged on a particular orthographic system.
+A chain is said to converge when two consecutive generations produce identical spellings for all 16 items (including the four unseen items), suggesting that the chain is unlikely to experience significant further change.
+We adopted this convergence-based approach because we were primarily interested in revealing what kinds of orthographic system are stable under the four conditions, but it was unclear a~priori how many generations would be required to reveal this information.
 
@@
 
-Participants were trained on the spoken and written forms through a combination of passive exposure trials and mini-test trials.
-In the passive exposure trials, the alien object appeared first, accompanied by the written and spoken forms after a 500~ms delay.
-In the mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects.
-The participant could freely type in a word and would then be given feedback on any spelling errors (additions shown with green text and deletions shown with red strikethrough text).
+Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials.
+In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms.
+The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word.
+In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects.
+The mini-tests were included to help participants actively engage in the learning process.
+After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text).
+The input box only accepted alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof.
 Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly.
+However, to receive the bonus, the word had to be submitted within 20~seconds.
+Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p).
+To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.
 
-\subsubsection{Test procedure}
+\subsubsection{Non-communicative test procedure}
 
-\subsubsection{Communication game procedure}
+After training, participants assigned to the non-communicative conditions completed a test phase, alternating between two types of test trial, production trials and comprehension trials.
+In production trials, the participant was shown an object and heard its associated word pronounced aloud.
+The task was to type~in the appropriate spelling under the same input restrictions as in mini-test trials and an additional restriction that the stem (i.e., the first three letters of the word) had to be spelled correctly to ensure that stems could not diverge from the spoken forms over time.
+Since participants heard the word pronounced aloud, typing the stem correctly should be trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly told the participant the correct spelling of the stem.
+In comprehension trials, the participant was shown a word and had to click on the matching object from all 16 objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice).
+In both types of test trial, the participant was awarded a bonus of 2p for every correct answer (unlike mini-tests, there was no 1p bonus for partially correct spellings).
+Each of the 16 object--word pairs (i.e., including the four unseen items) was tested once in production and once in comprehension, resulting in 32 trials and a maximum bonus of 64p.
+Binary feedback was provided on each trial in the form of a sound effect.
+
+\subsubsection{Communicative test procedure}
+
+After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant.
+The two participants both received training on the same orthographic system inherited from a previous generation.
+The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction.
+On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant.
+The second participant---the receiver---would then complete a comprehension trial (select an object for the transmitted word).
+The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme).
+However, the communication game differed from the non-communicative test in three ways.
+First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test.
+Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender.
+Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target.
+Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.
 
@@
 
+For the purpose of the results, we are primarily interested in the spelling of the suffixes, so we first needed to isolate the suffixes from the stems.
+Although the first three letters (<buv>, <zet>, <gaf>, <wop>) were fixed and unchanging throughout the experiment, it was possible for the stems to grow longer if additional material was inserted before the suffix (e.g., <gaff>, <wopb>, <zetch>).
+We consider the stem to be all material up to but not including the first vowel letter after the original three-letter stems, such that, for example, <zethyckoh> will be parsed into the stem <zeth> and suffix <yckoh>.
+Of course, it is not possible to know how participants themselves divided the strings (if at all), but since all words were phonetically identical from the /i/ vowel onward, it is reasonable to parse the strings this way on the basis that this yields the shortest grammar.
+
+To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 4×4 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings.
+For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
+\begin{equation}
+	S_\mathrm{degenerate} = \begin{bmatrix}
+	1 & 1 & 1 & 1 \\
+	1 & 1 & 1 & 1 \\
+	1 & 1 & 1 & 1 \\
+	1 & 1 & 1 & 1
+	\end{bmatrix}
+\end{equation}
+Here the number 1 represents the one and only suffix spelling.
+We define an ``expressive'' system as one that marks the color dimension using four distinct spellings, resulting in the following matrix:
+\begin{equation}
+	S_\mathrm{expressive} = \begin{bmatrix}
+	1 & 2 & 3 & 4 \\
+	1 & 2 & 3 & 4 \\
+	1 & 2 & 3 & 4 \\
+	1 & 2 & 3 & 4
+	\end{bmatrix}
+\end{equation}
+In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
+\begin{equation}
+	S_\mathrm{redundant} = \begin{bmatrix}
+	1 & 1 & 1 & 1 \\
+	2 & 2 & 2 & 2 \\
+	3 & 3 & 3 & 3 \\
+	4 & 4 & 4 & 4
+	\end{bmatrix}
+\end{equation}
+Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant.
+Such a system contains unnecessary complexity---different ways of writing /ikəʊ/ that carry no additional information.
+
+Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly.
+To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric.
+VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another set partition.
+So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information.
+The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.
+
+We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}).
+The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to.
+Most of the emergent system are either redundant or degenerate, with some straddling the border between the two.
+No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking.
+In one system, <icoe> is used for gray objects and <ikoe> for all other colors.
+In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>.
+Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the learning-only condition.
+
 \subsection{Summary}
+
+Our results suggest that the differentiation model is not a good explanation for the emergence of informative heterography.
+
+Three pathways
+
+- conscious design by communicators
+- emergence through learning (chance alignment with color)
+-
 
@@
 
-%    ORIG.    SC1      SC2      SC3
-
-% P: /ɪθə/           > /ɪθəʊ/ > /ɪθɛɪ/
-% G: /ɪsɛɪ/ > /ɪhɛɪ/          > /ɪθɛɪ/
-% B: /ɪʃəʊ/          > /ɪθəʊ/ > /ɪθɛɪ/
-% Y: /ɪhu/  > /ɪhɛɪ/          > /ɪθɛɪ/
-
 \subsection{Results}

### 2023-06-22 manuscript/wp4.tex
ms

@@
 
-The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography.
-For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).
+The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography.
+In English, for example, the meat--meet merger, which was complete by around 1700, resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated in response, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc.
+Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of the consonant cluster /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of the palatal fricative /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of the voiceless labial-velar fricative /ʍ/ into the voiced approximant /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).
 
@@
 French spellings
-double letters inn
+double letters in-inn, be-bee, for-fore-four, to-too-two, but-butt, by-bye-buy, more-moor, none-nun, we-wee, or-oar-ore, so-sew
 add -e
@@
 
+Why do we get redundant systems?
+
+- lots of variation available, why condition it on shape and not color?
+- strong cue from the spoken language hints that the language does not mark color, so the natural thing to think is that the spelling variation is based on shape
+
+Why do we not get expressive systems?
+
+- alignment problem - hard to align with someone when you can only send one word, not meta conversation
+- strong signal from the spoken language that it doesn't mark color, peolpe believe spelling shoud represent the truth of speech
+- social pressure to stick to the taught system - affraid to get in trouble or get partner in trouble
+- the goal is perceived as reproduction rather than invention
+
 \subsection{Summary}

### 2023-06-22 experiment/mission_control.py
Fix couple of bugs in mission control

@@ -50,6 +50,7 @@ STATUS_EMOJI = {
 	'available': '🟢',
 	'unavailable': '🔴',
 	'converged': '⚫️',
+	'completed': '⚫️',
 	'closed': '🟡',
 	'approval_needed': '🔵',
 }
@@ -94,6 +95,7 @@ class MissionControl:
 					'lexicon': lexicon,
 					'spoken_forms': spoken_forms,
 				})
+		self.db = DB[self.exp_id]
 		print('Launched task:', self.exp_id)
 
 	def erase(self, _=None):

### 2023-06-22 experiment/client/client.js
Restore English color word regexes

@@ -129,16 +129,17 @@ function showInputError(input_id) {
 	}, 500);
 }
 
+const STEM_LENGTH = 4; // set to 3 in Experiment 1
 function validateWord(label, expected_label=null) {
 	if (!label.match(/^[a-z]{4,9}$/))
 		return false;
-	// if (label.match(/(p|r|b|y|g)$/))
-	// 	return false;
-	// if (label.match(/(bc|bk|bl|gr|gy|pin|pk|pn|ppl|prp|pur|rd|re|ye|yl|yw)/))
-	// 	return false;
+	if (label.match(/(p|r|b|y|g)$/))
+		return false;
+	if (label.match(/(bc|bk|bl|gr|gy|pin|pk|pn|ppl|prp|pur|rd|re|ye|yl|yw)/))
+		return false;
 	if (expected_label) {
-		let expected_prefix = expected_label.slice(0, 3);
-		if (label.slice(0, 3) != expected_prefix) {
+		let expected_prefix = expected_label.slice(0, STEM_LENGTH);
+		if (label.slice(0, STEM_LENGTH) != expected_prefix) {
 			alert(`Please check your spelling. This word should begin with ${expected_prefix}...`);
 			return false;
 		}

### 2023-07-10 experiment/server.js
Add /prolific endpoint and automate the dropping of returns

@@ -29,11 +29,15 @@ const mongojs = require('mongojs');
 const socketio = require('socket.io');
 
 // ------------------------------------------------------------------
-// Load experiment config data
+// Load experiment config data and Prolific API token if available
 // ------------------------------------------------------------------
 
 const EXP_ID = process.argv[2];
 const EXP_CONFIG = JSON.parse(fs.readFileSync(`config/${EXP_ID}.json`));
+
+let PROLIFIC_API_TOKEN = null;
+if (fs.existsSync('prolific_api_token'))
+	PROLIFIC_API_TOKEN = fs.readFileSync('prolific_api_token').toString();
 
 // ------------------------------------------------------------------
 // Server setup
@@ -41,6 +45,13 @@ const EXP_CONFIG = JSON.parse(fs.readFileSync(`config/${EXP_ID}.json`));
 
 const app = express();
 app.use(express.static(`${__dirname}/client`));
+app.use(express.json());
+if (PROLIFIC_API_TOKEN)
+	app.post('/prolific', function(req, res) {
+		res.sendStatus(200);
+		if (req.body.event_type === 'submission.status.change')
+			return handleSubmissionStatusChange(req.body.resource_id);
+	});
 
 const config = {};
 if (PROTOCOL === 'https') {
@@ -377,6 +388,38 @@ function reportError(client, error_number, reason) {
 	const message = 'Error ' + error_number + ': ' + reason;
 	console.log(getCurrentTime() + ' ' + message);
 	return client.emit('report', {message});
+}
+
+function handleSubmissionStatusChange(submission_id) {
+	http.get(
+		`https://api.prolific.co/api/v1/submissions/${submission_id}/`,
+		{headers: {'Authorization': `Token ${PROLIFIC_API_TOKEN}`}},
+		res => {
+			let raw_data = '';
+			res.on('data', chunk => {raw_data += chunk;});
+			res.on('end', () => {
+				const data = JSON.parse(raw_data);
+				if (data['status'] === 'RETURNED') {
+					db.subjects.findAndModify({
+						query: {subject_id: data['participant']},
+						update: {$set: {status: 'dropout'}},
+						new: true,
+					}, function(err, subject, last_err) {
+						if (err || !subject || subject.chain_id === null)
+							return;
+						db.chains.findOne({chain_id: subject.chain_id}, function(err, chain) {
+							if (err || !chain)
+								return;
+							if (chain.subject_a === subject.subject_id)
[…]

### 2023-07-10 experiment/mission_control.py
Switch from the submission polling method to the Prolific hooks method

@@ -27,6 +27,16 @@ The general workflow is:
 - approve: approve the outcome and reopen the chain for a new generation
 - dump: pull down the data from all completed participants on a task
 
+Prior to running the experiment, create a subscription to the Prolific
+submission.status.change event by running:
+
+	create_subscription()
+
+Make a note of the subscription_id and at the end of the experiment,
+delete the subscription.
+
+	delete_subscription('<subscription_id>')
+
 '''
 
 import json
@@ -47,6 +57,18 @@ PORT = 27017
 
 DB = MongoClient(DOMAIN, PORT)
 
+try:
+	with open('prolific_api_token') as file:
+		PROLIFIC_API_TOKEN = file.read()
+except FileNotFoundError:
+	PROLIFIC_API_TOKEN = None
+
+try:
+	with open('prolific_workspace_id') as file:
+		PROLIFIC_WORKSPACE_ID = file.read()
+except FileNotFoundError:
+	PROLIFIC_WORKSPACE_ID = None
+
 STATUS_EMOJI = {
 	'available': '🟢',
 	'unavailable': '🔴',
@@ -60,10 +82,8 @@ EMPTY_SLOT = ' - ' * 8
 
 class MissionControl:
 
-	def __init__(self, exp_id, prolific_study_id=None, prolific_api_token=None):
+	def __init__(self, exp_id):
 		self.exp_id = exp_id
-		self.prolific_study_id = prolific_study_id
-		self.prolific_api_token = prolific_api_token
 		if self.exp_id in DB.list_database_names():
 			self.db = DB[self.exp_id]
 		else:
@@ -383,18 +403,6 @@ class MissionControl:
 				self.db.chains.update_one({'chain_id': subject['chain_id']}, {'$set':{'status': 'available', 'subject_b': None}})
 				print(f'Subject B slot opened on chain {chain["chain_id"]}')
 
-	def drop_returns(self):
-		'''
-		Use the Prolific API to get a list of returned participants. 
-		'''
-		api_url = f'https://api.prolific.co/api/v1/submissions/?study={self.prolific_study_id}'
-		response = requests.get(api_url, headers={'Authorization': f'Token {self.prolific_api_token}'}).json()
-		for submission in response['results']:
-			subject_id = submission['participant_id']
-			if submission['status'] == 'RETURNED' and subject_id not in self.returned_subject_IDs:
[…]

### 2023-07-10 code/training_model.py
Switch to multilevel model of learning curve

@@ -39,20 +39,38 @@ def get_correct(subject_id):
 	trial = list(range(len(correct)))
 	return trial, correct
 
-def fit_model(trial, correct):
+def fit_multilevel_model(trial, correct, subject_ids):
 	coords = {
-		'trial': trial,
+		'trial': list(range(1, 37)),
+		'subject': subject_ids,
 	}
 	with pm.Model(coords=coords) as model:
-		α = pm.Normal('α', 0, 10)
-		β = pm.Normal('β', 0, 10)
-		p = pm.Deterministic('p', pm.math.invlogit(α + β * trial), dims='trial')
-		y = pm.Bernoulli('y', p, observed=correct, dims='trial')
+
+		μ_α = pm.Normal('μ_α', 0, 10)
+		σ_α = pm.Exponential('σ_α', 1)
+
+		μ_β = pm.Normal('μ_β', 0, 10)
+		σ_β = pm.Exponential('σ_β', 1)
+
+		α = pm.Normal('α', μ_α, σ_α, dims='subject').T
+		β = pm.Normal('β', μ_β, σ_β, dims='subject').T
+		
+		θ = pm.Deterministic('θ', pm.math.invlogit(α + β * trial), dims=('trial', 'subject'))
+
+		c = pm.Bernoulli('c', θ, observed=correct, dims=('trial', 'subject'))
+
+		pm.Deterministic('θ_mean', pm.math.invlogit(μ_α + μ_β * trial[:, 0]))
+		
 		trace = pm.sample(1000, tune=2000)
+
 	return trace
 
-def make_plot(axis, trace):
-	post_mean = trace.posterior.p.mean(('chain', 'draw'))
+def make_plot(axis, trace, sub_i=None):
+	if sub_i is None:
+		post = trace.posterior.θ_
+	else:
+		post = trace.posterior.θ[:, :, :, sub_i]
+	post_mean = post.mean(('chain', 'draw'))
 	trial = range(1, len(post_mean) + 1)
 
 	if post_mean[-1] > 0.75:
@@ -62,7 +80,7 @@ def make_plot(axis, trace):
 
 	az.plot_hdi(
 	    trial,
-	    trace.posterior.p,
+	    post,
 	    hdi_prob=0.95,
 	    fill_kwargs={'alpha': 0.25, 'linewidth': 0},
 	    ax=axis,
@@ -75,30 +93,29 @@ def make_plot(axis, trace):
[…]

### 2023-07-10 manuscript/wp4.tex
ms 10 jul

--- 2023-06-21 manuscript/wp4.tex
@@
 \addbibresource{references.bib}
-\captionsetup[figure]{textfont=rm,font=small}
+\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
 \usepackage{amsmath}
@@
 \setmainfont{Times New Roman}
+% \setsansfont{Helvetica Neue}
 
@@
 For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
-In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, allowing the reader to get directly to the meaning.
+In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, giving the reader a headstart on parsing the upcoming syntax.
 This idiosyncratic spelling of words that are homophonous in speech may confer certain benefits in reading, particularly in terms of comprehension.
@@
 And if so, what might be the mechanism through which such selection could operate?
-\textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{berg_aronoff_models}.
+\textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
 The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form.
-For example, in (British) English, <cheque>, which the OED describes as a ``a differentiated spelling of \textit{check},'' allows the written language to distinguish the bank order from other meanings of the word \textit{check}.
-The word \textit{byte} is a respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
-In English, surnames that are homophonous with ordinary nouns often add <e> (Clarke, Greene, Wilde) or double the final consonant (Carr, Hogg, Mann), and it is common for brand names to adopt creative spellings, such as <flickr> (from <flicker>), <Froot Loops> (from <fruit>), or MVMT (from <movement>).
+For example, many monosyllabic words that are homophonous with function words have tended to adopt alternate spellings (usually by adding material): \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc.
+Other cases of heterography appear to be the result of a very conscious choice to disambiguate meanings; for example, in British English, the word \textit{cheque} was introduced to to distinguish the bank order from other meanings of the word \textit{check}, the word \textit{byte} is a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}, and the word \textit{lite} is a respelling of \textit{light} that is frequently used in food products to distinguish light-in-weight from light-in-color.
+Etymological (and folk-etymological) spellings introduced during the Renaissance also sometimes resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}.
+Heterography is also very common among proper nouns: Surnames that are homophonous with ordinary nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT} (movement).
 
-The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography.
-For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
+	\vspace*{2pt}
+	\caption{Two models of heterography.
+In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$, but by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
+In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
+Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
+	\label{models_of_heterography}
+	\end{figure}
 
-Innovation
-
-French spellings
-double letters inn
-add -e
+The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography.
+In English, for example, the meat--meet merger ultimately resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated accordingly, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc.
+Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of /ʍ/ into /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).
 
@@
 After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text).
-The input box only accepted alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof.
+The input box only accepted lowercase alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof.
 Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly.
-However, to receive the bonus, the word had to be submitted within 20~seconds.
+However, to receive the bonus, their response had to be submitted within 20~seconds.
 Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p).
@@
 After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant.
-The two participants both received training on the same orthographic system inherited from a previous generation.
+The two participants both received training on the same orthographic system inherited from a previous generation, albeit subject to independent bottlenecks on transmission.
 The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction.
 On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant.
-The second participant---the receiver---would then complete a comprehension trial (select an object for the transmitted word).
+The second participant---the receiver---would then complete a comprehension trial (select an object for the received word).
 The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme).
@@
 
-Our results suggest that the differentiation model is not a good explanation for the emergence of informative heterography.
+Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation.
+Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography.
+But this did not happen, even under the communicative pressure.
+Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings.
+This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.
+So, what was different about the present experiment?
+The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color.
+Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
+As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose.
+A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion} \textit{transmission}, and \textit{invention}).
 
-Three pathways
+Another pathway through which differentiation could have occurred is through conscious innovation, especially in the communicative conditions, where participants could have increased their bonuses substantially by innovating a system of color markers.
+However, there was no evidence of such innovation, and only four participants mentioned that they even \textit{considered} this possibility in their post-test comments (e.g., ``I wondered if I should use the color suffix rule even where I did not think it was the right word to try and help my partner get the right color object'').
+Why do participants choose not to innovate or even regularize an imperfect system?
+Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that we expected them to use the orthography learned during training.
+Participants who adopted this interpretation may have feared diverging from their training either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
+Secondly, even if participants are willing to diverge, they are presented with an alignment problem: How can they decide what system of color marking to use when there is no channel through which they can have a meta-discussion?
+They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.
+Interestingly, these are dynamics that also exist in the real world, where deviation from received, sacrosanct spellings is considered socially unacceptable or uneducated (e.g., the non-standard spelling <thru> has struggled for decades---or perhaps centuries---to make any headway against <through>), and even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If I decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take my spelling seriously or even understand the distinction I was trying to make?
 
-- conscious design by communicators
-- emergence through learning (chance alignment with color)
--
+Overall, our first experiment suggests that differentiation---be it through a self-organizing process in learning or through a conscious process of innovation in use---is not a good model of heterography, unless enforced by top-down authority.
+We now turn to consider the conservation model.
 
@@
 
-Our first experiment tests the ability of the conservation model to explain the emergence of informativeness in orthography.
+Our second experiment investigates the conservation model: Given than an informative system already exists (both in speech and in writing), will the informative system persist even after the degeneration of the spoken language towards homophony?
+And does this happen preferentially in the presence of a shared communicative task.
+How much homophony can an informative, heterographic system resist under an increasingly strong simplicity pressure.
 
@@
 
-The methods were mostly identical to Experiment~1 with three major exceptions.
-
-sound change
-
-non convergence
-
-no variation
-
-compositional start
+The methods were identical to Experiment~1 with the following major exceptions.
+Firstly, and most notably, the artificial language used to seed each chain starts out fully compositional (in both its spoken and written forms), since we are now interested in whether the orthographies will maintain informativeness rather than gain it.
+Secondly, and consequently, the convergence-based approach used in Experiment~1 to determine the number of generations is no longer appropriate, so we now use a fixed number of generations.
+Thirdly, we simulate regular sound mergers which result in the spoken forms of the suffixes becoming increasingly less informative over generational time (this contrasts with Experiment~1, where the spoken form of the suffixes were always fully uninformative).
 
 \subsubsection{Participants}
+
+We recruited xxx participants via the Prolific platform.
+The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.xxx).
+The median completion time was xxx:xxx, resulting in an hourly rate of £xxx (£xxx including bonus).
+Unlike Experiment~1 we restricted recruitment to native speakers of English because it was particularly important that participants would be able to distinguish the voiceless fricatives /θ s ʃ h/ and we found in piloting that non-native speakers of English frequently struggled with these distinctions.
+xxx participants were rejected because they did not pass the random audio checks, and a further xxx participants were removed because they had been partnered with one of the rejected participants.
+In addition, xxx participants were lost to communication-game pairing failures, resulting in a final dataset of 360 participants (organized into ten chains of twelve generations in each condition, with two participants per generation in Learning \& Communication).
 
@@
 Each iterated learning chain was run for 12 generations, divided into four epochs of three generations.
-In the first epoch, all suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color.
+In the first epoch, all spoken suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color.
 This was achieved through three sound changes occurring between the four epochs.

### 2023-07-20 experiment/server.js
Small adjustments

@@ -77,7 +77,7 @@ function submissionStatusChange(submission_id) {
 			});
 			res.on('end', () => {
 				const data = JSON.parse(raw_data);
-				if (data['status'] === 'RETURNED') {
+				if (data.status === 'RETURNED') {
 					console.log(`Subject ${data.participant} has returned; dropping.`);
 					db.subjects.findAndModify({
 						query: {subject_id: data.participant},
@@ -85,10 +85,10 @@ function submissionStatusChange(submission_id) {
 						new: false,
 					}, (err, subject, last_err) => {
 						if (err || !subject || subject.chain_id === null)
-							return console.log('- Subject not found');
+							return;
 						db.chains.findOne({chain_id: subject.chain_id}, (err, chain) => {
 							if (err || !chain)
-								return console.log('- Chain not found');
+								return;
 							if (chain.subject_a === subject.subject_id)
 								return db.chains.update({chain_id: chain.chain_id}, {'$set':{'status': 'available', 'subject_a': null}});
 							else if (chain.subject_b === subject.subject_id)
@@ -216,7 +216,7 @@ function itemsWithSameWord(lexicon, target_word) {
 }
 
 function generateTrialSequenceStub(progress=null) {
-	const initial_trial_estimate = 0.05;
+	const initial_trial_estimate = 0.02;
 	return [
 		{event: 'consent', payload: {
 			progress: 0,
@@ -292,7 +292,8 @@ function generateTrialSequence(task, words, training_items, lead_communicator, p
 			}
 		}
 		// second to last trial on each training rep is a catch trial
-		trial_sequence[trial_sequence.length - 2].payload.test_trial.catch_trial = true;
+		if (i % 2 === 0)
+			trial_sequence[trial_sequence.length - 2].payload.test_trial.catch_trial = true;
 	}
 	if (task.communication)
 		trial_sequence.push({event:'instructions', payload: {

### 2023-07-20 code/training_model.py
Generalize for other experiments

@@ -30,14 +30,22 @@ def find_subjects_with_regular_input():
 		if is_regular(data['input_lexicon'], data['training_items']):
 			print(data['subject_id'], data['chain_id'], data['generation'])
 
-def get_correct(subject_id):
-	data = json_load(f'../data/exp2/subject_{subject_id}.json')
+def get_correct(exp_id, subject_id):
+	data = json_load(f'../data/{exp_id}/subject_{subject_id}.json')
 	correct = [
 		int(trial['expected_label'] == trial['input_label'])
 		for trial in data['responses'] if trial['test_type'] == 'mini_test'
 	]
+	prod_correct = [
+		int(trial['expected_label'] == trial['input_label'])
+		for trial in data['responses'] if trial['test_type'] == 'test_production'
+	]
+	comp_correct = [
+		int(trial['selected_item'] in trial['items'])
+		for trial in data['responses'] if trial['test_type'] == 'test_comprehension'
+	]
 	trial = list(range(len(correct)))
-	return trial, correct
+	return trial, correct, (sum(prod_correct), sum(comp_correct))
 
 def fit_multilevel_model(trial, correct, subject_ids):
 	coords = {
@@ -93,29 +101,32 @@ def make_plot(axis, trace, sub_i=None):
 	axis.set_ylim(0, 1)
 	axis.set_xlim(1, 36)
 
-def run_multilevel(subject_ids):
+def run_multilevel(exp_id, subject_ids):
 	trial_matrix = []
 	correct_matrix = []
 	for subject_id in subject_ids:
-		trial, correct = get_correct(subject_id)
+		trial, correct, _ = get_correct(exp_id, subject_id)
 		trial_matrix.append(trial)
 		correct_matrix.append(correct)
 	trace = fit_multilevel_model(np.column_stack(trial_matrix), np.column_stack(correct_matrix), subject_ids)
-	trace.to_netcdf('../plots/exp2/training_trace2.netcdf')
+	trace.to_netcdf('../plots/pilot8/training_trace.netcdf')
 
-def plot_multilevel(subject_ids):
-	trace = az.from_netcdf('../plots/exp2/training_trace2.netcdf')
-	fig, axes = az.utils.plt.subplots(5, 10, figsize=(16, 9))
-	for sub_i, axis in zip(range(50), np.ravel(axes)):
+def plot_multilevel(exp_id, subject_ids):
+	trace = az.from_netcdf('../plots/pilot8/training_trace.netcdf')
[…]

### 2023-07-20 manuscript/wp4.tex
ms 20/07

@@
 \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
-The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form.
-For example, many monosyllabic words that are homophonous with function words have tended to adopt alternate spellings (usually by adding material): \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc.
-Other cases of heterography appear to be the result of a very conscious choice to disambiguate meanings; for example, in British English, the word \textit{cheque} was introduced to to distinguish the bank order from other meanings of the word \textit{check}, the word \textit{byte} is a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}, and the word \textit{lite} is a respelling of \textit{light} that is frequently used in food products to distinguish light-in-weight from light-in-color.
-Etymological (and folk-etymological) spellings introduced during the Renaissance also sometimes resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}.
-Heterography is also very common among proper nouns: Surnames that are homophonous with ordinary nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT} (movement).
+The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms.
+For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
+Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc.
+Heterography is also very common in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT}.
 

### 2023-07-20 experiment/server.js
Assign chains randomly rather than in numerical order

@@ -381,23 +381,25 @@ function assignToChain(chains, subject_id) {
 	// most urgent
 	for (let chain of chains) {
 		if (chain.task.communication && chain.subject_a === null && chain.subject_b)
-			return [chain, {$set: {status: 'unavailable', subject_a: subject_id}}];
+			return [ chain, {$set: {status: 'unavailable', subject_a: subject_id}} ];
 	}
 	// then look for a communicative chain where subject A has been
 	// assigned but we still need a subject B
 	for (let chain of chains) {
 		if (chain.task.communication && chain.subject_a && chain.subject_b === null)
-			return [chain, {$set: {status: 'unavailable', subject_b: subject_id}}];
+			return [ chain, {$set: {status: 'unavailable', subject_b: subject_id}} ];
 	}
 	// then look for a communicative chain where neither subject has been
-	// assigned
-	for (let chain of chains) {
-		if (chain.task.communication && chain.subject_a === null && chain.subject_b === null)
-			return [chain, {$set: {status: 'available', subject_a: subject_id}}];
-	}
-	// if no communication chains are available, return first chain in the
-	// list, which will be the one with the smallest generation count
-	return [chains[0], {$set: {status: 'unavailable', subject_a: subject_id}}];
+	// assigned, and, if one is available, pick one at random
+	const candidate_comm_chains = chains.filter((chain) => {
+		if (chain.task.communication)
+			return chain;
+	});
+	if (candidate_comm_chains.length > 0)
+		return [ candidate_comm_chains[randInt(candidate_comm_chains.length)], {$set: {status: 'available', subject_a: subject_id}} ];
+	// if no communication chains are available, return one of the chains at
+	// random
+	return [ chains[randInt(chains.length)], {$set: {status: 'unavailable', subject_a: subject_id}} ];
 }
 
 function getPartner(client, subject, callback, n_retries=9) {

### 2023-08-04 code/regmodel2.py
Update model

@@ -37,8 +37,8 @@ if __name__ == '__main__':
 	data1 = generate_random_dataset(n_chains, n_generations, 0.3, 0.3)
 	data2 = generate_random_dataset(n_chains, n_generations, 0.25, 0.3)
 
-	plot_dataset(data1, data2)
-	quit()
+	# plot_dataset(data1, data2)
+	# quit()
 
 	chain_ids = []
 	np.repeat
@@ -55,5 +55,5 @@ if __name__ == '__main__':
 	df = center_variable(df, 'generation')
 	df['chain_id'] = df[['condition', 'chain']].apply(lambda row: '_'.join(row.values.astype(str)), axis=1)
 
-	model = bmb.Model('cost ~ generation * condition + (generation | chain_id) + (1 | chain_id)', df)
+	model = bmb.Model('cost ~ generation * condition + (1 + generation | chain_id)', df)
 	results = model.fit(draws=1000, chains=4)

### 2023-08-04 manuscript/wp4.tex
Add example from Irish

@@
 Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart.
+
+Other languages
+
+Irish orthography has special features to handle lenition and eclipsis.
+For example, when the word /pɑʃti/ (children) is eclipsed, it becomes /bɑʃti/ (the initial consonant is vocalized), but the spelling system retains the letter <p> in addition to adding a <b>: <páistí> becomes <bpáistí> under eclipsis.
+This allows the reader quick access to the root word from which the eclipsed form is derived and, in some cases, results in ambiguity avoidance (in this case, if the <p> were not retained, the spelling <báistí> would be confusible with the word for rain).
 

### 2023-08-04 experiment/mission_control.py
Open and close can accept regex patterns to open/close many

@@ -39,6 +39,7 @@ delete the subscription.
 
 '''
 
+import re
 import json
 import random
 import time
@@ -178,22 +179,29 @@ class MissionControl:
 			with open(exp_dir / f'chain_{chain["chain_id"]}.json', 'w') as file:
 				file.write(dumps(chain, indent='\t'))
 
-	def open(self, chain_id=None):
+	def open(self, chain_id):
 		if chain_id is None:
 			raise ValueError('Chain ID must be specified')
-		chain = self.db.chains.find_one({'chain_id': chain_id})
-		if chain['status'] != 'closed':
-			raise ValueError('Cannot open: Chain not currently closed')
-		self.db.chains.update_one({'chain_id': chain['chain_id']}, {'$set': {'status': 'available'}})
+		pattern = re.compile(chain_id)
+		for chain in self.db.chains.find():
+			if pattern.fullmatch(chain['chain_id']):
+				if chain['status'] == 'closed':
+					self.db.chains.update_one({'chain_id': chain['chain_id']}, {'$set': {'status': 'available'}})
+					print(f'Opened {chain["chain_id"]}')
+				else:
+					print(f'Cannot open {chain["chain_id"]}: Chain not currently closed')
 
 	def close(self, chain_id=None):
 		if chain_id is None:
 			raise ValueError('Chain ID must be specified')
-		chain = self.db.chains.find_one({'chain_id': chain_id})
-		if chain['subject_a'] is None and chain['subject_b'] is None and chain['status'] == 'available':
-			self.db.chains.update_one({'chain_id': chain['chain_id']}, {'$set': {'status': 'closed'}})
-		else:
-			raise ValueError('Cannot close: Chain occupied')
+		pattern = re.compile(chain_id)
+		for chain in self.db.chains.find():
+			if pattern.fullmatch(chain['chain_id']):
+				if chain['subject_a'] is None and chain['subject_b'] is None and chain['status'] == 'available':
+					self.db.chains.update_one({'chain_id': chain['chain_id']}, {'$set': {'status': 'closed'}})
+					print(f'Closed {chain["chain_id"]}')
+				else:
+					print(f'Cannot close {chain["chain_id"]}: Chain occupied')
 
 	def review(self, chain_id=None):
 		if chain_id is None:
@@ -206,7 +214,7 @@ class MissionControl:
 		self.review_subject(chain['subject_a'])
 		self.review_subject(chain['subject_b'])
 
[…]

### 2023-08-04 experiment/mission_control.py
Add training score to review output

@@ -292,13 +292,14 @@ class MissionControl:
 		lexicon = {}
 		response_times = []
 		attention_checks_passed = 0
-		item_to_index_map = sorted(list(subject['input_lexicon'].keys()))
+		training_correct = []
 		for trial in subject['responses']:
 			match trial['test_type']:
 				case 'mini_test':
 					if trial['catch_trial']:
 						attention_checks_passed += trial['object_clicked']
 					response_times.append(trial['response_time'])
+					training_correct.append(trial['expected_label'] == trial['input_label'])
 				case 'test_production' | 'comm_production':
 					trial['item'] = f'{trial["shape"]}_{trial["color"]}'
 					lexicon[trial['item']] = trial['input_label']
@@ -315,6 +316,7 @@ class MissionControl:
 		print('Time taken:', f'{minutes}:{str(seconds).zfill(2)}')
 		print('Mean response time', round(sum(response_times) / len(response_times) / 1000, 2))
 		print('Attention checks:', attention_checks_passed)
+		print('Training score:', round(sum(training_correct[-12]) / 12 * 100, 0))
 		print('Total bonus:', subject['total_bonus'])
 		print('Comments:', subject['comments'])
 

### 2023-08-04 experiment/mission_control.py
Streamline review and approval pipeline

@@ -203,18 +203,16 @@ class MissionControl:
 				else:
 					print(f'Cannot close {chain["chain_id"]}: Chain occupied')
 
-	def review(self, chain_id=None):
-		if chain_id is None:
-			raise ValueError('Chain ID must be specified')
-		chain = self.db.chains.find_one({'chain_id': chain_id})
-		if chain is None:
-			raise ValueError('Chain not found')
-		if chain['status'] != 'approval_needed':
-			raise ValueError('Chain not awaiting approval')
-		self.review_subject(chain['subject_a'])
-		self.review_subject(chain['subject_b'])
+	def review(self):
+		for chain in self.db.chains.find({'status': 'approval_needed'}):
+			print(f'CHAIN: {chain["chain_id"]}')
+			print()
+			self.review_subject(chain['subject_a'])
+			self.review_subject(chain['subject_b'])
+			if input('Do you want to approve this generation? ') == 'yes':
+				self.approve(chain['chain_id'])
 
-	def approve(self, chain_id=None, do_not_reopen=True):
+	def approve(self, chain_id=None):
 		if chain_id is None:
 			raise ValueError('Chain ID must be specified')
 		chain = self.db.chains.find_one({'chain_id': chain_id})
@@ -228,7 +226,10 @@ class MissionControl:
 			next_lexicon = subject_a['lexicon']
 		else:
 			next_lexicon = subject_b['lexicon']
-		update_status = 'closed' if do_not_reopen else 'available'
+		if input('Do you want to reopen this chain? ') == 'yes':
+			update_status = 'available'
+		else:
+			update_status = 'closed'
 		if chain['current_gen'] + 1 >= chain['task']['max_gens']:
 			update_status = 'completed'
 			print('🎉 CHAIN COMPLETED!')
@@ -313,12 +314,14 @@ class MissionControl:
 			trained = '➡️ ' if item in subject['training_items'] else '  '
 			print(item, taught_word.ljust(9, ' '), trained, word.ljust(9, ' '), correct)
 		self.db.subjects.update_one({'subject_id': sub_id}, {'$set':{'status': 'reviewed', 'lexicon': lexicon}})
+		print('Subject ID:', subject['subject_id'])
 		print('Time taken:', f'{minutes}:{str(seconds).zfill(2)}')
 		print('Mean response time', round(sum(response_times) / len(response_times) / 1000, 2))
 		print('Attention checks:', attention_checks_passed)
[…]

### 2023-08-04 experiment/mission_control.py
Fix bug

@@ -318,7 +318,7 @@ class MissionControl:
 		print('Time taken:', f'{minutes}:{str(seconds).zfill(2)}')
 		print('Mean response time', round(sum(response_times) / len(response_times) / 1000, 2))
 		print('Attention checks:', attention_checks_passed)
-		print('Training score:', round(sum(training_correct[-12]) / 12 * 100, 0))
+		print('Training score:', round(sum(training_correct[-12:]) / 12 * 100, 0))
 		print('Total bonus:', subject['total_bonus'])
 		print('Comments:', subject['comments'])
 		print()

### 2023-08-04 experiment/mission_control.py
Show audio files in review output

@@ -312,7 +312,7 @@ class MissionControl:
 			taught_word = subject['input_lexicon'][item]
 			correct = '✅' if word == taught_word else '❌'
 			trained = '➡️ ' if item in subject['training_items'] else '  '
-			print(item, taught_word.ljust(9, ' '), trained, word.ljust(9, ' '), correct)
+			print(item, taught_word.ljust(9, ' '), trained, word.ljust(9, ' '), correct, subject['spoken_forms'][item])
 		self.db.subjects.update_one({'subject_id': sub_id}, {'$set':{'status': 'reviewed', 'lexicon': lexicon}})
 		print('Subject ID:', subject['subject_id'])
 		print('Time taken:', f'{minutes}:{str(seconds).zfill(2)}')

### 2023-08-04 code/exp1_analysis.py
Update analysis code

@@ -39,6 +39,10 @@ LABELS = {
 	'com_noVar': 'Learning & Communication, No variation',
 	'lrn': 'Learning-only',
 	'com': 'Learning & Communication',
+	'dif_lrn': 'Differentiation, learning',
+	'dif_com': 'Differentiation, communication',
+	'con_lrn': 'Conservation, learning',
+	'con_com': 'Conservation, communication',
 }
 
 
@@ -92,9 +96,10 @@ def perform_measures(exp_data_file, exp_csv_file):
 				print('    Generation', generation_i)
 				lexicon_a = convert_lexicon_meanings_to_tuple(subject_a['lexicon'])
 				error = transmission_error(lexicon_a, prev_lexicon) if prev_lexicon else None
-				cost = communicative_cost(lexicon_a, (4, 4))
+				cost = communicative_cost(lexicon_a, (3, 3))
 				# comp = complexity(lexicon_a, (4, 4))
-				struc = structure(lexicon_a, (4, 4))
+				# struc = structure(lexicon_a, (4, 4))
+				struc = 0
 				if subject_b:
 					lexicon_b = convert_lexicon_meanings_to_tuple(subject_b['lexicon'])
 					algn = transmission_error(lexicon_a, lexicon_b)
@@ -148,7 +153,7 @@ def plot_generational_change_by_condition(dataset, measure):
 		n_generations = 20
 	elif len(conditions) == 4:
 		fig, axes = plt.subplots(2, 2, figsize=(15, 10))
-		n_generations = 20
+		n_generations = 9
 	for axis, condition in zip(np.ravel(axes), conditions):
 		plot_generational_change(axis, dataset, condition, measure, n_generations, show_mean=False)
 		# axis.plot(range(1, 13), [0, 0, 0, 0.5, 0.5, 0.5, 1.0, 1.0, 1.0, 2.0, 2.0, 2.0], color='gray', linewidth=10, zorder=0)
@@ -321,15 +326,15 @@ def make_ternary_plot():
 
 if __name__ == '__main__':
 
-	exp_json_file = ROOT / 'data' / 'exp2.json'
-	exp_csv_file = ROOT / 'data' / 'exp2.csv'
+	exp_json_file = ROOT / 'data' / 'exp3.json'
+	exp_csv_file = ROOT / 'data' / 'exp3.csv'
 
 	# perform_measures(exp_json_file, exp_csv_file)
 
 	dataset_json = json_load(exp_json_file)
[…]

### 2023-08-17 manuscript/wp4.tex
ms 17/08/23

@@
 	\caption{Two models of heterography.
-In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$, but by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
+In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
 In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
@@
 
-variation
-
-communication
-
 Preregistered \url{https://aspredicted.org/YLG_CKD}
@@
 To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}.
-In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants through repeated learning and production.
-Participant $i$ in the transmission chain learns the system based on the linguistic output of participant $i-i$ and then produces new linguistic output for participant $i+1$ to learn from.
-Through several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
-
-To explore the hypotheses outlines above, we conducted the experiment under our different conditions: high variation vs.\@ no variation and communication vs.\@ no communication.
+In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production.
+Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
+Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
+To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission + Communication.
 
@@
 
-We recruited 538 participants via the Prolific platform.
-Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.64).
-The median completion time was 19:36, resulting in an hourly rate of £9.18 (£11.03 including bonus).
-We limited recruitment to participants who had English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied.
-The most common first languages were: English (59\%), Polish (9.7\%), Portuguese (6.1\%), Italian (5.4\%), Spanish (5\%), and Greek (3.7\%).
-Five participants were rejected because they did not pass the random audio checks, and a further three participants were removed because they had been partnered with one of the five rejected participants.
-In addition, 35 participants were lost to communication-game pairing failures, resulting in a final dataset of 495 participants.
+We recruited 287 participants via the Prolific platform.
+Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74).
+The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus).
+We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2).
+14 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks.
+A further three participants were lost to communication-game pairing failures.
+The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).
 
@@
 
-Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}).
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
+	\vspace*{2pt}
+	\caption{The nine alien object stimuli with their iconic stems (indicating shape) and suffixes (indicating color).
+The spoken forms of the suffixes are homophonous, but their associated spellings are free to evolve and may therefore take on differentiated forms.}
+	\label{stimuli}
+	\end{figure}
+
+Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}.
 The alien words had a spoken and written form composed of a stem and suffix.
-The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>.
-Since we are not interested in participants' learning of the stems, they were designed to be quick and easy to learn by being graphically and phonetically iconic of the shapes they represent, and they were fixed and unchanging throughout the experiment.
-The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to change over time.
-Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.
-The spoken forms were synthesized using the Apple text-to-speech synthesizer (the Tessa voice).
+The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus).
+These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent.
+Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time.
+Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.
 
-The transmission chains were seeded with an orthographic system that may exhibit high variation or no variation according to condition.
-In the no-variation conditions, the suffix had only one spelling, although each chain was initialized with a different randomly generated spelling.
-To generate a spelling we randomly choose a grapheme from each of the sets \{<ee>, <i>, <y>\}, \{<c>, <ck>, <k>\}, and \{<oe>, <oh>, <ow>\} and concatenate them.
-For example, in one chain the suffix might initially be spelled <icoe>, while in another chain it might initially be spelled <yckow>.
-In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects.
-Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow).
-These spellings were not systematic, however; for example, the <yckoh> spelling was not systematically used to represent yellow across different shapes.
-Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.
+Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system.
+This was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>.
+In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity.
+This models an initial state of high spelling variation (every object has a unique suffix spelling), which may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning).
+The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).
 
@@
 
-As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain.
-The first participant in a chain was taught the initial, randomly generated orthographic system, and this system was then free to evolve as it was culturally transmitted.
-This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color).
-Transmission continued until the chain converged on a particular orthographic system.
-A chain is said to converge when two consecutive generations produce identical spellings for all 16 items (including the four unseen items), suggesting that the chain is unlikely to experience significant further change.
-We adopted this convergence-based approach because we were primarily interested in revealing what kinds of orthographic system are stable under the four conditions, but it was unclear a~priori how many generations would be required to reveal this information.
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
+	\vspace*{2pt}
+	\caption{(A) Transmission-only procedure.
+Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items.
+These productions are then used as the training material for the next generation in the chain.
+The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent.
+(B) Transmission + Communication procedure.
+Each generation now consists of two participants who engage in a communicative task.
+Both participants receive training on the same system from the previous generation.
+Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
+	\label{transmission}
+	\end{figure}
+
+As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A).
+The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted.
+Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color).
+Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for the three unseen items.
+Transmission continued for nine generations in each of the ten independent chains.
 
@@
 
-Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials.
-In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms.
-The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word.
-In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects.
-The mini-tests were included to help participants actively engage in the learning process.
-After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text).
-The input box only accepted lowercase alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof.
-Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly.
-However, to receive the bonus, their response had to be submitted within 20~seconds.
-Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p).
-To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.
+All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials.
+Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
+In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each.
+In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type in the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text).
+The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s.
+Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p).
+To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.
 
-\subsubsection{Non-communicative test procedure}
+\subsubsection{Test procedure in Transmission-only}
 
-After training, participants assigned to the non-communicative conditions completed a test phase, alternating between two types of test trial, production trials and comprehension trials.
+After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials.
 In production trials, the participant was shown an object and heard its associated word pronounced aloud.
-The task was to type~in the appropriate spelling under the same input restrictions as in mini-test trials and an additional restriction that the stem (i.e., the first three letters of the word) had to be spelled correctly to ensure that stems could not diverge from the spoken forms over time.
-Since participants heard the word pronounced aloud, typing the stem correctly should be trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly told the participant the correct spelling of the stem.
-In comprehension trials, the participant was shown a word and had to click on the matching object from all 16 objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice).
-In both types of test trial, the participant was awarded a bonus of 2p for every correct answer (unlike mini-tests, there was no 1p bonus for partially correct spellings).
-Each of the 16 object--word pairs (i.e., including the four unseen items) was tested once in production and once in comprehension, resulting in 32 trials and a maximum bonus of 64p.
-Binary feedback was provided on each trial in the form of a sound effect.
+The participant's task was to type~in the appropriate spelling.
+The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial.
+Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.
+This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
+Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix.
+In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice).
+In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided.
+Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).
 
-\subsubsection{Communicative test procedure}
+\subsubsection{Test procedure in Transmission + Communication}
 
-After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant.
-The two participants both received training on the same orthographic system inherited from a previous generation, albeit subject to independent bottlenecks on transmission.
-The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction.
-On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant.
-The second participant---the receiver---would then complete a comprehension trial (select an object for the received word).
-The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme).
-However, the communication game differed from the non-communicative test in three ways.
-First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test.
-Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender.
-Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target.
-Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.
+Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items).
+The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction.
+On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array).
+The two participants then switched roles, resulting in the same overall trial structure as the transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).
+
+The framing and goal of the communication game was, however, quite different from the non-communicative test.
+In communication, the shared goal of the director and matcher is to have a successful communicative interaction, not necessarily to reproduce what they had learned in training.
+Both participants received the 2p bonus each time an interaction was successful; that is, the reward structure is not based on using the ``correct'' forms taught in training but based on accurately conveying meaning.
+The director is thus incentivized to produce a wordform that is unambiguous, and the matcher is incentivized to carefully interpret what the director has attempted to convey.
+The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target.
+Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.
 
 \subsection{Results}
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_lrn.eps}}
+	\vspace*{2pt}
+	\caption{Results from the Transmission-only condition in Experiment~1 (differentiation).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
+Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
+Spellings in bold-italic are the generalizations on unseen items.
+By the ninth generation, most systems are degenerate (e.g., Chains~A, B, and~I), redundant (e.g., Chains~D, E, and~H), or a mixture of the two (e.g., Chain~F).}
+	\label{dif_lrn}
+	\end{figure}
+
+The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}.
+Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}.
+For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension.
+We describe such a system as ``redundant'' because shape is consistently expressed by the stem, so the suffix spellings that emerged in this case convey no additional information beyond what is already conveyed by the stem.
+Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition.
+We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J.
+Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language.
+Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>.
+The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_com.eps}}
+	\vspace*{2pt}
+	\caption{Results from the Transmission + Communication condition in Experiment~1 (differentiation).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
+Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
+Spellings in bold-italic are the generalizations on unseen items.
+There are some isolated examples of differentiation through implicit generalization (K5, M1, P5, S2) and explicit innovation (K9, L7, M7, O4, Q7).}
+	\label{dif_com}
+	\end{figure}
+
+The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
+Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
+In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6).
+Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable.
+In particular, there are two main kinds of color-expressive system that emerge.
+The first are the generalization-based expressive systems: K5, M1, P5, and S2.
+In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape.
+For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>.
+In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->).
+We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.
+
+The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4.
+In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>.
+In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9.
+Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously.
+Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color.
+In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-blue>, <-oran>, <-pink>, <-red>, <-yell>), but these pairs were excluded and replaced with a new pair before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words.
+However, we allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red).
+Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.
 
@@
 
-For the purpose of the results, we are primarily interested in the spelling of the suffixes, so we first needed to isolate the suffixes from the stems.
-Although the first three letters (<buv>, <zet>, <gaf>, <wop>) were fixed and unchanging throughout the experiment, it was possible for the stems to grow longer if additional material was inserted before the suffix (e.g., <gaff>, <wopb>, <zetch>).
-We consider the stem to be all material up to but not including the first vowel letter after the original three-letter stems, such that, for example, <zethyckoh> will be parsed into the stem <zeth> and suffix <yckoh>.
-Of course, it is not possible to know how participants themselves divided the strings (if at all), but since all words were phonetically identical from the /i/ vowel onward, it is reasonable to parse the strings this way on the basis that this yields the shortest grammar.
-
-To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 4×4 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings.
+To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 3×3 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings.
 For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
@@
 	S_\mathrm{degenerate} = \begin{bmatrix}
-	1 & 1 & 1 & 1 \\
-	1 & 1 & 1 & 1 \\
-	1 & 1 & 1 & 1 \\
-	1 & 1 & 1 & 1
+	1 & 1 & 1 \\
+	1 & 1 & 1 \\
+	1 & 1 & 1
 	\end{bmatrix}
@@
 Here the number 1 represents the one and only suffix spelling.
-We define an ``expressive'' system as one that marks the color dimension using four distinct spellings, resulting in the following matrix:
+We define an ``expressive'' system as one that marks the color dimension using three distinct spellings, resulting in the following matrix:
 \begin{equation}
 	S_\mathrm{expressive} = \begin{bmatrix}
-	1 & 2 & 3 & 4 \\
-	1 & 2 & 3 & 4 \\
-	1 & 2 & 3 & 4 \\
-	1 & 2 & 3 & 4
+	1 & 2 & 3 \\
+	1 & 2 & 3 \\
+	1 & 2 & 3
 	\end{bmatrix}
@@
 	S_\mathrm{redundant} = \begin{bmatrix}
-	1 & 1 & 1 & 1 \\
-	2 & 2 & 2 & 2 \\
-	3 & 3 & 3 & 3 \\
-	4 & 4 & 4 & 4
+	1 & 1 & 1 \\
+	2 & 2 & 2 \\
+	3 & 3 & 3
 	\end{bmatrix}
@@
 Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant.
-Such a system contains unnecessary complexity---different ways of writing /ikəʊ/ that carry no additional information.
+Such a system contains unnecessary complexity---different ways of writing /kəʊ/ that carry no additional information.
 
@@
 To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric.
-VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another set partition.
+VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
 So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information.
@@
 In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>.
-Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the learning-only condition.
+Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.
 
@@
 
-Our second experiment investigates the conservation model: Given than an informative system already exists (both in speech and in writing), will the informative system persist even after the degeneration of the spoken language towards homophony?
-And does this happen preferentially in the presence of a shared communicative task.
-How much homophony can an informative, heterographic system resist under an increasingly strong simplicity pressure.
+We now turn our attention to the conservation model of heterographic homophones: Given that an informative system already exists (both in speech and in writing), does that informative system persist in writing even after the spoken language has degenerated into homophony?
+And, importantly, does this happen preferentially in the presence of a communicative pressure?
 
@@
 
-The methods were identical to Experiment~1 with the following major exceptions.
-Firstly, and most notably, the artificial language used to seed each chain starts out fully compositional (in both its spoken and written forms), since we are now interested in whether the orthographies will maintain informativeness rather than gain it.
-Secondly, and consequently, the convergence-based approach used in Experiment~1 to determine the number of generations is no longer appropriate, so we now use a fixed number of generations.
-Thirdly, we simulate regular sound mergers which result in the spoken forms of the suffixes becoming increasingly less informative over generational time (this contrasts with Experiment~1, where the spoken form of the suffixes were always fully uninformative).
+The methods were largely identical to Experiment~1 with two major exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.
 
@@
 
-We recruited xxx participants via the Prolific platform.
-The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.xxx).
-The median completion time was xxx:xxx, resulting in an hourly rate of £xxx (£xxx including bonus).
-Unlike Experiment~1 we restricted recruitment to native speakers of English because it was particularly important that participants would be able to distinguish the voiceless fricatives /θ s ʃ h/ and we found in piloting that non-native speakers of English frequently struggled with these distinctions.
-xxx participants were rejected because they did not pass the random audio checks, and a further xxx participants were removed because they had been partnered with one of the rejected participants.
-In addition, xxx participants were lost to communication-game pairing failures, resulting in a final dataset of 360 participants (organized into ten chains of twelve generations in each condition, with two participants per generation in Learning \& Communication).
+We recruited 297 native-English participants via the Prolific platform.
+The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78).
+The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus).
+17 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks.
+A further 10 participants were lost to communication-game pairing failures.
+Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).
 
@@
 The alien objects and word stems were identical to Experiment~1.
-Unlike Experiment~1, however, the chains were initialized with a fully compositional language that uses four distinct suffixes to systematically express each of the four colors.
-The suffixes took the form /ɪ\textit{CV}/ where $C$ is a voiceless fricative consonant from the set \{/θ/, /s/, /ʃ/, /h/\} and $V$ is a vowel from the set \{/ə/, /ɛɪ/, /əʊ/, /u/\}.
-The consonant and vowel sets were randomly permuted for each iterated learning chain, such that the suffix forms were unique to each chain.
-For example, under the permutations given above, the resulting suffixes would be /ɪθə/, /ɪsɛɪ/, /ɪʃəʊ/, and /ɪhu/, but another iterated learning chain might use the suffixes /ɪsəʊ/, /ɪʃɛɪ/, /ɪθu/, and /ɪhə/.
-The initial orthographic system used to seed the chain was generated based on the following phoneme–grapheme mapping \{/θ/→<th>, /s/→<s>, /ʃ/→<sh>, /h/→<x>, /ə/→<a>, /ɛɪ/→<e>, /əʊ/→<o>, /u/→<u>\}.
-The suffixes were designed to be distinctive (and therefore easy to memorize and recall), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /θ/ might be supplanted by /s/ in speech or that /θ/ might be spelled <x> in writing).
-We attempted to achieve this balance by combining consonants that are very similar (differing only in place of articulation) with vowels that are very dissimilar.
+Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors.
+These suffixes were made~up~of a consonant from the set \{/f/, /s/, /ʃ/\} and a vowel from the set \{/ə/, /ɛɪ/, /əʊ/\}, but the exact forms were randomized for each chain (e.g., one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/).
+The initial orthographic system was transparent and generated based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}.
+The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ in speech or that /ʃ/ might be spelled <s> in writing).
+We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems.
+The spoken forms were synthesized using the Apple text-to-speech synthesizer (Moira voice).
 
@@
 
-Each iterated learning chain was run for 12 generations, divided into four epochs of three generations.
-In the first epoch, all spoken suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color.
-This was achieved through three sound changes occurring between the four epochs.
-In the first sound change, two of the suffixes were chosen at random to be merged.
-To perform the merger, the consonant from one of the suffixes (chosen at random) was paired with the vowel from the other suffix, resulting in a new suffix form that would replace the original two.
-For example, assuming the language has the suffixes /ɪθə/ (pink), /ɪsɛɪ/ (gray), /ɪʃəʊ/ (blue), and /ɪhu/ (yellow), we might merge /ɪsɛɪ/ and /ɪhu/ into /ɪhɛɪ/, such that the spoken language is no longer able to mark a distinction between gray and yellow.
-In the second sound change, the same procedure is applied to the two suffixes that were not selected previously (e.g., /ɪθə/ and /ɪʃəʊ/ become /ɪθəʊ/).
-And in the third and final sound change, the two suffix forms created through the first and second sounds changes---/ɪhɛɪ/ (grey/yellow) and /ɪθəʊ/ (pink/blue)---merge into a single form (e.g., /ɪθɛɪ/).
-Importantly, the spellings do not change automatically following a sound change event; instead, the orthographic system is free to adapt (or not) in response to sound changes.
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/sound_change.eps}}
+	\vspace*{2pt}
+	\caption{Examples of the spoken suffixes in Experiment~2.
+During Epoch~I, color is represented by three distinct spoken suffixes.
+During Epoch~II, two of the suffixes are homophonous, reducing the informativeness of the spoken language.
+During Epoch~III, the spoken form of the language makes no color distinction.}
+	\label{sound_change}
+	\end{figure}
+
+As in Experiment~1, each transmission chain was run for nine generations.
+Additionally, the nine generations were divided into three epochs, each lasting three generations.
+During Epoch~I (Generations~1 to~3), the three spoken suffixes were distinct (as described above), allowing the spoken language to express all three colors without ambiguity.
+During Epoch~II (Generations~4 to~6), the spoken language had two distinct suffix forms, reducing its informativeness.
+During Epoch~III (Generations~7 to~9), all three spoken suffixes were homophonous, just as in Experiment~1, making the spoken language entirely uninformative about color.
+This was achieved through two sets of sound changes, the first occurring in the transition from Epoch~I to~II and the second occurring in the transition from Epoch~II to~III.
+An example is illustrated in Fig.~\ref{sound_change}.
+In the first sound change, two of the spoken suffix forms were chosen at random and the consonant from one (chosen at random) was paired with the vowel from the other, resulting in a new suffix form that replaced the original two.
+For example, a language that uses the suffixes /fəʊ/, /ʃə/, /sɛɪ/ in Epoch~I, might use the suffixes /fɛɪ/, /ʃə/, /fɛɪ/ in Epoch~II (i.e., /s/→/f/ and /əʊ/→/ɛɪ/).
+In the second sound change, the remaining two suffix forms were merged in the same way, resulting in full homophony.
+For example, in Epoch~III, the suffixes might become /fə/, /fə/, /fə/ (i.e., /ʃ/→/f/ and /ɛɪ/→/ə/).
+Crucially, the spellings do not automatically change following a sound change event; rather, the orthographic system is free to adapt (or not) in response to the sound changes.
 
 \subsection{Results}
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/con_lrn.eps}}
+	\vspace*{2pt}
+	\caption{Results from the Transmission-only condition in Experiment~2 (conservation).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
+Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
+Spellings in bold-italic are the generalizations on unseen items.}
+	\label{con_lrn}
+	\end{figure}
+
+The results for all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
+	\vspace*{2pt}
+	\caption{Results from the Transmission + Communication condition in Experiment~2 (conservation).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
+Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
+Spellings in bold-italic are the generalizations on unseen items.}
+	\label{con_com}
+	\end{figure}
+
+The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.
 

### 2023-08-17 code/demographics.py
Update analysis code

@@ -1,3 +1,4 @@
+import re
 from pathlib import Path
 import numpy as np
 import pandas as pd
@@ -7,20 +8,19 @@ from utils import json_load, json_save
 ROOT = Path(__file__).parent.parent.resolve()
 
 
-def calculate_rates(exp_id, with_bonus=False):
+def calculate_rates(exp_id, n_subjects, chain_filter='.+', with_bonus=False):
+	chain_filter = re.compile(chain_filter)
 	times = []
 	amounts = []
 	rates = []
-	comments_time = []
 
-	for i in range(1, 539):
+	for i in range(1, n_subjects + 1):
+
 		subject_id = str(i).zfill(3)
 		path = ROOT/'data'/exp_id/f'subject_{subject_id}.json'
-		try:
-			data = json_load(path)
-
-		except:
-			break
+		data = json_load(path)
+		if not chain_filter.fullmatch(data['chain_id']):
+			continue
 
 		time = data['modified_time'] - data['creation_time']
 		times.append(time)
@@ -33,6 +33,7 @@ def calculate_rates(exp_id, with_bonus=False):
 		rate = 3600 / time * amount
 		rates.append(rate)
 
+	print('Number of subjects', len(times))
 	print('Median completion time (mins):', np.median(times) / 60)
 	print('Median earnings (GBP):', np.median(amounts) / 100)
 	print('Median hourly rate (GBP):', np.median(rates) / 100)
@@ -48,5 +49,5 @@ def most_common_languages(exp_id):
 		print(i, lang, counts[lang], round(counts[lang] / len(langs) * 100, 2))
 
 
-calculate_rates('exp3', with_bonus=True)
-most_common_languages('exp3')
+calculate_rates('exp3', 584, chain_filter=r'dif.+', with_bonus=False)
+calculate_rates('exp3', 584, chain_filter=r'con.+', with_bonus=False)

### 2023-08-17 code/disttern.py

@@ -67,7 +67,7 @@ def draw_ternary_axes(axis):
 	# axis.plot([ref_r[0], centroid[0]], [ref_r[1], centroid[1]], color='gray')
 	# axis.plot([0.5, centroid[0]], [0, centroid[1]], color='gray')
 
-def make_ternary_plot(reference_objects, target_objects, distance_func, color='MediumSeaGreen', jitter=False):
+def make_ternary_plot(reference_objects, target_objects, distance_func, color='MediumSeaGreen', jitter=False, title=None):
 	'''
 	Given three reference objects (placed at the top, left, and right vertices
 	of the triangle), a collection of target objects, and a distance
@@ -88,6 +88,8 @@ def make_ternary_plot(reference_objects, target_objects, distance_func, color='M
 		points[:, 0] += (np.random.random(len(target_objects)) - 0.5) * 0.01
 		points[:, 1] += (np.random.random(len(target_objects)) - 0.5) * 0.01
 	axis.scatter(points[:, 0], points[:, 1], color=color, alpha=0.5, edgecolor='white')
+	if title:
+		axis.set_title(title)
 	axis.axis('off')
 	fig.tight_layout()
 	plt.show()

### 2023-08-17 code/exp1_analysis.py

@@ -155,7 +155,7 @@ def plot_generational_change_by_condition(dataset, measure):
 		fig, axes = plt.subplots(2, 2, figsize=(15, 10))
 		n_generations = 9
 	for axis, condition in zip(np.ravel(axes), conditions):
-		plot_generational_change(axis, dataset, condition, measure, n_generations, show_mean=True)
+		plot_generational_change(axis, dataset, condition, measure, n_generations, show_mean=False)
 		# axis.plot(range(1, 13), [0, 0, 0, 0.5, 0.5, 0.5, 1.0, 1.0, 1.0, 2.0, 2.0, 2.0], color='gray', linewidth=10, zorder=0)
 		# axis.plot(range(1, 13), [209, 209, 209, 180, 180, 180, 206, 206, 206, 123, 123, 123], color='gray', linewidth=10, zorder=0)
 		
@@ -299,10 +299,12 @@ def make_ternary_plot(exp_data):
 			scatter_objects.append(
 				matrix.make_matrix(chain[-1][0]['lexicon'], 3, 3)
 			)
-		disttern.make_ternary_plot(ref_objects, scatter_objects, matrix.voi.variation_of_information, jitter=True)
+		disttern.make_ternary_plot(ref_objects, scatter_objects, matrix.voi.variation_of_information, jitter=True, title=LABELS[condition])
 
 import visualize
 def make_panel_visualization(exp_data):
+	A_to_J = ['A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J']
+	K_to_T = ['K', 'L', 'M', 'N', 'O', 'P', 'Q', 'R', 'S', 'T']
 	for condition, data in exp_data.items():
 		print(condition)
 		panel = []
@@ -324,7 +326,8 @@ def make_panel_visualization(exp_data):
 				chain_matrices.append((mat, cp, ss, sounds, training_items))
 			panel.append(chain_matrices)
 		show_sounds = True if 'con' in condition else False
-		visualize.draw_panel(f'/Users/jon/Desktop/{condition}.pdf', panel, show_sounds=show_sounds)
+		chain_ids = A_to_J if 'lrn' in condition else K_to_T
+		visualize.draw_panel(f'/Users/jon/Desktop/{condition}.eps', panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
 
 
 

### 2023-08-17 code/visualize.py

@@ -1,24 +1,23 @@
 import numpy as np
 import cairocffi as cairo
 
-font_size = 50
+font_size = 25
 helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_NORMAL, cairo.FONT_WEIGHT_NORMAL)
 helvetica_sf = cairo.ScaledFont(helvetica, cairo.Matrix(xx=font_size, yy=font_size))
 
-seen_suffix_font_size = 8
+seen_suffix_font_size = 4
 seen_suffix_helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_NORMAL, cairo.FONT_WEIGHT_NORMAL)
 seen_suffix_helvetica_sf = cairo.ScaledFont(seen_suffix_helvetica, cairo.Matrix(xx=seen_suffix_font_size, yy=seen_suffix_font_size))
 
-unseen_suffix_font_size = 8
+unseen_suffix_font_size = 4
 unseen_suffix_helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_OBLIQUE, cairo.FONT_WEIGHT_BOLD)
 unseen_suffix_helvetica_sf = cairo.ScaledFont(unseen_suffix_helvetica, cairo.Matrix(xx=unseen_suffix_font_size, yy=unseen_suffix_font_size))
 
-sound_font_size = 10
+sound_font_size = 5
 sound_doulos = cairo.ToyFontFace('Doulos SIL', cairo.FONT_SLANT_NORMAL, cairo.FONT_WEIGHT_NORMAL)
 sound_doulos_sf = cairo.ScaledFont(sound_doulos, cairo.Matrix(xx=sound_font_size, yy=sound_font_size))
 
-chain_ids = ['A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J']
-padding = 8
+padding = 3
 
 
 def draw_char(surface, char, cell_size):
@@ -43,7 +42,7 @@ def draw_matrix(surface, generation, cell_size):
 		context.rectangle(j*square_size + padding, i*square_size + padding, square_size, square_size)
 		context.fill_preserve()
 		context.set_source_rgb(1, 1, 1)
-		context.set_line_width(2)
+		context.set_line_width(1)
 		context.stroke()
 		context.set_source_rgb(1, 1, 1)
 		if not sounds or f'{i}_{j}' in training_items:
[…]

### 2023-08-21 code/matrix.py
analysis

@@ -5,14 +5,16 @@ import re
 import Levenshtein
 
 
-RE_STEM_SUFFIX = re.compile(r'(buvi|zeti|gafi|wopi)(\w*)')
+RE_STEM_SUFFIX = re.compile(r'^(buvi|zeti|wopi|gafi)(\w*)$')
 
 
 def parse_stem_and_suffix(word):
 	word_match = RE_STEM_SUFFIX.match(word)
 	if word_match is None:
 		raise ValueError('Cannot parse stem and suffix')
-	return word_match.group(1), word_match.group(2)
+	stem = word_match.group(1)
+	suffix = word_match.group(2) or '∅'
+	return stem, suffix
 
 def get_suffix_spellings(lexicon):
 	suffixes = []
@@ -21,7 +23,7 @@ def get_suffix_spellings(lexicon):
 		suffixes.append(suffix)
 	return sorted(list(set(suffixes)))
 
-def make_matrix(lexicon, m=4, n=4):
+def make_matrix(lexicon, m=3, n=3):
 	suffix_spellings = get_suffix_spellings(lexicon)
 	matrix = np.zeros((m, n), dtype=int)
 	for i in range(m):
@@ -30,7 +32,7 @@ def make_matrix(lexicon, m=4, n=4):
 			matrix[i, j] = suffix_spellings.index(suffix)
 	return matrix
 
-def make_matrix_with_cp(lexicon, ss, m=4, n=4):
+def make_matrix_with_cp(lexicon, ss, m=3, n=3):
 	suffix_spellings = ss
 	matrix = np.zeros((m, n), dtype=int)
 	for i in range(m):
@@ -120,32 +122,6 @@ def cost(system):
 	U_size = np.product(system.shape)
 	return 1 / U_size * sum([-np.log2(1 / signals.count(s)) for m, s in np.ndenumerate(system)])
 
-# typology = {
-# 	'transparent': np.array([
-# 		[0, 0, 0, 0],
-# 		[0, 0, 0, 0],
-# 		[0, 0, 0, 0],
-# 		[0, 0, 0, 0],
-# 	], dtype=int),
-# 	'holistic': np.array([
-# 		[ 0, 1, 2, 3],
-# 		[ 4, 5, 6, 7],
-# 		[ 8, 9,10,11],
-# 		[12,13,14,15],
-# 	], dtype=int),
[…]

### 2023-08-21 code/visualize.py

@@ -104,8 +104,12 @@ def draw_panel(output_path, chains, figure_width=1000, chain_ids=None, show_gene
 	if output_path.endswith('.eps'):
 		surface = cairo.PSSurface(output_path, figure_width, figure_height)
 		surface.set_eps(True)
-	else:
+	elif output_path.endswith('.pdf'):
 		surface = cairo.PDFSurface(output_path, figure_width, figure_height)
+	elif output_path.endswith('.svg'):
+		surface = cairo.SVGSurface(output_path, figure_width, figure_height)
+	else:
+		raise ValueError('Unrecognized format')
 	context = cairo.Context(surface)
 	context.set_source_rgb(1, 1, 1)
 	context.paint()

### 2023-08-21 manuscript/wp4.tex
ms 20/08

@@
 Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
-To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission + Communication.
+To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.
 
@@
 We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2).
-14 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks.
+14 participants were excluded because they (or their partners) used English color words (8) or failed the auditory attention checks (6).
 A further three participants were lost to communication-game pairing failures.
-The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).
+The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).
 
@@
 The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent.
-(B) Transmission + Communication procedure.
+(B) Transmission~+ Communication procedure.
 Each generation now consists of two participants who engage in a communicative task.
@@
 Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color).
-Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for the three unseen items.
-Transmission continued for nine generations in each of the ten independent chains.
+Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items.
+Transmission continued for nine generations in each of ten independent chains.
+In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend).
+The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.
 
@@
 
-\subsubsection{Test procedure in Transmission + Communication}
+\subsubsection{Test procedure in Transmission~+ Communication}
 
@@
 On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array).
-The two participants then switched roles, resulting in the same overall trial structure as the transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).
+The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).
 
@@
 	\vspace*{2pt}
-	\caption{Results from the Transmission + Communication condition in Experiment~1 (differentiation).
+	\caption{Results from the Transmission~+ Communication condition in Experiment~1 (differentiation).
 Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
@@
 
-The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
 Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
@@
 Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color.
-In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-blue>, <-oran>, <-pink>, <-red>, <-yell>), but these pairs were excluded and replaced with a new pair before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words.
-However, we allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red).
+In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-pink> or <-red>, <-oran> or <-yell>, and <-blue>), but these pairs were excluded and replaced before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words.
+We allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red).
 Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.
@@
 Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography.
-But this did not happen, even under the communicative pressure.
+But this did not happen, even under communicative pressure.
 Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings.
@@
 As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose.
-A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion} \textit{transmission}, and \textit{invention}).
+A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).
 
@@
 
-The methods were largely identical to Experiment~1 with two major exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.
+The methods were identical to Experiment~1 with two main exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.
 
@@
 
-We recruited 297 native-English participants via the Prolific platform.
+The experiment was completed by 297 native-English participants recruited through Prolific.
 The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78).
 The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus).
-17 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks.
+17 participants were excluded because they (or their partners) failed the auditory attention checks (15) or used English color words (2).
 A further 10 participants were lost to communication-game pairing failures.
-Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).
+Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).
 
@@
 Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors.
-These suffixes were made~up~of a consonant from the set \{/f/, /s/, /ʃ/\} and a vowel from the set \{/ə/, /ɛɪ/, /əʊ/\}, but the exact forms were randomized for each chain (e.g., one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/).
-The initial orthographic system was transparent and generated based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}.
-The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ in speech or that /ʃ/ might be spelled <s> in writing).
+A separate set of suffixes was created for each chain by concatenating a randomly drawn consonant from \{/f/, /s/, /ʃ/\} and a randomly drawn vowel from \{/ə/, /ɛɪ/, /əʊ/\}, both without replacement.
+For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/.
+The initial orthographic system was transparent and based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}.
+The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ or /ʃ/ in speech or that /ʃ/ might be spelled <s> or <x> in writing).
 We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems.
@@
 In the first sound change, two of the spoken suffix forms were chosen at random and the consonant from one (chosen at random) was paired with the vowel from the other, resulting in a new suffix form that replaced the original two.
-For example, a language that uses the suffixes /fəʊ/, /ʃə/, /sɛɪ/ in Epoch~I, might use the suffixes /fɛɪ/, /ʃə/, /fɛɪ/ in Epoch~II (i.e., /s/→/f/ and /əʊ/→/ɛɪ/).
 In the second sound change, the remaining two suffix forms were merged in the same way, resulting in full homophony.
-For example, in Epoch~III, the suffixes might become /fə/, /fə/, /fə/ (i.e., /ʃ/→/f/ and /ɛɪ/→/ə/).
 Crucially, the spellings do not automatically change following a sound change event; rather, the orthographic system is free to adapt (or not) in response to the sound changes.
+Note also that an individual participant does not directly experience the sound changes; rather, a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
+In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.
 
@@
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
-Spellings in bold-italic are the generalizations on unseen items.}
+Spellings in bold-italic are the generalizations on unseen items.
+The final systems are characterized by the conservation of form without the conservation of expressivity.}
 	\label{con_lrn}
@@
 The results for all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.
+It is immediately clear that color-expressive suffixes (as indicated by vertical stripes) are maintained fairly reliably through the first epoch; perfectly in the case of Chains~A, B, D, F, and~I, and with some errors in the other five chains.
+Some of these errors are very minor, such as the use of <sha> instead of <xa> for one item in H3, while others are more catastrophic, such as the early loss of the <sei>/<xa> distinction in Chain~J.
+Generation~4 represents the first real test of the orthographic systems in the face of sound change, and, in most cases, the Generation~3 systems are preserved quite faithfully (Chains~A, B, D, E, H, and~I), but by the end of the epoch (Generation~6), many have degenerated into redundant systems that encode shape (Chains~B, G, and~J) or transparent systems that mirror the Epoch~II pattern of homophony (Chains~C, E, F, and~I).
+These processes continue into Epoch~III and by the ninth generation, all systems have become degenerate, redundant, or some mixture of the two.
+The one exception is Chain~D, whose original spellings---<sei>, <fa>, <xoe>---were conserved perfectly through to the final generation with only one generalization error in Generation~8, which was quickly reverted in Generation~9.
+
+Like Experiment~1, redundant systems are characteristic of the Transmission-only condition, especially in Epoch~III.
+As the spoken suffixes becomes more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings.
+Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost.
+For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Roman alphabet, but which trace their origins back to the original seed orthography.
+Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without the conservation of expressivity.
 
@@
 	\vspace*{2pt}
-	\caption{Results from the Transmission + Communication condition in Experiment~2 (conservation).
+	\caption{Results from the Transmission~+ Communication condition in Experiment~2 (conservation).
 Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
@@
 
-The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.
+Like Transmission-only, the seed systems are mostly maintained faithfully through Epoch~I; perfectly in the case of Chains~K, L, N, O, and~J, and with some errors in the other five chains, although some of these errors are non-destructive changes, such as <x> being replaced with <sh> in R1.
+Several systems were then maintained through Epoch~II, notably Chains~O, P, R, and~T, and, in one case, through to the end of Epoch~III (Chain~R, albeit with the original <xei> spelling replaced with <shei>).
+The Chain~O system was preserved up to Generation~7, Chain~P was maintained up to Generation~7 and almost to Generation~9 with two modifications (<sha> instead of <xa> and <fa> instead of <fei>), and Chain~T was conserved faithfully up to Generation~8.
+The final form of Chain~Q was also fully expressive, albeit through a combination of both conservation and differentiation: The <fei> form was conserved from the seed orthography, the <oxie> spelling appears to derive from a misremembering of <xoe> (partial conservation), and the <fe> spelling, which began as a typo introduced in Q6, seems to have been generalized across the blue items in Q7, perhaps to differentiate them from the yellow items (indeed, the participant's partner made the same generalization).
+A similar case of differentiation might also have occurred in L9, where the <sol> spelling (originally a typo on <so>) was generalized across the yellow items, resulting in a semi-expressive system (although the participant's partner generalized the <sol> spelling across shape).
+
+Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words in Generation~9 of Chain~K, although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English).
+Presumably, the general conservation of expressive spelling in Experiment~2 negated the need to innovate novel systems.
 
 \subsection{Summary}
+
+In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
 

### 2023-08-24 code/disttern.py
Deal with typological plots

@@ -57,17 +57,11 @@ def draw_ternary_axes(axis):
 	centroid = 0.5, height / 3
 	ref_l = proportion_to_point(0.5, right_axis=False)
 	ref_r = proportion_to_point(0.5, right_axis=True)
-	# axis.add_patch( plt.Polygon([(0.5, height), (0, 0), (1, 0)], color='black', fill=False, linewidth=2) )
-
 	axis.add_patch( plt.Polygon([(0.5, height), ref_l, centroid, ref_r], fill=True, color='black', alpha=0.4, linewidth=0) )
 	axis.add_patch( plt.Polygon([(0, 0), (0.5, 0), centroid, ref_l], fill=True, color='gray', alpha=0.4, linewidth=0) )
 	axis.add_patch( plt.Polygon([(1, 0), ref_r, centroid, (0.5, 0)], fill=True, color='lightgray', alpha=0.4, linewidth=0) )
 
-	# axis.plot([ref_l[0], centroid[0]], [ref_l[1], centroid[1]], color='green')
-	# axis.plot([ref_r[0], centroid[0]], [ref_r[1], centroid[1]], color='gray')
-	# axis.plot([0.5, centroid[0]], [0, centroid[1]], color='gray')
-
-def make_ternary_plot(reference_objects, target_objects, distance_func, color='MediumSeaGreen', jitter=False, title=None):
+def make_ternary_plot(axis, reference_objects, target_objects, distance_func, color='MediumSeaGreen', jitter=False, title=None):
 	'''
 	Given three reference objects (placed at the top, left, and right vertices
 	of the triangle), a collection of target objects, and a distance
@@ -75,14 +69,16 @@ def make_ternary_plot(reference_objects, target_objects, distance_func, color='M
 	different to an ordinary ternary plot; proximity from the three
 	vertices represents distance from three reference objects.
 	'''
+	print(len(target_objects))
+	if not isinstance(color, list):
+		color = [color] * len(target_objects)
 	props = calculate_proportions(reference_objects, target_objects, distance_func)
 	points = np.zeros((len(target_objects), 2), dtype=float)
 	for i, (prop_l, prop_r) in enumerate(props):
 		x1_l, y1_l = proportion_to_point(prop_l, right_axis=False)
 		x1_r, y1_r = proportion_to_point(prop_r, right_axis=True)
[…]

### 2023-08-24 code/exp1_analysis.py

@@ -16,14 +16,15 @@ from utils import json_load, json_save
 
 ROOT = Path(__file__).parent.parent.resolve()
 
+plt.rcParams.update({'font.sans-serif': 'Helvetica Neue', 'font.size': 7})
 
 COLORS = colormaps['tab10_r'].colors
 
 MEASURE_RANGES = {
-	'cost': (0, 2),
+	'cost': (0, 1.8),
 	'complexity': (100, 600),
 	'structure': (0, 10),
-	'error': (0, 3),
+	'error': (0, 2),
 	'alignment': (0, 3),
 }
 
@@ -93,6 +94,12 @@ def perform_measures(exp_data_file, exp_csv_file):
 			print('  Chain', chain_i)
 			prev_lexicon = None
 			for generation_i, (subject_a, subject_b) in enumerate(chain):
+				if generation_i < 4:
+					epoch_i = 1
+				elif generation_i < 7:
+					epoch_i = 2
+				else:
+					epoch_i = 3
 				print('    Generation', generation_i)
 				lexicon_a = convert_lexicon_meanings_to_tuple(subject_a['lexicon'])
 				error = transmission_error(lexicon_a, prev_lexicon) if prev_lexicon else None
@@ -109,6 +116,7 @@ def perform_measures(exp_data_file, exp_csv_file):
 					condition,
 					chain_i,
 					generation_i,
+					epoch_i,
 					error,
 					struc,
 					cost,
@@ -116,7 +124,7 @@ def perform_measures(exp_data_file, exp_csv_file):
 				])
 				prev_lexicon = lexicon_a
 
-	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'error', 'structure', 'cost', 'alignment'])
+	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'epoch', 'error', 'structure', 'cost', 'alignment'])
 	df.to_csv(exp_csv_file)
 
 
@@ -131,12 +139,10 @@ def plot_generational_change(axis, dataset, condition, measure, n_generations=20
 	for chain_i in sorted(condition_subset['chain'].unique()):
 		chain_subset = condition_subset[ condition_subset['chain'] == chain_i ]
 		chain_data.append(list(chain_subset[measure]))
-		if show_mean:
-			continue
 		axis.plot(chain_subset['generation'], chain_subset[measure], label=f'Chain {chain_i + 1}', color=COLORS[chain_i])
 	if show_mean:
[…]

### 2023-08-24 code/matrix.py

@@ -1,8 +1,6 @@
 import numpy as np
 import cairocffi as cairo
-import voi
 import re
-import Levenshtein
 
 
 RE_STEM_SUFFIX = re.compile(r'^(buvi|zeti|wopi|gafi)(\w*)$')
@@ -110,20 +108,8 @@ def draw(matrix, color_palette, output_path):
 			context.stroke()
 	surface.finish()
 
-def typological_distribution(matrix, probabilistic=False):
-	dist = [voi.variation_of_information(matrix, system) for name, system in typology.items()]
-	if probabilistic is False:
-		return dist
-	dist = np.array([np.exp2(-1 * val**2) for val in dist])
-	return dist / dist.sum()
 
-def cost(system):
-	signals = list(system.flatten())
-	U_size = np.product(system.shape)
-	return 1 / U_size * sum([-np.log2(1 / signals.count(s)) for m, s in np.ndenumerate(system)])
-
-
-typology = {
+reference_systems = {
 	'transparent': np.array([
 		[0, 0, 0],
 		[0, 0, 0],
@@ -149,9 +135,6 @@ typology = {
 
 if __name__ == '__main__':
 
-	for name, system in typology.items():
+	for name, system in reference_systems.items():
 		color_palette = generate_color_palette(system)
-		# print(color_palette)
 		draw(system, color_palette, f'/Users/jon/Desktop/{name}.pdf')
-
-		print(cost(system))

### 2023-08-24 code/ploymod_plot.py

@@ -1,0 +1,55 @@
+import matplotlib.pyplot as plt
+import arviz as az
+import numpy as np
+import pandas as pd
+
+
+def plot_chains(axis, dataset, var, show_mean=False):
+	chain_data = []
+	for chain_i in sorted(dataset['chain'].unique()):
+		chain_subset = dataset[ dataset['chain'] == chain_i ]
+		chain_data.append(list(chain_subset[var]))
+		axis.plot(chain_subset['generation'], chain_subset[var], label=f'Chain {chain_i + 1}')
+	if show_mean:
+		chain_data = np.array(chain_data)
+		axis.plot(chain_subset['generation'], chain_data.mean(axis=0), label=f'Chain {chain_i + 1}', color='black', linewidth=5)
+	# axis.set_xlim(0, n_generations)
+	# axis.set_ylim(*pad_range(*MEASURE_RANGES[measure]))
+	# axis.set_xticks(list(range(n_generations + 1)))
+	# axis.set_xlabel('Generation')
+	# axis.set_ylabel(LABELS[measure])
+	# axis.set_title(LABELS[condition])
+
+
+def plot_predictive(axis, trace, var, color):
+	generation = np.arange(0, 10)
+	az.plot_hdi(generation, trace.posterior[var], hdi_prob=0.95, smooth=False, color=color, fill_kwargs={'alpha': 0.25, 'linewidth': 0}, ax=axis)
+	axis.plot(generation, trace.posterior[var].mean(('chain', 'draw')), color=color)
+	# axis.set_ylim(0, 1.8)
+
+
+df = pd.read_csv('../data/exp3.csv')
+trace = az.from_netcdf('../data/exp3_con.netcdf')
+
+fig, axes = plt.subplots(2, 2, figsize=(6, 4))
+
+plot_chains(axes[0,0], df[ df['condition'] == 'con_lrn' ], 'cost', show_mean=True)
+plot_chains(axes[0,1], df[ df['condition'] == 'con_com' ], 'cost', show_mean=True)
+
+axes[0,0].set_ylim(0, 1.8)
+axes[0,1].set_ylim(0, 1.8)
+
+plot_predictive(axes[1,0], trace, 'pred_lrn', 'cadetblue')
+plot_predictive(axes[1,0], trace, 'pred_com', 'crimson')
+
+axes[1,0].set_ylim(0, 1.8)
+axes[1,1].set_ylim(0, 1.8)
+
[…]

### 2023-08-24 code/polymod.py

@@ -1,0 +1,67 @@
+import pandas as pd
+import numpy as np
+import pymc as pm
+import arviz as az
+
+
+def plot_predictive(axis, trace, var):
+	generation = np.arange(1, 10)
+	az.plot_hdi(generation, trace.posterior[var], hdi_prob=0.95, smooth=False, color='gray', ax=axis)
+	axis.plot(generation, trace.posterior[var].mean(('chain', 'draw')), color='black')
+	axis.set_ylim(0, 2)
+
+
+if __name__ == '__main__':
+
+	df = pd.read_csv('data/exp3.csv')
+
+	df_lrn = df[  (df['condition'].str.contains('^con_lrn'))  ]
+	cst_matrix_lrn = np.array(df_lrn['cost']).reshape((10, 10)).T
+	gen_matrix_lrn = np.repeat([range(10)], 10, axis=0).T
+	gen_matrix_lrn -= 5 # center generation
+
+	df_com = df[  (df['condition'].str.contains('^con_com'))  ]
+	cst_matrix_com = np.array(df_com['cost']).reshape((10, 10)).T
+	gen_matrix_com = np.repeat([range(10)], 10, axis=0).T
+	gen_matrix_com -= 5 # center generation
+
+	cst_matrix = np.stack((cst_matrix_lrn, cst_matrix_com), axis=2) + 0.0001
+	gen_matrix = np.stack((gen_matrix_lrn, gen_matrix_com), axis=2)
+
+	coords = {
+		'condition': ['lrn', 'com'],
+		'chain_id': list(range(10)),
+		'generation': list(range(10))
+	}
+
+	with pm.Model(coords=coords) as model:
+
+		b0_m = pm.Normal('b0_m', mu=0.1, sigma=0.5, dims='condition')
+		b1_m = pm.Normal('b1_m', mu=0.1, sigma=0.5, dims='condition')
+		# b2_m = pm.Normal('b2_m', mu=0.1, sigma=0.5, dims='condition')
+		# b3_m = pm.Normal('b3_m', mu=0.1, sigma=0.5, dims='condition')
+
+		b0_sd = pm.Exponential('b0_sd', lam=1, dims='condition')
+		b1_sd = pm.Exponential('b1_sd', lam=1, dims='condition')
[…]

### 2023-08-24 code/regmodel2.py

@@ -29,31 +29,57 @@ def plot_dataset(data1, data2):
 
 if __name__ == '__main__':
 
-	np.random.seed(117)
+	# np.random.seed(1177)
 
-	n_chains = 10
-	n_generations = 9
+	# n_chains = 10
+	# n_generations = 9
 
-	data1 = generate_random_dataset(n_chains, n_generations, 0.3, 0.3)
-	data2 = generate_random_dataset(n_chains, n_generations, 0.25, 0.3)
+	# data1 = generate_random_dataset(n_chains, n_generations, 0.3, 0.3)
+	# data2 = generate_random_dataset(n_chains, n_generations, 0.2, 0.3)
 
-	# plot_dataset(data1, data2)
-	# quit()
+	# # plot_dataset(data1, data2)
+	# # quit()
 
-	chain_ids = []
-	np.repeat
+	# chain_ids = []
+	# np.repeat
 
-	df = pd.DataFrame({
-		'cost': list(data1.flatten()) + list(data2.flatten()),
-		'condition': ['lrn'] * (n_chains * n_generations) + ['com'] * (n_chains * n_generations),
-		'generation': list(range(1, n_generations+1)) * (n_chains*2),
-		'chain': np.repeat(range(n_chains*2), n_generations),
-	})
+	# df = pd.DataFrame({
+	# 	'cost': list(data1.flatten()) + list(data2.flatten()),
+	# 	'condition': ['lrn'] * (n_chains * n_generations) + ['com'] * (n_chains * n_generations),
+	# 	'generation': list(range(1, n_generations+1)) * (n_chains*2),
+	# 	'chain': np.repeat(range(n_chains*2), n_generations),
+	# })
 
-	# df = pd.read_csv('../data/exp2.csv')
-	# df = df[ df['generation'] > 0 ]
+	df = pd.read_csv('data/exp3.csv')
+	df = df[ df['condition'].str.contains('^con') ]
+	df = df[ df['generation'] > 0 ]
+	df['cost'] = df['cost'] + 0.001
 	df = center_variable(df, 'generation')
 	df['chain_id'] = df[['condition', 'chain']].apply(lambda row: '_'.join(row.values.astype(str)), axis=1)
 
-	model = bmb.Model('cost ~ generation * condition + (1 + generation | chain_id)', df)
+	model = bmb.Model(
+		'cost ~ generation * epoch * condition + (1 + generation | chain_id)',
+		data=df,
[…]

### 2023-08-24 manuscript/wp4.tex
ms aug 24

@@
 
-\title{How does the writing system get informative? Differentiation or conservation?}
+\title{Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography}
 
-\shorttitle{Informativity in the writing system}
+\shorttitle{The cultural evolution of informative orthography}
 
@@
 
-\abstract{Abstract.}
+\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to acquire fully.
+It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading.
+For example, the heterographic homophones (e.g., <knight> vs.
+<night>) impose additional costs in learning but reduce ambiguity in reading.
+Could it be the case, then, that heterography---and therefore orthographic opacity---may sometimes be selected for in the evolution of writing systems, despite the cost in learning?
+We investigate this question by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity.
+In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises through exaptation as an epiphenomenon of sound change (e.g., <meat> vs.
+<meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift).
+Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, these systems can become informative, diverging from the spoken language in order to express meaning directly.}
 
@@
 
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
+Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
+In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6).
+Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable.
+In particular, there are two main kinds of color-expressive system that emerge.
+The first are the generalization-based expressive systems: K5, M1, P5, and S2.
+In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape.
+For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>.
+In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->).
+We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.
+
 	\begin{figure}
@@
 
-The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
-Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
-In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6).
-Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable.
-In particular, there are two main kinds of color-expressive system that emerge.
-The first are the generalization-based expressive systems: K5, M1, P5, and S2.
-In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape.
-For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>.
-In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->).
-We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.
-
 The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4.
@@
 
-Learnability, measured as transmission error (lower error = higher learnability).
-Mean Levenshtein edit distance between the words produced at generation i and the corresponding words produced at generation i-1.
+As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
+Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
+If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system.
+Communicative cost will take some larger value if the system contains ambiguity.
+Formally, it is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
+	% \begin{equation}
+	% \sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
+	% \end{equation}
+where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$.
+In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.
 
-Simplicity, measured as complexity (lower complexity = higher simplicity).
-The length (in bits) of the shortest grammar that losslessly compresses the lexicon, as computed by the Grammarette package: https://github.com/jwcarr/grammarette.
-Finding the shortest grammar for some lexicon is a non-trivial problem, and it is likely that we will need to adapt the method in response to the collected data.
+Fig.~\ref{cost} plots communicative cost over the nine generations of cultural transmission for both the Transmission-only condition (left) and Transmission + Communication condition (right).
+Each colored line represents an individual chain, and the thick black lines shows the mean across chains.
+The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling.
+In both conditions, cost increases with generation, implying that the languages become less informative over time.
+This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent.
+In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems, but it does align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produce differentiated spellings, be it through explicit innovation or implicit generalization.
+Note, however, that the cost plateaued at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication.
+This suggests that there is some tendency for the systems to remain more informative in the communicative condition.
 
-Communicative cost is given by
-\begin{equation}
-\mathrm{cost}(L) := \frac{1}{|U|} \sum_{m \in U} -\log \mathrm{Pr}(m|s_m),
-\end{equation}
-where $\mathrm{Pr}(m|s_m)$ is the probability that a listener would infer meaning $m$ given that a speaker produced signal $s$ for meaning $m$.
-Here we assume this probability to be $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$.
+To evaluate this argument statistically, we regressed cost on generation in a Bayesian hierarchical cubic regression model with a gamma likelihood and log link function.
+The posterior parameter estimates are shown in Fig.~\ref{stats1}.
+Essentially, the beta coefficients of the model (representing the shape of the polynomial) do not differ between conditions, but the intercepts do differ (generation was centered, such that the intercept corresponds to generation 5).
+This difference in intercepts reflects the fact that the cost plateaued at a lower level in the Transmission + Communication condition.
+Nevertheless, the posteriors for $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ do overlap to some extent, suggesting no definitive difference between conditions.
 
-To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 3×3 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings.
-For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
-\begin{equation}
-	S_\mathrm{degenerate} = \begin{bmatrix}
-	1 & 1 & 1 \\
-	1 & 1 & 1 \\
-	1 & 1 & 1
-	\end{bmatrix}
-\end{equation}
-Here the number 1 represents the one and only suffix spelling.
-We define an ``expressive'' system as one that marks the color dimension using three distinct spellings, resulting in the following matrix:
-\begin{equation}
-	S_\mathrm{expressive} = \begin{bmatrix}
-	1 & 2 & 3 \\
-	1 & 2 & 3 \\
-	1 & 2 & 3
-	\end{bmatrix}
-\end{equation}
-In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
-\begin{equation}
-	S_\mathrm{redundant} = \begin{bmatrix}
-	1 & 1 & 1 \\
-	2 & 2 & 2 \\
-	3 & 3 & 3
-	\end{bmatrix}
-\end{equation}
-Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant.
-Such a system contains unnecessary complexity---different ways of writing /kəʊ/ that carry no additional information.
+To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate, as illustrated in Fig.~\ref{}.
+These four typological categories form a continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning.
+The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
+To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content.
+To do this, we formalize the systems as set partitions (i.e., each system may be viewed as a partitioning of the universe of meanings into $n$ disjoint sets) and adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.}
 
-Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly.
-To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric.
-VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
-So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information.
-The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
+	\vspace*{2pt}
+	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+At generation~0, all 10 chains in both conditions use a holistic system (by design).
+In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning.
+In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way.
+Expressive systems were rare in both conditions.}
+	\label{typ_dist_dif}
+	\end{figure}
+
+Rather than simply classify each system into exactly one typological category, we computed probability distributions over the typological space according to distance.
+This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially redolent of one category but also redolent of another (for example, system T9 is part redundant and part degenerate).
+The probability distribution was calculated as $e^{-3\cdot\mathrm{VOI}(p, q)}$
+
+Fig.~\ref{typ_dist_dif} plots the typological distribution (i.e., across chains) for each generation and each condition.
 
@@
 No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking.
-In one system, <icoe> is used for gray objects and <ikoe> for all other colors.
-In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>.
-Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.
+In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.
 
@@
 As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose.
-A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).
+A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).
 
@@
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
-Spellings in bold-italic are the generalizations on unseen items.}
+Spellings in bold-italic are the generalizations on unseen items.
+Five chains (O, P, Q, R, T) remain fully expressive into the final epoch, in most cases conserving the original forms.}
 	\label{con_com}
@@
 
+As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost2}).
+Each colored line represents an individual chain, and the thick black lines shows the mean across chains.
+These mean lines reveal a fairly clear effect from epoch.
+During Epoch~I, there is a shallow increase in cost with generation in both conditions.
+This represents a kind of baseline level of noise when the spoken languages are fully informative.
+During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication.
+This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant.
+In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
+In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to simplify and the pressure to simplify has been ramped up.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_con.eps}}
+	\vspace*{2pt}
+	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+At generation~0, all 10 chains in both conditions use an expressive system (by design).
+In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission + Communication, they tend to remain expressive for longer and disfavor the redundant pattern.}
+	\label{typ_dist_con}
+	\end{figure}
+
 \subsection{Summary}
 
-In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
+In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, or six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
 

### 2023-08-29 manuscript/wp4.tex
ms 25/08

@@
 For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension.
-We describe such a system as ``redundant'' because shape is consistently expressed by the stem, so the suffix spellings that emerged in this case convey no additional information beyond what is already conveyed by the stem.
+We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information.
 Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition.
@@
 
+To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate.
+These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{}.
+The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
+To classify a participant's output into one of these four categories, we compute which of the reference systems is most similar in terms of its information content.
+To do this, we formalize the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
+Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content.
+The expressive and redundant systems are also considered very dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we compute a probability distribution over the typological space according to the respective distances of the four reference systems.
+This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate).
+The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
+\begin{equation}
+	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
+\end{equation}
+and averaged over chains for a given generation.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
+	\vspace*{2pt}
+	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+At Generation~0, all 10 chains in both conditions use a holistic system (by design).
+In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning.
+In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way.
+Expressive systems were rare in both conditions.}
+	\label{typ_dist_dif}
+	\end{figure}
+
+The results are plotted in Fig.~\ref{typ_dist_dif}.
+These probability distributions can be read as the proportion of chains that adopted each of the four configurations at each generation.
+At Generation~0, all chains are holistic by design.
+In Transmission-only, there is an initial rapid increase in the number of chains classified into the redundant typological category.
+The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy.
+In Transmission + Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common.
+There is, however, a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings.
+
 As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
@@
 Communicative cost will take some larger value if the system contains ambiguity.
-Formally, it is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
-	% \begin{equation}
-	% \sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
-	% \end{equation}
+Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
+	\begin{equation}
+	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
+	\end{equation}
 where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$.
@@
 Nevertheless, the posteriors for $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ do overlap to some extent, suggesting no definitive difference between conditions.
-
-To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate, as illustrated in Fig.~\ref{}.
-These four typological categories form a continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning.
-The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
-To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content.
-To do this, we formalize the systems as set partitions (i.e., each system may be viewed as a partitioning of the universe of meanings into $n$ disjoint sets) and adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.}
-
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
-	\vspace*{2pt}
-	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
-At generation~0, all 10 chains in both conditions use a holistic system (by design).
-In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning.
-In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way.
-Expressive systems were rare in both conditions.}
-	\label{typ_dist_dif}
-	\end{figure}
-
-Rather than simply classify each system into exactly one typological category, we computed probability distributions over the typological space according to distance.
-This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially redolent of one category but also redolent of another (for example, system T9 is part redundant and part degenerate).
-The probability distribution was calculated as $e^{-3\cdot\mathrm{VOI}(p, q)}$
-
-Fig.~\ref{typ_dist_dif} plots the typological distribution (i.e., across chains) for each generation and each condition.
-
-We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}).
-The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to.
-Most of the emergent system are either redundant or degenerate, with some straddling the border between the two.
-No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking.
-In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.
 

### 2023-08-29 code/exp1_analysis.py
Adjust the distance-similarity calculation

@@ -419,7 +419,11 @@ def make_typology_plot(exp_data, conditions, generations, output_path=None, prob
 					for ref_system in ref_systems
 				])
 				if probabilistic_classification:
-					unnorm_distribution = np.exp(-distances * 3) # 3 is arbitrary scaling constant
+					# -2 is weighting (free parameter), this value works well
+					# because it results in almost all prob mass in
+					# generation 0 being dedicated to the holistic category,
+					# which we know to be correct.
+					unnorm_distribution = np.exp(-2 * distances**2)
 					distribution += unnorm_distribution / unnorm_distribution.sum()
 				else:
 					classification = np.where(distances == distances.min())[0]
@@ -460,11 +464,11 @@ if __name__ == '__main__':
 
 
 	make_typology_plot(dataset_json, ['dif_lrn', 'dif_com'], list(range(10)),
-		output_path=ROOT / 'manuscript' / 'figs' / 'typ_dist_dif.pdf',
+		output_path=ROOT / 'manuscript' / 'figs' / 'typ_dist_dif.eps',
 		probabilistic_classification=True,
 	)
 	make_typology_plot(dataset_json, ['con_lrn', 'con_com'], list(range(10)),
-		output_path=ROOT / 'manuscript' / 'figs' / 'typ_dist_con.pdf',
+		output_path=ROOT / 'manuscript' / 'figs' / 'typ_dist_con.eps',
 		probabilistic_classification=True,
 	)
 

### 2023-08-30 code/comm_cost.py
Move analyses into separate figures script

@@ -1,0 +1,22 @@
+from collections import defaultdict
+from itertools import product
+import numpy as np
+
+
+def convert_lexicon_meanings_to_tuple(lexicon):
+	converted_lexicon = {}
+	for item, signal in lexicon.items():
+		meaning = tuple(map(int, item.split('_')))
+		converted_lexicon[meaning] = signal
+	return converted_lexicon
+
+
+def communicative_cost(lexicon, dims):
+	if isinstance(list(lexicon.keys())[0], str):
+		lexicon = convert_lexicon_meanings_to_tuple(lexicon)
+	reverse_lexicon = defaultdict(set)
+	for meaning, signal in lexicon.items():
+		reverse_lexicon[signal].add(meaning)
+	U = product(*[range(n_values) for n_values in dims])
+	U_size = np.product(dims)
+	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])

### 2023-08-30 code/exp1_analysis.py

@@ -365,83 +365,8 @@ def make_ternary_plot_by_generation(exp_data, conditions, generations, output_pa
 	plt.show()
 
 
-def make_panel_visualization(exp_data):
-	import visualize
 
-	A_to_J = ['A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J']
-	K_to_T = ['K', 'L', 'M', 'N', 'O', 'P', 'Q', 'R', 'S', 'T']
-	for condition, data in exp_data.items():
-		print(condition)
-		panel = []
-		for chain in data:
-			chain_matrices = []
-			ss, cp = matrix.generate_color_palette_many(chain)
-			for subject_a, _ in chain:
-				mat = matrix.make_matrix_with_cp(subject_a['lexicon'], ss, 3, 3)
-				sounds = []
-				if 'spoken_forms' in subject_a:
-					if 'con' in condition:
-						sounds = [get_sound(f'0_{i}', subject_a) for i in range(3)]
-					elif 'dif' in condition:
-						sounds = ['kəʊ', 'kəʊ', 'kəʊ']
-				if 'training_items' in subject_a:
-					training_items = subject_a['training_items']
-				else:
-					training_items = []
-				chain_matrices.append((mat, cp, ss, sounds, training_items))
-			panel.append(chain_matrices)
-		show_sounds = True if 'con' in condition else False
-		chain_ids = A_to_J if 'lrn' in condition else K_to_T
-		visualize.draw_panel(f'/Users/jon/Desktop/{condition}.eps', panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
 
-def make_typology_plot(exp_data, conditions, generations, output_path=None, probabilistic_classification=False):
-	import voi
-
-	ref_systems = [
-		matrix.reference_systems['holistic'],
-		matrix.reference_systems['expressive'],
-		matrix.reference_systems['redundant'],
-		matrix.reference_systems['transparent'],
-	]
-
-	distribution_colors = colormaps['viridis'](np.linspace(0, 1, 4))
-
-	fig, axes = plt.subplots(len(conditions), len(generations), figsize=(6, 2), squeeze=False, sharex=True, sharey=True)
-	
-	for i, condition in enumerate(conditions):
-		for j, generation in enumerate(generations):
-			distribution = np.zeros(len(ref_systems))
-			for chain in exp_data[condition]:
[…]

### 2023-08-30 code/figures.py

@@ -1,0 +1,275 @@
+from pathlib import Path
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import colormaps
+from utils import json_load
+
+
+plt.rcParams.update({'font.sans-serif': 'Helvetica Neue', 'font.size': 7})
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+DATA = ROOT / 'data'
+FIGS = ROOT / 'manuscript' / 'figs'
+
+
+def plot_transmission_chains(exp_data, condition, output_path):
+	import visualize
+	import matrix
+
+	cons = ['f', 's', 'ʃ']
+	vwls = ['əʊ', 'ə', 'ɛɪ']
+	def get_sound(item, data):
+		sound_file = data['spoken_forms'][item]
+		if '_' not in sound_file:
+			return 'kəʊ'
+		s, c, v = sound_file.split('.')[0].split('_')
+		return f'{cons[int(c)]}{vwls[int(v)]}'
+
+	A_to_J = ['A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J']
+	K_to_T = ['K', 'L', 'M', 'N', 'O', 'P', 'Q', 'R', 'S', 'T']
+	panel = []
+	for chain in exp_data[condition]:
+		chain_matrices = []
+		ss, cp = matrix.generate_color_palette_many(chain)
+		for subject_a, _ in chain:
+			mat = matrix.make_matrix_with_cp(subject_a['lexicon'], ss, 3, 3)
+			sounds = []
+			if 'spoken_forms' in subject_a:
+				if 'con' in condition:
+					sounds = [get_sound(f'0_{i}', subject_a) for i in range(3)]
+				elif 'dif' in condition:
+					sounds = ['kəʊ', 'kəʊ', 'kəʊ']
+			if 'training_items' in subject_a:
+				training_items = subject_a['training_items']
+			else:
+				training_items = []
+			chain_matrices.append((mat, cp, ss, sounds, training_items))
+		panel.append(chain_matrices)
+	show_sounds = True if 'con' in condition else False
+	chain_ids = A_to_J if 'lrn' in condition else K_to_T
+	visualize.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
+
+
+def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False):
[…]

### 2023-08-30 code/model_exp1.py

@@ -1,0 +1,66 @@
+from pathlib import Path
+import pandas as pd
+import numpy as np
+import pymc as pm
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+DATA = ROOT / 'data'
+
+
+if __name__ == '__main__':
+
+	n_chains = 10
+	generations = np.arange(1, 10)
+
+	df = pd.read_csv(DATA / 'exp3.csv')
+
+	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] > 0))  ]
+	cst_matrix_lrn = np.array(df_lrn['cost']).reshape((n_chains, -1)).T
+	gen_matrix_lrn = np.repeat([generations], n_chains, axis=0).T
+	gen_matrix_lrn -= 5 # center generation (gen 5 becomes intercept)
+
+	df_com = df[  (df['condition'].str.contains('^dif_com') & (df['generation'] > 0))  ]
+	cst_matrix_com = np.array(df_com['cost']).reshape((n_chains, -1)).T
+	gen_matrix_com = np.repeat([generations], n_chains, axis=0).T
+	gen_matrix_com -= 5 # center generation (gen 5 becomes intercept)
+
+	cst_matrix = np.stack((cst_matrix_lrn, cst_matrix_com), axis=2)
+	gen_matrix = np.stack((gen_matrix_lrn, gen_matrix_com), axis=2)
+
+	coords = {
+		'condition': ['lrn', 'com'],
+		'chain_id': list(range(n_chains)),
+		'generation': generations,
+	}
+
+	with pm.Model(coords=coords) as model:
+
+		α_m = pm.Normal('α_m', mu=0, sigma=2, dims='condition')
+		b1_m = pm.Normal('b1_m', mu=0, sigma=2, dims='condition')
+		b2_m = pm.Normal('b2_m', mu=0, sigma=2, dims='condition')
+		b3_m = pm.Normal('b3_m', mu=0, sigma=2, dims='condition')
+
+		α_sd = pm.Exponential('α_sd', lam=1, dims='condition')
+		b1_sd = pm.Exponential('b1_sd', lam=1, dims='condition')
+		b2_sd = pm.Exponential('b2_sd', lam=1, dims='condition')
[…]

### 2023-08-30 code/ploymod_plot.py

@@ -22,28 +22,31 @@ def plot_chains(axis, dataset, var, show_mean=False):
 
 
 def plot_predictive(axis, trace, var, color):
-	generation = np.arange(0, 10)
+	generation = np.arange(1, 10)
 	az.plot_hdi(generation, trace.posterior[var], hdi_prob=0.95, smooth=False, color=color, fill_kwargs={'alpha': 0.25, 'linewidth': 0}, ax=axis)
 	axis.plot(generation, trace.posterior[var].mean(('chain', 'draw')), color=color)
 	# axis.set_ylim(0, 1.8)
 
 
 df = pd.read_csv('../data/exp3.csv')
-trace = az.from_netcdf('../data/exp3_con.netcdf')
+trace = az.from_netcdf('../data/exp1.netcdf')
 
 fig, axes = plt.subplots(2, 2, figsize=(6, 4))
 
-plot_chains(axes[0,0], df[ df['condition'] == 'con_lrn' ], 'cost', show_mean=True)
-plot_chains(axes[0,1], df[ df['condition'] == 'con_com' ], 'cost', show_mean=True)
+plot_chains(axes[0,0], df[ df['condition'] == 'dif_lrn' ], 'cost', show_mean=True)
+plot_chains(axes[0,1], df[ df['condition'] == 'dif_com' ], 'cost', show_mean=True)
 
 axes[0,0].set_ylim(0, 1.8)
 axes[0,1].set_ylim(0, 1.8)
 
 plot_predictive(axes[1,0], trace, 'pred_lrn', 'cadetblue')
-plot_predictive(axes[1,0], trace, 'pred_com', 'crimson')
+plot_predictive(axes[1,1], trace, 'pred_com', 'crimson')
 
 axes[1,0].set_ylim(0, 1.8)
 axes[1,1].set_ylim(0, 1.8)
+
+axes[1,0].set_xlim(0, 9)
+axes[1,1].set_xlim(0, 9)
 
 axes[1,0].set_xlabel('Generation')
 axes[1,0].set_ylabel('Communicative cost (bits)')

### 2023-08-30 code/polymod.py

@@ -1,67 +1,0 @@
-import pandas as pd
-import numpy as np
-import pymc as pm
-import arviz as az
-
-
-def plot_predictive(axis, trace, var):
-	generation = np.arange(1, 10)
-	az.plot_hdi(generation, trace.posterior[var], hdi_prob=0.95, smooth=False, color='gray', ax=axis)
-	axis.plot(generation, trace.posterior[var].mean(('chain', 'draw')), color='black')
-	axis.set_ylim(0, 2)
-
-
-if __name__ == '__main__':
-
-	df = pd.read_csv('data/exp3.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^con_lrn'))  ]
-	cst_matrix_lrn = np.array(df_lrn['cost']).reshape((10, 10)).T
-	gen_matrix_lrn = np.repeat([range(10)], 10, axis=0).T
-	gen_matrix_lrn -= 5 # center generation
-
-	df_com = df[  (df['condition'].str.contains('^con_com'))  ]
-	cst_matrix_com = np.array(df_com['cost']).reshape((10, 10)).T
-	gen_matrix_com = np.repeat([range(10)], 10, axis=0).T
-	gen_matrix_com -= 5 # center generation
-
-	cst_matrix = np.stack((cst_matrix_lrn, cst_matrix_com), axis=2) + 0.0001
-	gen_matrix = np.stack((gen_matrix_lrn, gen_matrix_com), axis=2)
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'chain_id': list(range(10)),
-		'generation': list(range(10))
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		b0_m = pm.Normal('b0_m', mu=0.1, sigma=0.5, dims='condition')
-		b1_m = pm.Normal('b1_m', mu=0.1, sigma=0.5, dims='condition')
-		# b2_m = pm.Normal('b2_m', mu=0.1, sigma=0.5, dims='condition')
-		# b3_m = pm.Normal('b3_m', mu=0.1, sigma=0.5, dims='condition')
-
-		b0_sd = pm.Exponential('b0_sd', lam=1, dims='condition')
-		b1_sd = pm.Exponential('b1_sd', lam=1, dims='condition')
[…]

### 2023-08-30 code/visualize.py

@@ -101,13 +101,14 @@ def draw_panel(output_path, chains, figure_width=1000, chain_ids=None, show_gene
 	n_cols = len(chains[0])
 	figure_height = figure_width / n_cols * n_rows
 	cell_size = figure_width / n_cols
-	if output_path.endswith('.eps'):
-		surface = cairo.PSSurface(output_path, figure_width, figure_height)
+	filename = output_path.name
+	if filename.endswith('.eps'):
+		surface = cairo.PSSurface(str(output_path), figure_width, figure_height)
 		surface.set_eps(True)
-	elif output_path.endswith('.pdf'):
-		surface = cairo.PDFSurface(output_path, figure_width, figure_height)
-	elif output_path.endswith('.svg'):
-		surface = cairo.SVGSurface(output_path, figure_width, figure_height)
+	elif filename.endswith('.pdf'):
+		surface = cairo.PDFSurface(str(output_path), figure_width, figure_height)
+	elif filename.endswith('.svg'):
+		surface = cairo.SVGSurface(str(output_path), figure_width, figure_height)
 	else:
 		raise ValueError('Unrecognized format')
 	context = cairo.Context(surface)

### 2023-08-30 manuscript/wp4.tex
ms 30/08

@@
 
-Fig.~\ref{cost} plots communicative cost over the nine generations of cultural transmission for both the Transmission-only condition (left) and Transmission + Communication condition (right).
-Each colored line represents an individual chain, and the thick black lines shows the mean across chains.
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
+	\vspace*{2pt}
+	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
+The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
+The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI).
+Generation~0 is not included in the model, since it does not represent participant data.}
+	\label{cost_dif}
+	\end{figure}
+
+Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition.
+In the upper panels, each thin line represents an individual transmission chain, and the thick lines shows the mean across chains.
 The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling.
@@
 This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent.
-In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems, but it does align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produce differentiated spellings, be it through explicit innovation or implicit generalization.
-Note, however, that the cost plateaued at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication.
-This suggests that there is some tendency for the systems to remain more informative in the communicative condition.
+In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems.
+This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization.
+Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.
 
-To evaluate this argument statistically, we regressed cost on generation in a Bayesian hierarchical cubic regression model with a gamma likelihood and log link function.
-The posterior parameter estimates are shown in Fig.~\ref{stats1}.
-Essentially, the beta coefficients of the model (representing the shape of the polynomial) do not differ between conditions, but the intercepts do differ (generation was centered, such that the intercept corresponds to generation 5).
-This difference in intercepts reflects the fact that the cost plateaued at a lower level in the Transmission + Communication condition.
-Nevertheless, the posteriors for $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ do overlap to some extent, suggesting no definitive difference between conditions.
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_dif.eps}}
+	\vspace*{2pt}
+	\caption{Posterior parameter estimates for the Experiment~1 polynomial regression model.
+Blue curves show results for the Transmission-only condition and red curves show results for the Transmission + Communication condition.
+$\alpha$ represents the intercept of the polynomial (Generation~5) and the $\beta$ parameters characterize its shape.}
+	\label{posterior_dif}
+	\end{figure}
+
+To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model.
+The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}.
+Essentially, the $\beta$ parameters of the model (characterizing the shape of the polynomial) do not differ substantially between conditions, but the intercepts ($\alpha$) do appear to differ.
+Since generation was centered (for the purpose of fitting the model), the $\alpha$ parameters correspond to Generation~5 and therefore represent the main plateaus of the polynomials.
+The posterior difference between $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ was 0.301~bits (95\%~HDI: 0.014--0.597).
+This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus.
+We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage for systems that emerged in the communicative condition (i.e., systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%).
+Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would support a 100\% success rate.
 
@@
 
-As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost2}).
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_con.eps}}
+	\vspace*{2pt}
+	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+At generation~0, all 10 chains in both conditions use an expressive system (by design).
+In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission + Communication, they tend to remain expressive for longer and disfavor the redundant pattern.}
+	\label{typ_dist_con}
+	\end{figure}
+
+As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost_con}).
 Each colored line represents an individual chain, and the thick black lines shows the mean across chains.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_con.eps}}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
 	\vspace*{2pt}
-	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
-At generation~0, all 10 chains in both conditions use an expressive system (by design).
-In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission + Communication, they tend to remain expressive for longer and disfavor the redundant pattern.}
-	\label{typ_dist_con}
+	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
+The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
+The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI).
+Generation~0 is not included in the model, since it does not represent participant data.}
+	\label{cost_con}
 	\end{figure}

### 2023-09-06 code/figures.py
Analysis changes

@@ -145,7 +145,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 			import arviz as az
 			trace = az.from_netcdf(model_trace)
 			pred_var = f'pred_{condition.split("_")[1]}'
-			generations = np.arange(1, 10)
+			generations = np.arange(0, 10)
 			az.plot_hdi(generations, trace.posterior[pred_var], ax=axes[1,i], hdi_prob=0.95, smooth=False, color=color_light, fill_kwargs={'alpha': 1, 'linewidth': 0})
 			axes[1,i].plot(generations, trace.posterior[pred_var].mean(('chain', 'draw')), color=color)
 
@@ -254,15 +254,6 @@ if __name__ == '__main__':
 	# 	model_trace=DATA / 'exp1_cube.netcdf',
 	# )
 
-	plot_communicative_cost(exp_data,
-		conditions=['con_lrn', 'con_com'],
-		output_path=FIGS / 'cost_con.eps',
-		figsize=(6, 4),
-		show_mean=True,
-		add_jitter=True,
-		# show_model=DATA / 'exp1.netcdf',
-	)
-
 	# plot_posterior(DATA / 'exp1_cube.netcdf',
 	# 	variables=[
 	# 		{'var': 'α_m', 'label': '$α$'},
@@ -273,3 +264,23 @@ if __name__ == '__main__':
 	# 	output_path=FIGS / 'posterior_dif.eps',
 	# 	show_summary=True,
 	# )
+
+	plot_communicative_cost(exp_data,
+		conditions=['con_lrn', 'con_com'],
+		output_path=FIGS / 'cost_con.pdf',
+		figsize=(6, 4),
+		show_mean=True,
+		add_jitter=True,
+		model_trace=DATA / 'exp2.netcdf',
+	)
+
+	plot_posterior(DATA / 'exp2.netcdf',
+		variables=[
+			{'var': 'α_m', 'label': '$α$'},
+			{'var': 'β1_m', 'label': '$β_1$'},
+			{'var': 'β2_m', 'label': '$β_2$'},
+			{'var': 'β3_m', 'label': '$β_3$'},
+		],
+		output_path=FIGS / 'posterior_con.pdf',
+		show_summary=True,
+	)

### 2023-09-06 code/matrix.py

@@ -93,18 +93,18 @@ def generate_color_palette_many(chain):
 	return suffix_spellings, {suffix_spellings.index(suffix): hsv_to_rgb(hue, 0.7, 0.9) for suffix, hue in zip(suffix_spellings, hues)}
 
 
-def draw(matrix, color_palette, output_path):
+def draw(matrix, color_palette, output_path, square_size_pts=10):
 	m, n = matrix.shape
-	surface = cairo.PDFSurface(output_path, m, n)
+	surface = cairo.PDFSurface(output_path, m*square_size_pts, n*square_size_pts)
 	context = cairo.Context(surface)
 	for i, row in enumerate(matrix):
 		for j, cell in enumerate(row):
 			color = color_palette[cell]
 			context.set_source_rgb(*color)
-			context.rectangle(j, i, 1, 1)
+			context.rectangle(j*square_size_pts, i*square_size_pts, square_size_pts, square_size_pts)
 			context.fill_preserve()
 			context.set_source_rgb(1, 1, 1)
-			context.set_line_width(0.05)
+			context.set_line_width(1)
 			context.stroke()
 	surface.finish()
 
@@ -135,6 +135,12 @@ reference_systems = {
 
 if __name__ == '__main__':
 
+	from matplotlib import colormaps
+
 	for name, system in reference_systems.items():
-		color_palette = generate_color_palette(system)
+
+		color_palette = colormaps['viridis'](np.linspace(0, 1, len(np.unique(system))))
+		color_palette = [color[:3] for color in color_palette]
+		np.random.shuffle(color_palette)
+
 		draw(system, color_palette, f'/Users/jon/Desktop/{name}.pdf')

### 2023-09-06 code/model_exp1.py

@@ -61,6 +61,6 @@ if __name__ == '__main__':
 		pm.Deterministic('pred_com', α_m[1] + b1_m[1] * gen_matrix[:, 0, 0] + b2_m[1] * gen_matrix[:, 0, 0] ** 2 + b3_m[1] * gen_matrix[:, 0, 0] ** 3)
 		pm.Deterministic('α_diff', α_m[0] - α_m[1])
 
-		trace = pm.sample(1000, tune=2000, target_accept=0.9)
+		trace = pm.sample(2000, tune=2000, target_accept=0.95, chains=6, cores=6)
 
 		trace.to_netcdf(DATA / 'exp1_cube.netcdf')

### 2023-09-06 code/model_exp2.py

@@ -1,0 +1,73 @@
+from pathlib import Path
+import pandas as pd
+import numpy as np
+import pymc as pm
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+DATA = ROOT / 'data'
+
+
+if __name__ == '__main__':
+
+	n_chains = 10
+	# generations = np.arange(1, 10)
+
+	e1 = np.array([0, 1, 2, 3, 4, 5, 6, 7, 8, 9])
+	e2 = np.array([0, 0, 0, 0, 1, 2, 3, 4, 5, 6])
+	e3 = np.array([0, 0, 0, 0, 0, 0, 0, 1, 2, 3])
+
+	df = pd.read_csv(DATA / 'exp3.csv')
+
+	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] > -1))  ]
+	cst_lrn = np.array(df_lrn['cost']).reshape((n_chains, -1)).T
+
+	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] > -1))  ]
+	cst_com = np.array(df_com['cost']).reshape((n_chains, -1)).T
+
+	cst = np.stack((cst_lrn, cst_com), axis=2)
+
+	e1 = np.repeat([e1], n_chains, axis=0).T
+	e1 = np.stack((e1, e1), axis=2)
+
+	e2 = np.repeat([e2], n_chains, axis=0).T
+	e2 = np.stack((e2, e2), axis=2)
+
+	e3 = np.repeat([e3], n_chains, axis=0).T
+	e3 = np.stack((e3, e3), axis=2)
+
+	coords = {
+		'condition': ['lrn', 'com'],
+		'chain_id': list(range(n_chains)),
+		'generation': list(range(10)),
+	}
+
+	with pm.Model(coords=coords) as model:
+
+		α_m = pm.Normal('α_m', mu=0, sigma=1, dims='condition')
+		β1_m = pm.Normal('β1_m', mu=0, sigma=1, dims='condition')
+		β2_m = pm.Normal('β2_m', mu=0, sigma=1, dims='condition')
+		β3_m = pm.Normal('β3_m', mu=0, sigma=1, dims='condition')
+
+		α_sd = pm.Exponential('α_sd', lam=1, dims='condition')
+		β1_sd = pm.Exponential('β1_sd', lam=1, dims='condition')
[…]

### 2023-09-06 manuscript/wp4.tex
ms 31/08

@@
 spacing between words helps readers parse a sentence even if they are not technically required.
-Written languages can also have there own regularities such as spelling related words similarly (magician), past tense spellings directly conveying meaning, certain spellings conveying word class ous.
+Written languages can also have their own regularities such as spelling related words similarly (magician), past tense spellings directly conveying meaning, certain spellings conveying word class ous.
 These features abound in writing, especially in languages such as English that have not been subject to serious regulation.
 
-One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings.
-For example, <meat> and <meet> are both pronounced /miːt/.
-Furthermore, these spellings can be cognitively useful .
+One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/).
+Heterographic spellings such as there can be useful in reading.
 For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
-In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, giving the reader a headstart on parsing the upcoming syntax.
-This idiosyncratic spelling of words that are homophonous in speech may confer certain benefits in reading, particularly in terms of comprehension.
-Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart.
+In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax.
+Heterography may, therefore, confer certain benefits in reading, particularly in terms of comprehension.
+Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart, avoiding ambiguity.
 
-Other languages
+Examples from other languages Irish orthography has special features to handle eclipsis.
+For example, when the word /pɑʃti/ (children) is eclipsed, it becomes /bɑʃti/ (the initial consonant is vocalized), but the spelling system retains the letter <p> \textit{in addition to} adding a <b>: <páistí> becomes <bpáistí> under eclipsis.
+This allows the reader quick access to the root word from which the eclipsed form is derived, while also showing the reader how to pronounce the eclipsed form.
+Moreover, in some cases, it results in ambiguity avoidance---in this case, if the <p> were not retained, the spelling <báistí> would be confusable with the word for rain.
 
-Irish orthography has special features to handle lenition and eclipsis.
-For example, when the word /pɑʃti/ (children) is eclipsed, it becomes /bɑʃti/ (the initial consonant is vocalized), but the spelling system retains the letter <p> in addition to adding a <b>: <páistí> becomes <bpáistí> under eclipsis.
-This allows the reader quick access to the root word from which the eclipsed form is derived and, in some cases, results in ambiguity avoidance (in this case, if the <p> were not retained, the spelling <báistí> would be confusible with the word for rain).
-
-However, this additional informativeness comes at a cost.
+However, this all sounds great but additional informativeness comes at a cost.
 A system of words that conveys more information will---in general---be more complex than one that conveys less information.
@@
 A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
-The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Carr:2020, Kirby:2015} studies.
+The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2015} studies.
 Of particular note here is the idea that the pressure for simplicity in language structure derives principally from learning, while the pressure for informativeness derives from the need for precise communication.
+
+Unpack some of this literature a bit.
 
@@
 Is it possible, then, that heterography may be selected for in the cultural evolution of orthography?
-And if so, what might be the mechanism through which such selection could operate?
+And if so, what might the mechanism be through which such selection could operate?
 \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
@@
 
-Processes
+In this paper, we simulate the processes of differentiation and conservation experimentally.
 
-Sound change
-dialect merger (bury is the spelling from one dialect with the pronunciation from another dialect)
-Borrowing (English tends to retain spelling from the source language)
-Printers innovations
-- Etymological spellings (doubt)
-- Deliberate design of heterographs
-- Morphological spellings
-
-\subsection{Research questions}
-
-Q1) Does a learning pressure (induced through cultural transmission) result in the emergence of a simple, systematic, transparent orthography that is easier to learn?
-Hypotheses: In the high-variation conditions, where the orthography starts out chaotic (resulting in holistic systems), we expect simpler systems to emerge.
-Simplicity may be realized in a number of ways.
-Transparent systems use a one-to-one mapping to directly represent the spoken forms in writing.
-Redundant systems retain some spelling variation (with idiosyncratic spellings of the suffix for each shape) but become simpler by collapsing across colour.
-Expressive systems continue to express colour, but do so with a non-transparent but easy to learn compositional suffix system.
-The prevalence of these three types of system will be influenced by the presence of the communicative pressure.
-When communication is not present, transparent systems will tend to dominate, alongside some redundant systems and possibly some expressive systems.
-When communication is present, expressive systems will tend to dominate, perhaps with some redundant and holistic forms.
-The gradual emergence of simplicity, whatever form it takes, should be accompanied by increased learnability.
-
-Q2) Can a writing system evolve to become more informative than its uninformative spoken counterpart?
-Hypotheses: Yes, but preferentially when two vital ingredients are present: a communicative pressure for informativeness (which makes an expressive system desirable) and high variation (which provides the building blocks to make that happen).
-When both ingredients are present, expressive systems will tend to dominate.
-If only one of the ingredients is available (either communication or high variation), informative systems may emerge but are less likely.
-In the case of learning-only and high-variation, expressive systems may emerge if the learners – yearning for an explanation for the variation – assume that colour must somehow be involved and therefore impose (perhaps unconsciously) colour-expressive structure on the language (i.e.
-learners may, to a certain extent, have a preconceived notion that languages ought to be informative).
-In the case of communication and no variation (where, again, only one of the ingredients is met), expressive systems may emerge but this is highly dependent on participants being 'innovators' who are willing to develop novel spelling systems that diverge from the spoken system in order to express colour.
-If neither ingredient is present, expressive systems will not emerge.
+Preregistration: \url{https://aspredicted.org/YLG_CKD}
 
@@
 Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography. In particular we had two main hypotheses:
-
-Preregistered \url{https://aspredicted.org/YLG_CKD}
 
@@
 
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/typology.eps}}
+	\vspace*{2pt}
+	\caption{Four primary systems of interest (or typological categories) arranged along the cost/complexity continuum.
+A holistic system uses a unique suffix spelling for each shape--color combination.
+An expressive system only expresses color.
+A redundant system only expresses shape (which is already conveyed by the stem).
+A degenerate system expresses nothing.}
+	\label{typology}
+	\end{figure}
+
 To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate.
-These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{}.
+These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}.
 The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
@@
 
-As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
+As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
 Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
@@
 This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus.
-We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage for systems that emerged in the communicative condition (i.e., systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%).
-Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would support a 100\% success rate.
+We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage (in terms of communicative accuracy) for systems that emerged in the communicative condition.
+Specifically, systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%.
+Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.
 
@@
 Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation.
-Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography.
-But this did not happen, even under communicative pressure.
+Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or arbitrary silent letters.
+However, such differentiation was not very forthcoming, even under communicative pressure.
 Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings.
 This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.
+
 So, what was different about the present experiment?
 The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color.
-Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
+Faced with variation that could be conditioned on either dimension, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
 As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose.
@@
 
-Another pathway through which differentiation could have occurred is through conscious innovation, especially in the communicative conditions, where participants could have increased their bonuses substantially by innovating a system of color markers.
-However, there was no evidence of such innovation, and only four participants mentioned that they even \textit{considered} this possibility in their post-test comments (e.g., ``I wondered if I should use the color suffix rule even where I did not think it was the right word to try and help my partner get the right color object'').
-Why do participants choose not to innovate or even regularize an imperfect system?
-Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that we expected them to use the orthography learned during training.
-Participants who adopted this interpretation may have feared diverging from their training either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
-Secondly, even if participants are willing to diverge, they are presented with an alignment problem: How can they decide what system of color marking to use when there is no channel through which they can have a meta-discussion?
+It is particularly surprising that participants chose not to innovate novel systems in communication, despite their being a clear financial incentive to do so.
+Why might this be the case?
+Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
+Participants who adopted this interpretation may have feared diverging from their input, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
+Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
 They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.
-Interestingly, these are dynamics that also exist in the real world, where deviation from received, sacrosanct spellings is considered socially unacceptable or uneducated (e.g., the non-standard spelling <thru> has struggled for decades---or perhaps centuries---to make any headway against <through>), and even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If I decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take my spelling seriously or even understand the distinction I was trying to make?
 
-Overall, our first experiment suggests that differentiation---be it through a self-organizing process in learning or through a conscious process of innovation in use---is not a good model of heterography, unless enforced by top-down authority.
-We now turn to consider the conservation model.
+These are dynamics that also exist in the real world, especially in writing where the reader and writing are separated in time and space.
+Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>).
+And even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If you decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take your spelling seriously or even understand the distinction?
+Differentiation is hard in the written modality.
 
@@
 We now turn our attention to the conservation model of heterographic homophones: Given that an informative system already exists (both in speech and in writing), does that informative system persist in writing even after the spoken language has degenerated into homophony?
-And, importantly, does this happen preferentially in the presence of a communicative pressure?
+And, importantly, does this happen preferentially in the presence of communicative pressure?
 
@@
 As in Experiment~1, each transmission chain was run for nine generations.
-Additionally, the nine generations were divided into three epochs, each lasting three generations.
+These nine generations were divided into three epochs, each lasting three generations.
 During Epoch~I (Generations~1 to~3), the three spoken suffixes were distinct (as described above), allowing the spoken language to express all three colors without ambiguity.
@@
 
-As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost_con}).
-Each colored line represents an individual chain, and the thick black lines shows the mean across chains.
+Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition.
+At Generation~0, all systems are expressive (by design), but the dominance of this typological category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
+Notably, however, the loss of expressive systems appears to be slower in Transmission + Communication, and redundant systems were also less popular in Transmission + Communication.
+Overall, there is a general trend toward degeneracy---or, in more positive terms, a trend toward transparently mirroring the spoken language---in both conditions, but this degeneration occurs more slowly under communicative pressure.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
+	\vspace*{2pt}
+	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
+The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
+The lower panels show a regression model of the data (posterior mean and 95\% HDI).
+Generation~0 is not included in the model, since it does not represent participant data.}
+	\label{cost_con}
+	\end{figure}
+
+Fig.~\ref{cost_con} plots communicative cost.
 These mean lines reveal a fairly clear effect from epoch.
@@
 In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
-In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to simplify and the pressure to simplify has been ramped up.
+In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.
 
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_con.eps}}
 	\vspace*{2pt}
-	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
-The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
-The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI).
-Generation~0 is not included in the model, since it does not represent participant data.}
-	\label{cost_con}
+	\caption{Posterior parameter estimates for the Experiment~2 regression model.
+Blue curves show results for the Transmission-only condition and red curves show results for the Transmission + Communication condition.}
+	\label{posterior_con}
 	\end{figure}
+
+[N.B., I haven't completed this bit yet] To evaluate statistically the effect of generation (modulated by epoch) on cost, we fit the following Bayesian, hierarchical regression model
+    \begin{align*}
+     c_{i} & \sim \mathrm{Normal}(\mu, \sigma) \\
+       \mu & = \alpha + \beta_1 e_1 + \beta_2 e_2 + \beta_3 e_3 \\
+    \sigma & \sim \mathrm{Exponential}(1)
+    \end{align*}
+with the intercept and slope parameters fit by-chain.
+Essentially, $\beta_1$ represents the baseline effect of generation on cost, $\beta_2$ represents the additional effect during Epoch~II, and $\beta_3$ represents, again, the additional effect during Epoch~III.
+The results are shown in Fig.~\ref{posterior_con}.
+During the first epoch, the two conditions are identical.
+During the second epoch there's some difference.
+During the third epoch, there's a difference because Transmission-only plateaus and Communication continues.
+Are these slopes interpretable in any useful way?
 
@@
 
-In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, or six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
+Experiment~1 asked whether informative orthography can be \textit{created} under communicative pressure, and the evidence suggested that differentiation is a hard problem to solve in the written modality.
+In Experiment~2, by contrast, we asked whether informativeness can simply be \textit{maintained}, and preferentially so under communicative pressure.
+In the Transmission-only condition, only one chain remained expressive into Epoch~III, while in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, albeit not necessarily all the way to Generation~9.
+The fact that expressive spellings persisted longer and across more chains under communicative pressure suggests that an informative orthography---despite running contrary to the spoken language---may be maintained when it serves a useful purpose.
+
+Nevertheless, many of the chains did eschew informativeness in favor of greater transparency, and the inevitable consequence for all chains appears to be degeneracy in the long term.
+This is to be expected under Transmission-only, where the systems are adapting under a learnability pressure, but is somewhat surprising under the communicative condition.
+Participants were quite resistant to encoding into writing something that is not encoded in speech, even though they were---at the same time---content to conserve unusual suffix forms.
+This perhaps points to a general bias that reader/writers might possess in which the primary goal of writing is seen as representing speech---not representing meaning directly.
 
@@
 \section{Discussion}
+
+\begin{itemize}
+
+	\item Summary of theoretical idea... hard to test in the real world, but we can experimentally simulate the cultural evolution of spelling
+
+	\item Recapitulation of main findings
+
+	\item What can this tell us about real-world processes?
+
+	\item Cannot directly compare experiments because it's easier to maintain than create.
+However, this is also true of the real world, pointing to a general advantage of conservation over differentiation.
+
+	\item Limitations
+
+	\item Future directions. Remove the spoken forms from the test phase to simulate a silent-reader, literate society. This might make a dramatic difference to the results.
+
+	\item Conclusion: perhaps there are potentially some benefits to heterography.
+
+\end{itemize}
 

### 2023-09-07 manuscript/wp4.tex
ms 7/09, taking in Kathy's edits

@@
 \setmainfont{Times New Roman}
-% \setsansfont{Helvetica Neue}
 
@@
 We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2).
-14 participants were excluded because they (or their partners) used English color words (8) or failed the auditory attention checks (6).
+14 participants were excluded because they (or their communication partners) used English color words (8) or failed the auditory attention checks (6).
 A further three participants were lost to communication-game pairing failures.
-The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).
+The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 participants) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of participants).
 
@@
 	\vspace*{2pt}
-	\caption{The nine alien object stimuli with their iconic stems (indicating shape) and suffixes (indicating color).
-The spoken forms of the suffixes are homophonous, but their associated spellings are free to evolve and may therefore take on differentiated forms.}
+	\caption{The nine alien object stimuli with their stems and suffixes.
+The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
 	\label{stimuli}
@@
 
-Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system.
-This was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>.
+Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system, which was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>.
 In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity.
-This models an initial state of high spelling variation (every object has a unique suffix spelling), which may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning).
+This procedure models an initial state of high spelling variation (every object has a unique suffix spelling), but these spellings may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning).
 The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).
@@
 
-As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A).
-The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted.
-Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color).
+Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A).
+The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations.
+Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).
 Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items.
 Transmission continued for nine generations in each of ten independent chains.
-In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend).
+In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant; determined by whichever participant started the experiment first) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend).
 The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.
@@
 In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each.
-In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type in the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text).
+In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text).
 The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s.
@@
 After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials.
-In production trials, the participant was shown an object and heard its associated word pronounced aloud.
-The participant's task was to type~in the appropriate spelling.
+In production trials, the participant was shown an object and heard its associated pronunciation.
+The participant's task was to type the appropriate spelling.
 The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial.
@@
 
-Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items).
+Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit subject to a separate application of the bottleneck on transmission; i.e., the seen items were selected independently for each participant).
 The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction.
@@
 
-The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}.
-Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}.
+The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}.
+Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, just as in Fig.~\ref{stimuli}.
 For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension.
@@
 The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
-To classify a participant's output into one of these four categories, we compute which of the reference systems is most similar in terms of its information content.
-To do this, we formalize the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
+To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content.
+To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
 Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content.
-The expressive and redundant systems are also considered very dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we compute a probability distribution over the typological space according to the respective distances of the four reference systems.
-This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate).
+The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we computed a probability distribution over the typological space according to the respective distances from the four reference systems.
+This approach gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate).
 The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
@@
 
-As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
-Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
+As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}, which expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
 If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system.
@@
 Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition.
-In the upper panels, each thin line represents an individual transmission chain, and the thick lines shows the mean across chains.
+In the upper panels, each thin line represents an individual transmission chain, and the thick lines show the mean across chains.
 The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling.
@@
 To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model. 
+    \begin{align*}
+       c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+     \mu_i & =    \alpha_{\mathrm{CHAIN}[i]} + \beta_{\mathrm{CHAIN}[i]} g_i + \gamma_{\mathrm{CHAIN}[i]} g_i^2 + \delta_{\mathrm{CHAIN}[i]} g_i^3 \\
+    \sigma & \sim \mathrm{Exponential}(1)
+    \end{align*}
 The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}.
@@
 Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation.
-Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or arbitrary silent letters.
+Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue).
 However, such differentiation was not very forthcoming, even under communicative pressure.
@@
 Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
-Participants who adopted this interpretation may have feared diverging from their input, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
+Participants who adopted this interpretation may have feared diverging from what they learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
 Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
@@
 
-These are dynamics that also exist in the real world, especially in writing where the reader and writing are separated in time and space.
+These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space.
 Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>).
-And even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If you decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take your spelling seriously or even understand the distinction?
+Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the distinction?
 Differentiation is hard in the written modality.
@@
 A further 10 participants were lost to communication-game pairing failures.
-Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).
+Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 participants) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of participants).
 
@@
 The alien objects and word stems were identical to Experiment~1.
-Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors.
+Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that used three distinct suffixes to systematically express each of the colors.
 A separate set of suffixes was created for each chain by concatenating a randomly drawn consonant from \{/f/, /s/, /ʃ/\} and a randomly drawn vowel from \{/ə/, /ɛɪ/, /əʊ/\}, both without replacement.
-For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/.
+For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/ to represent the colors pink, yellow, and blue, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/ for those colors.
 The initial orthographic system was transparent and based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}.
@@
 
-As in Experiment~1, each transmission chain was run for nine generations.
-These nine generations were divided into three epochs, each lasting three generations.
+Each transmission chain was run for nine generations, which were divided into three epochs.
 During Epoch~I (Generations~1 to~3), the three spoken suffixes were distinct (as described above), allowing the spoken language to express all three colors without ambiguity.
@@
 In the second sound change, the remaining two suffix forms were merged in the same way, resulting in full homophony.
-Crucially, the spellings do not automatically change following a sound change event; rather, the orthographic system is free to adapt (or not) in response to the sound changes.
-Note also that an individual participant does not directly experience the sound changes; rather, a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
+Crucially, the spellings did not automatically change following a sound change event; rather, the orthographic system was free to adapt (or not) in response to the sound changes.
+Note also that an individual participant did not directly experience the sound changes; rather, a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
 In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.
@@
 
-The results for all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.
+The results for all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.
 It is immediately clear that color-expressive suffixes (as indicated by vertical stripes) are maintained fairly reliably through the first epoch; perfectly in the case of Chains~A, B, D, F, and~I, and with some errors in the other five chains.
@@
 Like Experiment~1, redundant systems are characteristic of the Transmission-only condition, especially in Epoch~III.
-As the spoken suffixes becomes more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings.
+As the spoken suffixes become more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings.
 Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost.
@@
 Fig.~\ref{cost_con} plots communicative cost.
-These mean lines reveal a fairly clear effect from epoch.
+The mean lines reveal a fairly clear effect of epoch.
 During Epoch~I, there is a shallow increase in cost with generation in both conditions.
-This represents a kind of baseline level of noise when the spoken languages are fully informative.
-During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication.
-This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant.
+This represents a baseline level of noise when the spoken languages are fully informative.
+During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in the Transmission + Communication condition.
+This difference is presumably due to the communicative pressure, which acts to slow down the rate at which the systems become transparent or redundant.
 In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
@@
     \end{align*}
+
+    \begin{align*}
+       c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+     \mu_i & =    \alpha_{\mathrm{CHAIN}[i]} + \beta_{\mathrm{CHAIN}[i]} g_i \\
+    \sigma & \sim \mathrm{Exponential}(1)
+    \end{align*}
+
 with the intercept and slope parameters fit by-chain.
@@
 
-Experiment~1 asked whether informative orthography can be \textit{created} under communicative pressure, and the evidence suggested that differentiation is a hard problem to solve in the written modality.
+Experiment~1 asked whether informative orthography can be \textit{created} under communicative pressure, and the evidence suggested that differentiation is hard to achieve in the written modality.
 In Experiment~2, by contrast, we asked whether informativeness can simply be \textit{maintained}, and preferentially so under communicative pressure.
@@
 Participants were quite resistant to encoding into writing something that is not encoded in speech, even though they were---at the same time---content to conserve unusual suffix forms.
-This perhaps points to a general bias that reader/writers might possess in which the primary goal of writing is seen as representing speech---not representing meaning directly.
+This perhaps points to a general preference that readers might possess (at least, readers who are accustomed to an alphabetic writing system) in which the primary goal of writing is seen as representing speech---not representing meaning directly.
 

### 2023-09-07 code/figures.py
Analysis changes

@@ -104,11 +104,13 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 	else:
 		plt.show()
 
+color_palette = colormaps['viridis'](np.linspace(0, 1, 5))
+color_palette = [color[:3] for color in color_palette]
 COLORS = {
-	'dif_lrn': ('cadetblue', '#AFD0D0'),
-	'dif_com': ('crimson', '#F58CA1'),
-	'con_lrn': ('cadetblue', '#AFD0D0'),
-	'con_com': ('crimson', '#F58CA1'),
+	'dif_lrn': ('darkorange', '#AFD0D0'),
+	'dif_com': ('darkslategray', '#F58CA1'),
+	'con_lrn': ('lightblue', '#AFD0D0'),
+	'con_com': ('lightcoral', '#F58CA1'),
 }
 def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 2), show_mean=True, add_jitter=False, model_trace=False):
 	import comm_cost
@@ -145,7 +147,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 			import arviz as az
 			trace = az.from_netcdf(model_trace)
 			pred_var = f'pred_{condition.split("_")[1]}'
-			generations = np.arange(0, 10)
+			generations = np.arange(1, 10)
 			az.plot_hdi(generations, trace.posterior[pred_var], ax=axes[1,i], hdi_prob=0.95, smooth=False, color=color_light, fill_kwargs={'alpha': 1, 'linewidth': 0})
 			axes[1,i].plot(generations, trace.posterior[pred_var].mean(('chain', 'draw')), color=color)
 
@@ -164,6 +166,20 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 		plt.show()
 
 
+def draw_hdi(axis, lower, upper, hdi_prob):
+	mn_y, mx_y  = axis.get_ylim()
+	padding = (mx_y - mn_y) * 0.1
+	axis.plot((lower, upper), (0, 0), color='MediumSeaGreen')
+	hdi_text = f'{int(hdi_prob*100)}% HDI'
+	hdi_width = round(upper - lower, 2)
+	axis.text((lower + upper)/2, mn_y + padding, hdi_text, ha='center', color='MediumSeaGreen', fontsize=6)
[…]

### 2023-09-07 code/model_exp2.py

@@ -4,42 +4,42 @@ import numpy as np
 import pymc as pm
 
 
+
 ROOT = Path(__file__).parent.parent.resolve()
 DATA = ROOT / 'data'
 
 
 if __name__ == '__main__':
 
-	n_chains = 10
-	# generations = np.arange(1, 10)
-
-	e1 = np.array([0, 1, 2, 3, 4, 5, 6, 7, 8, 9])
-	e2 = np.array([0, 0, 0, 0, 1, 2, 3, 4, 5, 6])
-	e3 = np.array([0, 0, 0, 0, 0, 0, 0, 1, 2, 3])
+	chains = list(range(0, 10))
+	generations = list(range(1, 10))
 
 	df = pd.read_csv(DATA / 'exp3.csv')
 
-	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] > -1))  ]
-	cst_lrn = np.array(df_lrn['cost']).reshape((n_chains, -1)).T
+	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
+	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
 
-	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] > -1))  ]
-	cst_com = np.array(df_com['cost']).reshape((n_chains, -1)).T
+	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] >= min(generations)))  ]
+	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
 
-	cst = np.stack((cst_lrn, cst_com), axis=2)
+	cost = np.stack((cost_lrn, cost_com), axis=2)
 
-	e1 = np.repeat([e1], n_chains, axis=0).T
+	e1 = np.array([1, 2, 3, 4, 5, 6, 7, 8, 9])
+	e1 = np.repeat([e1], len(chains), axis=0).T
 	e1 = np.stack((e1, e1), axis=2)
 
-	e2 = np.repeat([e2], n_chains, axis=0).T
+	e2 = np.array([0, 0, 0, 1, 2, 3, 4, 5, 6])
+	e2 = np.repeat([e2], len(chains), axis=0).T
 	e2 = np.stack((e2, e2), axis=2)
 
[…]

### 2023-09-11 code/comm_cost.py
Analysis updates

@@ -20,3 +20,74 @@ def communicative_cost(lexicon, dims):
 	U = product(*[range(n_values) for n_values in dims])
 	U_size = np.product(dims)
 	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
+
+
+if __name__ == '__main__':
+
+	degenerate = {
+		(0, 0): 'buvico',
+		(0, 1): 'buvico',
+		(0, 2): 'buvico',
+		(1, 0): 'zetico',
+		(1, 1): 'zetico',
+		(1, 2): 'zetico',
+		(2, 0): 'wopico',
+		(2, 1): 'wopico',
+		(2, 2): 'wopico',
+	}
+	holistic = {
+		(0, 0): 'buvico',
+		(0, 1): 'buviko',
+		(0, 2): 'buviqo',
+		(1, 0): 'zeticoe',
+		(1, 1): 'zetikoe',
+		(1, 2): 'zetiqoe',
+		(2, 0): 'wopicoh',
+		(2, 1): 'wopikoh',
+		(2, 2): 'wopiqoh',
+	}
+	redundant = {
+		(0, 0): 'buvico',
+		(0, 1): 'buvico',
+		(0, 2): 'buvico',
+		(1, 0): 'zetiko',
+		(1, 1): 'zetiko',
+		(1, 2): 'zetiko',
+		(2, 0): 'wopiqo',
+		(2, 1): 'wopiqo',
+		(2, 2): 'wopiqo',
+	}
+	expressive = {
+		(0, 0): 'buvico',
+		(0, 1): 'buviko',
+		(0, 2): 'buviqo',
+		(1, 0): 'zetico',
+		(1, 1): 'zetiko',
+		(1, 2): 'zetiqo',
+		(2, 0): 'wopico',
+		(2, 1): 'wopiko',
+		(2, 2): 'wopiqo',
+	}
+	semi_expressive = {
+		(0, 0): 'buvico',
+		(0, 1): 'buvico',
+		(0, 2): 'buviqo',
+		(1, 0): 'zetico',
+		(1, 1): 'zetico',
+		(1, 2): 'zetiqo',
+		(2, 0): 'wopico',
+		(2, 1): 'wopico',
+		(2, 2): 'wopiqo',
+	}
+
+	print(communicative_cost(degenerate, (3, 3)))
+	print(communicative_cost(holistic, (3, 3)))
[…]

### 2023-09-11 code/figures.py

@@ -13,6 +13,14 @@ DATA = ROOT / 'data'
 FIGS = ROOT / 'manuscript' / 'figs'
 
 
+COLORS = {
+	'dif_lrn': ('darkslategray', '#97A7A7'),
+	'dif_com': ('darkorange', '#FFC680'),
+	'con_lrn': ('darkslategray', '#97A7A7'),
+	'con_com': ('darkorange', '#FFC680'),
+}
+
+
 def plot_transmission_chains(exp_data, condition, output_path):
 	import visualize
 	import matrix
@@ -104,21 +112,17 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 	else:
 		plt.show()
 
-color_palette = colormaps['viridis'](np.linspace(0, 1, 5))
-color_palette = [color[:3] for color in color_palette]
-COLORS = {
-	'dif_lrn': ('darkorange', '#AFD0D0'),
-	'dif_com': ('darkslategray', '#F58CA1'),
-	'con_lrn': ('lightblue', '#AFD0D0'),
-	'con_com': ('lightcoral', '#F58CA1'),
-}
-def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 2), show_mean=True, add_jitter=False, model_trace=False):
-	import comm_cost
 
-	n_rows = 2 if model_trace else 1
-	fig, axes = plt.subplots(n_rows, len(conditions), figsize=figsize, squeeze=False, sharex=True, sharey=True)
+def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 2), show_mean=True, add_jitter=False, model_trace=False, spoken_cost=None):
+	import comm_cost
+	if model_trace:
+		import arviz as az
+		trace = az.from_netcdf(model_trace)
+
+	fig, axes = plt.subplots(1, len(conditions), figsize=figsize, squeeze=False, sharex=True, sharey=True)
 
 	for i, condition in enumerate(conditions):
+
 		cost_by_chain = []
 		for chain in exp_data[condition]:
 			cost_by_generation = []
@@ -130,12 +134,27 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
[…]

### 2023-09-11 code/model_exp1.py

@@ -10,57 +10,73 @@ DATA = ROOT / 'data'
 
 if __name__ == '__main__':
 
-	n_chains = 10
-	generations = np.arange(1, 10)
+	chains = list(range(0, 10))
+	generations = list(range(1, 10))
 
 	df = pd.read_csv(DATA / 'exp3.csv')
 
-	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] > 0))  ]
-	cst_matrix_lrn = np.array(df_lrn['cost']).reshape((n_chains, -1)).T
-	gen_matrix_lrn = np.repeat([generations], n_chains, axis=0).T
-	gen_matrix_lrn -= 5 # center generation (gen 5 becomes intercept)
+	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] >= min(generations)))  ]
+	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
 
-	df_com = df[  (df['condition'].str.contains('^dif_com') & (df['generation'] > 0))  ]
-	cst_matrix_com = np.array(df_com['cost']).reshape((n_chains, -1)).T
-	gen_matrix_com = np.repeat([generations], n_chains, axis=0).T
-	gen_matrix_com -= 5 # center generation (gen 5 becomes intercept)
+	df_com = df[  (df['condition'].str.contains('^dif_com') & (df['generation'] >= min(generations)))  ]
+	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
 
-	cst_matrix = np.stack((cst_matrix_lrn, cst_matrix_com), axis=2)
-	gen_matrix = np.stack((gen_matrix_lrn, gen_matrix_com), axis=2)
+	c = np.stack((cost_lrn, cost_com), axis=2)
+	
+	g = np.repeat([generations], len(chains), axis=0).T
+	g = np.stack((g, g), axis=2)
+	g -= 5 # center generation
 
 	coords = {
 		'condition': ['lrn', 'com'],
-		'chain_id': list(range(n_chains)),
+		'chain_id': chains,
 		'generation': generations,
 	}
 
 	with pm.Model(coords=coords) as model:
[…]

### 2023-09-11 code/model_exp2.py

@@ -4,72 +4,8 @@ import numpy as np
 import pymc as pm
 
 
-
 ROOT = Path(__file__).parent.parent.resolve()
 DATA = ROOT / 'data'
-
-
-if __name__ == '__main__':
-
-	chains = list(range(0, 10))
-	generations = list(range(1, 10))
-
-	df = pd.read_csv(DATA / 'exp3.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
-	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
-
-	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] >= min(generations)))  ]
-	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
-
-	cost = np.stack((cost_lrn, cost_com), axis=2)
-
-	e1 = np.array([1, 2, 3, 4, 5, 6, 7, 8, 9])
-	e1 = np.repeat([e1], len(chains), axis=0).T
-	e1 = np.stack((e1, e1), axis=2)
-
-	e2 = np.array([0, 0, 0, 1, 2, 3, 4, 5, 6])
-	e2 = np.repeat([e2], len(chains), axis=0).T
-	e2 = np.stack((e2, e2), axis=2)
-
-	e3 = np.array([0, 0, 0, 0, 0, 0, 1, 2, 3])
-	e3 = np.repeat([e3], len(chains), axis=0).T
-	e3 = np.stack((e3, e3), axis=2)
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'chain_id': chains,
-		'generation': generations,
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		α_m = pm.Normal('α_m', mu=0, sigma=1, dims='condition')
-		β1_m = pm.Normal('β1_m', mu=0, sigma=1, dims='condition')
-		β2_m = pm.Normal('β2_m', mu=0, sigma=1, dims='condition')
-		β3_m = pm.Normal('β3_m', mu=0, sigma=1, dims='condition')
-
-		α_sd = pm.Exponential('α_sd', lam=1, dims='condition')
[…]

### 2023-09-11 manuscript/wp4.tex
ms 10/09

@@
 	\vspace*{2pt}
-	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
-The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
-The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI).
-Generation~0 is not included in the model, since it does not represent participant data.}
+	\caption{Communicative cost by generation under Transmission-only (dark green; left) and Transmission + Communication (orange; right).
+The thin lines show each individual transmission chain (with a small amount of jitter added).
+The thick lines and shaded areas show the mean and 95\% HDI of a regression model fit to the data.
+Generation~0 is not included in the model, since it does not represent participant data.
+The black dotted lines show the communicative cost of the spoken language for comparison.
+In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.}
 	\label{cost_dif}
@@
 	\vspace*{2pt}
-	\caption{Posterior parameter estimates for the Experiment~1 polynomial regression model.
-Blue curves show results for the Transmission-only condition and red curves show results for the Transmission + Communication condition.
-$\alpha$ represents the intercept of the polynomial (Generation~5) and the $\beta$ parameters characterize its shape.}
+	\caption{Posterior parameter estimates from the Experiment~1 regression model.
+In the upper panels, the dark green curves show results for the Transmission-only condition and the yellow curves show results for the Transmission + Communication condition.
+The dotted line in the $\alpha$ panel shows the cost of the spoken language.
+The lower panels show the posterior differences between conditions along with the 95\% HDI credible interval.}
 	\label{posterior_dif}
@@
 
-To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model. 
+To evaluate this argument statistically, we regressed cost on generation in the following Bayesian third-degree polynomial regression model:
 	\begin{align*}
 	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-     \mu_i & =    \alpha_{\mathrm{CHAIN}[i]} + \beta_{\mathrm{CHAIN}[i]} g_i + \gamma_{\mathrm{CHAIN}[i]} g_i^2 + \delta_{\mathrm{CHAIN}[i]} g_i^3 \\
+	 \mu_i & =    \alpha + \beta_1 g_i + \beta_2 g_i^2 + \beta_3 g_i^3 \\
 	\sigma & \sim \mathrm{Exponential}(1)
 	\end{align*}
+where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~-4 through +4).
+The model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by a third-degree polynomial model.
+The intercept $\alpha$ corresponds to Generation~5 (the central generation), and the $\beta$ parameters determine the shape of the polynomial.
+$\sigma$ represents an across-participant noise term and is given a regularizing exponential prior.
+The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
+For the full model specification, see the Supplementary Material.
+
 The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}.
-Essentially, the $\beta$ parameters of the model (characterizing the shape of the polynomial) do not differ substantially between conditions, but the intercepts ($\alpha$) do appear to differ.
-Since generation was centered (for the purpose of fitting the model), the $\alpha$ parameters correspond to Generation~5 and therefore represent the main plateaus of the polynomials.
-The posterior difference between $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ was 0.301~bits (95\%~HDI: 0.014--0.597).
-This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus.
-We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage (in terms of communicative accuracy) for systems that emerged in the communicative condition.
-Specifically, systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%.
-Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.
+Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do appear to differ.
+This may be interpreted as the polynomial regression models adopting the same shape but plateauing at different levels.
+Indeed, the posterior difference in $\alpha$, and its credible interval, is less than zero, pointing to an effect of condition on the level at which cost plateaus.
+
+To give greater meaning to these parameter estimates, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$.
+If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-1.298} = 40.7$\% in Transmission-only and $2^{-0.985} = 50.5$\% in Transmission + Communication, representing a 9.8 percentage point advantage for the communicative condition.
+However, while the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.
 
@@
 	\vspace*{2pt}
-	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red).
-The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines).
-The lower panels show a regression model of the data (posterior mean and 95\% HDI).
-Generation~0 is not included in the model, since it does not represent participant data.}
+	\caption{Communicative cost by generation under Transmission-only (dark green; left) and Transmission + Communication (orange; right).
+The thin lines show each individual transmission chain (with a small amount of jitter added).
+The thick lines and shaded areas show the mean and 95\% HDI of a regression model fit to the data.
+Generation~0 is not included in the model, since it does not represent participant data.
+The black dotted lines show the communicative cost of the spoken language, which increases with epoch.
+In general, the orthographic systems tend to remain more informative under communicative pressure.}
 	\label{cost_con}
@@
 	\vspace*{2pt}
-	\caption{Posterior parameter estimates for the Experiment~2 regression model.
-Blue curves show results for the Transmission-only condition and red curves show results for the Transmission + Communication condition.}
+	\caption{Posterior parameter estimates from the Experiment~2 regression model.
+In the upper panels, the dark green curves show results for the Transmission-only condition and the yellow curves show results for the Transmission + Communication condition.
+The dotted lines in the $\alpha$ panels show the cost of the spoken language in the corresponding epoch.
+The lower panels show the posterior differences between conditions along with the 95\% HDI credible interval.}
 	\label{posterior_con}
@@
 
-[N.B., I haven't completed this bit yet] To evaluate statistically the effect of generation (modulated by epoch) on cost, we fit the following Bayesian, hierarchical regression model
+To evaluate statistically the effect of generation on cost across the two conditions, while also taking epoch into account, we fit the following Bayesian regression model:
 	\begin{align*}
-     c_{i} & \sim \mathrm{Normal}(\mu, \sigma) \\
-       \mu & = \alpha + \beta_1 e_1 + \beta_2 e_2 + \beta_3 e_3 \\
+	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+	 \mu_i & =    \alpha_{\mathrm{EPOCH}[i]} + \beta_{\mathrm{EPOCH}[i]} g_i \\
 	\sigma & \sim \mathrm{Exponential}(1)
 	\end{align*}
+where $g$ is an epoch-centered generation number (i.e., for Generations~1, 4, and~7, $g=-1$, for Generations~2, 5, and~8, $g=0$, and for Generations~3, 6, and~9, $g=1$).
+Overall, the model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by an epoch-specific linear model (i.e., each epoch has an independent intercept $\alpha$ and slope $\beta$).
+$\sigma$ represents an across-participant noise term and is given a regularizing exponential prior.
+The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
+For the full model specification, see the Supplementary Material.
 
-    \begin{align*}
-       c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-     \mu_i & =    \alpha_{\mathrm{CHAIN}[i]} + \beta_{\mathrm{CHAIN}[i]} g_i \\
-    \sigma & \sim \mathrm{Exponential}(1)
-    \end{align*}
-
-with the intercept and slope parameters fit by-chain.
-Essentially, $\beta_1$ represents the baseline effect of generation on cost, $\beta_2$ represents the additional effect during Epoch~II, and $\beta_3$ represents, again, the additional effect during Epoch~III.
-The results are shown in Fig.~\ref{posterior_con}.
-During the first epoch, the two conditions are identical.
-During the second epoch there's some difference.
-During the third epoch, there's a difference because Transmission-only plateaus and Communication continues.
-Are these slopes interpretable in any useful way?
+Fig.~\ref{posterior_con} shows the posterior estimates of the parameters of the statistical model (upper panels), along with the posterior differences between conditions (lower panels).
+The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), show a gradual divergence with the difference between conditions becoming more pronounced with each epoch.
+Nevertheless, even by Epoch~III, the posterior difference in intercepts, $\Delta(\alpha_3)$, does not conclusively exclude zero.
+Similarly, the $\beta$ parameters, which represent the slopes within each of the three epochs are not conclusively different between conditions, although there is some evidence to suggest that the slope is shallower in the communicative condition during Epoch~II (e.g., 89\% of the posterior probability mass on $\Delta(\beta_2)$ is below zero).
+In the case of $\beta_3$ the effect is reversed, with Transmission + Communication having a greater slope.
+Note, however, that this appears to be because cost has, by this stage, plateaued in the Transmission-only condition, so the increased homophony pressure introduced in Epoch~III has little effect in Transmission-only, but continues to exert an effect in Transmission + Communication.
 
@@
 
+\section{Appendix}
+
+\subsection{Model 1}
+
+	\begin{align*}
+	   c_i                   & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+	 \mu_i                   & = \alpha_{\mathrm{CHAIN}[i]} + \beta_{1,\mathrm{CHAIN}[i]} g_i + \beta_{2,\mathrm{CHAIN}[i]} g_i^2 + \beta_{3,\mathrm{CHAIN}[i]} g_i^3 \\
+	\sigma                   & \sim \mathrm{Exponential}(1) \\
+	\alpha_{\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\alpha_\mu, \alpha_\sigma) \\
+	\beta_{d,\mathrm{CHAIN}} & \sim \mathrm{Normal}(\beta_{d_\mu}, \beta_{d_\sigma}) \\
+	\alpha_\mu               & \sim \mathrm{Normal}(1, 1) \\
+	\alpha_\sigma            & \sim \mathrm{Exponential}(1) \\
+	\beta_{d_\mu}            & \sim \mathrm{Normal}(0, 1) \\
+	\beta_{d_\sigma}         & \sim \mathrm{Exponential}(1)
+	\end{align*}
+
+\subsection{Model 2}
+
+	\begin{align*}
+	   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+	 \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
+	\sigma                                 & \sim \mathrm{Exponential}(1) \\
+	\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
+	\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
+	\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
+	\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
+	\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
+	\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
+	\end{align*}
+
 %~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

### 2023-09-14 code/figures.py
Update analyses

@@ -1,6 +1,8 @@
 from pathlib import Path
 import numpy as np
+import arviz as az
 import matplotlib.pyplot as plt
+import matplotlib.gridspec as gridspec
 from matplotlib import colormaps
 from utils import json_load
 
@@ -59,9 +61,14 @@ def plot_transmission_chains(exp_data, condition, output_path):
 	visualize.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
 
 
-def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False):
+def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False, model_trace=None):
 	import voi
 	import matrix
+	if model_trace:
+		trace = az.from_netcdf(model_trace)
+		n_rows = len(conditions) + 1
+	else:
+		n_rows = len(conditions)
 
 	ref_systems = [
 		matrix.reference_systems['holistic'],
@@ -72,7 +79,6 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 
 	distribution_colors = colormaps['viridis'](np.linspace(0, 1, 4))
 
-	fig, axes = plt.subplots(len(conditions), len(generations), figsize=figsize, squeeze=False, sharex=True, sharey=True)
 	
 	for i, condition in enumerate(conditions):
 		for j, generation in enumerate(generations):
@@ -91,38 +97,64 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 					# which we know to be correct.
 					unnorm_distribution = np.exp(-2 * distances**2)
 					distribution += unnorm_distribution / unnorm_distribution.sum()
+					distribution /= distribution.sum()
+					distribution *= 10
 				else:
-					classification = np.where(distances == distances.min())[0]
-					distribution[classification] += 1 / len(classification)
-			distribution /= distribution.sum()
+					category = np.argmin(distances)
[…]

### 2023-09-14 code/model_exp1.py

@@ -79,4 +79,4 @@ if __name__ == '__main__':
 		# Sampling
 
 		trace = pm.sample(10000, tune=2000, target_accept=0.99, chains=6, cores=6)
-		trace.to_netcdf(DATA / 'exp1.netcdf')
+		trace.to_netcdf(DATA / 'exp1_cost.netcdf')

### 2023-09-14 code/model_exp2.py

@@ -8,86 +8,6 @@ ROOT = Path(__file__).parent.parent.resolve()
 DATA = ROOT / 'data'
 
 
-# if __name__ == '__main__':
-
-# 	chains = list(range(0, 10))
-# 	generations = list(range(1, 10))
-
-# 	df = pd.read_csv(DATA / 'exp3.csv')
-
-# 	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
-# 	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
-
-# 	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] >= min(generations)))  ]
-# 	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
-
-# 	cost = np.stack((cost_lrn, cost_com), axis=2)
-
-# 	e1 = np.array([1, 2, 3, 4, 5, 6, 7, 8, 9])
-# 	e1 = np.repeat([e1], len(chains), axis=0).T
-# 	e1 = np.stack((e1, e1), axis=2)
-# 	e1 -= 5
-
-# 	e2 = np.array([0, 0, 0, -1, 0, 1, 2, 3, 4])
-# 	e2 = np.repeat([e2], len(chains), axis=0).T
-# 	e2 = np.stack((e2, e2), axis=2)
-
-# 	e3 = np.array([0, 0, 0, 0, 0, 0, 2, 3, 4])
-# 	e3 = np.repeat([e3], len(chains), axis=0).T
-# 	e3 = np.stack((e3, e3), axis=2)
-
-# 	coords = {
-# 		'condition': ['lrn', 'com'],
-# 		'chain_id': chains,
-# 		'generation': generations,
-# 	}
-
-# 	with pm.Model(coords=coords) as model:
-
-# 		# Hyperpriors
-
-# 		α_m = pm.Normal('α_m', mu=0, sigma=1, dims='condition')
-# 		β1_m = pm.Normal('β1_m', mu=0, sigma=1, dims='condition')
-# 		β2_m = pm.Normal('β2_m', mu=0, sigma=1, dims='condition')
-# 		β3_m = pm.Normal('β3_m', mu=0, sigma=1, dims='condition')
-
[…]

### 2023-09-14 code/model_typo.py

@@ -1,16 +1,18 @@
 from pathlib import Path
 import numpy as np
 import pymc as pm
-import arviz as az
-from utils import json_load
 
 
 ROOT = Path(__file__).parent.parent.resolve()
+DATA = ROOT / 'data'
 
 
-def construct_t_matrix(dataset, exp_id):
+def construct_t_matrix(data_path, experiment):
+	from utils import json_load
 	import voi
 	import matrix
+
+	dataset = json_load(data_path)
 
 	ref_systems = [
 		matrix.reference_systems['holistic'],
@@ -20,8 +22,7 @@ def construct_t_matrix(dataset, exp_id):
 	]
 
 	t = np.zeros((2, 9, 4), dtype=int)
-
-	for i, condition in enumerate([f'{exp_id}_lrn', f'{exp_id}_com']):
+	for i, condition in enumerate([f'{experiment}_lrn', f'{experiment}_com']):
 		for j, generation in enumerate(range(1, 10)):
 			for chain in dataset[condition]:
 				subject_a = chain[generation][0]
@@ -37,9 +38,11 @@ def construct_t_matrix(dataset, exp_id):
 
 if __name__ == '__main__':
 
-	dataset = json_load(ROOT / 'data' / 'exp3.json')
+	data_path = DATA / 'exp3.json'
+	experiment = 'con'
+	output_path = DATA / 'exp2_typo.netcdf'
 
-	t = construct_t_matrix(dataset, 'dif')
+	t = construct_t_matrix(data_path, experiment)
 
 	coords = {
 		'category': ['H', 'E', 'R', 'D'],
@@ -59,8 +62,5 @@ if __name__ == '__main__':
 		pm.Deterministic('diff_θ', θ[1, :, :] - θ[0, :, :], dims=('generation', 'category'))
 
 		# Sampling
-		trace = pm.sample(10000, tune=1000)
-		# trace.to_netcdf()
-
-		az.plot_posterior(trace, var_names=['diff_θ'], hdi_prob=0.9)
-		az.utils.plt.show()
+		trace = pm.sample(10000, tune=2000, chains=6, cores=6)
+		trace.to_netcdf(output_path)

### 2023-09-14 manuscript/wp4.tex
ms 14/09

@@
 
-In this paper, we simulate the processes of differentiation and conservation experimentally.
+In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}.
+In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production.
+Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
+Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
+Iterated learning has also been explored in a wide range of domains, such as gesture \parencite{}, population structure \parencite{}, and blah \parencite{}.
+For reviews, see \textcite{}.
 
@@
 Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
-Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
 To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.
+
+[make this more specific to the present experiment - what is passed down the chain - preview the methods coming up]
 
@@
 The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}.
-Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, just as in Fig.~\ref{stimuli}.
-For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension.
-We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information.
-Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition.
+Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{stimuli}.
+The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation.
+We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simple repeats whichever shape was marked by the stem.
+Redundant suffix systems are characteristic of the Transmission-only condition, with similar outcomes occurring in Chains~C, E, F, and~H.
 We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J.
@@
 The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.
-
-The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
-Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
-In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6).
-Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable.
-In particular, there are two main kinds of color-expressive system that emerge.
-The first are the generalization-based expressive systems: K5, M1, P5, and S2.
-In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape.
-For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>.
-In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->).
-We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.
 
@@
 
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
+Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
+In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6).
+Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable.
+In particular, there are two main kinds of color-expressive system that emerge.
+The first are the generalization-based expressive systems: K5, M1, P5, and S2.
+In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape.
+For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>.
+In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->).
+We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner---resulting in low communicative accuracy---and not sustained or elaborated~on in subsequent generations.
+
 The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4.
@@
 In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9.
-Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously.
+Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that were too difficult to learn for subsequent generations.
 Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color.
@@
 
-To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate.
-These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}.
-The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing.
-To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content.
-To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.
+To analyze more formally what types of system tend to emerge under different conditions, we first designated four suffix systems that are of particular interest: holistic, expressive, redundant, and degenerate.
+These four typological categories may be positioned along the simplicity--informativeness continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}.
+The holistic\footnote{Our use of the term \textit{holistic} is slightly unusual here, owing to the fact that we are focused on the suffix level and not the word level.
+By holistic, we only mean that each shape--color combination has a unique suffix form, so the nine suffixes must all be learned as separate cases.
+We do not distinguish between truly holistic suffixes (nine unique suffixes with no structure in how they relate to each other) and compositional suffixes (nine unique suffixes that can be generated from compositional rules, as was the case in, for example, Q8).} system has nine unique forms each of which expresses a particular color--shape combination.
+This is complex to learn, but the suffix alone can pick out exactly one meaning.
+The expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant).
+These systems are easier to learn, but only the expressive system is fully informative (when acting in combination with the stem).
+The degenerate system uses just one suffix form.
+This makes it trivial to learn, but entirely uninformative.
+To classify a participant's output into one of these four typological categories, we computed which of the reference systems was most similar in terms of its information content.
+To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite{Meila:2007} defined as the distance metric between any two such partitions.\footnote{Variation of information is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one partition into another.
 Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content.
-The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we computed a probability distribution over the typological space according to the respective distances from the four reference systems.
-This approach gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate).
-The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
-\begin{equation}
-	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
-\end{equation}
-and averaged over chains for a given generation.
+The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} A given participant's output system is then classified into one of the four typological categories based on whichever reference system is closest.
 
@@
 	\vspace*{2pt}
-	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
-At Generation~0, all 10 chains in both conditions use a holistic system (by design).
-In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning.
-In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way.
-Expressive systems were rare in both conditions.}
+	\caption{Upper panels show the typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+Bottom panels show posterior differences (mean and 90\% HDI) in the proportion of chains falling into each typological category; positive values signify greater use of a given category in the communicative condition.}
 	\label{typ_dist_dif}
@@
 
-The results are plotted in Fig.~\ref{typ_dist_dif}.
-These probability distributions can be read as the proportion of chains that adopted each of the four configurations at each generation.
-At Generation~0, all chains are holistic by design.
-In Transmission-only, there is an initial rapid increase in the number of chains classified into the redundant typological category.
+The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation.
+In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2.
 The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy.
-In Transmission + Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common.
-There is, however, a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings.
+In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common.
+There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7).
+For a statistical comparison across conditions, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
+	\begin{equation}
+	\begin{aligned}
+	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
+	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
+	\end{aligned}
+	\end{equation}
+and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$.
+The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}.
+In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7).
+Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.
 
-As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}, which expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
+As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
+This measure expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
 If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system.
 Communicative cost will take some larger value if the system contains ambiguity.
-Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
+It is given by
 	\begin{equation}
@@
 	\vspace*{2pt}
-	\caption{Communicative cost by generation under Transmission-only (dark green; left) and Transmission + Communication (orange; right).
-The thin lines show each individual transmission chain (with a small amount of jitter added).
-The thick lines and shaded areas show the mean and 95\% HDI of a regression model fit to the data.
+	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+The thin lines show each individual transmission chain (with a small amount of added jitter).
+The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
 Generation~0 is not included in the model, since it does not represent participant data.
 The black dotted lines show the communicative cost of the spoken language for comparison.
-In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.}
+In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_dif}
@@
 
-Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition.
-In the upper panels, each thin line represents an individual transmission chain, and the thick lines show the mean across chains.
+Fig.~\ref{cost_dif}A plots communicative cost over the nine generations of cultural transmission in each condition.
 The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling.
@@
 This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent.
-In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems.
+In contrast, this is not the expected result in Transmission~+ Communication, where the communicative pressure should promote the emergence of informative systems.
 This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization.
-Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.
+Interestingly, however, cost plateaus at a slightly lower level in Transmission~+ Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.
 
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_dif.eps}}
-	\vspace*{2pt}
-	\caption{Posterior parameter estimates from the Experiment~1 regression model.
-In the upper panels, the dark green curves show results for the Transmission-only condition and the yellow curves show results for the Transmission + Communication condition.
-The dotted line in the $\alpha$ panel shows the cost of the spoken language.
-The lower panels show the posterior differences between conditions along with the 95\% HDI credible interval.}
-	\label{posterior_dif}
-	\end{figure}
-
-To evaluate this argument statistically, we regressed cost on generation in the following Bayesian third-degree polynomial regression model:
-	\begin{align*}
+To evaluate this argument statistically, we fit the following regression model:
+	\begin{equation}
+	\begin{aligned}
 	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
 	 \mu_i & =    \alpha + \beta_1 g_i + \beta_2 g_i^2 + \beta_3 g_i^3 \\
-	\sigma & \sim \mathrm{Exponential}(1)
-	\end{align*}
-where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~-4 through +4).
+	\sigma & \sim \mathrm{Exponential}(1),
+	\end{aligned}
+	\end{equation}
+where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~$-4$ through $+4$).
 The model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by a third-degree polynomial model.
@@
 For the full model specification, see the Supplementary Material.
+The posterior parameter estimates of the model are shown in Fig.~\ref{cost_dif}B.
+Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do differ.
+This may be interpreted as the polynomial regression models adopting approximately the same shape but plateauing at different levels of cost.
+Additionally, we find that the $\alpha$ parameters themselves are lower than the cost of the spoken language (indicated by the dotted line), although it is unclear how this situation might unfold beyond Generation~9, especially in Transmission-only where the model predicts that cost will continue to increase further until all orthographic systems become aligned with the spoken language.
 
-The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}.
-Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do appear to differ.
-This may be interpreted as the polynomial regression models adopting the same shape but plateauing at different levels.
-Indeed, the posterior difference in $\alpha$, and its credible interval, is less than zero, pointing to an effect of condition on the level at which cost plateaus.
-
-To give greater meaning to these parameter estimates, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$.
-If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-1.298} = 40.7$\% in Transmission-only and $2^{-0.985} = 50.5$\% in Transmission + Communication, representing a 9.8 percentage point advantage for the communicative condition.
-However, while the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.
+Finally, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$.
+If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: 35.1--46.3) in Transmission-only and 50.6\% (90\% HDI: 45.9--54.9) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition.
+Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that an expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use, while also being fairly easy to learn.
 
@@
 
-Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation.
+Our first experiment tested whether informative, heterographic orthography could emerge through differentiation and whether it would emerge preferentially under communicative pressure.
+Overall, there was little evidence of systematic orthographic differentiation.
 Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue).
-However, such differentiation was not very forthcoming, even under communicative pressure.
-Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings.
-This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.
+Such differentiation was, however, not very forthcoming, even under communicative pressure.
+Instead, when spelling variation was conditioned on anything, it was usually conditioned on shape, resulting in redundant suffix spellings.
+This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.
 
 So, what was different about the present experiment?
-The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color.
-Faced with variation that could be conditioned on either dimension, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
-As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose.
-A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).
+The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions.
+Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness.
+The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color.
+Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
+As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
+A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}).
+As we will see in the following experiment, it is indeed possible for participants to learn expressive systems, even under full homophony; the lack of such outcomes in the present experiment suggests a reluctance to defy the spoken language.
 
-It is particularly surprising that participants chose not to innovate novel systems in communication, despite their being a clear financial incentive to do so.
+It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite their being a clear financial incentive to do so.
 Why might this be the case?
 Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
-Participants who adopted this interpretation may have feared diverging from what they learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
+Participants who adopted this interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
 Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
 They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.
-
 These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space.
@@
 
-The methods were identical to Experiment~1 with two main exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.
+The methods were identical to Experiment~1 with two exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly homophonous and uninformative over time.
 
@@
 Crucially, the spellings did not automatically change following a sound change event; rather, the orthographic system was free to adapt (or not) in response to the sound changes.
-Note also that an individual participant did not directly experience the sound changes; rather, a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
+Note also that an individual participant did not directly experience the sound changes; a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
 In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.
@@
 For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Roman alphabet, but which trace their origins back to the original seed orthography.
-Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without the conservation of expressivity.
+Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without conserving how form patterns with meaning.
 
@@
 
-Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words in Generation~9 of Chain~K, although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English).
+Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words (Generation~9 of Chain~K), although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English).
 Presumably, the general conservation of expressive spelling in Experiment~2 negated the need to innovate novel systems.
@@
 At generation~0, all 10 chains in both conditions use an expressive system (by design).
-In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission + Communication, they tend to remain expressive for longer and disfavor the redundant pattern.}
+In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission~+ Communication, they tend to remain expressive for longer and disfavor the redundant pattern.
+Posterior difference (mean and 90\% HDI) in the proportion of chains falling into each typological category.
+Positive values signify greater use of a given typological category in the communicative condition (and vice versa for negative values).}
 	\label{typ_dist_con}
@@
 
-Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition.
+Following the methods used in Experiment~1, Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition, and Fig.~\ref{typo_posteriors_con} plots the posterior differences between conditions.
 At Generation~0, all systems are expressive (by design), but the dominance of this typological category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
-Notably, however, the loss of expressive systems appears to be slower in Transmission + Communication, and redundant systems were also less popular in Transmission + Communication.
+Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular in Transmission~+ Communication.
 Overall, there is a general trend toward degeneracy---or, in more positive terms, a trend toward transparently mirroring the spoken language---in both conditions, but this degeneration occurs more slowly under communicative pressure.
@@
 	\vspace*{2pt}
-	\caption{Communicative cost by generation under Transmission-only (dark green; left) and Transmission + Communication (orange; right).
-The thin lines show each individual transmission chain (with a small amount of jitter added).
-The thick lines and shaded areas show the mean and 95\% HDI of a regression model fit to the data.
+	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+The thin lines show each individual transmission chain (with a small amount of added jitter).
+The thick lines and shaded regions show the mean and 90\% HDI of a regression model fit to the data.
 Generation~0 is not included in the model, since it does not represent participant data.
 The black dotted lines show the communicative cost of the spoken language, which increases with epoch.
-In general, the orthographic systems tend to remain more informative under communicative pressure.}
+In general, the orthographic systems tend to remain more informative under communicative pressure.
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_con}
@@
 
-Fig.~\ref{cost_con} plots communicative cost.
+Fig.~\ref{cost_con}A plots communicative cost.
 The mean lines reveal a fairly clear effect of epoch.
@@
 This represents a baseline level of noise when the spoken languages are fully informative.
-During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in the Transmission + Communication condition.
+During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in the Transmission~+ Communication condition.
 This difference is presumably due to the communicative pressure, which acts to slow down the rate at which the systems become transparent or redundant.
 In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
-In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.
+In Transmission~+ Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.
 
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_con.eps}}
-	\vspace*{2pt}
-	\caption{Posterior parameter estimates from the Experiment~2 regression model.
-In the upper panels, the dark green curves show results for the Transmission-only condition and the yellow curves show results for the Transmission + Communication condition.
-The dotted lines in the $\alpha$ panels show the cost of the spoken language in the corresponding epoch.
-The lower panels show the posterior differences between conditions along with the 95\% HDI credible interval.}
-	\label{posterior_con}
-	\end{figure}
-
-To evaluate statistically the effect of generation on cost across the two conditions, while also taking epoch into account, we fit the following Bayesian regression model:
-	\begin{align*}
+To evaluate statistically the effect of generation on cost across the two conditions, while also taking epoch into account, we fit the following regression model:
+	\begin{equation}
+	\begin{aligned}
 	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
 	 \mu_i & =    \alpha_{\mathrm{EPOCH}[i]} + \beta_{\mathrm{EPOCH}[i]} g_i \\
-	\sigma & \sim \mathrm{Exponential}(1)
-	\end{align*}
+	\sigma & \sim \mathrm{Exponential}(1),
+	\end{aligned}
+	\end{equation}
 where $g$ is an epoch-centered generation number (i.e., for Generations~1, 4, and~7, $g=-1$, for Generations~2, 5, and~8, $g=0$, and for Generations~3, 6, and~9, $g=1$).
@@
 
-Fig.~\ref{posterior_con} shows the posterior estimates of the parameters of the statistical model (upper panels), along with the posterior differences between conditions (lower panels).
-The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), show a gradual divergence with the difference between conditions becoming more pronounced with each epoch.
-Nevertheless, even by Epoch~III, the posterior difference in intercepts, $\Delta(\alpha_3)$, does not conclusively exclude zero.
+The posterior parameter estimates are shown in Fig.~\ref{cost_con}B.
+The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), reveal the gradual divergence of the systems across the two conditions: By Epoch~III, the difference is xxx~bits (90\% HDI: [xxx, yyy]).
+Further, by Epoch~III, the orthographic systems in both conditions are more informative than the spoken language: xxx () in Transmission-only and xxx () in Transmission~+ Communication.
+
+ in terms of both (a) the difference between conditions, which becomes more pronounced with epoch, and the difference from the spoken language
+
 Similarly, the $\beta$ parameters, which represent the slopes within each of the three epochs are not conclusively different between conditions, although there is some evidence to suggest that the slope is shallower in the communicative condition during Epoch~II (e.g., 89\% of the posterior probability mass on $\Delta(\beta_2)$ is below zero).
-In the case of $\beta_3$ the effect is reversed, with Transmission + Communication having a greater slope.
-Note, however, that this appears to be because cost has, by this stage, plateaued in the Transmission-only condition, so the increased homophony pressure introduced in Epoch~III has little effect in Transmission-only, but continues to exert an effect in Transmission + Communication.
+In the case of $\beta_3$ the effect is reversed, with Transmission~+ Communication having a greater slope.
+Note, however, that this appears to be because cost has, by this stage, plateaued in the Transmission-only condition, so the increased homophony pressure introduced in Epoch~III has little effect in Transmission-only, but continues to exert an effect in Transmission~+ Communication.
 
@@
 
+	\item The fact that the results are less "significant" in conservation does not mean that conservation is a worse model; it means that, under the conservation mechanism, communicative pressure is less important.
+This might suggest an accident-of-history explanation: Learning alone (i.e., habitual, mindless replication) can explain the emergence of informative orthography.
+Informative orthography does not arise through adaption, but may instead be the result of mindlessly replicated historic residue following sound change.
+
 	\item Limitations
@@
 
-\subsection{Model 1}
+\subsection{Experiment 1 regression model}
 
-	\begin{align*}
+	\begin{equation}
+	\begin{aligned}
 	   c_i                   & \sim \mathrm{Normal}(\mu_i, \sigma) \\
@@
 	\alpha_{\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\alpha_\mu, \alpha_\sigma) \\
-	\beta_{d,\mathrm{CHAIN}} & \sim \mathrm{Normal}(\beta_{d_\mu}, \beta_{d_\sigma}) \\
+	\beta_{j,\mathrm{CHAIN}} & \sim \mathrm{Normal}(\beta_{j_\mu}, \beta_{j_\sigma}) \\
 	\alpha_\mu               & \sim \mathrm{Normal}(1, 1) \\
 	\alpha_\sigma            & \sim \mathrm{Exponential}(1) \\
-	\beta_{d_\mu}            & \sim \mathrm{Normal}(0, 1) \\
-	\beta_{d_\sigma}         & \sim \mathrm{Exponential}(1)
-	\end{align*}
+	\beta_{j_\mu}            & \sim \mathrm{Normal}(0, 1) \\
+	\beta_{j_\sigma}         & \sim \mathrm{Exponential}(1)
+	\end{aligned}
+	\end{equation}
 
-\subsection{Model 2}
+\subsection{Experiment 2 regression model}
 
-	\begin{align*}
+	\begin{equation}
+	\begin{aligned}
 	   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
@@
 	\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
-	\end{align*}
+	\end{aligned}
+	\end{equation}
 

### 2023-09-15 code/figures.py
Add summary table

@@ -22,6 +22,8 @@ COLORS = {
 	'con_com': ('darkorange', '#FFC680'),
 }
 
+HDI_PROB = 0.9
+
 
 def plot_transmission_chains(exp_data, condition, output_path):
 	import visualize
@@ -120,7 +122,7 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 
 				samples = trace.posterior.diff_θ.sel(generation=generation, category=category).to_numpy().flatten()
 
-				az_hdi = az.hdi(samples, hdi_prob=0.9)
+				az_hdi = az.hdi(samples, hdi_prob=HDI_PROB)
 				lower, upper = float(az_hdi[0]), float(az_hdi[1])
 				mean = samples.mean()
 				
@@ -180,8 +182,8 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 		pred_var = f'pred_{condition.split("_")[1]}'
 		
 		pred = [trace.posterior[pred_var+f'_{i}'].mean(('chain', 'draw')) for i in range(1, 10)]
-		pred_upper = [float(az.hdi(trace.posterior[pred_var+f'_{i}'].to_numpy().flatten(), hdi_prob=0.9)[1]) for i in range(1, 10)]
-		pred_lower = [float(az.hdi(trace.posterior[pred_var+f'_{i}'].to_numpy().flatten(), hdi_prob=0.9)[0]) for i in range(1, 10)]
+		pred_upper = [float(az.hdi(trace.posterior[pred_var+f'_{i}'].to_numpy().flatten(), hdi_prob=HDI_PROB)[1]) for i in range(1, 10)]
+		pred_lower = [float(az.hdi(trace.posterior[pred_var+f'_{i}'].to_numpy().flatten(), hdi_prob=HDI_PROB)[0]) for i in range(1, 10)]
 		
 		axis.fill_between(range(1, 10), pred_lower, pred_upper, color=color_light, zorder=1.5)
 		axis.plot(range(1, 10), pred, color=color, linewidth=4)
@@ -220,7 +222,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 			samples = trace.posterior[variable['var']].sel(epoch=variable['sub']).to_numpy().flatten()
 		else:
 			samples = trace.posterior[variable['var']].to_numpy().flatten()
[…]

### 2023-09-15 manuscript/wp4.tex
ms 15/09

@@
 
-In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}.
+In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008}.
 In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production.
 Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
-Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
-Iterated learning has also been explored in a wide range of domains, such as gesture \parencite{}, population structure \parencite{}, and blah \parencite{}.
-For reviews, see \textcite{}.
+Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
+For reviews, see \textcite{Tamariz:2017, Kirby:2014, ScottPhillips:2010}.
+
+Bottleneck
 
@@
 The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the full model specification, see the Supplementary Material.
+For the complete model specification, see the Supplementary Material.
+
 The posterior parameter estimates of the model are shown in Fig.~\ref{cost_dif}B.
@@
 Finally, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$.
-If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: 35.1--46.3) in Transmission-only and 50.6\% (90\% HDI: 45.9--54.9) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition.
+If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: [35.1, 46.3]) in Transmission-only and 50.6\% (90\% HDI: [45.9, 54.9]) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition.
 Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that an expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use, while also being fairly easy to learn.
@@
 Why might this be the case?
-Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
+Firstly, although innovation was not expressly forbidden in the instructions, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
 Participants who adopted this interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
@@
 The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.
-Like Transmission-only, the seed systems are mostly maintained faithfully through Epoch~I; perfectly in the case of Chains~K, L, N, O, and~J, and with some errors in the other five chains, although some of these errors are non-destructive changes, such as <x> being replaced with <sh> in R1.
+Like Transmission-only, the seed systems are mostly maintained faithfully through Epoch~I; perfectly in the case of Chains~K, L, N, O, and~T, and with some errors in the other five chains, although some of these errors are non-destructive changes, such as <x> being replaced with <sh> in R1.
 Several systems were then maintained through Epoch~II, notably Chains~O, P, R, and~T, and, in one case, through to the end of Epoch~III (Chain~R, albeit with the original <xei> spelling replaced with <shei>).
@@
 
-Following the methods used in Experiment~1, Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition, and Fig.~\ref{typo_posteriors_con} plots the posterior differences between conditions.
-At Generation~0, all systems are expressive (by design), but the dominance of this typological category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
-Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular in Transmission~+ Communication.
-Overall, there is a general trend toward degeneracy---or, in more positive terms, a trend toward transparently mirroring the spoken language---in both conditions, but this degeneration occurs more slowly under communicative pressure.
+Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition.
+Initially, all systems are expressive, but the dominance of this typological category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
+Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular under communicative pressure.
+The clearest difference between conditions comes at Generation~7, at which point there were more expressive systems and fewer redundant systems in the communicative condition.
 
@@
 Fig.~\ref{cost_con}A plots communicative cost.
-The mean lines reveal a fairly clear effect of epoch.
-During Epoch~I, there is a shallow increase in cost with generation in both conditions.
-This represents a baseline level of noise when the spoken languages are fully informative.
-During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in the Transmission~+ Communication condition.
-This difference is presumably due to the communicative pressure, which acts to slow down the rate at which the systems become transparent or redundant.
-In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
-In Transmission~+ Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.
+During Epoch~I, there is a shallow increase in cost in both conditions, representing a baseline level of noise when the spoken language is fully informative.
+During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a comparatively shallow rise in the Transmission~+ Communication condition, presumably due to the communicative pressure, which slows the rate at which the systems become transparent or redundant.
+In Epoch~III, cost plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
+In Transmission~+ Communication, by contrast, cost continues to increase at much the same rate as in Epoch~II.
 
@@
 The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the full model specification, see the Supplementary Material.
+For the complete model specification, see the Supplementary Material.
 
 The posterior parameter estimates are shown in Fig.~\ref{cost_con}B.
-The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), reveal the gradual divergence of the systems across the two conditions: By Epoch~III, the difference is xxx~bits (90\% HDI: [xxx, yyy]).
-Further, by Epoch~III, the orthographic systems in both conditions are more informative than the spoken language: xxx () in Transmission-only and xxx () in Transmission~+ Communication.
-
- in terms of both (a) the difference between conditions, which becomes more pronounced with epoch, and the difference from the spoken language
-
-Similarly, the $\beta$ parameters, which represent the slopes within each of the three epochs are not conclusively different between conditions, although there is some evidence to suggest that the slope is shallower in the communicative condition during Epoch~II (e.g., 89\% of the posterior probability mass on $\Delta(\beta_2)$ is below zero).
+The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), reveal the gradual divergence of the conditions: By Epoch~III, the difference is $-0.43$~bits (90\% HDI: [$-0.83$, $-0.04$]), meaning that the communicative condition is about half a bit more informative.
+Further, after Epoch~I, the orthographic systems tend to remain more informative than the spoken language, especially under the communicative condition.
+The $\beta$ parameters, which represent the slopes within each of the three epochs are not conclusively different between conditions, although there is some evidence to suggest that the slope is shallower in the communicative condition during Epoch~II (e.g., 89\% of the posterior probability mass on $\Delta(\beta_2)$ is below zero).
 In the case of $\beta_3$ the effect is reversed, with Transmission~+ Communication having a greater slope.
@@
 
+At Generation~7---the point at which the two conditions differ most substantially---the difference in cost is $-0.6$~bits (90\% HDI: [$-1.04$, $-0.14$]).
+Converting cost to the probability of communicative success, we obtain an estimate of 46.5\% (90\% HDI: [38.1, 54.7]) in Transmission-only and 70.9\% (90\% HDI: [52.8, 87.8]) in Transmission~+ Communication, representing an advantage of around 25 percentage points for the communicative condition.
+The equivalent value for the spoken language at Generation~7 is 33.3\% (i.e., because in Epoch~III all three colors are expressed homophonously).
+Thus, not only are the two conditions different from each other, but also the two conditions are more informative than the spoken language, especially under communicative pressure.
+
 \subsection{Summary}
 
-Experiment~1 asked whether informative orthography can be \textit{created} under communicative pressure, and the evidence suggested that differentiation is hard to achieve in the written modality.
-In Experiment~2, by contrast, we asked whether informativeness can simply be \textit{maintained}, and preferentially so under communicative pressure.
-In the Transmission-only condition, only one chain remained expressive into Epoch~III, while in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, albeit not necessarily all the way to Generation~9.
+Experiment~1 asked whether an informative, heterographic orthography may be \textit{created} de~novo under homophony.
+Experiment~2, by contrast, asked whether an informative, heterographic orthography can simply be \textit{maintained}, even under the same levels of homophony encountered in Experiment~1.
+In the Transmission-only condition, only one chain (Chain~D) remained expressive into the fully homophonous Epoch~III, while in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, albeit not necessarily all the way to Generation~9.
 The fact that expressive spellings persisted longer and across more chains under communicative pressure suggests that an informative orthography---despite running contrary to the spoken language---may be maintained when it serves a useful purpose.
+That being said, the fact that informativeness could, in~principle, be maintained without communicative pressure (most notably in Chain~D) suggests that a strong communicative pressure is not a strictly necessary condition for conservation: Learning alone can, to a limited extent, maintain informative heterography.
 
-Nevertheless, many of the chains did eschew informativeness in favor of greater transparency, and the inevitable consequence for all chains appears to be degeneracy in the long term.
-This is to be expected under Transmission-only, where the systems are adapting under a learnability pressure, but is somewhat surprising under the communicative condition.
-Participants were quite resistant to encoding into writing something that is not encoded in speech, even though they were---at the same time---content to conserve unusual suffix forms.
-This perhaps points to a general preference that readers might possess (at least, readers who are accustomed to an alphabetic writing system) in which the primary goal of writing is seen as representing speech---not representing meaning directly.
+Many of the chains did, however, eschew informativeness entirely in favor of greater transparency, and the inevitable long-term consequence for all chains appears to be degeneracy.
+This is to be expected under Transmission-only, where the systems are adapting under learnability pressure, but is somewhat surprising in Transmission~+ Communication.
+Our findings ultimately suggest that, in the long term, (alphabetic) orthographic systems favor the faithful encoding of speech over the useful encoding of meaning, although there may exist short windows of time during which informative heterography can resist the spoken language.
+This points to a preference that readers might possess---those accustomed to an alphabetic writing system, at least---in which the primary goal of writing is seen as representing speech, not representing meaning directly.
+Interestingly, although participants were quite resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual.
+Tradition has a powerful
 
@@
 
+\begin{table}
+\begin{center}
+\begin{threeparttable}
+\caption{Experiment 1 posterior parameter estimates}
+\footnotesize
+\begin{tabular}{lrrrrrr}
+\toprule
+ &   Mean &     SD &  HDI (5\%) &  HDI (95\%) &  ESS &  $\hat{R}$ \\
+\midrule
+\multicolumn{7}{c}{Transmission-only} \\
+$\alpha_\mu$  &  1.298 &  0.123 &   1.098 &    1.498 &   55213 &    1.0 \\
+$\beta_{1_\mu}$ &  0.003 &  0.030 &  -0.048 &    0.051 &   19478 &    1.0 \\
+$\beta_{2_\mu}$ & -0.021 &  0.008 &  -0.034 &   -0.009 &   49158 &    1.0 \\
+$\beta_{3_\mu}$ &  0.007 &  0.002 &   0.004 &    0.010 &   13927 &    1.0 \\
+$\sigma$    &  0.192 &  0.020 &   0.160 &    0.225 &   26492 &    1.0 \\
+\midrule
+\multicolumn{7}{c}{Transmission + Communication} \\
+$\alpha_\mu$  &  0.985 &  0.079 &   0.854 &    1.112 &   22089 &    1.0 \\
+$\beta_{1_\mu}$ &  0.003 &  0.059 &  -0.091 &    0.100 &   19115 &    1.0 \\
+$\beta_{2_\mu}$ & -0.017 &  0.007 &  -0.030 &   -0.005 &   15955 &    1.0 \\
+$\beta_{3_\mu}$ &  0.004 &  0.004 &  -0.002 &    0.010 &   12773 &    1.0 \\
+$\sigma$    &  0.370 &  0.034 &   0.313 &    0.425 &   27848 &    1.0 \\
+\midrule
+\multicolumn{7}{c}{Difference} \\
+$\Delta(\alpha_\mu)$    & -0.313 &  0.147 &  -0.551 &   -0.072 &   38539 &    1.0 \\
+$\Delta(\beta_{1_\mu})$   & -0.000 &  0.066 &  -0.109 &    0.107 &   19664 &    1.0 \\
+$\Delta(\beta_{2_\mu})$   &  0.004 &  0.011 &  -0.014 &    0.022 &   24478 &    1.0 \\
+$\Delta(\beta_{3_\mu})$   & -0.003 &  0.004 &  -0.010 &    0.003 &   12532 &    1.0 \\
+$\Delta(\sigma)$    &  0.178 &  0.040 &   0.112 &    0.242 &   28446 &    1.0 \\
+\bottomrule
+\end{tabular}
+\label{exp1_parameter_estimates}
+\end{threeparttable}
+\end{center}
+\end{table}
+
 \subsection{Experiment 2 regression model}
@@
 
+\begin{table}
+\begin{center}
+\begin{threeparttable}
+\caption{Experiment 2 posterior parameter estimates}
+\footnotesize
+\begin{tabular}{lrrrrrr}
+\toprule
+ &   Mean &     SD &  HDI (5\%) &  HDI (95\%) &  ESS &  $\hat{R}$ \\
+\midrule
+\multicolumn{7}{c}{Transmission-only} \\
+$\alpha_{1_\mu}$ &  0.099 &  0.066 &  -0.009 &    0.205 &   50911 &    1.0 \\
+$\alpha_{2_\mu}$ &  0.530 &  0.134 &   0.320 &    0.748 &   58969 &    1.0 \\
+$\alpha_{3_\mu}$ &  1.113 &  0.155 &   0.869 &    1.370 &   55884 &    1.0 \\
+$\beta_{1_\mu}$ &  0.048 &  0.044 &  -0.021 &    0.121 &   32489 &    1.0 \\
+$\beta_{2_\mu}$ &  0.216 &  0.074 &   0.096 &    0.336 &   47941 &    1.0 \\
+$\beta_{3_\mu}$ & -0.000 &  0.040 &  -0.065 &    0.065 &   32801 &    1.0 \\
+$\sigma$      &  0.145 &  0.018 &   0.116 &    0.173 &   15490 &    1.0 \\
+\midrule
+\multicolumn{7}{c}{Transmission + Communication} \\
+$\alpha_{1_\mu}$ &  0.052 &  0.042 &  -0.018 &    0.118 &   30904 &    1.0 \\
+$\alpha_{2_\mu}$ &  0.297 &  0.102 &   0.138 &    0.469 &   56334 &    1.0 \\
+$\alpha_{3_\mu}$ &  0.679 &  0.184 &   0.386 &    0.981 &   63206 &    1.0 \\
+$\beta_{1_\mu}$ &  0.033 &  0.044 &  -0.040 &    0.102 &   18241 &    1.0 \\
+$\beta_{2_\mu}$ &  0.093 &  0.073 &  -0.027 &    0.209 &   39754 &    1.0 \\
+$\beta_{3_\mu}$ &  0.167 &  0.121 &  -0.033 &    0.359 &   53767 &    1.0 \\
+$\sigma$      &  0.175 &  0.020 &   0.141 &    0.207 &   17902 &    1.0 \\
+\midrule
+\multicolumn{7}{c}{Difference} \\
+$\Delta(\alpha_{1_\mu})$     & -0.047 &  0.079 &  -0.178 &    0.077 &   40911 &    1.0 \\
+$\Delta(\alpha_{2_\mu})$     & -0.233 &  0.168 &  -0.504 &    0.040 &   55907 &    1.0 \\
+$\Delta(\alpha_{3_\mu})$     & -0.434 &  0.241 &  -0.825 &   -0.037 &   57860 &    1.0 \\
+$\Delta(\beta_{1_\mu})$     & -0.016 &  0.062 &  -0.118 &    0.084 &   23610 &    1.0 \\
+$\Delta(\beta_{2_\mu})$     & -0.124 &  0.105 &  -0.295 &    0.047 &   35386 &    1.0 \\
+$\Delta(\beta_{3_\mu})$     &  0.167 &  0.128 &  -0.045 &    0.371 &   50705 &    1.0 \\
+$\Delta(σ)$      &  0.030 &  0.027 &  -0.014 &    0.074 &   16452 &    1.0 \\
+\bottomrule
+\end{tabular}
+\label{exp2_parameter_estimates}
+\end{threeparttable}
+\end{center}
+\end{table}
+
 %~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%

### 2023-09-19 manuscript/wp4.tex
ms 19/09

@@
 
-Examples from other languages Irish orthography has special features to handle eclipsis.
-For example, when the word /pɑʃti/ (children) is eclipsed, it becomes /bɑʃti/ (the initial consonant is vocalized), but the spelling system retains the letter <p> \textit{in addition to} adding a <b>: <páistí> becomes <bpáistí> under eclipsis.
-This allows the reader quick access to the root word from which the eclipsed form is derived, while also showing the reader how to pronounce the eclipsed form.
-Moreover, in some cases, it results in ambiguity avoidance---in this case, if the <p> were not retained, the spelling <báistí> would be confusable with the word for rain.
+Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages.
+For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis.
+The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>.
+Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.
+
+Even the most orthographically transparent of languages have at least some techniques for disambiguating homophones in writing.
+Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).
+Many other languages employ various diacritics for the same purpose, such as the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).
 
@@
 Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time.
-Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.
+Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.
 
@@
 There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7).
-For a statistical comparison across conditions, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
+To quantify our uncertainty about these distributions and compare conditions statistically, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
 	\begin{equation}
@@
 
-Finally, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$.
+Finally, cost may be converted into the expected probability of communicative success\footnote{Note that we can compute this quantity regardless of which condition a participant was in.
+By expected probability of communicative success, we mean the level of success a pair of interlocutors could hypothetically achieve using a given system, regardless of whether that system emerged in the Transmission-only or Transmission~+ Communication condition.
+This also assumes that the hypothetical interlocutors have a perfect understanding of the same shared system.} with the equation $p=2^{-c}$.
 If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: [35.1, 46.3]) in Transmission-only and 50.6\% (90\% HDI: [45.9, 54.9]) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition.
-Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that an expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use, while also being fairly easy to learn.
+Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that a degenerate system would result in a communicative success rate of~33.3\% (i.e., chance-level performance), while an expressive system would result in a~100\% success rate, while also being fairly easy to learn.
 
@@
 A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}).
-As we will see in the following experiment, it is indeed possible for participants to learn expressive systems, even under full homophony; the lack of such outcomes in the present experiment suggests a reluctance to defy the spoken language.
 
@@
 Interestingly, although participants were quite resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual.
-Tradition has a powerful
+Tradition has a powerful hold over writing systems.
 
@@
 
-\begin{itemize}
+Writing---particularly writing in alphabetic scripts---is usually viewed as a visual representation of speech: Each visual letter corresponds to a sound.
+However, writing systems are known to diverge from their spoken counterparts in various ways.
+The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech.
+It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality.
+In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
+In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language, if that divergence confers some benefit in reading.
 
-	\item Summary of theoretical idea... hard to test in the real world, but we can experimentally simulate the cultural evolution of spelling
+We tested this hypothesis by focusing on the particular case of heterographic homophones---words that sound the same but which are spelled differently.
+Heterographic homophones permit the written language to be more informative than the spoken language.
+The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information, either in the form of context or paraphrasis.
+In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}).
+Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist, such as x, y, z, and, moreover, the etymological cases of heterography may, in part, owe their own success to the differentiation they inadvertently provide.
 
-	\item Recapitulation of main findings
+We put these ideas to the test by experimentally simulating the cultural evolution of orthography.
+In Experiment~1, we focused on the differentiation mechanism: Might variant spellings be used to differentiate meanings that are otherwise identical in speech?
+If so, we would expect levels of spelling differentiation to be greater when there is greater pressure for disambiguation, which we induced through the addition of a communication game.
+Importantly, the initial randomly generated orthographic systems that we used to seed the transmission chains contained high variation---that is, multiple ways of spelling the same sound.
+We included this variation because, for the differentiation mechanism to be viable, the writing system has to be receptive to spelling variation.
+A writing system that does not permit a one-to-many phoneme-to-grapheme mapping would not be capable of differentiating homophonous words.
+Only when variation is permitted and available can variant spellings be conditioned on meaning.
 
-	\item What can this tell us about real-world processes?
+Although we found evidence to suggest that the written forms were slightly more informative than the corresponding spoken forms, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
+This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but conditioned on shape rather than color.
+Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems.
+Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors: First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech.
+Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
+Secondly, even when participants appreciate the need to differentiate the written forms to be successful in the task, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
+It is notable, for example, that in the communication games, many attempts to differentiate with English color letters were not reciprocated.
 
-	\item Cannot directly compare experiments because it's easier to maintain than create.
-However, this is also true of the real world, pointing to a general advantage of conservation over differentiation.
+In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation.
+Under this mechanism, an expressive orthography does not emerge to fulfill a role but rather it is simply maintained from an earlier form of the spoken language that was expressive.
+This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}.
+That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may, at least in part, have been preserved precisely because of the informativeness they provide in reading.
+If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
 
-	\item The fact that the results are less "significant" in conservation does not mean that conservation is a worse model; it means that, under the conservation mechanism, communicative pressure is less important.
-This might suggest an accident-of-history explanation: Learning alone (i.e., habitual, mindless replication) can explain the emergence of informative orthography.
-Informative orthography does not arise through adaption, but may instead be the result of mindlessly replicated historic residue following sound change.
+Our findings showed that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation.
+There are two important caveats, however.
+First, we found that cultural transmission alone---that is, blind learning and reproduction---will result in at~least some conservation, not only in form but also in the conditioning of form on meaning.
+Cases such as Chain~D correspond to the ``accident of history'' explanation: Expressive orthography is preserved not because it serves any useful purpose (recall that in Transmission-only there is no functional need for the language to be informative), but because participants are simply reproducing what they learned, and what they learned has not (yet) placed a significant enough burden on learning for simplification to kick~in.
+The additional level of conservation that occurs in Transmission~+ Communication corresponds to the exaptation/repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly).
+The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure.
+Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely that all chains will ultimately undergo the same transformation eventually.
+Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.
 
-	\item Limitations
+It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effect to keep the two as close as possible.
+Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2.
+The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm.
+Similarly, we cannot draw direct conclusions from the relative strength of the results across the two experiments.
+For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate.
+Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation.
+For differentiation to operate, numerous hurdles must be overcome: participants must recognize the problem they are faced with, they must be willing to introduce and tolerate spelling variation, they must able to align with an interlocutor separated in time and space, and they must be willing to rebel against their input.
+Conserving an expressive orthography is, by comparison, a relatively simple affair.
+If conservation is a simpler (and therefore, by Occam's razor, more plausible) mechanism in our experimental setup, then presumably the same is true of the real world.
 
-	\item Future directions. Remove the spoken forms from the test phase to simulate a silent-reader, literate society. This might make a dramatic difference to the results.
+Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of the relevant real-world processes.
+Other than the general scaling-down of the x to an experimentally tractable
 
-	\item Conclusion: perhaps there are potentially some benefits to heterography.
+Communication game not a good model of written communication. link to texting which is written but near-synchronous.
 
-\end{itemize}
+color morphs a bit weird
+
+do not consider top-down authority
+
+unclear whether disambiguation matters in the real world. context usually makes things clear.
+
+Nevertheless, even very small biases can be amplified by cultural transmission.
+
+We foresee two interesting directions in which this work might be taken in the future.
+The first is a relatively simple extension of the present experiments.
+By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality.
+This minor change might have a dramatic effect on the results: Participants may be less inclined to simply transcribe how a word sounds and more inclined to differentiate and conserve.
+Such an experiment is not without pitfalls, however.
+Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it.
+Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.
+
+A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on language change.
+For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism.
+Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that ultimately become heterographic.
+
+As a final note, heterography is often derided as a source of unnecessary complexity that language learners must contend with.
+However, our results show that, when the communicative need for disambiguation is strong enough, language users may prefer to tradeoff some simplicity for greater informativeness in the writing system.
+Perhaps, then, efforts to reform and simplify spelling may, to some extent, be a little hasty, and more recognition should be given to the benefits that opaque orthographies provide in the context of reading.
+
+%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%
+
+\section{Acknowledgments}
+
+\noindent This work was funded by a grant from the Leverhulme Trust.
+
+\printbibliography
+
+%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%
+
+\clearpage
 
@@
 	\end{equation}
+
+\subsection{Experiment 2 regression model}
+
+	\begin{equation}
+	\begin{aligned}
+	   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+	 \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
+	\sigma                                 & \sim \mathrm{Exponential}(1) \\
+	\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
+	\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
+	\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
+	\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
+	\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
+	\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
+	\end{aligned}
+	\end{equation}
+
+\clearpage
 
@@
 
-\subsection{Experiment 2 regression model}
-
-	\begin{equation}
-	\begin{aligned}
-	   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-	 \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
-	\sigma                                 & \sim \mathrm{Exponential}(1) \\
-	\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
-	\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
-	\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
-	\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
-	\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
-	\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
-	\end{aligned}
-	\end{equation}
+\clearpage
 
@@
 
-%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%~%
-
-\section{Acknowledgments}
-
-\noindent This work was funded by a grant from the Leverhulme Trust.
-
-\printbibliography
-
 \end{document}

### 2023-09-27 manuscript/wp4.tex
ms 20/09

@@
 \noindent
-Written languages represent spoken languages.
-However, there are many ways in which written languages diverge.
-E.g.
-spacing between words helps readers parse a sentence even if they are not technically required.
-Written languages can also have their own regularities such as spelling related words similarly (magician), past tense spellings directly conveying meaning, certain spellings conveying word class ous.
-These features abound in writing, especially in languages such as English that have not been subject to serious regulation.
+Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each written letter corresponding to a distinct sound.
+However, writing systems are known to diverge from their spoken counterparts.
+The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech.
+It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality.
+In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
+In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading.
 
 One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/).
-Heterographic spellings such as there can be useful in reading.
+Heterographic spellings such as these can serve a valuable function in reading.
 For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
 In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax.
-Heterography may, therefore, confer certain benefits in reading, particularly in terms of comprehension.
-Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart, avoiding ambiguity.
+Heterography may, therefore, confer certain benefits in reading, permitting a writing system to be---at least in some areas---more informative than its spoken counterpart.
+Although such divergence is common in the more orthographically opaque languages, such as English, even the most transparent of languages have at least some techniques for disambiguating homophones in writing.
+Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).
 
-Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages.
+% The same is true of, for example, the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).
+
+% Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages.
 For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis.
@@
 
-Even the most orthographically transparent of languages have at least some techniques for disambiguating homophones in writing.
-Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).
-Many other languages employ various diacritics for the same purpose, such as the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).
-
-However, this all sounds great but additional informativeness comes at a cost.
+However, the pursuit of additional informativeness in orthography comes at a cost.
 A system of words that conveys more information will---in general---be more complex than one that conveys less information.
@@
 A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
-The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2015} studies.
-Of particular note here is the idea that the pressure for simplicity in language structure derives principally from learning, while the pressure for informativeness derives from the need for precise communication.
+An informative spelling system, on the other hand, is one that precisely conveys meaning.
+The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2008, Kirby:2015} studies.
+Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication.
+An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness in use that is outweighed by the cost of learning arbitrary spelling distinctions.
 
-Unpack some of this literature a bit.
-
-Any explanation for heterographic homophones must, therefore, demonstrate that heterographic spellings convey a benefit in terms of informativeness that is outweighed by the cost of learning the (often arbitrary) spelling distinction.
-What are the benefits?
-There seem to be broadly two kinds of benefit.
+% There seem to be broadly two kinds of benefit.
 (1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently.
@@
 
-Is it possible, then, that heterography may be selected for in the cultural evolution of orthography?
-And if so, what might the mechanism be through which such selection could operate?
-\textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
+This leads to the intriguing question of whether heterography might be selected for in the cultural evolution of orthography and, if so, the mechanisms through which this selection could occur.
+\textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
 The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms.
@@
 Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc.
-Heterography is also very common in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT}.
+Heterography is also commonly employed in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}).
 
@@
 
-In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008}.
-In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production.
-Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
+In this paper, we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}.
+In this paradigm, an artificially constructed language is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language.
+Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from.
 Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
@@
 
-Bottleneck
+One important component of the iterated learning paradigm is the \textit{bottleneck on transmission}, a limit on the amount of information flowing from one generation to the next.
+Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language.
+An important aspect of our studies is how learners respond to this bottleneck---how they choose to generalize.
 
-Preregistration: \url{https://aspredicted.org/YLG_CKD}
+Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that the heterographic representation of meanings expressed homophonously in speech arises preferentially under a communicative need for disambiguation.
+All data and code is available from \url{https://osf.io}.
+Both experiments were preregistered at \url{https://aspredicted.org/YLG_CKD}.
 
@@
 To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}.
-In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production.
-Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from.
-To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.
-
-[make this more specific to the present experiment - what is passed down the chain - preview the methods coming up]
+Each participant is first asked to learn a simple ``alien'' language---consisting of words for colored shapes---and is then asked to reproduce what they learned in a test phase.
+To explore the hypotheses outlined above, we conducted the experiment under two different conditions: Transmission-only, in which the test phases emphasizes simple reproduction, and Transmission~+ Communication, in which the test phase encourages disambiguation.
 
@@
 
-The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation.
+The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, revealing the proportion of the 10 chains that fall into each typological category at each generation.
 In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2.
@@
 
-Writing---particularly writing in alphabetic scripts---is usually viewed as a visual representation of speech: Each visual letter corresponds to a sound.
-However, writing systems are known to diverge from their spoken counterparts in various ways.
-The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech.
-It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality.
-In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
-In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language, if that divergence confers some benefit in reading.
+It is widely understood that the written and spoken forms of a language are not identical; they diverge in many ways as a result of the differing constraints relevant to each modality.
+Spacing between words for example does not exist in speech, but constitutes a useful innovation in writing that permits rapid reading.
+Might it be the case, then, that the written form of a language will become, over time, better adapted to the needs of writers and readers to the detriment of alignment with the spoken form of the language?
+Do writing systems adapt to the reading niche?
 
-We tested this hypothesis by focusing on the particular case of heterographic homophones---words that sound the same but which are spelled differently.
+We addressed these questions by focusing on the particular case of heterographic homophones---words that sound the same in speech but which are spelled differently.
 Heterographic homophones permit the written language to be more informative than the spoken language.
-The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information, either in the form of context or paraphrasis.
+The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
 In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}).
-Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist, such as x, y, z, and, moreover, the etymological cases of heterography may, in part, owe their own success to the differentiation they inadvertently provide.
+Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist---words such as \textit{byte}, \textit{cheque}, and \textit{lite}---and, moreover, the etymologically motivated cases of heterography may, in part, owe their continued success to the differentiation they inadvertently provide.
 
@@
 
-Although we found evidence to suggest that the written forms were slightly more informative than the corresponding spoken forms, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
-This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but conditioned on shape rather than color.
+Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
+This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color.
 Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems.
@@
 Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
-Secondly, even when participants appreciate the need to differentiate the written forms to be successful in the task, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
-It is notable, for example, that in the communication games, many attempts to differentiate with English color letters were not reciprocated.
+Secondly, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
+It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
 
 In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation.
-Under this mechanism, an expressive orthography does not emerge to fulfill a role but rather it is simply maintained from an earlier form of the spoken language that was expressive.
+Under this mechanism, an expressive orthography does not emerge to fulfill an adaptive role but rather it is simply maintained from an earlier form of the spoken language that was expressive.
 This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}.
-That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may, at least in part, have been preserved precisely because of the informativeness they provide in reading.
+That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may have been preserved, at least in part, precisely because of the informativeness they provide in reading.
 If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
 
-Our findings showed that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation.
+Our findings did indeed show that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation.
 There are two important caveats, however.
@@
 
-Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of the relevant real-world processes.
-Other than the general scaling-down of the x to an experimentally tractable
+Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of real-world processes.
+Other than the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality.
+We follow prior studies by using a real-time communication game, but this is perhaps not very representative of the dynamics involved in asynchronous written communication (although, cf.\@ text messaging).
+Perhaps more importantly, it is unclear whether lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes things clear.
+If \textit{knight} and \textit{night} were spelled the same way, it is hard to imagine a context in which they might be confused.
+That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography.
+An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place adaptive pressure on a writing system that is primarily shaped by the needs and preferences of writers.
 
-Communication game not a good model of written communication. link to texting which is written but near-synchronous.
+% color morphs a bit weird
 
-color morphs a bit weird
-
-do not consider top-down authority
-
-unclear whether disambiguation matters in the real world. context usually makes things clear.
-
-Nevertheless, even very small biases can be amplified by cultural transmission.
+% do not consider top-down authority
 
 We foresee two interesting directions in which this work might be taken in the future.
-The first is a relatively simple extension of the present experiments.
+The first is a simple extension of the present experiments.
 By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality.
@@
 For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism.
-Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that ultimately become heterographic.
+Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that have ultimately become heterographic in the present-day language.
 

### 2023-10-30 manuscript/wp4.tex
ms 27/10

@@
 \PassOptionsToPackage{hidelinks}{hyperref}
+
 \documentclass[doc,biblatex]{apa7}
+
+\usepackage{amsmath}
+\usepackage{mathspec}
+
 \DeclareLanguageMapping{american}{american-apa}
 \addbibresource{references.bib}
+
 \captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
-\usepackage{amsmath}
-\usepackage{mathspec}
 \setallmainfonts{Times New Roman}
@@
 
-\title{Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography}
+\title{Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography}
 
-\shorttitle{The cultural evolution of informative orthography}
+\shorttitle{Why do languages tolerate heterography?}
 
-\authorsnames[1,1]{Jon W. Carr, Kathleen Rastle}
+% PREPRINT VERSION
+\authorsnames{Jon W. Carr, Kathleen Rastle}
 
@@
 
-% \authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023). Informativity in the writing system. \textit{OSF Preprints}. Version~0.}
+\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023).
+Why do languages tolerate heterography?
+An experimental investigation into the emergence of informative orthography.
+\textit{OSF Preprints}.
+Version~1.
+https://doi.org/10.31219/osf.io/r5n3g}
+
+% SUBMISSION VERSION
+% \authorsnames{Anonymous}
+
+% \authorsaffiliations{{Anonymous}}
 
@@
 It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading.
-For example, the heterographic homophones (e.g., <knight> vs.
+For example, heterographic homophones (e.g., <knight> vs.
 <night>) impose additional costs in learning but reduce ambiguity in reading.
-Could it be the case, then, that heterography---and therefore orthographic opacity---may sometimes be selected for in the evolution of writing systems, despite the cost in learning?
-We investigate this question by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity.
-In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises through exaptation as an epiphenomenon of sound change (e.g., <meat> vs.
+If correct, this would suggest that heterography---and therefore orthographic opacity---might sometimes be selected for in the evolution of writing systems, despite the cost in learning.
+We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity.
+In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> vs.
 <meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift).
@@
 \noindent
-Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each written letter corresponding to a distinct sound.
+Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each letter corresponding to a distinct sound.
 However, writing systems are known to diverge from their spoken counterparts.
-The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech.
-It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality.
+The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}.
+It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}.
 In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
-In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading.
+In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading \parencite{Ulicheva:2020}.
 
-One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/).
-Heterographic spellings such as these can serve a valuable function in reading.
-For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
-In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax.
-Heterography may, therefore, confer certain benefits in reading, permitting a writing system to be---at least in some areas---more informative than its spoken counterpart.
-Although such divergence is common in the more orthographically opaque languages, such as English, even the most transparent of languages have at least some techniques for disambiguating homophones in writing.
+One potential case of functional divergence is heterographic homophones---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}).
+Heterographic spellings such as these may serve a valuable function in reading.
+For example, faced with a spoken sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr kat/, a form of the verb \textit{to~be}, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
+In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content.
+Heterography permits the written form of the language to be more informative than the spoken form.
+Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing.
 Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).
 
-% The same is true of, for example, the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).
+Heterographic homophony can also occur at the morpheme level.
+The suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing.
+Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English morphemes \parencite{Berg:2017} and it has been shown that English readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.
 
-% Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages.
-For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis.
-The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>.
-Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.
+However, heterography comes with two main costs.
+Firstly, by definition, heterography implies that a single sound can be spelled multiple ways.
+The heterographic spellings <meat> and <meet>, for example, imply that the /i/ sound can be spelled <ea> or <ee>.
+Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}.
+Secondly, the mapping between heterographic forms and meaning must also be learned.
+From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around.
+Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}.
 
-However, the pursuit of additional informativeness in orthography comes at a cost.
-A system of words that conveys more information will---in general---be more complex than one that conveys less information.
-There exists a tradeoff between the simplicity of the spelling system and its ability to be expressive.
-A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
-An informative spelling system, on the other hand, is one that precisely conveys meaning.
-The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2008, Kirby:2015} studies.
-Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication.
-An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness in use that is outweighed by the cost of learning arbitrary spelling distinctions.
+An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness that outweighs its associated costs.
+In general, a system of words that conveys more information will be more complex than one that conveys less information; there exists a tradeoff between the simplicity of the spelling system and its ability to be expressive.
+A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
+An informative spelling system, on the other hand, would be one that precisely conveys meaning.
+The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in various typological \parencite[e.g.,][]{Kemp:2018} and experimental \parencite[e.g.,][]{Kirby:2015} studies.
+Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication, a point we return to shortly.
 
-% There seem to be broadly two kinds of benefit.
-(1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently.
-(2) permitting the cognitive system to rapidly and accurately attain the meaning of a word.
-In this paper we limit ourselves to informativeness of the first kind, which is more amenable to experimental testing.
-
-This leads to the intriguing question of whether heterography might be selected for in the cultural evolution of orthography and, if so, the mechanisms through which this selection could occur.
-\textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}.
-The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms.
-For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
-Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc.
-Heterography is also commonly employed in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}).
+First, however, it is useful to consider the mechanisms by which selection could occur if it is indeed the case that heterography emerges for functional reasons.
+\textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography.
+The first model, \textit{the differentiation model} (Fig.~\ref{models_of_heterography}A), explains heterography through the creation of new orthographic forms.
+For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
+Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by adding <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}.
+Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}--- and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}.
+One important way in which differentiation can occur---especially in a language like English that has historically contained a lot of spelling variation \parencite{Nevalainen:2012, Stenroos:2016}---is by the conditioning of variant spellings on meaning; pairs like \textit{discreet}--\textit{discrete}, \textit{flour}--\textit{flower}, and \textit{plain}--\textit{plane}, which were once variant spellings of the same word, have taken on distinct meanings over time \parencite[§~5.4]{Carney:1994}.
+\textcite[p.~58]{Berg:2017} have referred to this as the ``functionalization of leftovers'': The spelling variants that survive are those that ``can find distributional or functional niches.''
 
@@
 	\caption{Two models of heterography.
-In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
-In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
+\textbf{(A)} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
+\textbf{(B)} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
 Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
@@
 
-The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography.
-In English, for example, the meat--meet merger ultimately resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated accordingly, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc.
-Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of /ʍ/ into /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).
+The second model, \textit{the conservation model} (Fig.~\ref{models_of_heterography}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography.
+For example, the \textit{meat}--\textit{meet} merger that occurred during the Great Vowel Shift ultimately resulted in Middle English /ɛː/ (spelled <ea>) and /eː/ (spelled <ee>) being pronounced /iː/ in Early Modern English \parencite{Lass:2000}, but the spellings were never changed accordingly, thus giving rise to a set of heterographic homophones that persist in present-day English \parencite[pp.~140--141]{Wells:1982}: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}.
+The same is true of the \textit{pain}--\textit{pane} merger \parencite[pp.~141--142]{Wells:1982}: \textit{maid}--\textit{made}, \textit{main}--\textit{mane}, \textit{pain}--\textit{pane}, \textit{raise}--\textit{raze}, \textit{sail}--\textit{sale}, \textit{vain}--\textit{vane}.
+Sound changes involving consonants have also resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (e.g., \textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}), the loss of /ç/ (e.g., \textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}), and the merger of /ʍ/ into /w/ (e.g., \textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}).\footnote{Although we can never be entirely certain how words were pronounced before the advent of sound recording technology, historical linguists have compiled persuasive evidence by a variety of methods.
+Comparison to modern Dutch and German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with German \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with Dutch and German \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these words continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.
 
-In this paper, we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}.
-In this paradigm, an artificially constructed language is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language.
-Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from.
-Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}.
-For reviews, see \textcite{Tamariz:2017, Kirby:2014, ScottPhillips:2010}.
+In some cases, it is debatable whether a given case of heterography was delivered by the differentiation or conservation mechanism.
+For example, while the etymological (and folk-etymological) respellings introduced during the Renaissance might appear to be cases of conservation \parencite[the most notorious example being the replacement of <dout> with <doubt> to indicate the word's Latin derivation from \textit{dubitare};][p.~268]{Crystal:2005}, it has also been argued that such respellings were motivated in~part by a desire to differentiate homophones such as \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, and \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}.
+Nevertheless, regardless of the particular mechanism behind specific cases in English or any other language, our primary contention here is that both of these mechanisms provide adaptive, functional explanations for heterography.
+Differentiated spellings that prove communicatively useful will be more likely to survive; likewise, conserved spellings that prove communicatively useful will be more likely to survive.
 
-One important component of the iterated learning paradigm is the \textit{bottleneck on transmission}, a limit on the amount of information flowing from one generation to the next.
+Our aims in this paper are twofold.
+First, we test the idea that heterography emerges in response to a functional pressure to disambiguate meaning in writing.
+Second, we seek to understand how the emergence of heterography plays~out under the two mechanisms of differentiation and conservation.
+Is one of these a better candidate explanation than the other?
+Approaching these evolutionary questions using data from natural languages is challenging.
+In particular, the available diachronic data (for any language) will necessarily be limited and impoverished---languages do not fossilize well, especially in their spoken forms.
+In addition, any answer derived from such datasets will have to rely on correlational, as opposed to causal, evidence---we cannot rerun history many times under different conditions.
+
+We therefore turn to a different approach.
+Here we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}.
+In this paradigm, an artificially constructed language (or so-called ``alien language'') is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language.
+Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from, although the participants themselves are not aware of this generational structure.
+It has been demonstrated in a wide variety of studies that, after several generations of cultural transmission, artificial languages can gradually adapt to the biases of the human learners and the environments in which they are used, yielding emergent linguistic phenomena, such as compositionality \parencite{Beckner:2017, Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, semantic category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, regularization \parencite{Ferdinand:2019, Smith:2010}, and argument marking \parencite{Motamedi:2021}, among many other things.
+For reviews, see \textcite{Bailes:2023, Kirby:2014, Kirby:2017, Smith:2022, Tamariz:2017}.
+
+\textcite{Kirby:2008} described the first experimental application of the iterated learning framework (which had previously been confined to computational modeling), showing that compositional structure---a systematic relationship between recombinant linguistic units and meaning---could spontaneously emerge under a \textit{bottleneck on transmission}.
+This ``bottleneck'' defines a limit on the amount of information that can flow from one generation to the next \parencite{Brighton:2002}.
 Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language.
-An important aspect of our studies is how learners respond to this bottleneck---how they choose to generalize.
+Generalization from limited input is a big driver behind systematic structure in the language as a whole, since human learners tend to generalize in ways that increase the simplicity (i.e., systematicity) of the system, albeit unconsciously \parencite{Culbertson:2016}.
+However, left unchecked, this bias for simplicity would ultimately result in the emergence of maximally simple, degenerate languages.
+\textcite{Kirby:2015} therefore extended the framework by including a communicative task; instead of each generation consisting of a single participant, each generation now consisted of a pair of participants engaged in a shared task requiring communicative precision.
+Crucially, this communicative component prevented the artificial languages from degenerating; instead, the languages find a tradeoff between simplicity on the one hand and informativeness on the other, just as in natural language \parencite{Mollica:2021, KempRegier:2012, Kemp:2018, Regier:2015, Zaslavsky:2018}.
+We follow this experimental approach in order to compare what happens to orthography when a pressure for informativeness \textit{is} and \textit{is~not} present.
+[KATHY: I feel that someone could wrongly think that we're just replicating what's already been done but with spelling.
+Is there a way to remind readers here that we're looking at what pressures a spelling system to diverge from spoken language?
+the key in our experiment is that diverging from spoken language increases pressure on learning.]
 
-Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that the heterographic representation of meanings expressed homophonously in speech arises preferentially under a communicative need for disambiguation.
+Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation.
 All data and code is available from \url{https://osf.io}.
-Both experiments were preregistered at \url{https://aspredicted.org/YLG_CKD}.
+Both experiments were preregistered at \url{https://aspredicted.org/YDV_MXV}.
+Although we did not specify a precise statistical protocol (due to the more exploratory nature of our experiments), our preregistration does cover hypotheses, conditions, measured variables, sample size, and exclusion criteria.
 
@@
 
-Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography. In particular we had two main hypotheses:
+Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography [KATHY: Possible to remind readers that we mean divergence from spoken language i.e.
+making the orthography more complex?] (i.e., in the face of homophony).
+We had two main hypotheses:
+\begin{enumerate}
+	\item Under pressure from learning alone, we expect to see the emergence of an increasingly transparent orthography.
+	\item Under additional pressure for disambiguation, we expect to see greater use of differentiated, non-transparent spellings.
+\end{enumerate}
 
@@
 
-To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}.
-Each participant is first asked to learn a simple ``alien'' language---consisting of words for colored shapes---and is then asked to reproduce what they learned in a test phase.
+Our methods follow the experimental iterated learning literature, as described above, with one main difference: The artificial language has both spoken and written forms that may diverge or converge over time.
+Broadly, participants are first asked to learn a simple alien language (consisting of words for colored shapes) and are then asked to reproduce what they learned in a test phase.
+The written form of the language may change over time, since the orthographic output of participant $i$ becomes the input to participant $i+1$ in a transmission chain design, but the spoken form of the language remains fixed and under experimenter control.
 To explore the hypotheses outlined above, we conducted the experiment under two different conditions: Transmission-only, in which the test phases emphasizes simple reproduction, and Transmission~+ Communication, in which the test phase encourages disambiguation.
@@
 	\vspace*{2pt}
-	\caption{The nine alien object stimuli with their stems and suffixes.
+	\caption{The nine object stimuli with their stems and suffixes.
 The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
@@
 
-Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}.
+Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}.
 The alien words had a spoken and written form composed of a stem and suffix.
@@
 Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system, which was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>.
-In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity.
+In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contained no particular regularity.
 This procedure models an initial state of high spelling variation (every object has a unique suffix spelling), but these spellings may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning).
@@
 	\vspace*{2pt}
-	\caption{(A) Transmission-only procedure.
+	\caption{\textbf{(A)} Transmission-only procedure.
 Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items.
@@
 The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent.
-(B) Transmission~+ Communication procedure.
+\textbf{(B)} Transmission~+ Communication procedure.
 Each generation now consists of two participants who engage in a communicative task.
@@
 Transmission continued for nine generations in each of ten independent chains.
-In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant; determined by whichever participant started the experiment first) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend).
-The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.
+In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consisted of a pair of participants, but the productions of only one of the two (the primary participant; determined by whichever participant started the experiment first) were iterated to the next generation.
+The productions of the secondary participant were not iterated any further and were thus a cultural deadend; the role of the secondary participant was to act as a genuine communicative partner for the primary participant, inducing pressure for the language to become informative.
 
@@
 The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target.
-Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.
+Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different.
+In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.
 
@@
 The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation.
-We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simple repeats whichever shape was marked by the stem.
+We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simply repeats whichever shape was marked by the stem.
 Redundant suffix systems are characteristic of the Transmission-only condition, with similar outcomes occurring in Chains~C, E, F, and~H.
@@
 The holistic\footnote{Our use of the term \textit{holistic} is slightly unusual here, owing to the fact that we are focused on the suffix level and not the word level.
-By holistic, we only mean that each shape--color combination has a unique suffix form, so the nine suffixes must all be learned as separate cases.
+By holistic, we only mean that each shape--color combination has a unique suffix form.
 We do not distinguish between truly holistic suffixes (nine unique suffixes with no structure in how they relate to each other) and compositional suffixes (nine unique suffixes that can be generated from compositional rules, as was the case in, for example, Q8).} system has nine unique forms each of which expresses a particular color--shape combination.
@@
 
-As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}.
+As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}.
 This measure expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
@@
 	\vspace*{2pt}
-	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+	\caption{\textbf{(A)} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
 The thin lines show each individual transmission chain (with a small amount of added jitter).
@@
 In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+\textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_dif}
@@
 Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue).
-Such differentiation was, however, not very forthcoming, even under communicative pressure.
+Such differentiation was, however, not forthcoming, even under communicative pressure.
 Instead, when spelling variation was conditioned on anything, it was usually conditioned on shape, resulting in redundant suffix spellings.
@@
 As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
-A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}).
+A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
 
@@
 These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space.
-Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>).
-Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the distinction?
+Additionally, deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>).
+Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the intention?
 Differentiation is hard in the written modality.
@@
 And, importantly, does this happen preferentially in the presence of communicative pressure?
+Our hypotheses were as follows:
+\begin{enumerate}
+	\item Under pressure from learning alone, we expect to find that orthography will track the spoken form of the language, becoming increasingly degenerate as homophony increases.
+	\item Under additional pressure for disambiguation, we expect the orthography to conserve archaic (but informative) spelling distinctions even after these distinctions cease to exist in the spoken form of the language.
+\end{enumerate}
 
@@
 Crucially, the spellings did not automatically change following a sound change event; rather, the orthographic system was free to adapt (or not) in response to the sound changes.
-Note also that an individual participant did not directly experience the sound changes; a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
+Note also that the participants did not directly experience the sound changes; a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
 In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.
@@
 These processes continue into Epoch~III and by the ninth generation, all systems have become degenerate, redundant, or some mixture of the two.
-The one exception is Chain~D, whose original spellings---<sei>, <fa>, <xoe>---were conserved perfectly through to the final generation with only one generalization error in Generation~8, which was quickly reverted in Generation~9.
+The one exception is Chain~D, whose original spellings were conserved perfectly through to the final generation with only one generalization error in Generation~8, which was quickly reverted in Generation~9.
 
@@
 In general, the orthographic systems tend to remain more informative under communicative pressure.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+\textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_con}
@@
 
-Experiment~1 asked whether an informative, heterographic orthography may be \textit{created} de~novo under homophony.
+Experiment~1 asked whether an informative, heterographic orthography may be \textit{created} de~novo under pressure from homophony.
 Experiment~2, by contrast, asked whether an informative, heterographic orthography can simply be \textit{maintained}, even under the same levels of homophony encountered in Experiment~1.
@@
 It is widely understood that the written and spoken forms of a language are not identical; they diverge in many ways as a result of the differing constraints relevant to each modality.
-Spacing between words for example does not exist in speech, but constitutes a useful innovation in writing that permits rapid reading.
+Spacing between words for example does not exist in speech, but constitutes a useful innovation in writing that permits rapid reading \parencite{Rastle:2019, RaynerFischer:1998, Sainio:2007, Zang:2013}.
 Might it be the case, then, that the written form of a language will become, over time, better adapted to the needs of writers and readers to the detriment of alignment with the spoken form of the language?
-Do writing systems adapt to the reading niche?
+Do writing systems adapt to the affordances and constraints of the written modality?
 
-We addressed these questions by focusing on the particular case of heterographic homophones---words that sound the same in speech but which are spelled differently.
-Heterographic homophones permit the written language to be more informative than the spoken language.
-The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
-In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}).
+We addressed these questions by focusing on the particular case of heterographic homophones---words that sound the same but are spelled differently.
+Heterographic homophones permit the written language to be more informative than the spoken language; the written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
+In English, at~least, which has a tendency to preserve etymology in spelling, heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, \textit{wright}, and \textit{write} are spelled differently due to their separate derivations (Old English \textit{riht}, Latin \textit{ritus}, Old English \textit{wryhta}, and Old English \textit{writan}).
 Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist---words such as \textit{byte}, \textit{cheque}, and \textit{lite}---and, moreover, the etymologically motivated cases of heterography may, in part, owe their continued success to the differentiation they inadvertently provide.
@@
 Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems.
-Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors: First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech.
+Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors.
+First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech.
 Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
-Secondly, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
+Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
 It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
@@
 
+[JON: something about the cultural ratchet.
+with each successive sound change, new heterography appears, but once it's present, it's hard to let go of it.
+Thus, the informativeness of the writing system ratchets up with successive sound changes.]
+
 Our findings did indeed show that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation.
@@
 The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure.
-Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely that all chains will ultimately undergo the same transformation eventually.
+Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely [KATHY: on what basis?] that all chains will ultimately undergo the same transformation eventually.
 Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.
 
-It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effect to keep the two as close as possible.
+It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible.
 Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2.
@@
 Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation.
-For differentiation to operate, numerous hurdles must be overcome: participants must recognize the problem they are faced with, they must be willing to introduce and tolerate spelling variation, they must able to align with an interlocutor separated in time and space, and they must be willing to rebel against their input.
-Conserving an expressive orthography is, by comparison, a relatively simple affair.
-If conservation is a simpler (and therefore, by Occam's razor, more plausible) mechanism in our experimental setup, then presumably the same is true of the real world.
+For differentiation to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming various social stigmas in the process (the social problem).
+In~comparison, the conservation of an expressive orthography is plain~sailing.
+The only real hurdle is that the system of non-transparent spellings must be sufficiently learnable (the learnability problem).
+In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.
 
-Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of real-world processes.
-Other than the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality.
-We follow prior studies by using a real-time communication game, but this is perhaps not very representative of the dynamics involved in asynchronous written communication (although, cf.\@ text messaging).
-Perhaps more importantly, it is unclear whether lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes things clear.
+These conclusions must be interpreted within the limitations of these experiments, which are, after all, highly simplified simulacra of real-world processes.
+Besides the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality.
+We follow a large body of recent studies by using a real-time communication game \parencite[e.g.,][]{Carr:2017, Kanwal:2017, Kirby:2015, Raviv:2018, Saldana:2019, Silvey:2019, Winters:2015}, but such games are not very representative of the dynamics involved in asynchronous written communication \parencite[although see][for some approaches]{Winters:2019}.
+That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as repair, stress, and emotion, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.
+
+Another limitation of this work is the extent to which lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes the meaning clear.
 If \textit{knight} and \textit{night} were spelled the same way, it is hard to imagine a context in which they might be confused.
 That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography.
-An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place adaptive pressure on a writing system that is primarily shaped by the needs and preferences of writers.
+An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers.
 
-% color morphs a bit weird
-
-% do not consider top-down authority
+It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many.
+For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space.
+Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects.
 
@@
 
-A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on language change.
-For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism.
-Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that have ultimately become heterographic in the present-day language.
+A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on change in natural languages.
+For example, by tracing the development of heterographic homophones in various languages, we might be able to determine which of the models offer a better fit to empirical data, or indeed classify individual cases as being derived through the differentiation or conservation mechanism.
+Perhaps the best way to proceed here would be to develop formal models of the two mechanisms and test them against diachronic data on the pronunciation and spelling of words that have ultimately become heterographic in the present-day.
 
 As a final note, heterography is often derided as a source of unnecessary complexity that language learners must contend with.
-However, our results show that, when the communicative need for disambiguation is strong enough, language users may prefer to tradeoff some simplicity for greater informativeness in the writing system.
+However, our results show that, when the communicative need for disambiguation is strong enough, language users may prefer to trade some simplicity for greater informativeness in the writing system.
 Perhaps, then, efforts to reform and simplify spelling may, to some extent, be a little hasty, and more recognition should be given to the benefits that opaque orthographies provide in the context of reading.
@@
 
-\noindent This work was funded by a grant from the Leverhulme Trust.
+\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034) awarded to K.R.
 

### 2023-10-30 code/comm_cost.py
Make figures with a,b labels

@@ -18,7 +18,7 @@ def communicative_cost(lexicon, dims):
 	for meaning, signal in lexicon.items():
 		reverse_lexicon[signal].add(meaning)
 	U = product(*[range(n_values) for n_values in dims])
-	U_size = np.product(dims)
+	U_size = np.prod(dims)
 	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
 
 

### 2023-10-30 code/figures.py

@@ -3,6 +3,7 @@ import numpy as np
 import arviz as az
 import matplotlib.pyplot as plt
 import matplotlib.gridspec as gridspec
+import matplotlib.transforms as transforms
 from matplotlib import colormaps
 from utils import json_load
 
@@ -202,8 +203,10 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 		axis.set_xticks(range(10))
 		axis.set_xticks(range(10))
 		axis.set_ylim(-0.15, 1.75)
+		
 		if i == 0:
 			axis.set_ylabel('Communicative cost (bits)')
+			axis.text(0.1, 1.0, '(A)', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis.transAxes), size=8, font='Arial', weight='bold')
 		else:
 			axis.set_yticklabels([])
 
@@ -252,6 +255,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 		if i == 0:
 			axis_post.set_ylabel('Posterior')
 			axis_diff.set_ylabel('Posterior difference')
+			axis_post.text(0.1, 1.0, '(B)', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis_post.transAxes), size=8, font='Arial', weight='bold')
 
 		if variable['diff']:
 
@@ -357,7 +361,7 @@ if __name__ == '__main__':
 
 	# plot_transmission_chains(exp_data, 
 	# 	condition='con_com',
-	# 	output_path=FIGS / 'con_com.eps',
+	# 	output_path=FIGS / 'con_com.svg',
 	# )
 
 	# plot_typological_distribution(exp_data,
@@ -418,11 +422,11 @@ if __name__ == '__main__':
 	# 	latex_table=True,
 	# )
 
-	model_summary(
-		model_trace=DATA / 'exp2_cost.netcdf',
-		variables=['α_m', 'β_m', 'σ', 'diff_α1', 'diff_α2', 'diff_α3', 'diff_β1', 'diff_β2', 'diff_β3', 'diff_σ'],
-		latex_table=True,
-	)
+	# model_summary(
+	# 	model_trace=DATA / 'exp2_cost.netcdf',
[…]

### 2023-10-30 code/comm_cost.py
Clean up old cruft

@@ -80,14 +80,8 @@ if __name__ == '__main__':
 		(2, 2): 'wopiqo',
 	}
 
-	print(communicative_cost(degenerate, (3, 3)))
-	print(communicative_cost(holistic, (3, 3)))
-	print(communicative_cost(redundant, (3, 3)))
-	print(communicative_cost(expressive, (3, 3)))
-	print(communicative_cost(semi_expressive, (3, 3)))
-
-
-	
-	
-
-
+	print('degenerate', communicative_cost(degenerate, (3, 3)))
+	print('holistic', communicative_cost(holistic, (3, 3)))
+	print('redundant', communicative_cost(redundant, (3, 3)))
+	print('expressive', communicative_cost(expressive, (3, 3)))
+	print('semi_expressive', communicative_cost(semi_expressive, (3, 3)))

### 2023-10-30 code/complexity.py

@@ -1,134 +1,0 @@
-transparent = {
-	"0_0": "buviko",
-	"0_1": "buviko",
-	"0_2": "buviko",
-	"0_3": "buviko",
-	"1_0": "zetiko",
-	"1_1": "zetiko",
-	"1_2": "zetiko",
-	"1_3": "zetiko",
-	"2_0": "gafiko",
-	"2_1": "gafiko",
-	"2_2": "gafiko",
-	"2_3": "gafiko",
-	"3_0": "wopiko",
-	"3_1": "wopiko",
-	"3_2": "wopiko",
-	"3_3": "wopiko"
-}
-redundant = {
-	"0_0": "buvikoh",
-	"0_1": "buvikoh",
-	"0_2": "buvikoh",
-	"0_3": "buvikoh",
-	"1_0": "zeteeco",
-	"1_1": "zeteeco",
-	"1_2": "zeteeco",
-	"1_3": "zeteeco",
-	"2_0": "gafykoe",
-	"2_1": "gafykoe",
-	"2_2": "gafykoe",
-	"2_3": "gafykoe",
-	"3_0": "wopycow",
-	"3_1": "wopycow",
-	"3_2": "wopycow",
-	"3_3": "wopycow"
-}
-expressive = {
-	"0_0": "buvikop",
-	"0_1": "buvikog",
-	"0_2": "buvikob",
-	"0_3": "buvikoy",
-	"1_0": "zetikop",
-	"1_1": "zetikog",
-	"1_2": "zetikob",
-	"1_3": "zetikoy",
-	"2_0": "gafikop",
-	"2_1": "gafikog",
-	"2_2": "gafikob",
-	"2_3": "gafikoy",
-	"3_0": "wopikop",
-	"3_1": "wopikog",
-	"3_2": "wopikob",
-	"3_3": "wopikoy",
-}
-expressive = {
-	"0_0": "buvikoh",
-	"0_1": "buveeco",
-	"0_2": "buvykoe",
-	"0_3": "buvycow",
-	"1_0": "zetikoh",
-	"1_1": "zeteeco",
-	"1_2": "zetykoe",
-	"1_3": "zetycow",
-	"2_0": "gafikoh",
-	"2_1": "gafeeco",
-	"2_2": "gafykoe",
-	"2_3": "gafycow",
-	"3_0": "wopikoh",
-	"3_1": "wopeeco",
-	"3_2": "wopykoe",
-	"3_3": "wopycow"
-}
-
-
-
[…]

### 2023-10-30 code/make_audio_labels.py

@@ -4,6 +4,9 @@ the Mac OS text-to-speech synthesizer, so must be run on a Mac.
 '''
 
 import subprocess
+from pathlib import Path
+
+ROOT = Path(__file__).parent.parent.resolve()
 
 
 def say(utterance, voice, output_path):
@@ -13,35 +16,21 @@ def say(utterance, voice, output_path):
 # Audio check sentences
 
 for i, sentence in enumerate(['the quick cat', 'the green bird', 'the hungry dog', 'the angry lion']):
-	path = f'../experiment/client/sounds/test{i}.m4a'
+	path = ROOT/'experiment'/'client'/'sounds'/f'test{i}.m4a'
 	say(sentence, 'Tessa', path)
 
-say('This is an attention check; please click on the object.', 'Tessa', '../experiment/client/sounds/catch_instruction.m4a')
-say('Great. Please continue.', 'Tessa', '../experiment/client/sounds/catch_acknowledge.m4a')
+say('This is an attention check; please click on the object.', 'Tessa', ROOT/'experiment'/'client'/'sounds'/'catch_instruction.m4a')
+say('Great. Please continue.', 'Tessa', ROOT/'experiment'/'client'/'sounds'/'catch_acknowledge.m4a')
 
-
-# Experiment 1 spoken word stimuli
-
-for i, word in enumerate(['booveekoe', 'zeteekoe', 'gafeekoe', 'woppeekoe']):
-	say(word, 'Tessa', f'../experiment/client/words/{i}.m4a')
-
-
-# Experiment 2 spoken word stimuli
-
-for i, stem in enumerate(['boovy', 'zetty', 'gaffy', 'woppy']):
-	for c, cons in enumerate(['th', 's', 'sh', 'h']):
-		for v, vowl in enumerate(['a', 'ay', 'oe', 'u']):
-			word = f'{stem}{cons}{vowl}'
-			say(word, 'Tessa', f'../experiment/client/words/{i}_{c}_{v}.m4a')
-
-
-# Experiment 3 spoken word stimuli
+# Experiment 1, differentiation
 
 for i, word in enumerate(['boovikoe', 'zettikoe', 'woppikoe']):
-	say(word, 'Tessa', f'../experiment/client/words/{i}.m4a')
+	say(word, 'Tessa', ROOT/'experiment'/'client'/'words'/f'{i}.m4a')
+
+# Experiment 2, conservation
[…]

### 2023-10-30 code/make_shape_stimuli.py

@@ -3,6 +3,7 @@ import cairosvg
 
 ROOT = Path(__file__).parent.parent.resolve()
 
+
 header = '''<?xml version="1.0" encoding="UTF-8" standalone="no"?>
 <!DOCTYPE svg PUBLIC "-//W3C//DTD SVG 1.1//EN" "http://www.w3.org/Graphics/SVG/1.1/DTD/svg11.dtd">
 <svg width="100%" height="100%" viewBox="0 0 1000 1000" version="1.1" xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink" xml:space="preserve" xmlns:serif="http://www.serif.com/" style="fill-rule:evenodd;clip-rule:evenodd;stroke-linejoin:round;stroke-miterlimit:2;">
@@ -12,12 +13,12 @@ footer = '\n</svg>'
 
 shapes = [
 	'<g transform="matrix(1.60121,0,0,1.40577,-296.42,-151.265)"><path d="M497.387,173.739L764.67,394.929L662.577,752.823L332.197,752.823L230.104,394.929L497.387,173.739Z" style="fill:%s;"/></g>',
-	'<g transform="matrix(1.2861,0,0,1.36861,-160.848,-168.265)"><path d="M513.838,159.48L570.681,323.881L680.222,273.078L662.657,386.676L846.607,386.676L697.788,488.281L783.054,570.482L662.657,589.886L719.501,754.287L570.681,652.682L513.838,754.287L456.994,652.682L308.175,754.287L365.019,589.886L244.621,570.482L329.887,488.281L181.068,386.676L365.019,386.676L347.453,273.078L456.994,323.881L513.838,159.48Z" style="fill:%s;"/></g>',
-	'<g transform="matrix(1.55578,0,0,1.79159,-300.984,-369.947)"><path d="M583.618,280.079L721.162,366.857L652.39,449.055L763.665,480.452L711.128,620.862L599.853,589.465L599.853,691.067L429.839,691.067L429.839,589.465L318.563,620.862L266.026,480.452L377.302,449.055L308.53,366.857L446.074,280.079L514.846,362.277L583.618,280.079Z" style="fill:%s;"/></g>',
[…]

### 2023-10-30 code/measures.py

@@ -1,413 +1,0 @@
-from collections import defaultdict
-from itertools import product
-from pathlib import Path
-import json
-import numpy as np
-from scipy import stats
-import matplotlib.pyplot as plt
-import matplotlib.patches as patches
-import mantel
-import matrix
-import grammarette
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-
-
-def json_load(filepath):
-	with open(filepath) as file:
-		data = json.load(file)
-	return data
-
-def levenshtein_distance(s1, s2, normalize=False):
-	if len(s1) > len(s2):
-		s1, s2 = s2, s1
-	distances = range(len(s1) + 1)
-	for index2, char2 in enumerate(s2):
-		newDistances = [index2 + 1]
-		for index1, char1 in enumerate(s1):
-			if char1 == char2:
-				newDistances.append(distances[index1])
-			else:
-				newDistances.append(
-					1 + min((distances[index1], distances[index1+1], newDistances[-1]))
-				)
-		distances = newDistances
-	if normalize:
-		return distances[-1] / max(len(s1), len(s2))
-	return distances[-1]
-
-def transmission_error(lexicon1, lexicon2):
-	sum_norm_lev_dist = 0.0
-	for item, word1 in lexicon1.items():
-		word2 = lexicon2[item]
-		sum_norm_lev_dist += levenshtein_distance(word1, word2, normalize=False)
-	return sum_norm_lev_dist / len(lexicon1)
-
-def expressivity(lexicon):
-	return len(set(lexicon.values()))
-
-def structure(lexicon):
-	meaning_dists = []
-	string_dists = []
-	meanings, strings = zip(*lexicon.items())
-	for i in range(len(meanings)):
-		for j in range(i + 1, len(meanings)):
-			meaning_dists.append(levenshtein_distance(meanings[i], meanings[j]))
-			string_dists.append(levenshtein_distance(strings[i], strings[j]))
-	meantel_result = mantel.test(meaning_dists, string_dists)
-	return meantel_result.z
-
-def plot_response_time(dataset):
-	res = defaultdict(list)
-	for chain, generations in dataset.items():
-		for subject_a, subject_b in generations:
[…]

### 2023-10-30 code/model.py

@@ -1,44 +1,0 @@
-import numpy as np
-import voi
-
-
-transparent = np.array([
-	[0, 0, 0, 0],
-	[0, 0, 0, 0],
-	[0, 0, 0, 0],
-	[0, 0, 0, 0],
-], dtype=int)
-
-holistic = np.array([
-	[ 0, 1, 2, 3],
-	[ 7, 6, 5, 4],
-	[ 8, 9,10,11],
-	[15,14,13,12],
-], dtype=int)
-
-redundant = np.array([
-	[1, 1, 1, 1],
-	[2, 2, 2, 2],
-	[3, 3, 3, 3],
-	[0, 0, 0, 0],
-], dtype=int)
-
-expressive = np.array([
-	[0, 1, 2, 3],
-	[0, 1, 2, 3],
-	[0, 1, 2, 3],
-	[0, 1, 2, 3],
-], dtype=int)
-
-typology = [transparent, holistic, redundant, expressive]
-
-
-output = np.array([
-	[0, 0, 0, 0],
-	[0, 0, 0, 0],
-	[1, 1, 1, 1],
-	[1, 1, 1, 1],
-], dtype=int)
-
-
-print([voi.variation_of_information(output, grammar) for grammar in typology])

### 2023-10-30 code/ploymod_plot.py

@@ -1,58 +1,0 @@
-import matplotlib.pyplot as plt
-import arviz as az
-import numpy as np
-import pandas as pd
-
-
-def plot_chains(axis, dataset, var, show_mean=False):
-	chain_data = []
-	for chain_i in sorted(dataset['chain'].unique()):
-		chain_subset = dataset[ dataset['chain'] == chain_i ]
-		chain_data.append(list(chain_subset[var]))
-		axis.plot(chain_subset['generation'], chain_subset[var], label=f'Chain {chain_i + 1}')
-	if show_mean:
-		chain_data = np.array(chain_data)
-		axis.plot(chain_subset['generation'], chain_data.mean(axis=0), label=f'Chain {chain_i + 1}', color='black', linewidth=5)
-	# axis.set_xlim(0, n_generations)
-	# axis.set_ylim(*pad_range(*MEASURE_RANGES[measure]))
-	# axis.set_xticks(list(range(n_generations + 1)))
-	# axis.set_xlabel('Generation')
-	# axis.set_ylabel(LABELS[measure])
-	# axis.set_title(LABELS[condition])
-
-
-def plot_predictive(axis, trace, var, color):
-	generation = np.arange(1, 10)
-	az.plot_hdi(generation, trace.posterior[var], hdi_prob=0.95, smooth=False, color=color, fill_kwargs={'alpha': 0.25, 'linewidth': 0}, ax=axis)
-	axis.plot(generation, trace.posterior[var].mean(('chain', 'draw')), color=color)
-	# axis.set_ylim(0, 1.8)
-
-
-df = pd.read_csv('../data/exp3.csv')
-trace = az.from_netcdf('../data/exp1.netcdf')
-
-fig, axes = plt.subplots(2, 2, figsize=(6, 4))
-
-plot_chains(axes[0,0], df[ df['condition'] == 'dif_lrn' ], 'cost', show_mean=True)
-plot_chains(axes[0,1], df[ df['condition'] == 'dif_com' ], 'cost', show_mean=True)
-
-axes[0,0].set_ylim(0, 1.8)
-axes[0,1].set_ylim(0, 1.8)
-
-plot_predictive(axes[1,0], trace, 'pred_lrn', 'cadetblue')
-plot_predictive(axes[1,1], trace, 'pred_com', 'crimson')
-
-axes[1,0].set_ylim(0, 1.8)
[…]

### 2023-10-30 code/regmodel2.py

@@ -1,85 +1,0 @@
-import numpy as np
-import pandas as pd
-import bambi as bmb
-import arviz as az
-
-
-def center_variable(df, var):
-	width = max(df[var]) - min(df[var])
-	midpoint = min(df[var]) + width / 2
-	df[var] = df[var] - midpoint
-	return df
-
-def generate_random_dataset(n_chains, n_generations, beta, noise_sd):
-	gen = np.array(beta * np.arange(1, n_generations + 1))
-	chains = np.array([gen + np.random.normal(0, noise_sd, n_generations) for _ in range(n_chains)])
-	chains = np.clip(chains, 0, 3.16)
-	return chains
-
-def plot_dataset(data1, data2):
-	fig, axes = az.utils.plt.subplots(1, 2)
-	for chain in data1:
-		axes[0].plot(range(1, 10), chain)
-	for chain in data2:
-		axes[1].plot(range(1, 10), chain)
-	axes[0].set_ylim(0, 3)
-	axes[1].set_ylim(0, 3)
-	az.utils.plt.show()
-
-
-if __name__ == '__main__':
-
-	# np.random.seed(1177)
-
-	# n_chains = 10
-	# n_generations = 9
-
-	# data1 = generate_random_dataset(n_chains, n_generations, 0.3, 0.3)
-	# data2 = generate_random_dataset(n_chains, n_generations, 0.2, 0.3)
-
-	# # plot_dataset(data1, data2)
-	# # quit()
-
-	# chain_ids = []
-	# np.repeat
-
-	# df = pd.DataFrame({
-	# 	'cost': list(data1.flatten()) + list(data2.flatten()),
-	# 	'condition': ['lrn'] * (n_chains * n_generations) + ['com'] * (n_chains * n_generations),
-	# 	'generation': list(range(1, n_generations+1)) * (n_chains*2),
-	# 	'chain': np.repeat(range(n_chains*2), n_generations),
-	# })
-
-	df = pd.read_csv('data/exp3.csv')
-	df = df[ df['condition'].str.contains('^con') ]
-	df = df[ df['generation'] > 0 ]
-	df['cost'] = df['cost'] + 0.001
-	df = center_variable(df, 'generation')
-	df['chain_id'] = df[['condition', 'chain']].apply(lambda row: '_'.join(row.values.astype(str)), axis=1)
-
-	model = bmb.Model(
-		'cost ~ generation * epoch * condition + (1 + generation | chain_id)',
[…]

### 2023-10-30 code/stat_model.py

@@ -1,47 +1,0 @@
-import pymc as pm
-import arviz as az
-import numpy as np
-
-def generate_systems(n_systems, mu, sigma):
-	return np.random.normal(mu, sigma, n_systems)
-
-def unpaired_t_test(condition1, condition2, measure):
-	condition1 = np.array(condition1)
-	condition2 = np.array(condition2)
-
-	with pm.Model(coords={'condition': [0, 1]}) as model:
-		if measure == 'complexity':
-			μ = pm.Uniform('μ', np.array([100, 100]), np.array([700, 700]), dims='condition')
-			σ = pm.Exponential('σ', lam=np.array([0.01, 0.01]), dims='condition')
-		else:
-			μ = pm.Uniform('μ', np.array([0, 0]), np.array([2, 2]), dims='condition')
-			σ = pm.Exponential('σ', lam=np.array([1, 1]), dims='condition')
-
-		x1 = pm.TruncatedNormal('x1', mu=μ[0], sigma=σ[0], lower=0, upper=2, observed=condition1)
-		x2 = pm.TruncatedNormal('x2', mu=μ[1], sigma=σ[1], lower=0, upper=2, observed=condition2)
-		
-		Δ = pm.Deterministic('Δ', μ[0] - μ[1])
-		
-		trace = pm.sample(5000, tune=1000, chains=4, cores=4, return_inferencedata=True)
-	return trace
-
-
-if __name__ == '__main__':
-
-	# H1
-	condition1 = generate_systems(10, 200, 30)
-	condition2 = generate_systems(10, 600, 8)
-	measure = 'complexity'
-
-	# H2
-	condition1 = np.array([0, 0, 0, 0, 0.1, 0.5, 0.4, 0.1, 0, 0.05])
-	condition2 = np.array([2, 2, 2, 2, 2, 2, 2, 2, 2, 2])
-	measure = 'cost'
-
-	print(condition1.mean())
-	print(condition2.mean())
-
-	trace = unpaired_t_test(condition1, condition2, measure=measure)
-
-	az.plot_posterior(trace, hdi_prob=0.99, var_names=['μ', 'σ', 'Δ'])
-	az.utils.plt.show()

### 2023-10-30 code/build_csv.py
Tidy up main scripts

@@ -1,0 +1,84 @@
+from pathlib import Path
+from collections import defaultdict
+from itertools import product
+import numpy as np
+import pandas as pd
+import Levenshtein
+from utils import json_load
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+
+
+def convert_lexicon_meanings_to_tuple(lexicon):
+	converted_lexicon = {}
+	for item, signal in lexicon.items():
+		meaning = tuple(map(int, item.split('_')))
+		converted_lexicon[meaning] = signal
+	return converted_lexicon
+
+def transmission_error(lexicon1, lexicon2):
+	sum_lev_dist = 0.0
+	for item, word1 in lexicon1.items():
+		word2 = lexicon2[item]
+		sum_lev_dist += Levenshtein.distance(word1, word2)
+	return sum_lev_dist / len(lexicon1)
+
+def communicative_cost(lexicon, dims):
+	reverse_lexicon = defaultdict(set)
+	for meaning, signal in lexicon.items():
+		reverse_lexicon[signal].add(meaning)
+	U = product(*[range(n_values) for n_values in dims])
+	U_size = np.product(dims)
+	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
+
+def build_csv(exp_data_file, exp_csv_file):
+	exp_data = json_load(exp_data_file)
+	table = []
+	for condition, data in exp_data.items():
+		print(condition.upper())
+		for chain_i, chain in enumerate(data):
+			print('  Chain', chain_i)
+			prev_lexicon = None
+			for generation_i, (subject_a, subject_b) in enumerate(chain):
+				if condition.startswith('con'):
+					if generation_i == 0:
+						epoch_i = 0
+					elif generation_i < 4:
+						epoch_i = 1
+					elif generation_i < 7:
+						epoch_i = 2
+					else:
+						epoch_i = 3
+				else:
+					epoch_i = 0
+				print('    Generation', generation_i)
+				lexicon_a = convert_lexicon_meanings_to_tuple(subject_a['lexicon'])
+				error = transmission_error(lexicon_a, prev_lexicon) if prev_lexicon else None
+				cost = communicative_cost(lexicon_a, (3, 3))
+				struc = 0
+				if subject_b:
+					lexicon_b = convert_lexicon_meanings_to_tuple(subject_b['lexicon'])
[…]

### 2023-10-30 code/build_figures.py

@@ -27,7 +27,7 @@ HDI_PROB = 0.9
 
 
 def plot_transmission_chains(exp_data, condition, output_path):
-	import visualize
+	import draw_chains
 	import matrix
 
 	cons = ['f', 's', 'ʃ']
@@ -61,7 +61,7 @@ def plot_transmission_chains(exp_data, condition, output_path):
 		panel.append(chain_matrices)
 	show_sounds = True if 'con' in condition else False
 	chain_ids = A_to_J if 'lrn' in condition else K_to_T
-	visualize.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
+	draw_chains.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
 
 
 def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False, model_trace=None):
@@ -82,6 +82,7 @@ def plot_typological_distribution(exp_data, conditions, generations, output_path
 
 	distribution_colors = colormaps['viridis'](np.linspace(0, 1, 4))
 
+	fig, axes = plt.subplots(3, 9, figsize=figsize, sharex=True)
 	
 	for i, condition in enumerate(conditions):
 		for j, generation in enumerate(generations):
@@ -344,112 +345,112 @@ if __name__ == '__main__':
 
 	exp_data = json_load(ROOT / 'data' / 'exp3.json')
 
-	# plot_transmission_chains(exp_data, 
-	# 	condition='dif_lrn',
-	# 	output_path=FIGS / 'dif_lrn.eps',
-	# )
+	plot_transmission_chains(exp_data, 
+		condition='dif_lrn',
+		output_path=FIGS / 'dif_lrn.eps',
+	)
 
-	# plot_transmission_chains(exp_data, 
-	# 	condition='dif_com',
-	# 	output_path=FIGS / 'dif_com.eps',
-	# )
+	plot_transmission_chains(exp_data, 
+		condition='dif_com',
[…]

### 2023-10-30 code/exp1_analysis.py

@@ -1,402 +1,0 @@
-from pathlib import Path
-from collections import defaultdict
-from itertools import product
-
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-import matplotlib.patches as patches
-from matplotlib import colormaps
-import Levenshtein
-import mantel
-import grammarette
-
-from utils import json_load, json_save
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-
-plt.rcParams.update({'font.sans-serif': 'Helvetica Neue', 'font.size': 7})
-
-COLORS = colormaps['tab10_r'].colors
-
-MEASURE_RANGES = {
-	'cost': (0, 1.8),
-	'complexity': (100, 600),
-	'structure': (0, 10),
-	'error': (0, 2),
-	'alignment': (0, 3),
-}
-
-LABELS = {
-	'cost': 'Communicative cost (bits)',
-	'complexity': 'Complexity (bits)',
-	'structure': 'Structure (z-score)',
-	'error': 'Transmission error (edits)',
-	'alignment': 'Alignment (edits)',
-	'lrn_hiVar': 'Learning-only',
-	'lrn_noVar': 'Learning-only, No variation',
-	'com_hiVar': 'Learning & Communication',
-	'com_noVar': 'Learning & Communication, No variation',
-	'lrn': 'Learning-only',
-	'com': 'Learning & Communication',
-	'dif_lrn': 'Differentiation, learning',
-	'dif_com': 'Differentiation, communication',
-	'con_lrn': 'Conservation, learning',
-	'con_com': 'Conservation, communication',
-}
-
-
-def convert_lexicon_meanings_to_tuple(lexicon):
-	converted_lexicon = {}
-	for item, signal in lexicon.items():
-		meaning = tuple(map(int, item.split('_')))
-		converted_lexicon[meaning] = signal
-	return converted_lexicon
-
-def transmission_error(lexicon1, lexicon2):
-	sum_lev_dist = 0.0
-	for item, word1 in lexicon1.items():
-		word2 = lexicon2[item]
-		sum_lev_dist += Levenshtein.distance(word1, word2)
-	return sum_lev_dist / len(lexicon1)
-
-def complexity(lexicon, dims):
-	grammar = grammarette.induce(lexicon, dims)
-	return grammar.codelength
-
-def structure(lexicon, dims):
-	meaning_dists = []
-	string_dists = []
[…]

### 2023-10-30 code/print_word_chains.py

@@ -1,0 +1,41 @@
+from pathlib import Path
+from itertools import product
+from utils import json_load
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+
+
+cons = ['f', 's', 'ʃ']
+vwls = ['əʊ', 'ə', 'ɛɪ']
+def get_sound(item, data):
+	sound_file = data['spoken_forms'][item]
+	if '_' not in sound_file:
+		return 'kəʊ'
+	s, c, v = sound_file.split('.')[0].split('_')
+	return f'{cons[int(c)]}{vwls[int(v)]}'
+
+def print_word_chains(dataset):
+	for condition, data in dataset.items():
+		print(condition.upper())
+		for chain_i, chain in enumerate(data):
+			print('  Chain', chain_i)
+			table = [[] for _ in range(9)]
+			for item_i, (shape, color) in enumerate(product(range(3), range(3))):
+				item = f'{shape}_{color}'
+				word = chain[0][0]['lexicon'][item]
+				table[item_i].append(word.ljust(9, ' '))
+				for subject_a, subject_b in chain[1:]:
+					bottleneck = '➤ ' if item in subject_a['training_items'] else '  '
+					word = subject_a['lexicon'][item]
+					sound = get_sound(item, subject_a).ljust(4) if 'spoken_forms' in subject_a else ''
+					table[item_i].append(bottleneck + sound + word.ljust(9, ' '))
+			print(''.join([str(gen_i).ljust(16, ' ') for gen_i in range(len(table[0]))]).strip())
+			for row in table:
+				print(' '.join(row).strip())
+
+
+if __name__ == '__main__':
+
+	dataset = json_load(ROOT / 'data' / 'exp3.json')
+	print_word_chains(dataset)

### 2023-10-30 code/build_csv.py
Change exp3 to exp

@@ -78,7 +78,7 @@ def build_csv(exp_data_file, exp_csv_file):
 
 if __name__ == '__main__':
 
-	exp_json_file = ROOT / 'data' / 'exp3.json'
-	exp_csv_file = ROOT / 'data' / 'exp3.csv'
+	exp_json_file = ROOT / 'data' / 'exp.json'
+	exp_csv_file = ROOT / 'data' / 'exp.csv'
 
 	build_csv(exp_json_file, exp_csv_file)

### 2023-10-30 code/build_figures.py

@@ -343,7 +343,7 @@ def diff_predictions(model_trace):
 
 if __name__ == '__main__':
 
-	exp_data = json_load(ROOT / 'data' / 'exp3.json')
+	exp_data = json_load(ROOT / 'data' / 'exp.json')
 
 	plot_transmission_chains(exp_data, 
 		condition='dif_lrn',

### 2023-10-30 code/demographics.py

@@ -49,5 +49,5 @@ def most_common_languages(exp_id):
 		print(i, lang, counts[lang], round(counts[lang] / len(langs) * 100, 2))
 
 
-calculate_rates('exp3', 584, chain_filter=r'dif.+', with_bonus=False)
-calculate_rates('exp3', 584, chain_filter=r'con.+', with_bonus=False)
+calculate_rates('exp', 584, chain_filter=r'dif.+', with_bonus=False)
+calculate_rates('exp', 584, chain_filter=r'con.+', with_bonus=False)

### 2023-10-30 code/model_exp1.py

@@ -13,7 +13,7 @@ if __name__ == '__main__':
 	chains = list(range(0, 10))
 	generations = list(range(1, 10))
 
-	df = pd.read_csv(DATA / 'exp3.csv')
+	df = pd.read_csv(DATA / 'exp.csv')
 
 	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] >= min(generations)))  ]
 	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T

### 2023-10-30 code/model_exp2.py

@@ -13,7 +13,7 @@ if __name__ == '__main__':
 	chains = list(range(0, 10))
 	generations = list(range(1, 10))
 
-	df = pd.read_csv(DATA / 'exp3.csv')
+	df = pd.read_csv(DATA / 'exp.csv')
 
 	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
 	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T

### 2023-10-30 code/model_typo.py

@@ -38,29 +38,31 @@ def construct_t_matrix(data_path, experiment):
 
 if __name__ == '__main__':
 
-	data_path = DATA / 'exp3.json'
-	experiment = 'con'
-	output_path = DATA / 'exp2_typo.netcdf'
+	data_path = DATA / 'exp.json'
 
-	t = construct_t_matrix(data_path, experiment)
+	for exp_i, condition in enumerate(['dif', 'con'], 1):
 
-	coords = {
-		'category': ['H', 'E', 'R', 'D'],
-		'generation': list(range(1, 10)),
-		'condition': ['lrn', 'com'],
-	}
+		output_path = DATA / f'exp{exp_i}_typo.netcdf'
 
-	with pm.Model(coords=coords) as model:
+		t = construct_t_matrix(data_path, condition)
 
-		# Prior
-		θ = pm.Dirichlet('θ', a=np.ones(4), dims=('condition', 'generation', 'category'))
+		coords = {
+			'category': ['H', 'E', 'R', 'D'],
+			'generation': list(range(1, 10)),
+			'condition': ['lrn', 'com'],
+		}
 
-		# Likelihood
-		pm.Multinomial('c', n=10, p=θ, observed=t, dims=('condition', 'generation', 'category'))
+		with pm.Model(coords=coords) as model:
 
-		# Deterministic parameter
-		pm.Deterministic('diff_θ', θ[1, :, :] - θ[0, :, :], dims=('generation', 'category'))
+			# Prior
+			θ = pm.Dirichlet('θ', a=np.ones(4), dims=('condition', 'generation', 'category'))
 
-		# Sampling
-		trace = pm.sample(10000, tune=2000, chains=6, cores=6)
-		trace.to_netcdf(output_path)
+			# Likelihood
+			pm.Multinomial('c', n=10, p=θ, observed=t, dims=('condition', 'generation', 'category'))
+
+			# Deterministic parameter
+			pm.Deterministic('diff_θ', θ[1, :, :] - θ[0, :, :], dims=('generation', 'category'))
+
+			# Sampling
+			trace = pm.sample(10000, tune=2000, chains=6, cores=6)
+			trace.to_netcdf(output_path)

### 2023-10-30 code/print_word_chains.py

@@ -37,5 +37,5 @@ def print_word_chains(dataset):
 
 if __name__ == '__main__':
 
-	dataset = json_load(ROOT / 'data' / 'exp3.json')
+	dataset = json_load(ROOT / 'data' / 'exp.json')
 	print_word_chains(dataset)

### 2023-10-30 code/process_exp_data.py

@@ -298,10 +298,10 @@ def create_abridged_data_file(exp_id, conditions, n_chains):
 
 if __name__ == '__main__':
 
-	individual_check('exp3', 584)
+	individual_check('exp', 584)
 
-	iteration_check('exp3', ['dif_lrn', 'dif_com', 'con_lrn', 'con_com'], 10)
+	iteration_check('exp', ['dif_lrn', 'dif_com', 'con_lrn', 'con_com'], 10)
 
-	communication_check('exp3', ['dif_com', 'con_com'], 10)
+	communication_check('exp', ['dif_com', 'con_com'], 10)
 
-	create_abridged_data_file('exp3', ['dif_lrn', 'dif_com', 'con_lrn', 'con_com'], 10)
+	create_abridged_data_file('exp', ['dif_lrn', 'dif_com', 'con_lrn', 'con_com'], 10)

### 2023-11-02 experiment/mission_control.py
Parameterize prolific endpoint url

@@ -55,6 +55,7 @@ DATA_DIR = Path('../data')
 DOMAIN = 'localhost'
 PORT = 27017
 PROLIFIC_CREDENTIALS_FILE = Path.home() / '.prolific_credentials.json'
+PROLIFIC_ENDPOINT = 'https://joncarr.net:8080/prolific'
 
 DB = MongoClient(DOMAIN, PORT)
 
@@ -608,7 +609,7 @@ def create_subscription():
 	response = requests.post(
 		'https://api.prolific.co/api/v1/hooks/subscriptions/',
 		headers={'Authorization': f'Token {PROLIFIC_CREDENTIALS["api_token"]}'},
-		json={'workspace_id': PROLIFIC_CREDENTIALS['workspace_id'], 'event_type': 'submission.status.change', 'target_url': 'https://joncarr.net:8080/prolific'},
+		json={'workspace_id': PROLIFIC_CREDENTIALS['workspace_id'], 'event_type': 'submission.status.change', 'target_url': PROLIFIC_ENDPOINT},
 	)
 	subscription_id = response.json()['id']
 	x_hook_secret = response.headers['X-Hook-Secret']
@@ -617,7 +618,7 @@ def create_subscription():
 	response = requests.post(
 		f'https://api.prolific.co/api/v1/hooks/subscriptions/{subscription_id}/',
 		headers={'Authorization': f'Token {PROLIFIC_CREDENTIALS["api_token"]}'},
-		json={'secret': x_hook_secret, 'workspace_id': PROLIFIC_CREDENTIALS['workspace_id'], 'event_type': 'submission.status.change', 'target_url': 'https://joncarr.net:8080/prolific'},
+		json={'secret': x_hook_secret, 'workspace_id': PROLIFIC_CREDENTIALS['workspace_id'], 'event_type': 'submission.status.change', 'target_url': PROLIFIC_ENDPOINT},
 	)
 	print(response)
 

### 2023-11-02 code/matrix.py
Remove hardcoded path

@@ -143,4 +143,4 @@ if __name__ == '__main__':
 		color_palette = [color[:3] for color in color_palette]
 		np.random.shuffle(color_palette)
 
-		draw(system, color_palette, f'/Users/jon/Desktop/{name}.pdf')
+		draw(system, color_palette, f'../plots/{name}.pdf')

### 2023-11-02 manuscript/manuscript.tex
Final manuscript version before submission

--- 2023-10-30 manuscript/wp4.tex
@@
 
-% PREPRINT VERSION
-\authorsnames{Jon W. Carr, Kathleen Rastle}
+\authorsnames{Anonymous}
 
-\authorsaffiliations{{Department of Psychology, Royal Holloway, University of London, England}}
-
-\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023).
-Why do languages tolerate heterography?
-An experimental investigation into the emergence of informative orthography.
-\textit{OSF Preprints}.
-Version~1.
-https://doi.org/10.31219/osf.io/r5n3g}
-
-% SUBMISSION VERSION
-% \authorsnames{Anonymous}
-
-% \authorsaffiliations{{Anonymous}}
+\authorsaffiliations{{Anonymous}}
 
@@
 
-\keywords{communication; cultural evolution; informativity; iterated learning; language evolution; orthography; spelling; sound change; writing}
+\keywords{communication; iterated learning; language evolution; orthography; reading; writing}
 
@@
 Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language.
-Generalization from limited input is a big driver behind systematic structure in the language as a whole, since human learners tend to generalize in ways that increase the simplicity (i.e., systematicity) of the system, albeit unconsciously \parencite{Culbertson:2016}.
+Generalization from limited input is a major driver of systematic structure in the language as a whole, since human learners tend to generalize in ways that increase the simplicity (i.e., systematicity) of the system, albeit unconsciously \parencite{Culbertson:2016}.
 However, left unchecked, this bias for simplicity would ultimately result in the emergence of maximally simple, degenerate languages.
@@
 Crucially, this communicative component prevented the artificial languages from degenerating; instead, the languages find a tradeoff between simplicity on the one hand and informativeness on the other, just as in natural language \parencite{Mollica:2021, KempRegier:2012, Kemp:2018, Regier:2015, Zaslavsky:2018}.
-We follow this experimental approach in order to compare what happens to orthography when a pressure for informativeness \textit{is} and \textit{is~not} present.
-[KATHY: I feel that someone could wrongly think that we're just replicating what's already been done but with spelling.
-Is there a way to remind readers here that we're looking at what pressures a spelling system to diverge from spoken language?
-the key in our experiment is that diverging from spoken language increases pressure on learning.]
 
 Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation.
-All data and code is available from \url{https://osf.io}.
+All data and code is available from \url{https://osf.io/7auw6/?view_only=f8c2498b4add41839f51bdbefa03392e}.
 Both experiments were preregistered at \url{https://aspredicted.org/YDV_MXV}.
@@
 
-Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography [KATHY: Possible to remind readers that we mean divergence from spoken language i.e.
-making the orthography more complex?] (i.e., in the face of homophony).
+Our first experiment tests the ability of the differentiation model to explain the emergence of informative orthography.
+Can variant spellings become conditioned on meaning, such that the written form of the language diverges from the spoken form in a way that is expressive, despite the extra cost in learning?
 We had two main hypotheses:
@@
 The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the complete model specification, see the Supplementary Material.
+For the complete model specification, see Appendix~A.
 
@@
 The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the complete model specification, see the Supplementary Material.
+For the complete model specification, see Appendix~B.
 
@@
 
-[JON: something about the cultural ratchet.
-with each successive sound change, new heterography appears, but once it's present, it's hard to let go of it.
-Thus, the informativeness of the writing system ratchets up with successive sound changes.]
-
 Our findings did indeed show that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation.
@@
 Cases such as Chain~D correspond to the ``accident of history'' explanation: Expressive orthography is preserved not because it serves any useful purpose (recall that in Transmission-only there is no functional need for the language to be informative), but because participants are simply reproducing what they learned, and what they learned has not (yet) placed a significant enough burden on learning for simplification to kick~in.
-The additional level of conservation that occurs in Transmission~+ Communication corresponds to the exaptation/repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly).
+The additional level of conservation that occurs in Transmission~+ Communication corresponds to the repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly).
 The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure.
-Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely [KATHY: on what basis?] that all chains will ultimately undergo the same transformation eventually.
+Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and based on the trajectories of the communicative cost curves (Fig.~\ref{cost_con}), it seems likely that all chains will ultimately undergo the same transformation eventually.
 Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.
@@
 We follow a large body of recent studies by using a real-time communication game \parencite[e.g.,][]{Carr:2017, Kanwal:2017, Kirby:2015, Raviv:2018, Saldana:2019, Silvey:2019, Winters:2015}, but such games are not very representative of the dynamics involved in asynchronous written communication \parencite[although see][for some approaches]{Winters:2019}.
-That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as repair, stress, and emotion, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.
+That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as repair, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.
 
@@
 
-\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034) awarded to K.R.
+\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034).
 
@@
 
-\section{Appendix}
+\appendix
 
-\subsection{Experiment 1 regression model}
+\section{}
 
@@
 
-\subsection{Experiment 2 regression model}
-
-	\begin{equation}
-	\begin{aligned}
-	   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-	 \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
-	\sigma                                 & \sim \mathrm{Exponential}(1) \\
-	\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
-	\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
-	\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
-	\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
-	\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
-	\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
-	\end{aligned}
-	\end{equation}
-
-\clearpage
-
-\begin{table}
+\begin{table}[h]
 \begin{center}
@@
 
-\begin{table}
+\section{}
+
+\begin{equation}
+\begin{aligned}
+   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
+ \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
+\sigma                                 & \sim \mathrm{Exponential}(1) \\
+\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
+\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
+\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
+\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
+\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
+\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
+\end{aligned}
+\end{equation}
+
+\begin{table}[h]
 \begin{center}

### 2024-01-15 manuscript/manuscript.tex
Fix typo

@@
 
-It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite their being a clear financial incentive to do so.
+It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite there being a clear financial incentive to do so.
 Why might this be the case?

### 2024-01-15 manuscript/manuscript.tex
Add learnability analysis to ms

@@
 	\caption{Two models of heterography.
-\textbf{(A)} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
-\textbf{(B)} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
+\textbf{A} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing.
+\textbf{B} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model.
 Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
@@
 	\vspace*{2pt}
-	\caption{\textbf{(A)} Transmission-only procedure.
+	\caption{\textbf{A} Transmission-only procedure.
 Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items.
@@
 The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent.
-\textbf{(B)} Transmission~+ Communication procedure.
+\textbf{B} Transmission~+ Communication procedure.
 Each generation now consists of two participants who engage in a communicative task.
@@
 Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>.
-The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.
+The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated toward transparency.
 
@@
 
+\subsubsection{Typological distribution}
+
 To analyze more formally what types of system tend to emerge under different conditions, we first designated four suffix systems that are of particular interest: holistic, expressive, redundant, and degenerate.
@@
 
+\subsubsection{Communicative cost}
+
 As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}.
@@
 	\vspace*{2pt}
-	\caption{\textbf{(A)} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
 The thin lines show each individual transmission chain (with a small amount of added jitter).
@@
 In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.
-\textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_dif}
@@
 
+\subsubsection{Learnability}
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/te_dif.eps}}
+	\vspace*{2pt}
+	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+The thin lines show each individual transmission chain (with a small amount of added jitter).
+The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
+Generation~0 is not included in the model, since it does not represent participant data.
+Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher.
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+	\label{te_dif}
+	\end{figure}
+
+Aside from the informativeness of the orthographic systems, we also predicted that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
+In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form.
+In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---these were not particularly forthcoming).
+We operationalize learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$.
+Following other papers, we define transmission error as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations.
+
+The results are plotted in Fig.~\ref{te_dif}, along with the posterior parameter estimates of a regression model fit to the data (this is the same model outlined above, except that we use a 2nd degree polynomial which better characterized the data).
+In Transmission-only, transmission error tends toward zero as the systems become trivial to learn and reproduce.
+In Transmission~+ Communication, there is some slight evidence for an uptick in transmission error in the last few generations as participants start to introduce novel disambiguating forms (notably English color letters), which diverge from the previous generation.
+Although there is little evidence to suggest that the shapes of the curves differ between conditions, there is a relatively clear difference in the intercepts: Transmission error remains higher in the communicative condition.
+This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates some change to the system.
+Nevertheless, the effect size is rather small (STATS), suggesting again that the presence of the communicative task did not have a big impact on how the systems evolved.
+
 \subsection{Summary}
@@
 Overall, there was little evidence of systematic orthographic differentiation.
-Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue).
-Such differentiation was, however, not forthcoming, even under communicative pressure.
-Instead, when spelling variation was conditioned on anything, it was usually conditioned on shape, resulting in redundant suffix spellings.
+Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <x> for pink, <f> for yellow, and <m> for blue).
+Such differentiation was not forthcoming, however, even under communicative pressure.
+Instead, if spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings.
 This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.
 
-So, what was different about the present experiment?
+So, what was different about the present experiment compared the large body of prior experimental iterated learning studies?
 The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions.
@@
 The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color.
-Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
+Faced with orthographic variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
 As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
@@
 Why might this be the case?
-Firstly, although innovation was not expressly forbidden in the instructions, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training.
-Participants who adopted this interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules.
-Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
+Firstly, although innovation was not expressly forbidden in the instructions, it is possible that some participants may have inferred such a rule.
+Participants who adopted such an interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking these perceived rules or because they did not know whether their partner would feel comfortable breaking them.
+Secondly, even when participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
 They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.
@@
 
+\subsubsection{Typological distribution}
+
 Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition.
@@
 
+\subsubsection{Communicative cost}
+
 	\begin{figure}
@@
 In general, the orthographic systems tend to remain more informative under communicative pressure.
-\textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
 	\label{cost_con}
@@
 Thus, not only are the two conditions different from each other, but also the two conditions are more informative than the spoken language, especially under communicative pressure.
+
+\subsubsection{Learnability}
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[]{figs/te_con.eps}}
+	\vspace*{2pt}
+	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
+The thin lines show each individual transmission chain (with a small amount of added jitter).
+The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
+Generation~0 is not included in the model, since it does not represent participant data.
+Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher.
+\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
+	\label{te_con}
+	\end{figure}
+
+Fig.~\ref{te_con} plots results for transmission error following the same methods outlined previously.
+Overall there were no strong differences between conditions, although we did not particularly expect to find any.
+Descriptively, in Transmission-only transmission error follows an inverted-\textsc{u} shape; it first increases in response to the homophony introduced at Generation~4, but then decreases as the orthographic systems converge on easy to learn patterns.
+In Transmission~+ Communication, the transmission error tend to remain fairly stable through Epochs~II and~III, representing a kind of baseline level of error in response to homophony.
 

### 2024-01-15 code/build_csv.py
Calculate communicative success and put in main csv

@@ -32,6 +32,22 @@ def communicative_cost(lexicon, dims):
 	U_size = np.product(dims)
 	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
 
+def communicative_success(subject_a_id, subject_b_id):
+	subject_a = json_load(ROOT / 'data' / 'exp' / f'subject_{subject_a_id}.json')
+	subject_b = json_load(ROOT / 'data' / 'exp' / f'subject_{subject_b_id}.json')
+
+	n_correct = 0
+	for response_a, response_b in zip(subject_a['responses'], subject_b['responses']):
+		if response_a['test_type'] == 'mini_test':
+			continue
+		assert response_a['item'] == response_b['item']
+		if response_a['test_type'] == 'comm_production':
+			n_correct += response_a['item'] == response_b['selected_item']
+		else:
+			n_correct += response_b['item'] == response_a['selected_item']
+
+	return n_correct
+
 def build_csv(exp_data_file, exp_csv_file):
 	exp_data = json_load(exp_data_file)
 	table = []
@@ -59,9 +75,9 @@ def build_csv(exp_data_file, exp_csv_file):
 				struc = 0
 				if subject_b:
 					lexicon_b = convert_lexicon_meanings_to_tuple(subject_b['lexicon'])
-					algn = transmission_error(lexicon_a, lexicon_b)
+					success = communicative_success(subject_a['subject_id'], subject_b['subject_id'])
 				else:
-					algn = None
+					success = None
 				table.append([
 					condition,
 					chain_i,
@@ -69,10 +85,10 @@ def build_csv(exp_data_file, exp_csv_file):
 					epoch_i,
 					cost,
 					error,
-					algn,
+					success,
 				])
 				prev_lexicon = lexicon_a
-	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'epoch', 'cost', 'error', 'alignment'])
+	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'epoch', 'cost', 'error', 'comm_success'])
 	df.to_csv(exp_csv_file, index=False)
 
 

### 2024-01-15 code/draw_chains.py
Adjust sizing of chains

@@ -1,15 +1,15 @@
 import numpy as np
 import cairocffi as cairo
 
-font_size = 25
+font_size = 20
 helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_NORMAL, cairo.FONT_WEIGHT_NORMAL)
 helvetica_sf = cairo.ScaledFont(helvetica, cairo.Matrix(xx=font_size, yy=font_size))
 
-seen_suffix_font_size = 4
+seen_suffix_font_size = 5
 seen_suffix_helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_NORMAL, cairo.FONT_WEIGHT_NORMAL)
 seen_suffix_helvetica_sf = cairo.ScaledFont(seen_suffix_helvetica, cairo.Matrix(xx=seen_suffix_font_size, yy=seen_suffix_font_size))
 
-unseen_suffix_font_size = 4
+unseen_suffix_font_size = 5
 unseen_suffix_helvetica = cairo.ToyFontFace('Helvetica', cairo.FONT_SLANT_OBLIQUE, cairo.FONT_WEIGHT_BOLD)
 unseen_suffix_helvetica_sf = cairo.ScaledFont(unseen_suffix_helvetica, cairo.Matrix(xx=unseen_suffix_font_size, yy=unseen_suffix_font_size))
 

### 2024-01-15 code/build_figures.py
Prepare learnability analysis

@@ -24,6 +24,8 @@ COLORS = {
 }
 
 HDI_PROB = 0.9
+SINGLE_COLUMN = 3.54
+DOUBLE_COLUMN = 7.48
 
 
 def plot_transmission_chains(exp_data, condition, output_path):
@@ -61,7 +63,7 @@ def plot_transmission_chains(exp_data, condition, output_path):
 		panel.append(chain_matrices)
 	show_sounds = True if 'con' in condition else False
 	chain_ids = A_to_J if 'lrn' in condition else K_to_T
-	draw_chains.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=432)
+	draw_chains.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=538)
 
 
 def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False, model_trace=None):
@@ -209,7 +211,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 		
 		if i == 0:
 			axis.set_ylabel('Communicative cost (bits)')
-			axis.text(0.1, 1.0, '(A)', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis.transAxes), size=8, font='Arial', weight='bold')
+			axis.text(0.1, 1.0, 'A', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis.transAxes), size=10, font='Arial', weight='bold')
 		else:
 			axis.set_yticklabels([])
 
@@ -258,7 +260,7 @@ def plot_communicative_cost(exp_data, conditions, output_path=None, figsize=(6, 
 			if i == 0:
 				axis_post.set_ylabel('Posterior')
 				axis_diff.set_ylabel('Posterior difference')
[…]

### 2024-01-17 code/build_figures.py
Build smaller figures

@@ -301,7 +301,10 @@ def plot_transmission_error(exp_data, conditions, output_path=None, figsize=(6, 
 		trace = az.from_netcdf(model_trace)
 
 	fig = plt.figure(layout='tight', figsize=figsize)
-	gs = gridspec.GridSpec(2, 2, figure=fig, height_ratios=[2, 3])
+	if variables:
+		gs = gridspec.GridSpec(2, 2, figure=fig, height_ratios=[2, 3])
+	else:
+		gs = gridspec.GridSpec(1, 2, figure=fig)
 	gs_upper = gridspec.GridSpecFromSubplotSpec(1, 2, subplot_spec=gs[0:1, 0:2])
 
 	for i, condition in enumerate(conditions):
@@ -346,11 +349,12 @@ def plot_transmission_error(exp_data, conditions, output_path=None, figsize=(6, 
 		
 		if i == 0:
 			axis.set_ylabel('Transmission error ($n$ edits)')
-			axis.text(0.1, 1.0, 'A', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis.transAxes), size=10, font='Arial', weight='bold')
+			if variables:
+				axis.text(0.1, 1.0, 'A', in_layout=False, transform=transforms.blended_transform_factory(fig.dpi_scale_trans, axis.transAxes), size=10, font='Arial', weight='bold')
 		else:
 			axis.set_yticklabels([])
 
-	if model_trace:
+	if variables:
 		gs_lower = gridspec.GridSpecFromSubplotSpec(2, len(variables), subplot_spec=gs[1:2, 0:2])
 
 		for i, variable in enumerate(variables):
@@ -574,7 +578,7 @@ if __name__ == '__main__':
 
 	# plot_transmission_error(exp_data,
 	# 	conditions=['dif_lrn', 'dif_com'],
-	# 	output_path=FIGS / 'te_dif.eps',
+	# 	output_path=FIGS / 'te_dif.pdf',
 	# 	figsize=(DOUBLE_COLUMN, 4.8),
 	# 	show_mean=False,
 	# 	add_jitter=True,
@@ -585,6 +589,15 @@ if __name__ == '__main__':
 	# 		{'var': 'β2_m', 'label': '$β_2$', 'diff': 'diff_β2'},
 	# 	],
 	# )
+
+	plot_transmission_error(exp_data,
[…]

### 2024-01-17 manuscript/manuscript.tex
Add \newmaterial command

@@
 \usepackage{mathspec}
+
+% temp for revision %%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{xcolor}
+\newcommand\newmaterial[1]{\textcolor{blue}{#1}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Aside from the informativeness of the orthographic systems, we also predicted that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
+\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
 In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form.
@@
 We operationalize learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$.
-Following other papers, we define transmission error as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations.
+Following other papers, we define transmission error as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations.}
 
-The results are plotted in Fig.~\ref{te_dif}, along with the posterior parameter estimates of a regression model fit to the data (this is the same model outlined above, except that we use a 2nd degree polynomial which better characterized the data).
+\newmaterial{The results are plotted in Fig.~\ref{te_dif}, along with the posterior parameter estimates of a regression model fit to the data (this is the same model outlined above, except that we use a 2nd degree polynomial which better characterized the data).
 In Transmission-only, transmission error tends toward zero as the systems become trivial to learn and reproduce.
@@
 This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates some change to the system.
-Nevertheless, the effect size is rather small (STATS), suggesting again that the presence of the communicative task did not have a big impact on how the systems evolved.
+Nevertheless, the effect size is rather small (STATS), suggesting again that the presence of the communicative task did not have a big impact on how the systems evolved.}
 
@@
 
-Fig.~\ref{te_con} plots results for transmission error following the same methods outlined previously.
+\newmaterial{Fig.~\ref{te_con} plots results for transmission error following the same methods outlined previously.
 Overall there were no strong differences between conditions, although we did not particularly expect to find any.
 Descriptively, in Transmission-only transmission error follows an inverted-\textsc{u} shape; it first increases in response to the homophony introduced at Generation~4, but then decreases as the orthographic systems converge on easy to learn patterns.
-In Transmission~+ Communication, the transmission error tend to remain fairly stable through Epochs~II and~III, representing a kind of baseline level of error in response to homophony.
+In Transmission~+ Communication, the transmission error tend to remain fairly stable through Epochs~II and~III, representing a kind of baseline level of error in response to homophony.}
 

### 2024-01-19 manuscript/manuscript.tex
Respond to Reviewer 1 comments

@@
 In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content.
-Heterography permits the written form of the language to be more informative than the spoken form.
+Heterography permits the written form of the language to be more informative than the spoken form, \newmaterial{and this might be especially important given that written language often lacks extralinguistic cues to meaning, such as stress, context, and body language, as well as the fact that written language is often more sophisticated, increasing the pressure to avoid ambiguity}.
 Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing.
@@
 Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}.
-Secondly, the mapping between heterographic forms and meaning must also be learned.
+Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned.
 From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around.
 Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}.
+\newmaterial{Notably, opaque writing systems take longer to learn \parencite{} and present greater challenges for dyslexics \parencite{}.}
 
@@
 The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus).
-These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent.
+These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent \newmaterial{(e.g., the round torus shape is represented by ``round'' sounds/letters)}.
 Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time.
-Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.
+Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.
 
@@
 The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations.
-Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).
-Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items.
+Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).\footnote{\newmaterial{The 2/3 bottleneck parameter was chosen based on common practices in the field and some piloting.
+In general, increasing this parameter will lead to a slower evolutionary process, so the value was chosen to allow for sufficient change to occur within nine generations.}} Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items.
 Transmission continued for nine generations in each of ten independent chains.
@@
 
-All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials.
-Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
+All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, which lasted around 8~min}.
+Participants were also told explicitly that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
 In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each.
@@
 The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial.
-Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.
-This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
+Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem corrections were rarely encountered.
+81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it once, and 5\% encountered it two or three times.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
 Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix.
 In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice).
-In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided.
+In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \newmaterial{on the correctness of the signal or object selection}.
 Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).
@@
 
-Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit subject to a separate application of the bottleneck on transmission; i.e., the seen items were selected independently for each participant).
-The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction.
+Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant).
+We felt this was more ecological than giving both participants identical input.}} \newmaterial{The communication game, which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
 On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array).
@@
 The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target.
-Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different.
+\newmaterial{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.} Overall, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different.
 In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.
@@
 As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
-A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
+\newmaterial{Similar outcomes have been noted before in the context of artificial language learning \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
 
@@
 Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
-Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
+\newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).} Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
 It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
@@
 An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers.
+\newmaterial{In many respects, writing systems are cultural fossils that do not readily adapt to external pressures.}
 
@@
 Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects.
+\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; the emergence of informative orthography might be less forthcoming among a population accustomed to transparency (e.g., participants whose native language is Finnish, Italian, or Serbian).
+We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower (i.e., because the community first has to evolve the concept of directly encoding meaning).
+English, for example, was at one point much more transparent than it is now, but it has gradually switched to become more logographic.
+In this sense, our experiments jumpstart the process by adopting a participant population already used to heterography.}
 

### 2024-01-19 code/build_csv.py
Include structure score in main csv

@@ -5,6 +5,7 @@ import numpy as np
 import pandas as pd
 import Levenshtein
 from utils import json_load
+import mantel
 
 
 ROOT = Path(__file__).parent.parent.resolve()
@@ -31,6 +32,22 @@ def communicative_cost(lexicon, dims):
 	U = product(*[range(n_values) for n_values in dims])
 	U_size = np.product(dims)
 	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
+
+def structure(lexicon, dims):
+	strings = []
+	meanings = []
+	for meaning, signal in lexicon.items():
+		strings.append(signal)
+		meanings.append(meaning)
+	string_dists = []
+	for i in range(len(strings)):
+		for j in range(i+1, len(strings)):
+			string_dists.append(Levenshtein.distance(strings[i], strings[j]))
+	meaning_dists = []
+	for i in range(len(meanings)):
+		for j in range(i+1, len(meanings)):
+			meaning_dists.append(Levenshtein.distance(meanings[i], meanings[j]))
+	return mantel.test(string_dists, meaning_dists).z
 
 def communicative_success(subject_a_id, subject_b_id):
 	subject_a = json_load(ROOT / 'data' / 'exp' / f'subject_{subject_a_id}.json')
@@ -72,7 +89,7 @@ def build_csv(exp_data_file, exp_csv_file):
 				lexicon_a = convert_lexicon_meanings_to_tuple(subject_a['lexicon'])
 				error = transmission_error(lexicon_a, prev_lexicon) if prev_lexicon else None
 				cost = communicative_cost(lexicon_a, (3, 3))
-				struc = 0
+				struc = structure(lexicon_a, (3, 3))
 				if subject_b:
 					lexicon_b = convert_lexicon_meanings_to_tuple(subject_b['lexicon'])
 					success = communicative_success(subject_a['subject_id'], subject_b['subject_id'])
@@ -86,9 +103,10 @@ def build_csv(exp_data_file, exp_csv_file):
 					cost,
 					error,
 					success,
+					struc,
 				])
 				prev_lexicon = lexicon_a
-	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'epoch', 'cost', 'error', 'comm_success'])
+	df = pd.DataFrame(table, columns=['condition', 'chain', 'generation', 'epoch', 'cost', 'error', 'comm_success', 'structure'])
 	df.to_csv(exp_csv_file, index=False)
 
 

### 2024-01-23 manuscript/manuscript.tex
ms 22/01

@@
 The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial.
-Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem corrections were rarely encountered.
-81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it once, and 5\% encountered it two or three times.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
+Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem correction messages were rarely encountered.
+81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it on one trial (out of nine), and 5\% encountered it on two or three trials.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
 Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix.
@@
 Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
-\newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).} Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome.
+\newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).
+This highlights the important role of generational turnover, since new learners will be more receptive to new functions.} Second, even when participants \textit{do} appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome within so few trials.
 It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
@@
 For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate.
+
 Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation.
-For differentiation to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming various social stigmas in the process (the social problem).
-In~comparison, the conservation of an expressive orthography is plain~sailing.
-The only real hurdle is that the system of non-transparent spellings must be sufficiently learnable (the learnability problem).
-In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.
+For \textit{differentiation} to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming social stigma in the process (the social problem).
+\newmaterial{Furthermore, once a system has been created, it needs to be reliably transmitted across multiple generations, perhaps with extensive training (the learnability problem).} The \textit{conservation} of an expressive orthography is, by~comparison, plain~sailing---it is only the learnability problem that applies.
+More generally, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.
 
@@
 We follow a large body of recent studies by using a real-time communication game \parencite[e.g.,][]{Carr:2017, Kanwal:2017, Kirby:2015, Raviv:2018, Saldana:2019, Silvey:2019, Winters:2015}, but such games are not very representative of the dynamics involved in asynchronous written communication \parencite[although see][for some approaches]{Winters:2019}.
-That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as repair, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.
+That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as feedback, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.
 
@@
 For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space.
-Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects.
-\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; the emergence of informative orthography might be less forthcoming among a population accustomed to transparency (e.g., participants whose native language is Finnish, Italian, or Serbian).
-We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower (i.e., because the community first has to evolve the concept of directly encoding meaning).
-English, for example, was at one point much more transparent than it is now, but it has gradually switched to become more logographic.
-In this sense, our experiments jumpstart the process by adopting a participant population already used to heterography.}
+From this perspective, written forms of language resist change because they need to be accessible across decades or centuries and across different jurisdictions or dialect areas.
+\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be less forthcoming under other populations (i.e., participants with transparent native languages, such as Finnish, Italian, or Serbian).
+We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower because the community first has to evolve the concept of directly encoding meaning in the written form.
+English, for example, was at one point much more transparent than it is now, but it has gradually become more logographic.
+In this sense, our experiments jumpstart the process by adopting a participant population already accustomed to these ideas.}
 

### 2024-02-26 code/build_csv.py
Change information placed in csv

@@ -5,10 +5,19 @@ import numpy as np
 import pandas as pd
 import Levenshtein
 from utils import json_load
-# import mantel
+import voi
+import matrix
 
 
 ROOT = Path(__file__).parent.parent.resolve()
+
+
+ref_systems = [
+	matrix.reference_systems['holistic'],
+	matrix.reference_systems['expressive'],
+	matrix.reference_systems['redundant'],
+	matrix.reference_systems['transparent'],
+]
 
 
 def convert_lexicon_meanings_to_tuple(lexicon):
@@ -18,12 +27,13 @@ def convert_lexicon_meanings_to_tuple(lexicon):
 		converted_lexicon[meaning] = signal
 	return converted_lexicon
 
-def transmission_error(lexicon1, lexicon2):
-	sum_lev_dist = 0.0
-	for item, word1 in lexicon1.items():
-		word2 = lexicon2[item]
-		sum_lev_dist += Levenshtein.distance(word1, word2)
-	return sum_lev_dist / len(lexicon1)
+def typological_classification(lexicon):
+	system = matrix.make_matrix(lexicon, 3, 3)
+	distances = [
+		voi.variation_of_information(system, ref_system) for ref_system in ref_systems
+	]
+	classification = np.argmin(distances)
+	return ['H', 'E', 'R', 'D'][classification]
 
 def communicative_cost(lexicon, dims):
 	reverse_lexicon = defaultdict(set)
@@ -32,6 +42,33 @@ def communicative_cost(lexicon, dims):
 	U = product(*[range(n_values) for n_values in dims])
 	U_size = np.prod(dims)
 	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
+
+def transmission_error(lexicon1, lexicon2):
+	sum_lev_dist = 0.0
+	for item, word1 in lexicon1.items():
+		word2 = lexicon2[item]
+		sum_lev_dist += Levenshtein.distance(word1, word2)
+	return sum_lev_dist / len(lexicon1)
+
+def communicative_success(subject_a_id, subject_b_id):
+	subject_a = json_load(ROOT / 'data' / 'exp' / f'subject_{subject_a_id}.json')
+	subject_b = json_load(ROOT / 'data' / 'exp' / f'subject_{subject_b_id}.json')
+	n_correct = 0
+	n_trials = 0
[…]

### 2024-02-26 code/model_exp1.py
Remove old statistical models

@@ -1,82 +1,0 @@
-from pathlib import Path
-import pandas as pd
-import numpy as np
-import pymc as pm
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-DATA = ROOT / 'data'
-
-
-if __name__ == '__main__':
-
-	chains = list(range(0, 10))
-	generations = list(range(1, 10))
-
-	df = pd.read_csv(DATA / 'exp.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] >= min(generations)))  ]
-	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
-
-	df_com = df[  (df['condition'].str.contains('^dif_com') & (df['generation'] >= min(generations)))  ]
-	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
-
-	c = np.stack((cost_lrn, cost_com), axis=2)
-	
-	g = np.repeat([generations], len(chains), axis=0).T
-	g = np.stack((g, g), axis=2)
-	g -= 5 # center generation
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'chain_id': chains,
-		'generation': generations,
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		# Hyperpriors
-
-		α_m = pm.Normal('α_m', mu=1, sigma=1, dims='condition')
-		β1_m = pm.Normal('β1_m', mu=0, sigma=1, dims='condition')
-		β2_m = pm.Normal('β2_m', mu=0, sigma=1, dims='condition')
-		β3_m = pm.Normal('β3_m', mu=0, sigma=1, dims='condition')
-
-		α_sd = pm.Exponential('α_sd', lam=1, dims='condition')
-		β1_sd = pm.Exponential('β1_sd', lam=1, dims='condition')
-		β2_sd = pm.Exponential('β2_sd', lam=1, dims='condition')
-		β3_sd = pm.Exponential('β3_sd', lam=1, dims='condition')
-
-		# Priors
-
-		α = pm.Normal('α', mu=α_m, sigma=α_sd, dims=('chain_id', 'condition'))
-		β1 = pm.Normal('β1', mu=β1_m, sigma=β1_sd, dims=('chain_id', 'condition'))
[…]

### 2024-02-26 code/model_exp1_te.py

@@ -1,82 +1,0 @@
-from pathlib import Path
-import pandas as pd
-import numpy as np
-import pymc as pm
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-DATA = ROOT / 'data'
-
-
-if __name__ == '__main__':
-
-	chains = list(range(0, 10))
-	generations = list(range(1, 10))
-
-	df = pd.read_csv(DATA / 'exp.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^dif_lrn') & (df['generation'] >= min(generations)))  ]
-	error_lrn = np.array(df_lrn['error']).reshape((len(chains), len(generations))).T
-
-	df_com = df[  (df['condition'].str.contains('^dif_com') & (df['generation'] >= min(generations)))  ]
-	error_com = np.array(df_com['error']).reshape((len(chains), len(generations))).T
-
-	e = np.stack((error_lrn, error_com), axis=2)
-	
-	g = np.repeat([generations], len(chains), axis=0).T
-	g = np.stack((g, g), axis=2)
-	g -= 5 # center generation
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'chain_id': chains,
-		'generation': generations,
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		# Hyperpriors
-
-		α_m = pm.Normal('α_m', mu=1, sigma=1, dims='condition')
-		β1_m = pm.Normal('β1_m', mu=0, sigma=1, dims='condition')
-		β2_m = pm.Normal('β2_m', mu=0, sigma=1, dims='condition')
-		# β3_m = pm.Normal('β3_m', mu=0, sigma=1, dims='condition')
-
-		α_sd = pm.Exponential('α_sd', lam=1, dims='condition')
-		β1_sd = pm.Exponential('β1_sd', lam=1, dims='condition')
-		β2_sd = pm.Exponential('β2_sd', lam=1, dims='condition')
-		# β3_sd = pm.Exponential('β3_sd', lam=1, dims='condition')
-
-		# Priors
-
-		α = pm.Normal('α', mu=α_m, sigma=α_sd, dims=('chain_id', 'condition'))
-		β1 = pm.Normal('β1', mu=β1_m, sigma=β1_sd, dims=('chain_id', 'condition'))
[…]

### 2024-02-26 code/model_exp2.py

@@ -1,83 +1,0 @@
-from pathlib import Path
-import pandas as pd
-import numpy as np
-import pymc as pm
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-DATA = ROOT / 'data'
-
-
-if __name__ == '__main__':
-
-	chains = list(range(0, 10))
-	generations = list(range(1, 10))
-
-	df = pd.read_csv(DATA / 'exp.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
-	cost_lrn = np.array(df_lrn['cost']).reshape((len(chains), len(generations))).T
-
-	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] >= min(generations)))  ]
-	cost_com = np.array(df_com['cost']).reshape((len(chains), len(generations))).T
-
-	cost = np.stack((cost_lrn, cost_com), axis=2)
-
-	c = np.zeros((3, 10, 3, 2))
-	g = np.zeros((3, 10, 3, 2))
-	for (gn, ch, ep, cd) in np.ndindex(c.shape):
-		gen_index = ep * 3 + gn
-		c[gn, ch, ep, cd] = cost[gen_index, ch, cd]
-		g[gn, ch, ep, cd] = (gen_index % 3) - 1
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'epoch': [1, 2, 3],
-		'chain_id': chains,
-		'generation': [1, 2, 3],
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		# Hyperpriors
-
-		α_m = pm.Normal('α_m', mu=0, sigma=1, dims=('epoch', 'condition'))
-		β_m = pm.Normal('β_m', mu=0, sigma=1, dims=('epoch', 'condition'))
-
-		α_sd = pm.Exponential('α_sd', lam=1, dims=('epoch', 'condition'))
-		β_sd = pm.Exponential('β_sd', lam=1, dims=('epoch', 'condition'))
-
-		# Priors
-
-		α = pm.Normal('α', mu=α_m, sigma=α_sd, dims=('chain_id', 'epoch', 'condition'))
-		β = pm.Normal('β', mu=β_m, sigma=β_sd, dims=('chain_id', 'epoch', 'condition'))
[…]

### 2024-02-26 code/model_exp2_te.py

@@ -1,83 +1,0 @@
-from pathlib import Path
-import pandas as pd
-import numpy as np
-import pymc as pm
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-DATA = ROOT / 'data'
-
-
-if __name__ == '__main__':
-
-	chains = list(range(0, 10))
-	generations = list(range(1, 10))
-
-	df = pd.read_csv(DATA / 'exp.csv')
-
-	df_lrn = df[  (df['condition'].str.contains('^con_lrn') & (df['generation'] >= min(generations)))  ]
-	error_lrn = np.array(df_lrn['error']).reshape((len(chains), len(generations))).T
-
-	df_com = df[  (df['condition'].str.contains('^con_com') & (df['generation'] >= min(generations)))  ]
-	error_com = np.array(df_com['error']).reshape((len(chains), len(generations))).T
-
-	error = np.stack((error_lrn, error_com), axis=2)
-
-	e = np.zeros((3, 10, 3, 2))
-	g = np.zeros((3, 10, 3, 2))
-	for (gn, ch, ep, cd) in np.ndindex(e.shape):
-		gen_index = ep * 3 + gn
-		e[gn, ch, ep, cd] = error[gen_index, ch, cd]
-		g[gn, ch, ep, cd] = (gen_index % 3) - 1
-
-	coords = {
-		'condition': ['lrn', 'com'],
-		'epoch': [1, 2, 3],
-		'chain_id': chains,
-		'generation': [1, 2, 3],
-	}
-
-	with pm.Model(coords=coords) as model:
-
-		# Hyperpriors
-
-		α_m = pm.Normal('α_m', mu=0, sigma=1, dims=('epoch', 'condition'))
-		β_m = pm.Normal('β_m', mu=0, sigma=1, dims=('epoch', 'condition'))
-
-		α_sd = pm.Exponential('α_sd', lam=1, dims=('epoch', 'condition'))
-		β_sd = pm.Exponential('β_sd', lam=1, dims=('epoch', 'condition'))
-
-		# Priors
-
-		α = pm.Normal('α', mu=α_m, sigma=α_sd, dims=('chain_id', 'epoch', 'condition'))
-		β = pm.Normal('β', mu=β_m, sigma=β_sd, dims=('chain_id', 'epoch', 'condition'))
[…]

### 2024-02-26 code/model_typo.py

@@ -1,68 +1,0 @@
-from pathlib import Path
-import numpy as np
-import pymc as pm
-
-
-ROOT = Path(__file__).parent.parent.resolve()
-DATA = ROOT / 'data'
-
-
-def construct_t_matrix(data_path, experiment):
-	from utils import json_load
-	import voi
-	import matrix
-
-	dataset = json_load(data_path)
-
-	ref_systems = [
-		matrix.reference_systems['holistic'],
-		matrix.reference_systems['expressive'],
-		matrix.reference_systems['redundant'],
-		matrix.reference_systems['transparent'],
-	]
-
-	t = np.zeros((2, 9, 4), dtype=int)
-	for i, condition in enumerate([f'{experiment}_lrn', f'{experiment}_com']):
-		for j, generation in enumerate(range(1, 10)):
-			for chain in dataset[condition]:
-				subject_a = chain[generation][0]
-				subject_system = matrix.make_matrix(subject_a['lexicon'], 3, 3)
-				distances = [
-					voi.variation_of_information(subject_system, ref_system)
-					for ref_system in ref_systems
-				]
-				category = np.argmin(distances)
-				t[i, j, category] += 1
-	return t
-
-
-if __name__ == '__main__':
-
-	data_path = DATA / 'exp.json'
-
-	for exp_i, condition in enumerate(['dif', 'con'], 1):
-
-		output_path = DATA / f'exp{exp_i}_typo.netcdf'
-
-		t = construct_t_matrix(data_path, condition)
-
-		coords = {
-			'category': ['H', 'E', 'R', 'D'],
-			'generation': list(range(1, 10)),
-			'condition': ['lrn', 'com'],
-		}
-
-		with pm.Model(coords=coords) as model:
-
-			# Prior
-			θ = pm.Dirichlet('θ', a=np.ones(4), dims=('condition', 'generation', 'category'))
-
-			# Likelihood
-			pm.Multinomial('c', n=10, p=θ, observed=t, dims=('condition', 'generation', 'category'))
-
-			# Deterministic parameter
-			pm.Deterministic('diff_θ', θ[1, :, :] - θ[0, :, :], dims=('generation', 'category'))
-
-			# Sampling
-			trace = pm.sample(10000, tune=2000, chains=6, cores=6)
-			trace.to_netcdf(output_path)

### 2024-02-26 code/fit_models.py
Add new model code

@@ -1,0 +1,128 @@
+from pathlib import Path
+import numpy as np
+import pandas as pd
+import bambi as bmb
+import xarray as xr
+
+
+ROOT = Path(__file__).parent.parent.resolve()
+DATA = ROOT / 'data'
+
+
+def invlogit(a):
+	return 1 / (1 + np.exp(-a))
+
+def append_generational_model_to_trace(trace, apply_invlogit_transform=False):
+	x = xr.DataArray(np.arange(1, 10))
+	trace.add_groups({'constant_data': {'x1': x}})
+	trace.constant_data['x'] = x
+	theta = trace.posterior['α'] + trace.posterior['β'] * np.log(trace.constant_data['x'] + 1)
+	if apply_invlogit_transform:
+		trace.posterior['y_model'] = (('chain', 'draw', 'gen'), invlogit(theta).data)
+	else:
+		trace.posterior['y_model'] = (('chain', 'draw', 'gen'), theta.data)
+
+def append_epochal_model_to_trace(trace, apply_invlogit_transform=False):
+	x = xr.DataArray(np.arange(1, 10))
+	trace.add_groups({'constant_data': {'x1': x}})
+	trace.constant_data['x'] = x
+	theta = []
+	for epoch in range(1, 4):
+		for generation_in_epoch in range(1, 4):
+			theta.append(trace.posterior['α'] + \
+				trace.posterior['β'] * epoch + \
+				trace.posterior['γ'] * generation_in_epoch
+			)
+	theta = np.transpose(np.array(theta), (1, 2, 0))
+	if apply_invlogit_transform:
+		trace.posterior['y_model'] = (('chain', 'draw', 'gen'), invlogit(theta))
+	else:
+		trace.posterior['y_model'] = (('chain', 'draw', 'gen'), theta)
+
+
+if __name__ == '__main__':
+
+
+	df = pd.read_csv(DATA / 'exp.csv')
+
+
+	# Experiment 1
+
+	for condition in ['dif_lrn', 'dif_com', 'sil_com']:
+
+		# Probability informative
+
+		model = bmb.Model(
+			formula='informative ~ np.log(generation + 1) + (1 + np.log(generation + 1) | chain)',
+			data=df[ (df['condition'] == condition) & (df['generation'] >= 0) ],
+			family='bernoulli',
+		)
+		model.set_alias({'Intercept': 'α', 'np.log(generation + 1)': 'β'})
[…]

### 2024-02-26 code/print_model_tables.py

@@ -1,0 +1,108 @@
+import arviz as az
+import pandas as pd
+
+HDI_PROB = 0.95
+
+def append_difference1(trace1, trace2):
+	trace2.posterior['Δ(α)'] = trace2.posterior['α'] - trace1.posterior['α']
+	trace2.posterior['Δ(β)'] = trace2.posterior['β'] - trace1.posterior['β']
+
+def append_difference2(trace1, trace2):
+	trace2.posterior['Δ(α)'] = trace2.posterior['α'] - trace1.posterior['α']
+	trace2.posterior['Δ(β)'] = trace2.posterior['β'] - trace1.posterior['β']
+	trace2.posterior['Δ(γ)'] = trace2.posterior['γ'] - trace1.posterior['β']
+
+# Experiment 1
+
+for measure in ['pinf', 'cost', 'error']:
+
+	print(measure)
+
+	trace1 = az.from_netcdf(f'../data/models/dif_lrn_{measure}.netcdf')
+	trace2 = az.from_netcdf(f'../data/models/dif_com_{measure}.netcdf')
+
+	append_difference1(trace1, trace2)
+
+	table1 = az.summary(trace1, var_names=['α', 'β'], hdi_prob=HDI_PROB, round_to=2).drop(['mcse_mean', 'mcse_sd', 'ess_tail'], axis=1)
+	table1.rename(index={'α': r'$\alpha_\mathrm{trans}$', 'β': r'$\beta_\mathrm{trans}$'}, inplace=True)
+
+	table2 = az.summary(trace2, var_names=['α', 'β', 'Δ(α)', 'Δ(β)'], hdi_prob=HDI_PROB, round_to=2).drop(['mcse_mean', 'mcse_sd', 'ess_tail'], axis=1)
+	table2.rename(index={
+		'α': r'$\alpha_\mathrm{comm}$',
+		'β': r'$\beta_\mathrm{comm}$',
+		'Δ(α)': r'$\Delta(\alpha)$',
+		'Δ(β)': r'$\Delta(\beta)$',
+	}, inplace=True)
+
+	table = pd.concat((table1, table2))
+
+	table = table.astype({'ess_bulk':int})
+
+	print(table.to_latex().replace('0000 ', ' '))
+	print()
+
+
+# Experiment 2
+
+for measure in ['pinf', 'cost', 'error']:
+
+	print(measure)
[…]

### 2024-02-26 code/build_figures.py
Overhaul plotting code

@@ -6,6 +6,7 @@ import matplotlib.pyplot as plt
 import matplotlib.gridspec as gridspec
 import matplotlib.transforms as transforms
 from matplotlib import colormaps
+from scipy.stats import gaussian_kde
 from utils import json_load
 
 
@@ -16,17 +17,19 @@ ROOT = Path(__file__).parent.parent.resolve()
 DATA = ROOT / 'data'
 FIGS = ROOT / 'manuscript' / 'figs'
 
-
 COLORS = {
-	'dif_lrn': ('darkslategray', '#97A7A7'),
-	'dif_com': ('darkorange', '#FFC680'),
-	'con_lrn': ('darkslategray', '#97A7A7'),
-	'con_com': ('darkorange', '#FFC680'),
+	'dif_lrn': 'darkolivegreen',
+	'dif_com': 'darkorange',
+	'con_lrn': 'cadetblue',
+	'con_com': 'crimson',
+	'sil_com': 'darkslateblue',
 }
+DISTRIBUTION_COLORS = colormaps['viridis'](np.linspace(0, 1, 4))
 
-HDI_PROB = 0.9
 SINGLE_COLUMN = 3.54
 DOUBLE_COLUMN = 7.48
+
+HDI_PROB = 0.95
 
 
 def plot_transmission_chains(exp_data, condition, output_path):
@@ -63,319 +66,79 @@ def plot_transmission_chains(exp_data, condition, output_path):
 			chain_matrices.append((mat, cp, ss, sounds, training_items))
 		panel.append(chain_matrices)
 	show_sounds = True if 'con' in condition else False
-	chain_ids = A_to_J if 'lrn' in condition else K_to_T
+	chain_ids = A_to_J if 'lrn' in condition or 'sil' in condition else K_to_T
 	draw_chains.draw_panel(output_path, panel, chain_ids=chain_ids, show_sounds=show_sounds, figure_width=538)
 
 
-def plot_typological_distribution(exp_data, conditions, generations, output_path=None, figsize=(6, 2), probabilistic_classification=False, model_trace=None):
-	import voi
-	import matrix
-	if model_trace:
-		trace = az.from_netcdf(model_trace)
-		n_rows = len(conditions) + 1
-	else:
-		n_rows = len(conditions)
-
-	ref_systems = [
-		matrix.reference_systems['holistic'],
-		matrix.reference_systems['expressive'],
-		matrix.reference_systems['redundant'],
[…]

### 2024-02-26 code/draw_chains.py

@@ -31,7 +31,7 @@ def draw_char(surface, char, cell_size):
 	context.show_text(char)
 
 
-def draw_matrix(surface, generation, cell_size):
+def draw_matrix(surface, generation, cell_size, gen0=False):
 	matrix, color_palette, suffix_spellings, sounds, training_items = generation
 	square_size = (cell_size - (2 * padding)) / 3
 	context = cairo.Context(surface)
@@ -45,7 +45,7 @@ def draw_matrix(surface, generation, cell_size):
 		context.set_line_width(1)
 		context.stroke()
 		context.set_source_rgb(1, 1, 1)
-		if not sounds or f'{i}_{j}' in training_items:
+		if gen0 or f'{i}_{j}' in training_items:
 			context.set_font_face(seen_suffix_helvetica)
 			context.set_font_size(seen_suffix_font_size)
 			text_width = seen_suffix_helvetica_sf.text_extents(suffix)[4]
@@ -66,7 +66,7 @@ def draw_chain(surface, chain, cell_size):
 		if isinstance(generation, str):
 			draw_char(subsurface, generation, cell_size)
 		elif isinstance(generation, tuple):
-			draw_matrix(subsurface, generation, cell_size)
+			draw_matrix(subsurface, generation, cell_size, gen_i == 1)
 		x += cell_size
 
 

### 2024-02-26 code/matrix.py

@@ -26,7 +26,10 @@ def make_matrix(lexicon, m=3, n=3):
 	matrix = np.zeros((m, n), dtype=int)
 	for i in range(m):
 		for j in range(n):
-			stem, suffix = parse_stem_and_suffix(lexicon[ f'{i}_{j}' ])
+			try:
+				stem, suffix = parse_stem_and_suffix(lexicon[ f'{i}_{j}' ])
+			except KeyError:
+				stem, suffix = parse_stem_and_suffix(lexicon[ (i, j) ])
 			matrix[i, j] = suffix_spellings.index(suffix)
 	return matrix
 

### 2024-02-26 manuscript/manuscript.tex
Main update to the manuscript following review

@@
 \usepackage{mathspec}
+\usepackage{xeCJK} % for Chinese characters
 
@@
 
-\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to acquire fully.
-It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading.
-For example, heterographic homophones (e.g., <knight> vs.
-<night>) impose additional costs in learning but reduce ambiguity in reading.
-If correct, this would suggest that heterography---and therefore orthographic opacity---might sometimes be selected for in the evolution of writing systems, despite the cost in learning.
-We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity.
-In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> vs.
-<meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift).
-Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, these systems can become informative, diverging from the spoken language in order to express meaning directly.}
+\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to fully acquire.
+It is less widely recognized, however, that such opacity may offer certain benefits in the context of reading.
+For example, heterographic homophones such as <knight> and <night> (words that sound the same but which are spelled differently) impose additional costs in learning but reduce ambiguity in reading.
+Here, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness when there is functional pressure to do so.
+We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for ambiguity avoidance.
+In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate meaning (e.g., <lite> and <light>), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> and <meet>).
+Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, they tend to favor some additional informativeness.
+Nevertheless, our findings also suggest that, in the long term, simpler, transparent spellings will tend to be preferred.}
 
@@
 Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each letter corresponding to a distinct sound.
-However, writing systems are known to diverge from their spoken counterparts.
-The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}.
-It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}.
+However, it is also well understood that writing systems diverge from their spoken counterparts in important ways \parencite{Bolinger:1946, Coulmas:1991}.
+The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}, and it seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}.
 In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
@@
 
-One potential case of functional divergence is heterographic homophones---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}).
+One potential case of such functional divergence is heterographic homophony---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}).
 Heterographic spellings such as these may serve a valuable function in reading.
-For example, faced with a spoken sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr kat/, a form of the verb \textit{to~be}, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/.
+For example, an English speaker faced with a spoken sentence beginning /ðɛr.../ will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr~kat/, a form of the verb \textit{to~be}, as in /ðɛr~ɪz/, or the progressive form of a verb, as in /ðɛr~gəʊɪŋ/.
 In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content.
-Heterography permits the written form of the language to be more informative than the spoken form, \newmaterial{and this might be especially important given that written language often lacks extralinguistic cues to meaning, such as stress, context, and body language, as well as the fact that written language is often more sophisticated, increasing the pressure to avoid ambiguity}.
-Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing.
-Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).
+Heterographic homophony can also occur in affixes.
+The English suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing.
+Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English affixes \parencite{Berg:2017} and it has been shown that readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.
+\newmaterial{These features might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair.
+Furthermore, written language tends to be more sophisticated than spoken language (e.g., longer, more elaborate sentences), placing different pressures on ambiguity resolution.}
 
-Heterographic homophony can also occur at the morpheme level.
-The suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing.
-Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English morphemes \parencite{Berg:2017} and it has been shown that English readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.
+\newmaterial{Although heterography is particularly notable in English, it is also a feature of many other languages and writing systems.
+French is similar to English in having a large number of heterographic homophones; for example, the heterographic words \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}) are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison).
+In Danish, the words \textit{hver} (\textit{every}), \textit{vejr} (\textit{weather}), \textit{vær} (\textit{be}), and \textit{værd} (\textit{worth}) are all pronounced /vεˀɐ̯/.
+In Vietnamese, the graphemes <d>, <gi>, and <r> are homophonous, resulting in sets like \textit{dao} (\textit{knife}), \textit{giao} (\textit{delivery}), and \textit{rao} (\textit{advertise}), all pronounced /zāw/.
+In some cases, features of an orthography designed for other purposes can inadvertently result in homophone disambiguation: Noun capitalization in German contrasts \textit{Wagen} (\textit{car}) and \textit{wagen} (\textit{to dare}), both pronounced /vaːɡṇ/; eclipsis marking in Irish contrasts \textit{bpáistí} (\textit{children}) and \textit{báistí} (\textit{rain}), both pronounced /bɑʃti/; and the morphological principle of Russian orthography contrasts \textit{приступить} (\textit{to start}) and \textit{преступить} (\textit{to transgress}), both pronounced /pristupʲitʲ/.
+Even in the most transparent of orthographies it is possible to find some instances of heterography: In Italian, the residual~\textsc{h} inherited from Latin contrasts \textit{hanno} (\textit{have}) and \textit{anno} (\textit{year}), both pronounced /anno/, while the grave accent is sometimes used to distinguish common homophonous words, such as \textit{la} (\textit{the}) and \textit{là} (\textit{there}).}
 
-However, heterography comes with two main costs.
+\newmaterial{However, perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese script.
+The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic script---particularly useful \parencite{Katz:1992}.
+In Mandarin Chinese, the words 糖 (\textit{sugar}), 塘 (\textit{embankment}), 溏 (\textit{pond}), and 搪 (\textit{to block}) are homophonous in speech but heterographic in writing---the phonetic radical on the right (唐; /táŋ/) represents the spoken syllable, while the semantic radicals on the left differentiate the meanings.
+In addition, 唐 itself is a surname/dynasty (\textit{Tang}), and another unrelated word 堂 (\textit{hall}) is also pronounced /táŋ/, yielding at~least six ways to write the same sound depending on the meaning.
+This allows written Chinese to convey more information about meaning---to be more \textit{informative}---than its spoken counterpart.}
+
+Despite the benefits that heterography may provide in reading, it comes with two main costs.
 Firstly, by definition, heterography implies that a single sound can be spelled multiple ways.
-The heterographic spellings <meat> and <meet>, for example, imply that the /i/ sound can be spelled <ea> or <ee>.
-Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}.
+In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>.
+Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding} \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}.
 Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned.
-From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around.
-Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}.
-\newmaterial{Notably, opaque writing systems take longer to learn \parencite{} and present greater challenges for dyslexics \parencite{}.}
+From the point of view of a modern English speaker, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>.
+Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Frith:1979, Shankweiler:1992}.
 
-An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness that outweighs its associated costs.
-In general, a system of words that conveys more information will be more complex than one that conveys less information; there exists a tradeoff between the simplicity of the spelling system and its ability to be expressive.
-A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
+\newmaterial{In this paper, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness.} A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
 An informative spelling system, on the other hand, would be one that precisely conveys meaning.
-The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in various typological \parencite[e.g.,][]{Kemp:2018} and experimental \parencite[e.g.,][]{Kirby:2015} studies.
-Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication, a point we return to shortly.
+This idea of a tradeoff between simplicity and informativeness in the writing system has long been noted \parencite[e.g.,][]{Coulmas:1991}, and such a tradeoff has also been discussed within the study of language more broadly \parencite[e.g.,][]{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978}.
+Recent typological \parencite[e.g.,][]{Kemp:2018} and experimental \parencite[e.g.,][]{Kirby:2015} studies have also subjected these ideas to empirical investigation in various domains.
+Of particular note here is the finding that complex, systematic structure emerges under concurrent pressures to be both simple and informative, a point we return to shortly.
 
@@
 The first model, \textit{the differentiation model} (Fig.~\ref{models_of_heterography}A), explains heterography through the creation of new orthographic forms.
-For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
+For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
 Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by adding <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}.
-Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}--- and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}.
+Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}---and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}.
 One important way in which differentiation can occur---especially in a language like English that has historically contained a lot of spelling variation \parencite{Nevalainen:2012, Stenroos:2016}---is by the conditioning of variant spellings on meaning; pairs like \textit{discreet}--\textit{discrete}, \textit{flour}--\textit{flower}, and \textit{plain}--\textit{plane}, which were once variant spellings of the same word, have taken on distinct meanings over time \parencite[§~5.4]{Carney:1994}.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/models_of_heterography.pdf}}
 	\vspace*{2pt}
@@
 Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from, although the participants themselves are not aware of this generational structure.
-It has been demonstrated in a wide variety of studies that, after several generations of cultural transmission, artificial languages can gradually adapt to the biases of the human learners and the environments in which they are used, yielding emergent linguistic phenomena, such as compositionality \parencite{Beckner:2017, Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, semantic category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, regularization \parencite{Ferdinand:2019, Smith:2010}, and argument marking \parencite{Motamedi:2021}, among many other things.
+It has been demonstrated in a wide variety of studies that, after several generations of cultural transmission, artificial languages can gradually adapt to the biases of the human learners and the environments in which they are used, yielding emergent linguistic phenomena, such as compositionality \parencite{Beckner:2017, Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, semantic category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, regularization \parencite{Smith:2010}, and argument marking \parencite{Motamedi:2021}, among many other things.
 For reviews, see \textcite{Bailes:2023, Kirby:2014, Kirby:2017, Smith:2022, Tamariz:2017}.
@@
 All data and code is available from \url{https://osf.io/7auw6/?view_only=f8c2498b4add41839f51bdbefa03392e}.
-Both experiments were preregistered at \url{https://aspredicted.org/YDV_MXV}.
-Although we did not specify a precise statistical protocol (due to the more exploratory nature of our experiments), our preregistration does cover hypotheses, conditions, measured variables, sample size, and exclusion criteria.
+\newmaterial{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}.
+Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}
 
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/stimuli.pdf}}
 	\vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/transmission.pdf}}
 	\vspace*{2pt}
@@
 
-All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, which lasted around 8~min}.
-Participants were also told explicitly that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
+All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, lasting around 8~min}.
+Participants were also told explicitly before starting the training session that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
 In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each.
@@
 To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.
+\newmaterial{The instructions provided to participants are provided in Appendix~A in the supplementary material.}
 
@@
 
-Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant).
-We felt this was more ecological than giving both participants identical input.}} \newmaterial{The communication game, which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
+Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant.
+Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant).
+We felt this was more ecological than giving both participants identical input.}} The communication game, \newmaterial{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
 On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array).
-The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).
+The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials).
 
@@
 The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target.
-\newmaterial{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.} Overall, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different.
-In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.
+\newmaterial{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.}
+
+Overall, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different.
+In the non-communicative test, the goal and reward structure are based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure are based on successfully communicating a target object.
 
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_lrn.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/dif_lrn.pdf}}
 	\vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_com.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/dif_com.pdf}}
 	\vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/typology.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/typology.pdf}}
 	\vspace*{2pt}
@@
 	\end{figure}
-
-\subsubsection{Typological distribution}
 
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp1.pdf}}
 	\vspace*{2pt}
-	\caption{Upper panels show the typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
-Bottom panels show posterior differences (mean and 90\% HDI) in the proportion of chains falling into each typological category; positive values signify greater use of a given category in the communicative condition.}
-	\label{typ_dist_dif}
+	\caption{Results of Experiment~1.
+\textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+\textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation.
+The dots show the observed proportions and the curves show logistic regression models fit to the data.
+\textbf{C} Communicative cost by generation along with regression models fit to the data.
+\textbf{D} Transmission error by generation along with regression models fit to the data.
+The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$) parameters by condition as well as the posterior differences between conditions.
+The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
+	\label{results_exp1}
 	\end{figure}
 
-The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, revealing the proportion of the 10 chains that fall into each typological category at each generation.
+The typological distributions are plotted in Fig.~\ref{results_exp1}A, revealing the proportion of the 10 chains that fall into each typological category at each generation.
 In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2.
@@
 There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7).
-To quantify our uncertainty about these distributions and compare conditions statistically, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
-	\begin{equation}
-	\begin{aligned}
-	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
-	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
-	\end{aligned}
-	\end{equation}
-and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$.
-The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}.
-In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7).
-Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.
+Although there was not much evidence of expressive systems emerging in either condition, it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.
 
-\subsubsection{Communicative cost}
+\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate).
+We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain).
+Bernoulli likelihood with logit link function, weakly informative default priors, six chains of 12000 samples.
+All statistical models were fit using Bambi 0.13 \parencite{Capretto:2022}.
+Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}.
+Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, representing the intercept of the regression model, and $\beta$ representing its slope.
+To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero.
+Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero.
+We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision; the further the probability mass is from zero, the stronger the evidence of an effect.
+The results are shown in Fig.~\ref{results_exp1}B.
+In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure.
+The difference in slopes, $\Delta(\beta) = 4.53$ (95\%~HDI: 1.48, 7.88), clearly rejects zero, pointing to a meaningful difference between conditions.}
 
-As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}.
-This measure expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
-If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system.
+\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are.
+In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0).
+We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
+If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero~bits---that is, zero information will be lost during each attempt to communicate using that system.
 Communicative cost will take some larger value if the system contains ambiguity.
-It is given by
-	\begin{equation}
-	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
-	\end{equation}
-where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$.
+It is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$, where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$.
 In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.
+\newmaterial{The results are shown in Fig.~\ref{results_exp1}C.
+We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian).
+In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure.
+However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$).
+Although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}
 
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
-	\vspace*{2pt}
-	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
-The thin lines show each individual transmission chain (with a small amount of added jitter).
-The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
-Generation~0 is not included in the model, since it does not represent participant data.
-The black dotted lines show the communicative cost of the spoken language for comparison.
-In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
-	\label{cost_dif}
-	\end{figure}
-
-Fig.~\ref{cost_dif}A plots communicative cost over the nine generations of cultural transmission in each condition.
-The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling.
-In both conditions, cost increases with generation, implying that the languages become less informative over time.
-This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent.
-In contrast, this is not the expected result in Transmission~+ Communication, where the communicative pressure should promote the emergence of informative systems.
-This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization.
-Interestingly, however, cost plateaus at a slightly lower level in Transmission~+ Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.
-
-To evaluate this argument statistically, we fit the following regression model:
-	\begin{equation}
-	\begin{aligned}
-	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-	 \mu_i & =    \alpha + \beta_1 g_i + \beta_2 g_i^2 + \beta_3 g_i^3 \\
-	\sigma & \sim \mathrm{Exponential}(1),
-	\end{aligned}
-	\end{equation}
-where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~$-4$ through $+4$).
-The model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by a third-degree polynomial model.
-The intercept $\alpha$ corresponds to Generation~5 (the central generation), and the $\beta$ parameters determine the shape of the polynomial.
-$\sigma$ represents an across-participant noise term and is given a regularizing exponential prior.
-The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the complete model specification, see Appendix~A.
-
-The posterior parameter estimates of the model are shown in Fig.~\ref{cost_dif}B.
-Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do differ.
-This may be interpreted as the polynomial regression models adopting approximately the same shape but plateauing at different levels of cost.
-Additionally, we find that the $\alpha$ parameters themselves are lower than the cost of the spoken language (indicated by the dotted line), although it is unclear how this situation might unfold beyond Generation~9, especially in Transmission-only where the model predicts that cost will continue to increase further until all orthographic systems become aligned with the spoken language.
-
-Finally, cost may be converted into the expected probability of communicative success\footnote{Note that we can compute this quantity regardless of which condition a participant was in.
-By expected probability of communicative success, we mean the level of success a pair of interlocutors could hypothetically achieve using a given system, regardless of whether that system emerged in the Transmission-only or Transmission~+ Communication condition.
-This also assumes that the hypothetical interlocutors have a perfect understanding of the same shared system.} with the equation $p=2^{-c}$.
-If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: [35.1, 46.3]) in Transmission-only and 50.6\% (90\% HDI: [45.9, 54.9]) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition.
-Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that a degenerate system would result in a communicative success rate of~33.3\% (i.e., chance-level performance), while an expressive system would result in a~100\% success rate, while also being fairly easy to learn.
-
-\subsubsection{Learnability}
-
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/te_dif.eps}}
-	\vspace*{2pt}
-	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
-The thin lines show each individual transmission chain (with a small amount of added jitter).
-The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
-Generation~0 is not included in the model, since it does not represent participant data.
-Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
-	\label{te_dif}
-	\end{figure}
-
-\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
+\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted in our preregistration that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
 In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form.
-In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---these were not particularly forthcoming).
-We operationalize learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$.
-Following other papers, we define transmission error as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations.}
-
-\newmaterial{The results are plotted in Fig.~\ref{te_dif}, along with the posterior parameter estimates of a regression model fit to the data (this is the same model outlined above, except that we use a 2nd degree polynomial which better characterized the data).
-In Transmission-only, transmission error tends toward zero as the systems become trivial to learn and reproduce.
-In Transmission~+ Communication, there is some slight evidence for an uptick in transmission error in the last few generations as participants start to introduce novel disambiguating forms (notably English color letters), which diverge from the previous generation.
-Although there is little evidence to suggest that the shapes of the curves differ between conditions, there is a relatively clear difference in the intercepts: Transmission error remains higher in the communicative condition.
-This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates some change to the system.
-Nevertheless, the effect size is rather small (STATS), suggesting again that the presence of the communicative task did not have a big impact on how the systems evolved.}
+In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---expressive systems rarely emerged).
+Following prior work, we operationalized learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$; transmission error is defined as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations \parencite[see e.g.,][]{Kirby:2008}.
+These results are plotted in Fig.~\ref{results_exp1}D.
+In both conditions, the estimates of the $\beta$ parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time.
+There was some weak evidence for a difference in the intercepts, with transmission error remaining higher in the communicative condition.
+This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates more change to the system that was learned.}
 
@@
 
-Our first experiment tested whether informative, heterographic orthography could emerge through differentiation and whether it would emerge preferentially under communicative pressure.
-Overall, there was little evidence of systematic orthographic differentiation.
-Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <x> for pink, <f> for yellow, and <m> for blue).
+Our first experiment tested whether informative, heterographic orthography could emerge through spelling differentiation and whether it would emerge preferentially under communicative pressure.
+Although there was evidence to suggest that the orthographic systems remain informative for longer under communicative pressure, both conditions ultimately converged on degenerate, uninformative systems and there was little evidence of systematic differentiation in spelling.
+Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <ko> for pink, <co> for yellow, and <qo> for blue), but also through less obvious strategies such as the use of length (e.g., <ko> for pink, <koo> for yellow, and <kooo> for blue) or the use of arbitrary silent letters (e.g., <kox> for pink, <kof> for yellow, and <kom> for blue).
 Such differentiation was not forthcoming, however, even under communicative pressure.
@@
 The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color.
+\newmaterial{In support of this explanation, we conducted an additional experiment during review, which showed that when homophony is removed, the systems tend to resist degeneration in line with prior work.
+We discuss this experiment in more detail in the Discussion section and in Appendix~C of the supplementary material.}
+
 Faced with orthographic variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language.
 As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
-\newmaterial{Similar outcomes have been noted before in the context of artificial language learning \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
+\newmaterial{Similar outcomes have been noted before in the context of artificial language learning experiments \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
+\newmaterial{Interestingly, however, these redundant systems were relatively uncommon under communicative pressure, suggesting that communicating participants recognized the futility of using the suffix to mark shape.}
 
-It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite there being a clear financial incentive to do so.
-Why might this be the case?
-Firstly, although innovation was not expressly forbidden in the instructions, it is possible that some participants may have inferred such a rule.
-Participants who adopted such an interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking these perceived rules or because they did not know whether their partner would feel comfortable breaking them.
-Secondly, even when participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners?
-They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.
-These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space.
-Additionally, deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>).
-Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the intention?
-Differentiation is hard in the written modality.
+Overall, although the orthographic systems tended to remain slightly more informative under communicative pressure, the emergent orthographies ultimately preferred to transparently encode sound rather than meaning.
+This finding seems to align with our general experience of the world: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the intention?
+Without top-down diktat, it is hard to get spelling differentiation off~the~ground in the written modality.
 
@@
 The methods were identical to Experiment~1 with two exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly homophonous and uninformative over time.
+This was designed to model the historical processes of sound change and conservation described in the Introduction.
 
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/sound_change.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/sound_change.pdf}}
 	\vspace*{2pt}
@@
 Crucially, the spellings did not automatically change following a sound change event; rather, the orthographic system was free to adapt (or not) in response to the sound changes.
-Note also that the participants did not directly experience the sound changes; a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
+Note also that individual participants did not directly experience the sound changes; a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds).
 In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/con_lrn.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/con_lrn.pdf}}
 	\vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/con_com.pdf}}
 	\vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_con.eps}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp2.pdf}}
 	\vspace*{2pt}
-	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
-At generation~0, all 10 chains in both conditions use an expressive system (by design).
-In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission~+ Communication, they tend to remain expressive for longer and disfavor the redundant pattern.
-Posterior difference (mean and 90\% HDI) in the proportion of chains falling into each typological category.
-Positive values signify greater use of a given typological category in the communicative condition (and vice versa for negative values).}
-	\label{typ_dist_con}
+	\caption{Results of Experiment~2.
+\textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+\textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation.
+The dots show the observed proportions and the curves show logistic regression models fit to the data.
+\textbf{C} Communicative cost by generation along with regression models fit to the data.
+\textbf{D} Transmission error by generation along with regression models fit to the data.
+The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$ and $\gamma$) parameters by condition as well as the posterior differences between conditions.
+$\beta$ represents the effect of epoch and $\gamma$ represents the effect of generation number within epoch.
+The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
+	\label{results_exp2}
 	\end{figure}
 
-\subsubsection{Typological distribution}
+\newmaterial{Our quantitative analysis of Experiment~2 is identical to that of Experiment~1 with a slight change to the statistical model.
+Rather than predict the outcome variables as a function of generation, we now predict the outcome variables as a function of the epoch number (1, 2, or 3) and the generation number within the epoch (1, 2, or 3).\footnote{Model: dependent $\sim$ epoch + (1 + epoch | chain) + generation\_in\_epoch + (1 + generation\_in\_epoch | chain).} This yields two slope parameters: $\beta$, which represents the effect of epoch, and $\gamma$, which represents the additional effect of generational turnover.
+This model is more appropriate to the Experiment~2 setup, where the experimentally induced homophony results in discontinuities from one epoch to the next, and it allows us to separate out the effect of the homophony pressure from the more general effect of generational turnover.}
 
-Fig.~\ref{typ_dist_con} plots the typological distributions by generation and condition.
-Initially, all systems are expressive, but the dominance of this typological category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
+Fig.~\ref{results_exp2}A plots the typological distributions by generation and condition.
+Initially, all systems are expressive, but the dominance of this category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
 Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular under communicative pressure.
-The clearest difference between conditions comes at Generation~7, at which point there were more expressive systems and fewer redundant systems in the communicative condition.
+\newmaterial{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time.
+The results, shown in Fig.~\ref{results_exp2}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition.
+Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this measure.}
 
-\subsubsection{Communicative cost}
+\newmaterial{The results in terms of the more nuanced communicative cost measure are presented in Fig.~\ref{results_exp2}C.
+Here we find a nonzero effect of both epoch and generation on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
+Like the previous analysis, the $\gamma$ slopes were in close alignment.
+This suggests that the effect of generational turnover is very similar between conditions and that the difference between conditions is mostly driven by the increases in homophony induced in each epoch.
+The overall result is that, in Transmission~+ Communication, the increase in cost is linear across the nine generations, whereas in the Transmission-only condition, the increase in cost follows something more akin to a step function, with sudden increases in cost in response to each additional bout of homophony.
+In other words, in the Transmission-only condition, the orthographic systems respond rapidly to the changing spoken forms, while in the Transmission~+ Communication condition, the orthographic systems are more resistant to the homophony.}
 
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
-	\vspace*{2pt}
-	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
-The thin lines show each individual transmission chain (with a small amount of added jitter).
-The thick lines and shaded regions show the mean and 90\% HDI of a regression model fit to the data.
-Generation~0 is not included in the model, since it does not represent participant data.
-The black dotted lines show the communicative cost of the spoken language, which increases with epoch.
-In general, the orthographic systems tend to remain more informative under communicative pressure.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
-	\label{cost_con}
-	\end{figure}
-
-Fig.~\ref{cost_con}A plots communicative cost.
-During Epoch~I, there is a shallow increase in cost in both conditions, representing a baseline level of noise when the spoken language is fully informative.
-During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a comparatively shallow rise in the Transmission~+ Communication condition, presumably due to the communicative pressure, which slows the rate at which the systems become transparent or redundant.
-In Epoch~III, cost plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change.
-In Transmission~+ Communication, by contrast, cost continues to increase at much the same rate as in Epoch~II.
-
-To evaluate statistically the effect of generation on cost across the two conditions, while also taking epoch into account, we fit the following regression model:
-	\begin{equation}
-	\begin{aligned}
-	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
-	 \mu_i & =    \alpha_{\mathrm{EPOCH}[i]} + \beta_{\mathrm{EPOCH}[i]} g_i \\
-	\sigma & \sim \mathrm{Exponential}(1),
-	\end{aligned}
-	\end{equation}
-where $g$ is an epoch-centered generation number (i.e., for Generations~1, 4, and~7, $g=-1$, for Generations~2, 5, and~8, $g=0$, and for Generations~3, 6, and~9, $g=1$).
-Overall, the model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by an epoch-specific linear model (i.e., each epoch has an independent intercept $\alpha$ and slope $\beta$).
-$\sigma$ represents an across-participant noise term and is given a regularizing exponential prior.
-The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors.
-For the complete model specification, see Appendix~B.
-
-The posterior parameter estimates are shown in Fig.~\ref{cost_con}B.
-The $\alpha$ parameters, which represent the middle generation of each epoch (Generations~2, 5, and 8 respectively), reveal the gradual divergence of the conditions: By Epoch~III, the difference is $-0.43$~bits (90\% HDI: [$-0.83$, $-0.04$]), meaning that the communicative condition is about half a bit more informative.
-Further, after Epoch~I, the orthographic systems tend to remain more informative than the spoken language, especially under the communicative condition.
-The $\beta$ parameters, which represent the slopes within each of the three epochs are not conclusively different between conditions, although there is some evidence to suggest that the slope is shallower in the communicative condition during Epoch~II (e.g., 89\% of the posterior probability mass on $\Delta(\beta_2)$ is below zero).
-In the case of $\beta_3$ the effect is reversed, with Transmission~+ Communication having a greater slope.
-Note, however, that this appears to be because cost has, by this stage, plateaued in the Transmission-only condition, so the increased homophony pressure introduced in Epoch~III has little effect in Transmission-only, but continues to exert an effect in Transmission~+ Communication.
-
-At Generation~7---the point at which the two conditions differ most substantially---the difference in cost is $-0.6$~bits (90\% HDI: [$-1.04$, $-0.14$]).
-Converting cost to the probability of communicative success, we obtain an estimate of 46.5\% (90\% HDI: [38.1, 54.7]) in Transmission-only and 70.9\% (90\% HDI: [52.8, 87.8]) in Transmission~+ Communication, representing an advantage of around 25 percentage points for the communicative condition.
-The equivalent value for the spoken language at Generation~7 is 33.3\% (i.e., because in Epoch~III all three colors are expressed homophonously).
-Thus, not only are the two conditions different from each other, but also the two conditions are more informative than the spoken language, especially under communicative pressure.
-
-\subsubsection{Learnability}
-
-	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[]{figs/te_con.eps}}
-	\vspace*{2pt}
-	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange).
-The thin lines show each individual transmission chain (with a small amount of added jitter).
-The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data.
-Generation~0 is not included in the model, since it does not represent participant data.
-Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher.
-\textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
-	\label{te_con}
-	\end{figure}
-
-\newmaterial{Fig.~\ref{te_con} plots results for transmission error following the same methods outlined previously.
-Overall there were no strong differences between conditions, although we did not particularly expect to find any.
-Descriptively, in Transmission-only transmission error follows an inverted-\textsc{u} shape; it first increases in response to the homophony introduced at Generation~4, but then decreases as the orthographic systems converge on easy to learn patterns.
-In Transmission~+ Communication, the transmission error tend to remain fairly stable through Epochs~II and~III, representing a kind of baseline level of error in response to homophony.}
+\newmaterial{For completeness, Fig.~\ref{results_exp2}D also plots transmission error; however, we did not hypothesize any particular differences in learnability in Experiment~2, either over time or by condition.
+The expressive orthographic systems used to initialize the chains start out very easy to learn, and learnability remains fairly consistent throughout the experiment in both conditions, albeit with some constant level of change over time as the systems gradually come into alignment with the spoken forms.}
 
@@
 This is to be expected under Transmission-only, where the systems are adapting under learnability pressure, but is somewhat surprising in Transmission~+ Communication.
-Our findings ultimately suggest that, in the long term, (alphabetic) orthographic systems favor the faithful encoding of speech over the useful encoding of meaning, although there may exist short windows of time during which informative heterography can resist the spoken language.
-This points to a preference that readers might possess---those accustomed to an alphabetic writing system, at least---in which the primary goal of writing is seen as representing speech, not representing meaning directly.
-Interestingly, although participants were quite resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual.
+Our findings ultimately suggest that, in the long term, alphabetic orthographic systems might favor the faithful encoding of speech over the useful encoding of meaning, although there may exist brief windows of time during which informative heterography can resist the spoken language.
+Interestingly, although participants were resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual.
 Tradition has a powerful hold over writing systems.
@@
 
-It is widely understood that the written and spoken forms of a language are not identical; they diverge in many ways as a result of the differing constraints relevant to each modality.
-Spacing between words for example does not exist in speech, but constitutes a useful innovation in writing that permits rapid reading \parencite{Rastle:2019, RaynerFischer:1998, Sainio:2007, Zang:2013}.
-Might it be the case, then, that the written form of a language will become, over time, better adapted to the needs of writers and readers to the detriment of alignment with the spoken form of the language?
-Do writing systems adapt to the affordances and constraints of the written modality?
+The written and spoken forms of a language are never perfectly identical; they diverge in many ways as a result of the differing constraints relevant to each.
+Spacing between words for example does not exist in speech but constitutes a useful innovation in writing that permits rapid reading \parencite{RaynerFischer:1998, Sainio:2007, Zang:2013}.
+Similarly, the consistent spelling of affixes, such as the English past-tense marker <-ed>, which diverges from its spoken realization (/-d/, /-t/, or /-ɪd/ depending on the preceding sound), permits faster access to meaning \parencite{}.
+Might it be the case that---left to evolve freely---the written form of a language will become better adapted to the needs of writers and readers to the detriment of its alignment with the spoken form of the language?
+Do writing systems adapt to the affordances and constraints of the written modality \parencite{Rastle:2019}?
 
-We addressed these questions by focusing on the particular case of heterographic homophones---words that sound the same but are spelled differently.
+We addressed these questions by focusing on the particular case of heterographic homophones---morphemes that sound the same but that are spelled differently.
 Heterographic homophones permit the written language to be more informative than the spoken language; the written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
@@
 
-Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
-This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color.
-Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems.
-Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors.
-First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech.
-Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express.
-\newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).
-This highlights the important role of generational turnover, since new learners will be more receptive to new functions.} Second, even when participants \textit{do} appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome within so few trials.
+Although the emergent orthographies tended to be slightly more informative under communicative pressure and the informativeness persisted for more generations, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
+This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were retained but ineffectually conditioned on shape rather than color.
+Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn and reproduce such systems.
+Instead, it seems that, in Experiment~1, differentiation could not get off the ground.
+We see two reasons for this.
+First, participants seemed disinclined to directly encode meaning in how they chose to spell, preferring instead to ``write by ear'' \parencite{Frith:1979}.
+When asked to type in a word for a pink pentagon called /buvɪkəʊ/, participants were inclined to type a sequence of graphemes that reflected the sound they heard, without encoding meaning.
+\newmaterial{This behavior might be connected to the concept of ``functional fixedness'' \parencite{German:2000}, which states that learners find it difficult to adduce a new function (e.g., writing meaning) when they are accustomed to another function (e.g., writing sound), which highlights a potentially important role for generational turnover in the development of writing systems, since new learners will be more receptive to new functions.} Second, even when participants \textit{did} appreciate the need to differentiate the written forms to be successful, they often appeared reluctant or unable to do so, perhaps because they viewed the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---was too difficult to overcome \newmaterial{without the ability to coordinate over an extended period of time}.
 It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
 
+\newmaterial{This conclusion is in partial agreement with work by \textcite{Treiman:2015}.
+In a series of experiments, participants were asked to provide spellings for novel English words (e.g., /hæf/ meaning \textit{alehouse}) that were homophonous with preexisting English words (in this case, \textit{half}).
+Participants tended to provide the same spelling as the preexisting word (i.e., <half>) rather than other possible alternatives that would have had the benefit of differentiating meaning (e.g., <haf>, <haff>, <haph>).
+The authors argue that participants prefer the ``lesser effort that is required to use a familiar whole-word orthographic form compared to that needed for assembling a novel spelling'' (p.~544), which aligns with our findings.
+\textcite{Treiman:2015} also found, however, that, when given two alternatives to choose from (e.g., <half> vs.\@ <haff>), participants generally did prefer the novel spelling.
+This runs contrary to our first experiment, since our participants are similarly provided with multiple possible spellings of the sound /kəʊ/ (<coe>, <koh>, <qo>, etc.), but they nevertheless tended not to condition these on the color dimension, even under communicative pressure with financial incentive.}
+
+\newmaterial{During review, a concern was raised that our participants might not fully understand the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling.
+Alternatively, it was posited that our experiments might suffer from insufficient power.
+Our position, as outlined above, was that the lack of differentiation was due to the very strong homophony pressure.
+To test whether the lack of informativeness might be attributed to the homophony and to check that the design of our experiment and implementation of the communicative pressure was, in~principle, sufficient to promote more informative systems, we ran an additional experiment that entirely removed the spoken form, thereby replicating previous iterated learning experiments, which are generally only orthographic in nature \parencite[e.g.,][]{Kirby:2008, Kirby:2015, Beckner:2017}.
+This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness, and a commensurate increase in communicative success (see Appendix~D).
+This suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred the spellings to encoded sound rather than meaning.
+Or, rather, the cultural evolutionary process ultimately tended to favor simplicity over informativeness in this particular domain, amplifying the participants' (potentially relatively weak) preference for ``lessor effort.''}
+
 In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation.
-Under this mechanism, an expressive orthography does not emerge to fulfill an adaptive role but rather it is simply maintained from an earlier form of the spoken language that was expressive.
-This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}.
-That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may have been preserved, at least in part, precisely because of the informativeness they provide in reading.
-If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
+Under this mechanism, an expressive orthography does not emerge to fulfill an adaptive role but is simply a fossil representing an earlier form of the spoken language that \textit{was} expressive of a particular meaning distinction.
+\newmaterial{Over time, the orthography may experience a ratcheting effect, in which heterographic forms accumulate (due to successive sound changes) but rarely recede (due to the informativeness they provide).
+Over longer periods of time, this mechanism might even shift an orthography from a phonographic principle to a logographic one.} This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into written forms \parencite{Berg:2021}, \newmaterial{and which are sometimes argued to give English a semi-logographic character \parencite{Chomsky:1968, Coulmas:1991, DeFransis:1989, Zachrisson:1931}}.
+To be clear, this is not to say that such heterography is nonadaptive or an accident of history; rather, such heterographs may have been preserved precisely \textit{because} of the informativeness they inadvertently provide in reading.
+\newmaterial{Thus, just as the spoken language avoids \textit{sound} mergers that increase ambiguity \parencite[e.g.,][]{Wedel:2013}, so the written language might likewise avoid \textit{spelling} mergers that increase ambiguity.} If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
 
@@
 The additional level of conservation that occurs in Transmission~+ Communication corresponds to the repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly).
-The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure.
-Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and based on the trajectories of the communicative cost curves (Fig.~\ref{cost_con}), it seems likely that all chains will ultimately undergo the same transformation eventually.
+The second caveat is that, in the long term, it appears that transparency might ultimately win the day, even under communicative pressure.
+Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and based on the trajectories of the communicative cost curves (Fig.~\ref{results_exp2}C), it seems likely that all chains will ultimately undergo the same transformation eventually.
 Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.
 
-It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible.
+\newmaterial{The results of both experiments were, however, relatively weak statistically, with the differences in slopes often only just (or not quite) meeting the 95\% criterion.
+This perhaps reflects a situation of low power, with only ten chains---ten independent sampling units---in each condition.
+This is a general issue with iterated learning experiments, however, and our use of ten chains follows norms in the field.} It is also important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible.
 Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2.
 The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm.
-Similarly, we cannot draw direct conclusions from the relative strength of the results across the two experiments.
-For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate.
+Similarly, we cannot draw direct conclusions from the relative strength of the statistical results across the two experiments.
 
@@
 For \textit{differentiation} to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming social stigma in the process (the social problem).
-\newmaterial{Furthermore, once a system has been created, it needs to be reliably transmitted across multiple generations, perhaps with extensive training (the learnability problem).} The \textit{conservation} of an expressive orthography is, by~comparison, plain~sailing---it is only the learnability problem that applies.
+Furthermore, once a system has been created, it needs to be reliably transmitted across multiple generations (the learnability problem).
+The \textit{conservation} of an expressive orthography is, by~comparison, plain~sailing---it is only the learnability problem that applies.
 More generally, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.
@@
 That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography.
-An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers.
-\newmaterial{In many respects, writing systems are cultural fossils that do not readily adapt to external pressures.}
-
-It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many.
+An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers, \newmaterial{especially given that writing systems are often fixed cultural fossils that do not readily adapt to external pressures.} It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many.
 For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space.
 From this perspective, written forms of language resist change because they need to be accessible across decades or centuries and across different jurisdictions or dialect areas.
-\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be less forthcoming under other populations (i.e., participants with transparent native languages, such as Finnish, Italian, or Serbian).
-We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower because the community first has to evolve the concept of directly encoding meaning in the written form.
-English, for example, was at one point much more transparent than it is now, but it has gradually become more logographic.
-In this sense, our experiments jumpstart the process by adopting a participant population already accustomed to these ideas.}
+
+\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to transparent writing systems.
+This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage---whether that baggage is for the encoding of sound or the encoding of meaning.
+Ideally, our experiments would be performed with participants who have no writing experience at all, but since that would be very difficult to achieve, perhaps the second-best option is a participant population that is relatively open-minded to both types of writing systems.
+In this sense, our use of English speakers is actually quite appropriate, since the English writing system is neither fully phonographic nor fully logographic.}
 
@@
 Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it.
-Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.
+Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies \newmaterial{and our Experiment~3}---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.
 
@@
 
-As a final note, heterography is often derided as a source of unnecessary complexity that language learners must contend with.
-However, our results show that, when the communicative need for disambiguation is strong enough, language users may prefer to trade some simplicity for greater informativeness in the writing system.
-Perhaps, then, efforts to reform and simplify spelling may, to some extent, be a little hasty, and more recognition should be given to the benefits that opaque orthographies provide in the context of reading.
+As a final note, heterography is often derided as a source of unnecessary complexity, \newmaterial{and as a consequence, the orthographic reforms implemented in many languages have tended to focus on its elimination.} However, our results suggest that, when the communicative need for disambiguation is strong enough, language users may prefer to trade some simplicity for greater informativeness in the writing system.
+Perhaps, then, efforts to reform and simplify spelling may be a little hasty, since, ultimately, the writing system should help the reader get to the meaning of a word rather than its sound.
 
@@
 
-\section{}
+\section{Participant instructions}
 
-\begin{equation}
-\begin{aligned}
-   c_i                   & \sim \mathrm{Normal}(\mu_i, \sigma) \\
- \mu_i                   & = \alpha_{\mathrm{CHAIN}[i]} + \beta_{1,\mathrm{CHAIN}[i]} g_i + \beta_{2,\mathrm{CHAIN}[i]} g_i^2 + \beta_{3,\mathrm{CHAIN}[i]} g_i^3 \\
-\sigma                   & \sim \mathrm{Exponential}(1) \\
-\alpha_{\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\alpha_\mu, \alpha_\sigma) \\
-\beta_{j,\mathrm{CHAIN}} & \sim \mathrm{Normal}(\beta_{j_\mu}, \beta_{j_\sigma}) \\
-\alpha_\mu               & \sim \mathrm{Normal}(1, 1) \\
-\alpha_\sigma            & \sim \mathrm{Exponential}(1) \\
-\beta_{j_\mu}            & \sim \mathrm{Normal}(0, 1) \\
-\beta_{j_\sigma}         & \sim \mathrm{Exponential}(1)
-\end{aligned}
-\end{equation}
+\begin{figure}[h!]
+\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/instructions.pdf}}
+\vspace*{2pt}
+\caption{\textbf{A} Training instructions.
+\textbf{B} Explicit prefix training.
+\textbf{C} Test instructions in the Transmission-only condition.
+\textbf{D} Communication game instructions in the Transmission + Communication condition (an animation, which can be viewed in the OSF repository, showed examples of successful and unsuccessful trials).}
+\label{supp_training}
+\end{figure}
+
+\clearpage
+
+\section{Posterior parameter estimates and diagnostics}
+
+These tables provide detailed posterior parameter estimates for each of the three experiments, including the posterior differences between conditions.
+The means and standard deviations characterize the central tendency and width of the posterior.
+The 95\% HDI is the narrowest interval that includes 95\% of the posterior probability mass.
+The effective sample size (ESS) is the effective number of samples drawn from the posterior after accounting for autocorrelation in the MCMC chains; values above 10,000 are recommended for numerical stability.
+$\hat{R}$ is a measure of convergence across the independent MCMC chains used to sample from the posterior; values close to 1 indicate high convergence.
 
@@
 \begin{threeparttable}
-\caption{Experiment 1 posterior parameter estimates}
+\caption{Experiment 1}
 \footnotesize
-\begin{tabular}{lrrrrrr}
+\begin{tabular}{lrrrlrr}
 \toprule
- & Mean & SD & HDI (5\%) & HDI (95\%) & ESS & $\hat{R}$ \\
+ & Mean & SD & \multicolumn{2}{c}{95\% HDI} & ESS & $\hat{R}$ \\
 \midrule
-\multicolumn{7}{c}{Transmission-only} \\
-$\alpha_\mu$    &  1.298 &  0.123 &   1.098 &    1.498 &   55213 &    1.0 \\
-$\beta_{1_\mu}$ &  0.003 &  0.030 &  -0.048 &    0.051 &   19478 &    1.0 \\
-$\beta_{2_\mu}$ & -0.021 &  0.008 &  -0.034 &   -0.009 &   49158 &    1.0 \\
-$\beta_{3_\mu}$ &  0.007 &  0.002 &   0.004 &    0.010 &   13927 &    1.0 \\
-$\sigma$        &  0.192 &  0.020 &   0.160 &    0.225 &   26492 &    1.0 \\
+\multicolumn{7}{c}{Probability informative} \\
+$\alpha_\mathrm{trans}$ & 6.77 & 1.79 & 3.51 & 10.43 & 51930 & 1.00 \\
+$\beta_\mathrm{trans}$ & -6.47 & 1.56 & -9.61 & -3.64 & 36904 & 1.00 \\
+$\alpha_\mathrm{comm}$ & 2.23 & 0.76 & 0.78 & 3.74 & 65662 & 1.00 \\
+$\beta_\mathrm{comm}$ & -1.94 & 0.55 & -3.05 & -0.91 & 34929 & 1.00 \\
+$\Delta(\alpha)$ & -4.54 & 1.95 & -8.41 & -0.82 & 51523 & 1.00 \\
+$\Delta(\beta)$ & 4.53 & 1.65 & 1.48 & 7.88 & 35250 & 1.00 \\
 \midrule
-\multicolumn{7}{c}{Transmission + Communication} \\
-$\alpha_\mu$    &  0.985 &  0.079 &   0.854 &    1.112 &   22089 &    1.0 \\
-$\beta_{1_\mu}$ &  0.003 &  0.059 &  -0.091 &    0.100 &   19115 &    1.0 \\
-$\beta_{2_\mu}$ & -0.017 &  0.007 &  -0.030 &   -0.005 &   15955 &    1.0 \\
-$\beta_{3_\mu}$ &  0.004 &  0.004 &  -0.002 &    0.010 &   12773 &    1.0 \\
-$\sigma$        &  0.370 &  0.034 &   0.313 &    0.425 &   27848 &    1.0 \\
+\multicolumn{7}{c}{Communicative cost} \\
+$\alpha_\mathrm{trans}$ & 0.14 & 0.08 & -0.03 & 0.30 & 68812 & 1.00 \\
+$\beta_\mathrm{trans}$ & 0.60 & 0.07 & 0.47 & 0.73 & 30143 & 1.00 \\
+$\alpha_\mathrm{comm}$ & 0.15 & 0.11 & -0.07 & 0.36 & 75737 & 1.00 \\
+$\beta_\mathrm{comm}$ & 0.43 & 0.07 & 0.28 & 0.57 & 57053 & 1.00 \\
+$\Delta(\alpha)$ & 0.01 & 0.14 & -0.26 & 0.28 & 74870 & 1.00 \\
+$\Delta(\beta)$ & -0.17 & 0.10 & -0.36 & 0.02 & 40473 & 1.00 \\
 \midrule
-\multicolumn{7}{c}{Difference} \\
-$\Delta(\alpha_\mu)$    & -0.313 &  0.147 &  -0.551 &   -0.072 &   38539 &    1.0 \\
-$\Delta(\beta_{1_\mu})$ & -0.000 &  0.066 &  -0.109 &    0.107 &   19664 &    1.0 \\
-$\Delta(\beta_{2_\mu})$ &  0.004 &  0.011 &  -0.014 &    0.022 &   24478 &    1.0 \\
-$\Delta(\beta_{3_\mu})$ & -0.003 &  0.004 &  -0.010 &    0.003 &   12532 &    1.0 \\
-$\Delta(\sigma)$        &  0.178 &  0.040 &   0.112 &    0.242 &   28446 &    1.0 \\
+\multicolumn{7}{c}{Transmission error} \\
+$\alpha_\mathrm{trans}$ & 0.94 & 0.08 & 0.77 & 1.11 & 105363 & 1.00 \\
+$\beta_\mathrm{trans}$ & -0.39 & 0.05 & -0.48 & -0.29 & 101977 & 1.00 \\
+$\alpha_\mathrm{comm}$ & 1.17 & 0.15 & 0.86 & 1.47 & 95553 & 1.00 \\
+$\beta_\mathrm{comm}$ & -0.35 & 0.10 & -0.54 & -0.17 & 58199 & 1.00 \\
+$\Delta(\alpha)$ & 0.23 & 0.18 & -0.12 & 0.57 & 98993 & 1.00 \\
+$\Delta(\beta)$ & 0.03 & 0.11 & -0.18 & 0.24 & 64129 & 1.00 \\
 \bottomrule
@@
 
-\clearpage
-
-\section{}
-
-\begin{equation}
-\begin{aligned}
-   c_i                                 & \sim \mathrm{Normal}(\mu_i, \sigma) \\
- \mu_i                                 & = \alpha_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} + \beta_{\mathrm{EPOCH}[i],\mathrm{CHAIN}[i]} g_i \\
-\sigma                                 & \sim \mathrm{Exponential}(1) \\
-\alpha_{\mathrm{EPOCH},\mathrm{CHAIN}} & \sim \mathrm{Normal}(\alpha_{\mathrm{EPOCH}_\mu}, \alpha_{\mathrm{EPOCH}_\sigma}) \\
-\beta_{\mathrm{EPOCH},\mathrm{CHAIN}}  & \sim \mathrm{Normal}(\beta_{\mathrm{EPOCH}_\mu}, \beta_{\mathrm{EPOCH}_\sigma}) \\
-\alpha_{\mathrm{EPOCH}_\mu}            & \sim \mathrm{Normal}(0, 1) \\
-\alpha_{\mathrm{EPOCH}_\sigma}         & \sim \mathrm{Exponential}(1) \\
-\beta_{\mathrm{EPOCH}_\mu}             & \sim \mathrm{Normal}(0, 1) \\
-\beta_{\mathrm{EPOCH}_\sigma}          & \sim \mathrm{Exponential}(1)
-\end{aligned}
-\end{equation}
-
 \begin{table}[h]
@@
 \begin{threeparttable}
-\caption{Experiment 2 posterior parameter estimates}
+\caption{Experiment 2}
 \footnotesize
@@
 \toprule
- &   Mean &     SD &  HDI (5\%) &  HDI (95\%) &  ESS &  $\hat{R}$ \\
+ & Mean & SD & \multicolumn{2}{c}{95\% HDI} & ESS & $\hat{R}$ \\
 \midrule
-\multicolumn{7}{c}{Transmission-only} \\
-$\alpha_{1_\mu}$ &  0.099 &  0.066 &  -0.009 & 0.205 &   50911 &    1.0 \\
-$\alpha_{2_\mu}$ &  0.530 &  0.134 &   0.320 & 0.748 &   58969 &    1.0 \\
-$\alpha_{3_\mu}$ &  1.113 &  0.155 &   0.869 & 1.370 &   55884 &    1.0 \\
-$\beta_{1_\mu}$  &  0.048 &  0.044 &  -0.021 & 0.121 &   32489 &    1.0 \\
-$\beta_{2_\mu}$  &  0.216 &  0.074 &   0.096 & 0.336 &   47941 &    1.0 \\
-$\beta_{3_\mu}$  & -0.000 &  0.040 &  -0.065 & 0.065 &   32801 &    1.0 \\
-$\sigma$         &  0.145 &  0.018 &   0.116 & 0.173 &   15490 &    1.0 \\
+\multicolumn{7}{c}{Probability informative} \\
+$\alpha_\mathrm{trans}$ & 14.14 & 3.45 & 7.75 & 20.96 & 75580 & 1.00 \\
+$\beta_\mathrm{trans}$ & -5.89 & 1.35 & -8.60 & -3.43 & 79605 & 1.00 \\
+$\gamma_\mathrm{trans}$ & -0.22 & 0.68 & -1.57 & 1.14 & 67339 & 1.00 \\
+$\alpha_\mathrm{comm}$ & 12.52 & 3.36 & 6.45 & 19.25 & 42630 & 1.00 \\
+$\beta_\mathrm{comm}$ & -3.24 & 0.98 & -5.22 & -1.39 & 56677 & 1.00 \\
+$\gamma_\mathrm{comm}$ & -0.98 & 0.64 & -2.25 & 0.28 & 70323 & 1.00 \\
+$\Delta(\alpha)$ & -1.62 & 4.82 & -11.03 & 7.90 & 47446 & 1.00 \\
+$\Delta(\beta)$ & 2.65 & 1.67 & -0.58 & 5.95 & 66432 & 1.00 \\
+$\Delta(\gamma)$ & 4.91 & 1.49 & 2.13 & 7.92 & 75152 & 1.00 \\
 \midrule
-\multicolumn{7}{c}{Transmission + Communication} \\
-$\alpha_{1_\mu}$ &  0.052 &  0.042 &  -0.018 & 0.118 &   30904 &    1.0 \\
-$\alpha_{2_\mu}$ &  0.297 &  0.102 &   0.138 & 0.469 &   56334 &    1.0 \\
-$\alpha_{3_\mu}$ &  0.679 &  0.184 &   0.386 & 0.981 &   63206 &    1.0 \\
-$\beta_{1_\mu}$  &  0.033 &  0.044 &  -0.040 & 0.102 &   18241 &    1.0 \\
-$\beta_{2_\mu}$  &  0.093 &  0.073 &  -0.027 & 0.209 &   39754 &    1.0 \\
-$\beta_{3_\mu}$  &  0.167 &  0.121 &  -0.033 & 0.359 &   53767 &    1.0 \\
-$\sigma$         &  0.175 &  0.020 &   0.141 & 0.207 &   17902 &    1.0 \\
+\multicolumn{7}{c}{Communicative cost} \\
+$\alpha_\mathrm{trans}$ & -0.63 & 0.14 & -0.89 & -0.35 & 64515 & 1.00 \\
+$\beta_\mathrm{trans}$ & 0.52 & 0.07 & 0.38 & 0.66 & 26030 & 1.00 \\
+$\gamma_\mathrm{trans}$ & 0.09 & 0.04 & 0.02 & 0.16 & 98284 & 1.00 \\
+$\alpha_\mathrm{comm}$ & -0.50 & 0.13 & -0.75 & -0.26 & 65698 & 1.00 \\
+$\beta_\mathrm{comm}$ & 0.32 & 0.07 & 0.18 & 0.46 & 24609 & 1.00 \\
+$\gamma_\mathrm{comm}$ & 0.10 & 0.04 & 0.02 & 0.18 & 89149 & 1.00 \\
+$\Delta(\alpha)$ & 0.13 & 0.18 & -0.24 & 0.49 & 66441 & 1.00 \\
+$\Delta(\beta)$ & -0.20 & 0.10 & -0.40 & -0.00 & 24592 & 1.00 \\
+$\Delta(\gamma)$ & -0.42 & 0.08 & -0.57 & -0.25 & 31416 & 1.00 \\
 \midrule
-\multicolumn{7}{c}{Difference} \\
-$\Delta(\alpha_{1_\mu})$ & -0.047 &  0.079 &  -0.178 &    0.077 &   40911 &    1.0 \\
-$\Delta(\alpha_{2_\mu})$ & -0.233 &  0.168 &  -0.504 &    0.040 &   55907 &    1.0 \\
-$\Delta(\alpha_{3_\mu})$ & -0.434 &  0.241 &  -0.825 &   -0.037 &   57860 &    1.0 \\
-$\Delta(\beta_{1_\mu})$  & -0.016 &  0.062 &  -0.118 &    0.084 &   23610 &    1.0 \\
-$\Delta(\beta_{2_\mu})$  & -0.124 &  0.105 &  -0.295 &    0.047 &   35386 &    1.0 \\
-$\Delta(\beta_{3_\mu})$  &  0.167 &  0.128 &  -0.045 &    0.371 &   50705 &    1.0 \\
-$\Delta(σ)$              &  0.030 &  0.027 &  -0.014 &    0.074 &   16452 &    1.0 \\
+\multicolumn{7}{c}{Transmission error} \\
+$\alpha_\mathrm{trans}$ & 0.18 & 0.16 & -0.14 & 0.50 & 133362 & 1.00 \\
+$\beta_\mathrm{trans}$ & 0.11 & 0.06 & -0.00 & 0.23 & 101042 & 1.00 \\
+$\gamma_\mathrm{trans}$ & -0.02 & 0.06 & -0.13 & 0.09 & 110157 & 1.00 \\
+$\alpha_\mathrm{comm}$ & 0.10 & 0.18 & -0.25 & 0.44 & 87045 & 1.00 \\
+$\beta_\mathrm{comm}$ & 0.16 & 0.06 & 0.05 & 0.28 & 95395 & 1.00 \\
+$\gamma_\mathrm{comm}$ & -0.02 & 0.06 & -0.14 & 0.09 & 95801 & 1.00 \\
+$\Delta(\alpha)$ & -0.08 & 0.24 & -0.55 & 0.39 & 110229 & 1.00 \\
+$\Delta(\beta)$ & 0.05 & 0.08 & -0.12 & 0.21 & 97655 & 1.00 \\
+$\Delta(\gamma)$ & -0.14 & 0.08 & -0.30 & 0.03 & 98052 & 1.00 \\
 \bottomrule
@@
 
+\begin{table}[h]
+\begin{center}
+\begin{threeparttable}
+\caption{Experiment 3 (compared to Exp. 1, Comm.)}
+\footnotesize
+\begin{tabular}{lrrrlrr}
+\toprule
+ & Mean & SD & \multicolumn{2}{c}{95\% HDI} & ESS & $\hat{R}$ \\
+\midrule
+\multicolumn{7}{c}{Probability informative} \\
+$\alpha_\mathrm{exp1}$ & 2.23 & 0.76 & 0.78 & 3.74 & 65662 & 1.00 \\
+$\beta_\mathrm{exp1}$ & -1.94 & 0.55 & -3.05 & -0.91 & 34929 & 1.00 \\
+$\alpha_\mathrm{exp3}$ & 1.46 & 0.66 & 0.19 & 2.78 & 74546 & 1.00 \\
+$\beta_\mathrm{exp3}$ & -0.25 & 0.38 & -1.02 & 0.49 & 73672 & 1.00 \\
+$\Delta(\alpha)$ & -0.76 & 1.01 & -2.76 & 1.21 & 63934 & 1.00 \\
+$\Delta(\beta)$ & 1.69 & 0.67 & 0.40 & 3.04 & 39662 & 1.00 \\
+\midrule
+\multicolumn{7}{c}{Communicative cost} \\
+$\alpha_\mathrm{exp1}$ & 0.15 & 0.11 & -0.07 & 0.36 & 75737 & 1.00 \\
+$\beta_\mathrm{exp1}$ & 0.43 & 0.07 & 0.28 & 0.57 & 57053 & 1.00 \\
+$\alpha_\mathrm{exp3}$ & 0.36 & 0.13 & 0.12 & 0.61 & 101323 & 1.00 \\
+$\beta_\mathrm{exp3}$ & 0.10 & 0.08 & -0.06 & 0.25 & 72628 & 1.00 \\
+$\Delta(\alpha)$ & 0.21 & 0.17 & -0.11 & 0.54 & 89412 & 1.00 \\
+$\Delta(\beta)$ & -0.33 & 0.11 & -0.54 & -0.12 & 62222 & 1.00 \\
+\midrule
+\multicolumn{7}{c}{Transmission error} \\
+$\alpha_\mathrm{exp1}$ & 1.17 & 0.15 & 0.86 & 1.47 & 95553 & 1.00 \\
+$\beta_\mathrm{exp1}$ & -0.35 & 0.10 & -0.54 & -0.17 & 58199 & 1.00 \\
+$\alpha_\mathrm{exp3}$ & 1.35 & 0.20 & 0.97 & 1.75 & 82673 & 1.00 \\
+$\beta_\mathrm{exp3}$ & -0.28 & 0.12 & -0.50 & -0.05 & 75632 & 1.00 \\
+$\Delta(\alpha)$ & 0.19 & 0.25 & -0.30 & 0.69 & 87400 & 1.00 \\
+$\Delta(\beta)$ & 0.08 & 0.15 & -0.22 & 0.37 & 66566 & 1.00 \\
+\bottomrule
+\end{tabular}
+\label{exp3_parameter_estimates}
+\end{threeparttable}
+\end{center}
+\end{table}
+
+\clearpage
+
+\section{Experiment 3}
+
+Our Experiment~1 results were surprising in that they failed to find much evidence for the emergence of clear, systematic, informative systems, even under communicative pressure with financial incentive.
+One possible concern put forward during review was that there might be some other spurious reason why participants failed to differentiate the spellings in the communicative condition.
+The Experiment~1 finding might perhaps have been explained by participants not understanding the communicative goal of the study or some other subtle methodological issue.
+Our position was that the lack of differentiation was due to the very strong homophony pressure; participants found it difficult to learn, create, and produce spellings that contradicted the spoken form.
+To test whether the lack of informativeness might be attributed to the homophony pressure and to check that the communicative pressure was, in~principle, sufficient to promote more informative systems, we ran a third experiment that entirely removed the spoken modality.
+
+\subsection{Methods}
+
+The methods were identical to the Transmission~+ Communication condition of Experiment~1 with two exceptions.
+First, no spoken forms were provided to participants during training or test.
+This removes any homophony pressure and brings the experimental design into alignment with most prior studies, which are generally only orthographic in nature \parencite[e.g.,][]{Kirby:2008, Kirby:2015, Beckner:2017}.
+Second, a new set of spellings was selected to initialize the chains.
+This was because our goal was to remove all traces of homophony but the Experiment~1 seed spellings (<coh>, <koe>, <co>, etc.) were likely to be perceived as homophonous, even if they were not explicitly given any phonological value.
+
+\subsubsection{Participants}
+
+The experiment was completed by 230 native-English participants, who were recruited through Prolific and paid £2.50 for participation.
+The median bonus was £0.70 and the median completion time was 17~m, resulting in a median hourly rate of £8.78 (£11.08 including bonus).
+26 participants were excluded because they (or their partners) used English color words, and a further 24 participants were lost to communication-game pairing failures.
+The final dataset comprises 180 participants (10 chains of 9 pairs of participants).
+
+\subsubsection{Stimuli}
+
+The suffixes used to initialize the chains consisted of a consonant grapheme from the set \{<k>, <m>, <s>\} concatenated with a vowel grapheme from the set \{<a>, <ei>, <oe>\}.
+The nine resulting suffixes were mapped to the nine possible objects randomly (without replacement).
+This procedure was identical to Experiment~1, except for the change in grapheme identities, which were chosen because their conventional phonological values are very different from each other (i.e., different places and manners of articulation).
+
+\subsection{Results}
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/sil_com.pdf}}
+	\vspace*{2pt}
+	\caption{Results from Experiment~3.
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns).
+Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
+Spellings in bold-italic are the generalizations on unseen items.}
+	\label{sil_com}
+	\end{figure}
+
+Fig.~\ref{sil_com} plots all suffix spellings by chain and generation.
+The most notable difference in comparison to Experiment~1 was the large number of times participants used English to signal color.
+Like the main experiments, we rejected participants who used full color words.
+The proportion of participants rejected for this reason was almost four times higher (10.4\% in Experiment~3 vs.\@ 2.8\% in Experiment~1).
+In addition, participants were much more likely to append English color letters: All ten chains converged on this solution, typically around Generation~4 or~5.
+In comparison, only five of the chains in Experiment~1 adopted this solution.
+In the absence of any spoken form and in the absence of any explicit instruction not to use English, participants appear to find the use of English to be the simplest strategy for marking color.
+Importantly, the widespread use of English suggests that participants did indeed understand the communicative need to mark color.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp3.pdf}}
+	\vspace*{2pt}
+	\caption{Results of Experiment~3.
+\textbf{A} Typological distribution by generation and condition over the four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow).
+\textbf{B} Proportion of systems classified as informative (holistic or expressive) by generation.
+The dots show the observed proportions and the curves show logistic regression models fit to the data.
+\textbf{C} Communicative cost by generation along with regression models fit to the data.
+\textbf{D} Transmission error by generation along with regression models fit to the data.
+The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$) parameters by condition as well as the posterior differences between conditions.
+The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
+	\label{results_exp3}
+	\end{figure}
+
+Aside from the use of English, there were also some instances of expressive systems emerging, notably E8 (<kei>, <yel>, <kol>) and H9 (<may>, <mei>, <mab>), but also B4, D3, E2 (semi-expressive systems).
+In addition, many of the cases of English color marking mentioned above can be analyzed as expressive systems if you reconsider which part of the word is the stem.
+For example, the system that emerged in Chain~F seems to treat the stems as <buvikoe->, <zetimoe->, <wopikoe-> and the suffixes as $\emptyset$ (pink), <-y> (yellow), <-b> (blue).
+
+The quantitative results are plotted in Fig.~\ref{results_exp3} and are compared against the equivalent results from Experiment~1 (the Transmission~+ Communication condition).
+Overall, the effect of removing the homophony pressure was to prevent the systems from degenerating over time---both measures of informativeness show a clear difference in slopes, and informativeness levels remain fairly constant over time in Experiment~3, mainly in the form of holistic systems.
+
+\subsection{Summary}
+
+Our first aim with Experiment~3 was to perform an additional check on whether participants understood the nature of the communicative task.
+By removing the homophony, Experiment~3 becomes equivalent to prior iterated learning studies, meaning that we should have been able to replicate the finding that informativeness remains high under communicative need.
+The results showed that participants were much more willing to innovate and differentiate when unconstrained by the presence of spoken forms.
+In particular, the widespread use of English as a strategy for marking color suggests that participants did indeed understand the need for disambiguation and were willing to diverge from their training material if it seemed advantageous to do so.
+Our second aim with Experiment~3 was to better understand whether the presence of homophonous spoken forms in Experiment~1 prevented the emergence of informative orthographic systems.
+The results suggested that this is indeed the case: The systems tended not to degenerate in Experiment~3 and instead tended to remain informative usually by adopting some form of English-based suffixes.
+
+Like Experiment~1, however, there was still a notable lack of expressive systems in Experiment~3.
+Why might this be the case?
+One possibility is that there was insufficient simplicity pressure to produce expressive systems (both in this experiment and the main experiments).
+Recall that the holistic and expressive systems are both informative about color; the expressive system is just a simpler way of marking color.
+With this in mind, expressive systems require not only sufficient pressure for informativeness (induced through the communicative task), but also sufficient pressure for simplicity (induced through learning).
+Given the small scale of our stimulus space---just nine items---it is perhaps the case that the simplicity pressure is not strong enough to give rise to the expressive solution; participants can achieve satisfactory performance without it.
+Consider also that, although expressive systems will be advantageous in the long run---making learning and communication easier for future generations of participant---it may not be particularly advantageous for the pair of participants at a given generation to radically alter the system; it may instead be more expedient to simply append English-based color markers onto the preexisting forms.
+
+\section{Communicative success}
+
+Fig.~\ref{success} compares the three experiments in terms of communicative success, which is defined as the proportion of trials in the communication game in which the comprehending participant selected the correct target item.
+In Experiment~1, communicative success typically remains around chance; the orthographic systems tend to remain uninformative about color over time, mirroring the spoken form of the language.
+In Experiment~3, by comparison, which removed the homophony pressure, communicative success does tend to increase over time, as the participants find communicative strategies.
+In Experiment~2, the orthographic systems started out expressive, and as a consequence communicative success is initially high.
+However, success drops over time as the systems become less informative, gradually mirroring the homophony in the spoken form.
+Nevertheless, success does remain high in some chains where the expressive systems were maintained.
+
+	\begin{figure}[h]
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/success.pdf}}
+	\vspace*{2pt}
+	\caption{Communicative success by generation in the communicative conditions of all three experiments.
+The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
+	\label{success}
+	\end{figure}
+
 \end{document}

### 2024-02-26 code/comm_cost.py
Remove redundant code

@@ -1,87 +1,0 @@
-from collections import defaultdict
-from itertools import product
-import numpy as np
-
-
-def convert_lexicon_meanings_to_tuple(lexicon):
-	converted_lexicon = {}
-	for item, signal in lexicon.items():
-		meaning = tuple(map(int, item.split('_')))
-		converted_lexicon[meaning] = signal
-	return converted_lexicon
-
-
-def communicative_cost(lexicon, dims):
-	if isinstance(list(lexicon.keys())[0], str):
-		lexicon = convert_lexicon_meanings_to_tuple(lexicon)
-	reverse_lexicon = defaultdict(set)
-	for meaning, signal in lexicon.items():
-		reverse_lexicon[signal].add(meaning)
-	U = product(*[range(n_values) for n_values in dims])
-	U_size = np.prod(dims)
-	return 1 / U_size * sum([-np.log2(1 / len(reverse_lexicon[lexicon[m]])) for m in U])
-
-
-if __name__ == '__main__':
-
-	degenerate = {
-		(0, 0): 'buvico',
-		(0, 1): 'buvico',
-		(0, 2): 'buvico',
-		(1, 0): 'zetico',
-		(1, 1): 'zetico',
-		(1, 2): 'zetico',
-		(2, 0): 'wopico',
-		(2, 1): 'wopico',
-		(2, 2): 'wopico',
-	}
-	holistic = {
-		(0, 0): 'buvico',
-		(0, 1): 'buviko',
-		(0, 2): 'buviqo',
-		(1, 0): 'zeticoe',
-		(1, 1): 'zetikoe',
-		(1, 2): 'zetiqoe',
-		(2, 0): 'wopicoh',
-		(2, 1): 'wopikoh',
-		(2, 2): 'wopiqoh',
-	}
-	redundant = {
-		(0, 0): 'buvico',
-		(0, 1): 'buvico',
-		(0, 2): 'buvico',
-		(1, 0): 'zetiko',
-		(1, 1): 'zetiko',
-		(1, 2): 'zetiko',
-		(2, 0): 'wopiqo',
-		(2, 1): 'wopiqo',
-		(2, 2): 'wopiqo',
-	}
-	expressive = {
-		(0, 0): 'buvico',
-		(0, 1): 'buviko',
-		(0, 2): 'buviqo',
-		(1, 0): 'zetico',
-		(1, 1): 'zetiko',
-		(1, 2): 'zetiqo',
-		(2, 0): 'wopico',
[…]

### 2024-02-26 code/disttern.py

@@ -1,89 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-
-
-def proportion_to_point(prop, right_axis=False):
-	'''
-	Given a proportion along the left or right triangle axis, return its
-	cartesian coordinates.
-	'''
-	h = (1 - prop)
-	x = h * np.cos(60 * np.pi / 180)
-	y = h * np.sin(60 * np.pi / 180)
-	if right_axis:
-		x = 1 - x
-	return x, y
-
-def intersection(data):
-	'''
-	Given two line segments, return the point at which they intersect.
-	'''
-	L = len(data)
-	x1, y1, x2, y2 = data.reshape(L * 2, -1).T
-	R = np.full([L, 2], np.nan)
-	X = np.concatenate([
-		(y2 - y1).reshape(L * 2, -1), 
-		(x1 - x2).reshape(L * 2, -1)], 
-		axis=1
-	).reshape(L, 2, 2)
-	B = (x1 * y2 - x2 * y1).reshape(L, 2)
-	I = np.isfinite(np.linalg.cond(X))
-	R[I] = np.matmul(np.linalg.inv(X[I]), B[I][:,:,None]).squeeze(-1)
-	return R[0]
-
-def calculate_proportions(reference_objects, target_objects, distance_func):
-	'''
-	Given three reference objects (placed at the top, left, and right vertices
-	of the triangle) and a collection of target objects, compute the
-	proportions along the left and right axes that describe where the
-	target objects should be located.
-	'''
-	assert len(reference_objects) == 3
-	props = []
-	for target in target_objects:
-		dist_to_top = distance_func(target, reference_objects[0])
-		dist_to_left = distance_func(target, reference_objects[1])
-		dist_to_right = distance_func(target, reference_objects[2])
-		prop_left = dist_to_top / (dist_to_top + dist_to_left)
-		prop_right = dist_to_top / (dist_to_top + dist_to_right)
-		props.append((prop_left, prop_right))
-	return props
-
-def draw_ternary_axes(axis):
-	'''
-	Draw the ternary plot axes and grid lines.
-	'''
-	height = np.sqrt(3) / 2
-	centroid = 0.5, height / 3
-	ref_l = proportion_to_point(0.5, right_axis=False)
-	ref_r = proportion_to_point(0.5, right_axis=True)
-	axis.add_patch( plt.Polygon([(0.5, height), ref_l, centroid, ref_r], fill=True, color='black', alpha=0.4, linewidth=0) )
[…]

### 2024-02-26 code/training_model.py

@@ -1,132 +1,0 @@
-import pymc as pm
-import numpy as np
-import pandas as pd
-import arviz as az
-from utils import json_load
-
-
-def is_regular(lexicon, training_items):
-	suffixes = {'0': set(), '1': set(), '2': set(), '3': set()}
-	for item, word in lexicon.items():
-		if item not in training_items:
-			continue
-		color = item.split('_')[1]
-		suffixes[color].add(word[4:])
-	if sum([len(suffixes[s]) == 1 for s in suffixes]) == 4:
-		print(suffixes)
-		return True
-	return False
-
-def find_subjects_with_regular_input():
-	for i in range(1, 398):
-		subject_id = str(i).zfill(3)
-		if subject_id in gen_ones:
-			continue
-		data = json_load(f'../data/exp2/subject_{subject_id}.json')
-		if data['generation'] > 3:
-			continue
-		if data['status'] != 'approved':
-			continue
-		if is_regular(data['input_lexicon'], data['training_items']):
-			print(data['subject_id'], data['chain_id'], data['generation'])
-
-def get_correct(exp_id, subject_id):
-	data = json_load(f'../data/{exp_id}/subject_{subject_id}.json')
-	correct = [
-		int(trial['expected_label'] == trial['input_label'])
-		for trial in data['responses'] if trial['test_type'] == 'mini_test'
-	]
-	prod_correct = [
-		int(trial['expected_label'] == trial['input_label'])
-		for trial in data['responses'] if trial['test_type'] == 'test_production'
-	]
-	comp_correct = [
-		int(trial['selected_item'] in trial['items'])
-		for trial in data['responses'] if trial['test_type'] == 'test_comprehension'
-	]
-	trial = list(range(len(correct)))
-	return trial, correct, (sum(prod_correct), sum(comp_correct))
-
-def fit_multilevel_model(trial, correct, subject_ids):
-	coords = {
-		'trial': list(range(1, 37)),
-		'subject': subject_ids,
-	}
-	with pm.Model(coords=coords) as model:
-
-		μ_α = pm.Normal('μ_α', 0, 10)
-		σ_α = pm.Exponential('σ_α', 1)
-
-		μ_β = pm.Normal('μ_β', 0, 10)
-		σ_β = pm.Exponential('σ_β', 1)
-
[…]

### 2024-02-26 code/build_figures.py
Some tidying up

@@ -322,32 +322,37 @@ if __name__ == '__main__':
 
 	# Create detailed transmission chain figures
 
-	# data = json_load(ROOT / 'data' / 'exp.json')
+	data = json_load(DATA / 'exp.json')
 	
-	# plot_transmission_chains(data, 
-	# 	condition='dif_lrn',
-	# 	output_path=FIGS / 'dif_lrn.pdf',
-	# )
+	# Experiment 1 (differentiation) - Tranmission-only
+	plot_transmission_chains(data, 
+		condition='dif_lrn',
+		output_path=FIGS / 'dif_lrn.pdf',
+	)
 
-	# plot_transmission_chains(data, 
-	# 	condition='dif_com',
-	# 	output_path=FIGS / 'dif_com.pdf',
-	# )
+	# Experiment 1 (differentiation) - Transmission + Communication
+	plot_transmission_chains(data, 
+		condition='dif_com',
+		output_path=FIGS / 'dif_com.pdf',
+	)
 
-	# plot_transmission_chains(data, 
-	# 	condition='con_lrn',
-	# 	output_path=FIGS / 'con_lrn.pdf',
-	# )
+	# Experiment 2 (conservation) - Tranmission-only
+	plot_transmission_chains(data, 
+		condition='con_lrn',
+		output_path=FIGS / 'con_lrn.pdf',
+	)
 
-	# plot_transmission_chains(data, 
-	# 	condition='con_com',
-	# 	output_path=FIGS / 'con_com.pdf',
-	# )
+	# Experiment 2 (conservation) - Transmission + Communication
+	plot_transmission_chains(data, 
+		condition='con_com',
+		output_path=FIGS / 'con_com.pdf',
+	)
 
-	# plot_transmission_chains(data, 
-	# 	condition='sil_com',
-	# 	output_path=FIGS / 'sil_com.pdf',
-	# )
+	# Experiment 3 (no sound)
+	plot_transmission_chains(data, 
+		condition='sil_com',
+		output_path=FIGS / 'sil_com.pdf',
+	)
 
 
 	# Plot quantitative results
@@ -355,35 +360,31 @@ if __name__ == '__main__':
 	data = pd.read_csv(DATA / 'exp.csv')
 
 	# Experiment 1
+	plot_results(data,
+		conditions=('dif_lrn', 'dif_com'),
+		condition_names=('Transmission-only', 'Transmission + Communication'),
+		params=['α', 'β'],
[…]

### 2024-02-26 code/demographics.py

@@ -49,6 +49,11 @@ def most_common_languages(exp_id):
 		print(i, lang, counts[lang], round(counts[lang] / len(langs) * 100, 2))
 
 
+print('Experiment 1')
 calculate_rates('exp', 814, chain_filter=r'dif.+', with_bonus=False)
+
+print('Experiment 2')
 calculate_rates('exp', 814, chain_filter=r'con.+', with_bonus=False)
+
+print('Experiment 3')
 calculate_rates('exp', 814, chain_filter=r'sil.+', with_bonus=False, basic_pay=250)

### 2024-02-26 code/print_word_chains.py

@@ -9,6 +9,8 @@ ROOT = Path(__file__).parent.parent.resolve()
 cons = ['f', 's', 'ʃ']
 vwls = ['əʊ', 'ə', 'ɛɪ']
 def get_sound(item, data):
+	if data['spoken_forms'] is None:
+		return '___'
 	sound_file = data['spoken_forms'][item]
 	if '_' not in sound_file:
 		return 'kəʊ'

### 2024-03-16 code/build_figures.py
Add missing figure labels

@@ -310,8 +310,9 @@ def plot_success(data, conditions, titles, output_path):
 		axis.set_ylim(0, 1)
 		axis.set_xticks(range(1, 10))
 		axis.set_title(title, fontsize=7)
-		if title == 'Experiment 1':
+		if condition == 'dif_com':
 			axis.set_ylabel('Proportion of trials successful')
+		axis.set_xlabel('Generation')
 		axis.axhline(1/9, color='gray', linestyle=':')
 		axis.axhline(1/3, color='gray', linestyle='--')
 		axis.set_ylim(-0.05, 1.05)

### 2024-03-16 manuscript/manuscript.tex
Final revisions before submission

@@
 We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for ambiguity avoidance.
-In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate meaning (e.g., <lite> and <light>), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> and <meet>).
+In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate meaning (e.g., <lite> vs.~<light>), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> vs.~<meet>).
 Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, they tend to favor some additional informativeness.
-Nevertheless, our findings also suggest that, in the long term, simpler, transparent spellings will tend to be preferred.}
+Nevertheless, our findings also suggest that, in the long term, simpler, transparent spellings may be preferred in the absence of top-down explicit teaching.}
 
@@
 \noindent
-Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each letter corresponding to a distinct sound.
-However, it is also well understood that writing systems diverge from their spoken counterparts in important ways \parencite{Bolinger:1946, Coulmas:1991}.
-The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}, and it seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}.
+Writing systems, particularly those employing alphabetic scripts, are commonly regarded as providing a visual representation of speech, with letters or chunks of letters corresponding to distinct sounds.
+However, it is also well understood that writing systems diverge from their spoken counterparts in important ways \parencite{Biber:1988, Bolinger:1946, Coulmas:1991}.
+The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}.
+It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}.
 In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality.
@@
 
-One potential case of such functional divergence is heterographic homophony---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}).
+One potential case of such functional divergence is heterographic homophony---words that sound alike but which are written differently (e.g., <meat> and <meet> for /miːt/).
 Heterographic spellings such as these may serve a valuable function in reading.
@@
 In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content.
-Heterographic homophony can also occur in affixes.
+Heterographic homophony is also common below the word level, since many orthographies forego the phonological principle in favor of the morphological principle in the spelling of affixes \parencite{Sandra:2024}.
 The English suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing.
 Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English affixes \parencite{Berg:2017} and it has been shown that readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.
-\newmaterial{These features might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair.
-Furthermore, written language tends to be more sophisticated than spoken language (e.g., longer, more elaborate sentences), placing different pressures on ambiguity resolution.}
+\newmaterial{Heterography might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair.
+Furthermore, written language has richer vocabulary and more complex syntax than spoken language \parencite{Biber:1988, Korochkina:2024, Nation:2022}, placing different pressures on ambiguity resolution.}
 
-\newmaterial{Although heterography is particularly notable in English, it is also a feature of many other languages and writing systems.
-French is similar to English in having a large number of heterographic homophones; for example, the heterographic words \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}) are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison).
+\newmaterial{Heterography is particularly notable in English, but it is also a feature of many other languages and writing systems.
+French is similar to English in having a large number of heterographic homophones: \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}), for example, are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison).
 In Danish, the words \textit{hver} (\textit{every}), \textit{vejr} (\textit{weather}), \textit{vær} (\textit{be}), and \textit{værd} (\textit{worth}) are all pronounced /vεˀɐ̯/.
@@
 
-\newmaterial{However, perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese script.
-The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic script---particularly useful \parencite{Katz:1992}.
-In Mandarin Chinese, the words 糖 (\textit{sugar}), 塘 (\textit{embankment}), 溏 (\textit{pond}), and 搪 (\textit{to block}) are homophonous in speech but heterographic in writing---the phonetic radical on the right (唐; /táŋ/) represents the spoken syllable, while the semantic radicals on the left differentiate the meanings.
+\newmaterial{Perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese orthography.
+The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic writing system---particularly useful \parencite{Frost:2012}.
+In Mandarin Chinese, the words 糖 (\textit{sugar}), 塘 (\textit{embankment}), 溏 (\textit{pond}), and 搪 (\textit{to block}) are homophonous in speech but heterographic in writing---the phonetic radical on the right (<唐>, /táŋ/) represents the spoken syllable, while the semantic radicals on the left differentiate the meanings \parencite[p.~101]{Coulmas:1991}.
 In addition, 唐 itself is a surname/dynasty (\textit{Tang}), and another unrelated word 堂 (\textit{hall}) is also pronounced /táŋ/, yielding at~least six ways to write the same sound depending on the meaning.
-This allows written Chinese to convey more information about meaning---to be more \textit{informative}---than its spoken counterpart.}
+This property allows the written form of Chinese to convey more information about meaning---to be more \textit{informative}---than its spoken counterpart.}
 
@@
 In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>.
-Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding} \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}.
+Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}.
 Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned.
@@
 Sound changes involving consonants have also resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (e.g., \textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}), the loss of /ç/ (e.g., \textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}), and the merger of /ʍ/ into /w/ (e.g., \textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}).\footnote{Although we can never be entirely certain how words were pronounced before the advent of sound recording technology, historical linguists have compiled persuasive evidence by a variety of methods.
-Comparison to modern Dutch and German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with German \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with Dutch and German \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these words continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.
+Comparison to modern German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these word pairs continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.
 
@@
 All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, lasting around 8~min}.
-Participants were also told explicitly before starting the training session that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
+Participants were also told explicitly before starting the training session that the stems looked and sounded like the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
 In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each.
@@
 Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant.
-Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant).
+Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{The seen items were selected independently for each participant.
 We felt this was more ecological than giving both participants identical input.}} The communication game, \newmaterial{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
@@
 Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero.
-We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision; the further the probability mass is from zero, the stronger the evidence of an effect.
+We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision.
 The results are shown in Fig.~\ref{results_exp1}B.
@@
 In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure.
-However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$).
-Although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}
+However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$); although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}
 
@@
 Although there was evidence to suggest that the orthographic systems remain informative for longer under communicative pressure, both conditions ultimately converged on degenerate, uninformative systems and there was little evidence of systematic differentiation in spelling.
-Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <ko> for pink, <co> for yellow, and <qo> for blue), but also through less obvious strategies such as the use of length (e.g., <ko> for pink, <koo> for yellow, and <kooo> for blue) or the use of arbitrary silent letters (e.g., <kox> for pink, <kof> for yellow, and <kom> for blue).
+Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <ko> for pink, <co> for yellow, and <qo> for blue), but also through less obvious strategies such as the use of length (e.g., <ko> for pink, <kko> for yellow, and <kkko> for blue) or the use of arbitrary silent letters (e.g., <kox> for pink, <kof> for yellow, and <kom> for blue).
 Such differentiation was not forthcoming, however, even under communicative pressure.
@@
 
-So, what was different about the present experiment compared the large body of prior experimental iterated learning studies?
+So, what was different about the present experiment compared to the large body of prior experimental iterated learning studies?
 The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions.
@@
 
-\newmaterial{The results in terms of the more nuanced communicative cost measure are presented in Fig.~\ref{results_exp2}C.
+\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{results_exp2}C.
 Here we find a nonzero effect of both epoch and generation on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
@@
 Spacing between words for example does not exist in speech but constitutes a useful innovation in writing that permits rapid reading \parencite{RaynerFischer:1998, Sainio:2007, Zang:2013}.
-Similarly, the consistent spelling of affixes, such as the English past-tense marker <-ed>, which diverges from its spoken realization (/-d/, /-t/, or /-ɪd/ depending on the preceding sound), permits faster access to meaning \parencite{}.
+Similarly, the consistent spelling of affixes, such as the English past-tense marker <-ed>, which diverges from its spoken realization (/-d/, /-t/, or /-ɪd/ depending on the preceding sound), permits faster access to meaning \parencite{Ulicheva:2020}.
 Might it be the case that---left to evolve freely---the written form of a language will become better adapted to the needs of writers and readers to the detriment of its alignment with the spoken form of the language?
@@
 We addressed these questions by focusing on the particular case of heterographic homophones---morphemes that sound the same but that are spelled differently.
-Heterographic homophones permit the written language to be more informative than the spoken language; the written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
-In English, at~least, which has a tendency to preserve etymology in spelling, heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, \textit{wright}, and \textit{write} are spelled differently due to their separate derivations (Old English \textit{riht}, Latin \textit{ritus}, Old English \textit{wryhta}, and Old English \textit{writan}).
-Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist---words such as \textit{byte}, \textit{cheque}, and \textit{lite}---and, moreover, the etymologically motivated cases of heterography may, in part, owe their continued success to the differentiation they inadvertently provide.
+Heterographic homophones permit the written language to be more informative than the spoken language; the spellings <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information.
+We investigate whether heterography might arise for functional reasons by experimentally simulating the cultural evolution of orthography under two distinct mechanisms, differentiation and conservation, as described by \textcite{Berg:2021}.
 
-We put these ideas to the test by experimentally simulating the cultural evolution of orthography.
-In Experiment~1, we focused on the differentiation mechanism: Might variant spellings be used to differentiate meanings that are otherwise identical in speech?
+\subsection{Experiment 1: Differentiation}
+
+In our first experiment, we focused on the differentiation mechanism: Might variant spellings be used to differentiate meanings that are otherwise identical in speech?
 If so, we would expect levels of spelling differentiation to be greater when there is greater pressure for disambiguation, which we induced through the addition of a communication game.
@@
 A writing system that does not permit a one-to-many phoneme-to-grapheme mapping would not be capable of differentiating homophonous words.
-Only when variation is permitted and available can variant spellings be conditioned on meaning.
+Only when spelling variation is permitted and available can variants be conditioned on meaning.
 
-Although the emergent orthographies tended to be slightly more informative under communicative pressure and the informativeness persisted for more generations, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
+Although the emergent orthographies tended to be slightly more informative under communicative pressure, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
 This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were retained but ineffectually conditioned on shape rather than color.
@@
 \newmaterial{This conclusion is in partial agreement with work by \textcite{Treiman:2015}.
-In a series of experiments, participants were asked to provide spellings for novel English words (e.g., /hæf/ meaning \textit{alehouse}) that were homophonous with preexisting English words (in this case, \textit{half}).
+In this study, participants were asked to provide spellings for novel English words (e.g., /hæf/ meaning \textit{alehouse}) that were homophonous with preexisting English words (in this case, \textit{half}).
 Participants tended to provide the same spelling as the preexisting word (i.e., <half>) rather than other possible alternatives that would have had the benefit of differentiating meaning (e.g., <haf>, <haff>, <haph>).
@@
 \textcite{Treiman:2015} also found, however, that, when given two alternatives to choose from (e.g., <half> vs.\@ <haff>), participants generally did prefer the novel spelling.
-This runs contrary to our first experiment, since our participants are similarly provided with multiple possible spellings of the sound /kəʊ/ (<coe>, <koh>, <qo>, etc.), but they nevertheless tended not to condition these on the color dimension, even under communicative pressure with financial incentive.}
+This runs contrary to our first experiment, since our participants are similarly provided with multiple possible spellings of the sound /kəʊ/ (<coe>, <koh>, <qo>, etc.), but they nevertheless tended not to condition these on the color dimension, even under communicative pressure with financial incentive.
+Thus, although the preference for simplicity might be relatively weak at the individual level, it might nevertheless be amplified by the iterated learning process at the population level.}
 
-\newmaterial{During review, a concern was raised that our participants might not fully understand the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling.
-Alternatively, it was posited that our experiments might suffer from insufficient power.
+\newmaterial{During the review process, a concern was raised that our participants might not have fully understood the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling.
 Our position, as outlined above, was that the lack of differentiation was due to the very strong homophony pressure.
-To test whether the lack of informativeness might be attributed to the homophony and to check that the design of our experiment and implementation of the communicative pressure was, in~principle, sufficient to promote more informative systems, we ran an additional experiment that entirely removed the spoken form, thereby replicating previous iterated learning experiments, which are generally only orthographic in nature \parencite[e.g.,][]{Kirby:2008, Kirby:2015, Beckner:2017}.
-This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness, and a commensurate increase in communicative success (see Appendix~D).
-This suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred the spellings to encoded sound rather than meaning.
-Or, rather, the cultural evolutionary process ultimately tended to favor simplicity over informativeness in this particular domain, amplifying the participants' (potentially relatively weak) preference for ``lessor effort.''}
+To test whether the lack of informativeness might be attributed to the homophony and to check that the design of our experiment and implementation of the communicative pressure was sufficient to promote more informative systems, we ran an additional experiment.
+This experiment was identical to the communicative condition of Experiment~1, except that we removed the spoken forms (thereby replicating previous iterated learning experiments, which are generally only orthographic in nature; e.g., Beckner et al., 2017; Kirby et al., 2008, 2015) and altered the orthographic forms so that no homophony was implied in the spellings.
+This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness, and a commensurate increase in communicative success (a point we return to shortly).
+Importantly, this suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred their spellings to encode sound rather than meaning.
+Or, rather, the cultural evolutionary process ultimately tended to favor simplicity over informativeness in this particular domain.}
 
-In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation.
-Under this mechanism, an expressive orthography does not emerge to fulfill an adaptive role but is simply a fossil representing an earlier form of the spoken language that \textit{was} expressive of a particular meaning distinction.
+\subsection{Experiment 2: Conservation}
+
+In our second experiment, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation.
+Under this mechanism, expressive forms do not emerge but are simply fossils representing an earlier form of the spoken language that \textit{was} expressive of a particular meaning distinction.
 \newmaterial{Over time, the orthography may experience a ratcheting effect, in which heterographic forms accumulate (due to successive sound changes) but rarely recede (due to the informativeness they provide).
@@
 
-\newmaterial{The results of both experiments were, however, relatively weak statistically, with the differences in slopes often only just (or not quite) meeting the 95\% criterion.
-This perhaps reflects a situation of low power, with only ten chains---ten independent sampling units---in each condition.
-This is a general issue with iterated learning experiments, however, and our use of ten chains follows norms in the field.} It is also important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible.
+\subsection{Differentiation or conservation?}
+
+It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible.
 Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2.
 The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm.
-Similarly, we cannot draw direct conclusions from the relative strength of the statistical results across the two experiments.
-
-Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation.
+However, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation.
 For \textit{differentiation} to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming social stigma in the process (the social problem).
@@
 The \textit{conservation} of an expressive orthography is, by~comparison, plain~sailing---it is only the learnability problem that applies.
-More generally, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.
+
+	\begin{figure}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/success.pdf}}
+	\vspace*{2pt}
+	\caption{Communicative success by generation in the communicative conditions of all three experiments.
+The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
+	\label{success}
+	\end{figure}
+
+In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one \parencite[see also][]{Smith:2002}.
+\newmaterial{This is made particularly salient by Fig.~\ref{success}, which compares the experiments in terms of communicative success (the proportion of trials in which the comprehending participant selected the correct target item in response to their partner).
+In Experiment~1, communicative success remains around chance level (a one in three probability of selecting the right color) because the orthographic systems tend to become uninformative, mirroring the spoken form of the language.
+In Experiment~2, by comparison, communicative success remains high in several of the chains (i.e., those chains that preserved the expressive system).
+This suggests that, while it may be difficult for participants to establish an informative writing system, it is comparatively easy to preserve an informative system that offers a clear advantage.
+It is also interesting to observe what happens to communicative success in Experiment~3 when the homophony pressure is removed.
+Here, communicative success \textit{does} increase over time, as the participants---unencumbered by having to represent sound---find communicative strategies to differentiate meaning.
+Taken together, these results suggest that it is difficult for the differentiation mechanism to operate in the face of homophony, but comparatively easy for the conservation mechanism to operate under these same circumstances.}
+
+\subsection{Limitations}
 
@@
 
-\newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to transparent writing systems.
-This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage---whether that baggage is for the encoding of sound or the encoding of meaning.
+\newmaterial{The results of both experiments were relatively weak statistically, with the differences in terms of informativeness between the Transmission-only and Transmission~+ Communication conditions only just (or not quite) meeting the 95\% criterion.
+This is likely to be related to a combination of two factors: a relatively small effect size combined with a relatively small sample size, with only ten chains---ten independent sampling units---in each condition.
+This is a general issue with iterated learning experiments that results from their resource-intensive nature, and our use of ten chains follows or exceeds norms in the field \parencite[e.g.,][]{Kempe:2015, Kirby:2008, Kirby:2015, Raviv:2018, Roberts:2018, Smith:2010, Tamariz:2015}.
+Nevertheless, our study suggests that future work in this area could benefit from increased power.}
+
+\newmaterial{One important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems.
+This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage, whether that baggage is for the encoding of sound or the encoding of meaning.
 Ideally, our experiments would be performed with participants who have no writing experience at all, but since that would be very difficult to achieve, perhaps the second-best option is a participant population that is relatively open-minded to both types of writing systems.
@@
 
-We foresee two interesting directions in which this work might be taken in the future.
-The first is a simple extension of the present experiments.
-By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality.
-This minor change might have a dramatic effect on the results: Participants may be less inclined to simply transcribe how a word sounds and more inclined to differentiate and conserve.
-Such an experiment is not without pitfalls, however.
-Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it.
-Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies \newmaterial{and our Experiment~3}---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.
+\section{Conclusion}
 
-A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on change in natural languages.
-For example, by tracing the development of heterographic homophones in various languages, we might be able to determine which of the models offer a better fit to empirical data, or indeed classify individual cases as being derived through the differentiation or conservation mechanism.
-Perhaps the best way to proceed here would be to develop formal models of the two mechanisms and test them against diachronic data on the pronunciation and spelling of words that have ultimately become heterographic in the present-day.
-
-As a final note, heterography is often derided as a source of unnecessary complexity, \newmaterial{and as a consequence, the orthographic reforms implemented in many languages have tended to focus on its elimination.} However, our results suggest that, when the communicative need for disambiguation is strong enough, language users may prefer to trade some simplicity for greater informativeness in the writing system.
-Perhaps, then, efforts to reform and simplify spelling may be a little hasty, since, ultimately, the writing system should help the reader get to the meaning of a word rather than its sound.
+\newmaterial{It has long been known that heterography makes reading and learning to read difficult \parencite{Pexman:2001, Seymour:2003}.
+As a result, heterography has often been derided as a source of unnecessary complexity, and the orthographic reforms implemented in many languages have tended to focus on its elimination.
+However, recent research suggests that heterography may in some circumstances be functional because it permits rapid access to meaning \parencite{Rastle:2019, Ulicheva:2020}.
+The novel research presented in this article suggests that the cultural evolution of writing systems may prefer to trade some simplicity for greater informativeness when the communicative need for disambiguation is strong enough.
+These results imply that writing systems may, under some circumstances, evolve to fill a “reading niche.” However, our research also shows that creating heterography, and even maintaining it, is challenging given the demands it poses on learning.
+These findings raise the prospect of a third major issue relevant to the cultural evolution of writing systems: education.
+Instead of yielding to the pressure of learnability, orthographies like English and Chinese have developed and maintained a high degree of informativeness because those societies have invested in education systems that spend many years teaching children to read \parencite[e.g.,][]{Xinchun:1999}.
+Thus, informative writing systems that contribute to rapid, skilled reading may not only impose learning costs, but may also require ongoing economic investment.}
 
@@
 
-\section{Acknowledgments}
+% \section{Acknowledgments}
 
-\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034).
+% \noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034).
 
@@
 
-\section{Communicative success}
-
-Fig.~\ref{success} compares the three experiments in terms of communicative success, which is defined as the proportion of trials in the communication game in which the comprehending participant selected the correct target item.
-In Experiment~1, communicative success typically remains around chance; the orthographic systems tend to remain uninformative about color over time, mirroring the spoken form of the language.
-In Experiment~3, by comparison, which removed the homophony pressure, communicative success does tend to increase over time, as the participants find communicative strategies.
-In Experiment~2, the orthographic systems started out expressive, and as a consequence communicative success is initially high.
-However, success drops over time as the systems become less informative, gradually mirroring the homophony in the spoken form.
-Nevertheless, success does remain high in some chains where the expressive systems were maintained.
-
-	\begin{figure}[h]
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/success.pdf}}
-	\vspace*{2pt}
-	\caption{Communicative success by generation in the communicative conditions of all three experiments.
-The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
-	\label{success}
-	\end{figure}
-
 \end{document}

### 2024-03-19 manuscript/manuscript.tex
Rename figures numerically

@@
 \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography.
-The first model, \textit{the differentiation model} (Fig.~\ref{models_of_heterography}A), explains heterography through the creation of new orthographic forms.
+The first model, \textit{the differentiation model} (Fig.~\ref{fig01}A), explains heterography through the creation of new orthographic forms.
 For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/models_of_heterography.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig01.pdf}}
 	\vspace*{2pt}
@@
 Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
-	\label{models_of_heterography}
+	\label{fig01}
 	\end{figure}
 
-The second model, \textit{the conservation model} (Fig.~\ref{models_of_heterography}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography.
+The second model, \textit{the conservation model} (Fig.~\ref{fig01}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography.
 For example, the \textit{meat}--\textit{meet} merger that occurred during the Great Vowel Shift ultimately resulted in Middle English /ɛː/ (spelled <ea>) and /eː/ (spelled <ee>) being pronounced /iː/ in Early Modern English \parencite{Lass:2000}, but the spellings were never changed accordingly, thus giving rise to a set of heterographic homophones that persist in present-day English \parencite[pp.~140--141]{Wells:1982}: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/stimuli.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig02.pdf}}
 	\vspace*{2pt}
@@
 The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
-	\label{stimuli}
+	\label{fig02}
 	\end{figure}
 
-Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}.
+Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{fig02}.
 The alien words had a spoken and written form composed of a stem and suffix.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/transmission.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig03.pdf}}
 	\vspace*{2pt}
@@
 Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
-	\label{transmission}
+	\label{fig03}
 	\end{figure}
 
-Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A).
+Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{fig03}A).
 The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations.
@@
 Transmission continued for nine generations in each of ten independent chains.
-In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consisted of a pair of participants, but the productions of only one of the two (the primary participant; determined by whichever participant started the experiment first) were iterated to the next generation.
+In the Transmission~+ Communication condition (see Fig.~\ref{fig03}B), each generation consisted of a pair of participants, but the productions of only one of the two (the primary participant; determined by whichever participant started the experiment first) were iterated to the next generation.
 The productions of the secondary participant were not iterated any further and were thus a cultural deadend; the role of the secondary participant was to act as a genuine communicative partner for the primary participant, inducing pressure for the language to become informative.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/dif_lrn.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig04.pdf}}
 	\vspace*{2pt}
 	\caption{Results from the Transmission-only condition in Experiment~1 (differentiation).
-Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
-Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{fig02}).
+Chains are labeled A--J and generations are labeled 0--9 (0 is the randomly generated seed system).
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
@@
 By the ninth generation, most systems are degenerate (e.g., Chains~A, B, and~I), redundant (e.g., Chains~D, E, and~H), or a mixture of the two (e.g., Chain~F).}
-	\label{dif_lrn}
+	\label{fig04}
 	\end{figure}
 
-The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}.
-Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{stimuli}.
+The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{fig04}.
+Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{fig02}.
 The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/dif_com.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig05.pdf}}
 	\vspace*{2pt}
 	\caption{Results from the Transmission~+ Communication condition in Experiment~1 (differentiation).
-Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
-Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{fig02}).
+Chains are labeled K--T and generations are labeled 0--9 (0 is the randomly generated seed system).
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
@@
 There are some isolated examples of differentiation through implicit generalization (K5, M1, P5, S2) and explicit innovation (K9, L7, M7, O4, Q7).}
-	\label{dif_com}
+	\label{fig05}
 	\end{figure}
 
-The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}.
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{fig05}.
 Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O).
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/typology.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig06.pdf}}
 	\vspace*{2pt}
@@
 A degenerate system expresses nothing.}
-	\label{typology}
+	\label{fig06}
 	\end{figure}
@@
 To analyze more formally what types of system tend to emerge under different conditions, we first designated four suffix systems that are of particular interest: holistic, expressive, redundant, and degenerate.
-These four typological categories may be positioned along the simplicity--informativeness continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}.
+These four typological categories may be positioned along the simplicity--informativeness continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{fig06}.
 The holistic\footnote{Our use of the term \textit{holistic} is slightly unusual here, owing to the fact that we are focused on the suffix level and not the word level.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp1.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig07.pdf}}
 	\vspace*{2pt}
@@
 The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
-	\label{results_exp1}
+	\label{fig07}
 	\end{figure}
 
-The typological distributions are plotted in Fig.~\ref{results_exp1}A, revealing the proportion of the 10 chains that fall into each typological category at each generation.
+The typological distributions are plotted in Fig.~\ref{fig07}A, revealing the proportion of the 10 chains that fall into each typological category at each generation.
 In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2.
@@
 We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision.
-The results are shown in Fig.~\ref{results_exp1}B.
+The results are shown in Fig.~\ref{fig07}B.
 In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure.
@@
 In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.
-\newmaterial{The results are shown in Fig.~\ref{results_exp1}C.
+\newmaterial{The results are shown in Fig.~\ref{fig07}C.
 We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian).
@@
 Following prior work, we operationalized learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$; transmission error is defined as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations \parencite[see e.g.,][]{Kirby:2008}.
-These results are plotted in Fig.~\ref{results_exp1}D.
+These results are plotted in Fig.~\ref{fig07}D.
 In both conditions, the estimates of the $\beta$ parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/sound_change.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig08.pdf}}
 	\vspace*{2pt}
@@
 During Epoch~III, the spoken form of the language makes no color distinction.}
-	\label{sound_change}
+	\label{fig08}
 	\end{figure}
@@
 This was achieved through two sets of sound changes, the first occurring in the transition from Epoch~I to~II and the second occurring in the transition from Epoch~II to~III.
-An example is illustrated in Fig.~\ref{sound_change}.
+An example is illustrated in Fig.~\ref{fig08}.
 In the first sound change, two of the spoken suffix forms were chosen at random and the consonant from one (chosen at random) was paired with the vowel from the other, resulting in a new suffix form that replaced the original two.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/con_lrn.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig09.pdf}}
 	\vspace*{2pt}
 	\caption{Results from the Transmission-only condition in Experiment~2 (conservation).
-Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
-Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{fig02}).
+Chains are labeled A--J and generations are labeled 0--9 (0 is the randomly generated seed system).
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
@@
 The final systems are characterized by the conservation of form without the conservation of expressivity.}
-	\label{con_lrn}
+	\label{fig09}
 	\end{figure}
 
-The results for all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.
+The results for all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{fig09}.
 It is immediately clear that color-expressive suffixes (as indicated by vertical stripes) are maintained fairly reliably through the first epoch; perfectly in the case of Chains~A, B, D, F, and~I, and with some errors in the other five chains.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/con_com.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig10.pdf}}
 	\vspace*{2pt}
 	\caption{Results from the Transmission~+ Communication condition in Experiment~2 (conservation).
-Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}).
-Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system).
+Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{fig02}).
+Chains are labeled K--T and generations are labeled 0--9 (0 is the randomly generated seed system).
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
@@
 Five chains (O, P, Q, R, T) remain fully expressive into the final epoch, in most cases conserving the original forms.}
-	\label{con_com}
+	\label{fig10}
 	\end{figure}
 
-The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.
+The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{fig10}.
 Like Transmission-only, the seed systems are mostly maintained faithfully through Epoch~I; perfectly in the case of Chains~K, L, N, O, and~T, and with some errors in the other five chains, although some of these errors are non-destructive changes, such as <x> being replaced with <sh> in R1.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp2.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig11.pdf}}
 	\vspace*{2pt}
@@
 The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
-	\label{results_exp2}
+	\label{fig11}
 	\end{figure}
@@
 
-Fig.~\ref{results_exp2}A plots the typological distributions by generation and condition.
+Fig.~\ref{fig11}A plots the typological distributions by generation and condition.
 Initially, all systems are expressive, but the dominance of this category is gradually eroded over time, particularly during the second and third epochs once the spoken forms had become homophonous.
@@
 \newmaterial{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time.
-The results, shown in Fig.~\ref{results_exp2}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition.
+The results, shown in Fig.~\ref{fig11}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition.
 Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this measure.}
 
-\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{results_exp2}C.
+\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C.
 Here we find a nonzero effect of both epoch and generation on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
@@
 
-\newmaterial{For completeness, Fig.~\ref{results_exp2}D also plots transmission error; however, we did not hypothesize any particular differences in learnability in Experiment~2, either over time or by condition.
+\newmaterial{For completeness, Fig.~\ref{fig11}D also plots transmission error; however, we did not hypothesize any particular differences in learnability in Experiment~2, either over time or by condition.
 The expressive orthographic systems used to initialize the chains start out very easy to learn, and learnability remains fairly consistent throughout the experiment in both conditions, albeit with some constant level of change over time as the systems gradually come into alignment with the spoken forms.}
@@
 The second caveat is that, in the long term, it appears that transparency might ultimately win the day, even under communicative pressure.
-Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and based on the trajectories of the communicative cost curves (Fig.~\ref{results_exp2}C), it seems likely that all chains will ultimately undergo the same transformation eventually.
+Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and based on the trajectories of the communicative cost curves (Fig.~\ref{fig11}C), it seems likely that all chains will ultimately undergo the same transformation eventually.
 Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/success.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/fig12.pdf}}
 	\vspace*{2pt}
@@
 The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
-	\label{success}
+	\label{fig12}
 	\end{figure}
@@
 In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one \parencite[see also][]{Smith:2002}.
-\newmaterial{This is made particularly salient by Fig.~\ref{success}, which compares the experiments in terms of communicative success (the proportion of trials in which the comprehending participant selected the correct target item in response to their partner).
+\newmaterial{This is made particularly salient by Fig.~\ref{fig12}, which compares the experiments in terms of communicative success (the proportion of trials in which the comprehending participant selected the correct target item in response to their partner).
 In Experiment~1, communicative success remains around chance level (a one in three probability of selecting the right color) because the orthographic systems tend to become uninformative, mirroring the spoken form of the language.
@@
 \begin{figure}[h!]
-\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/instructions.pdf}}
+\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/supp1.pdf}}
 \vspace*{2pt}
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/sil_com.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/supp2.pdf}}
 	\vspace*{2pt}
@@
 Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns).
-Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system).
+Chains are labeled A--J and generations are labeled 0--9 (0 is the randomly generated seed system).
 Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings.
 Spellings in bold-italic are the generalizations on unseen items.}
-	\label{sil_com}
+	\label{supp2}
 	\end{figure}
 
-Fig.~\ref{sil_com} plots all suffix spellings by chain and generation.
+Fig.~\ref{supp2} plots all suffix spellings by chain and generation.
 The most notable difference in comparison to Experiment~1 was the large number of times participants used English to signal color.
@@
 	\begin{figure}
-	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/results_exp3.pdf}}
+	\makebox[\textwidth][c]{\includegraphics[scale=0.8]{figs/supp3.pdf}}
 	\vspace*{2pt}
@@
 The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
-	\label{results_exp3}
+	\label{supp3}
 	\end{figure}
@@
 
-The quantitative results are plotted in Fig.~\ref{results_exp3} and are compared against the equivalent results from Experiment~1 (the Transmission~+ Communication condition).
+The quantitative results are plotted in Fig.~\ref{supp3} and are compared against the equivalent results from Experiment~1 (the Transmission~+ Communication condition).
 Overall, the effect of removing the homophony pressure was to prevent the systems from degenerating over time---both measures of informativeness show a clear difference in slopes, and informativeness levels remain fairly constant over time in Experiment~3, mainly in the form of holistic systems.

### 2024-04-09 code/build_figures.py
Remove intercepts from results figures

@@ -117,6 +117,7 @@ def plot_pinf_data(axis, data, condition, condition_name, min_gen=0):
 	data = data[ (data['condition'] == condition) & (data['generation'] >= min_gen) ]
 	prop_inf = data.groupby('generation')['informative'].mean()
 	axis.scatter(np.arange(min_gen, 10), prop_inf, color=COLORS[condition], alpha=0.5, s=15, linewidth=0)
+	axis.set_ylim(-0.05, 1.05)
 
 
 def plot_cost_data(axis, data, variable, condition, condition_name, min_gen=0):
@@ -190,14 +191,14 @@ def plot_results(data, conditions, condition_names=None, params=[], show_epochs=
 	gs = gridspec.GridSpec(8, 10, figure=fig)
 	gs_typol_dist = gridspec.GridSpecFromSubplotSpec(2, 9, subplot_spec=gs[0:2, :])
 	
-	gs_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[2:4, 0:6])
-	gs_posterior = gridspec.GridSpecFromSubplotSpec(2, len(params), subplot_spec=gs[2:4, 6:10])
+	gs_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[2:4, 0:7])
+	gs_posterior = gridspec.GridSpecFromSubplotSpec(2, len(params), subplot_spec=gs[2:4, 7:10])
 
-	gs_cost_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[4:6, 0:6])
-	gs_cost_posterior = gridspec.GridSpecFromSubplotSpec(2, len(params), subplot_spec=gs[4:6, 6:10])
+	gs_cost_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[4:6, 0:7])
+	gs_cost_posterior = gridspec.GridSpecFromSubplotSpec(2, len(params), subplot_spec=gs[4:6, 7:10])
 
-	gs_error_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[6:8, 0:6])
-	gs_error_posterior = gridspec.GridSpecFromSubplotSpec(2, len(params), subplot_spec=gs[6:8, 6:10])
+	gs_error_model = gridspec.GridSpecFromSubplotSpec(1, 1, subplot_spec=gs[6:8, 0:7])
[…]

### 2024-04-09 manuscript/manuscript.tex
Second revision

@@
 \usepackage{xcolor}
-\newcommand\newmaterial[1]{\textcolor{blue}{#1}}
+\newcommand\newmaterial[1]{\textcolor{black}{#1}}
+\newcommand\secondrevision[1]{\textcolor{black}{#1}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\authorsnames{Anonymous}
+\authorsnames{Jon W. Carr, Kathleen Rastle}
 
-\authorsaffiliations{{Anonymous}}
+\authorsaffiliations{{Department of Psychology, Royal Holloway, University of London, England}}
+
+\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2024).
+Why do languages tolerate heterography?
+An experimental investigation into the emergence of informative orthography.
+\textit{PsyArXiv}.
+Version~3.
+\url{https://doi.org/10.31234/osf.io/qnz5k}}
 
@@
 Firstly, by definition, heterography implies that a single sound can be spelled multiple ways.
-In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>.
+In English, the heterographic spellings <meat> and <meet> imply that /iː/ can be spelled <ea> or <ee>.
 Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}.
@@
 For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
-Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by adding <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}.
+Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by appending <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}.
 Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}---and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}.
@@
 Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation.
-All data and code is available from \url{https://osf.io/7auw6/?view_only=f8c2498b4add41839f51bdbefa03392e}.
+All data and code is available from \url{https://osf.io/7auw6/}.
 \newmaterial{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}.
@@
 \textbf{D} Transmission error by generation along with regression models fit to the data.
-The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$) parameters by condition as well as the posterior differences between conditions.
+The panels on the right show the posterior estimates of the slope ($\beta$) parameters by condition as well as the posterior differences between conditions.
 The green, blue, and red bars indicate respectively the 95\%, 90\%, and 85\% HDIs (credible intervals).}
@@
 Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}.
-Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, representing the intercept of the regression model, and $\beta$ representing its slope.
+Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, which represents the intercept of the regression model \secondrevision{(i.e., the model estimate of the dependent variable at Generation~0)}, and $\beta$, which represents its slope \secondrevision{(i.e., the model estimate of how much the dependent variable changes per generation)}.
 To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero.
-Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero.
+\secondrevision{In other words, we test to see whether the dependent variable is changing over time more rapidly in one condition compared to the other.} Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero.
 We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision.
@@
 \newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are.
-In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0).
+In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts\secondrevision{; Table~B1 in the supplementary material}), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0).
 We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system.
@@
 These results are plotted in Fig.~\ref{fig07}D.
-In both conditions, the estimates of the $\beta$ parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time.
-There was some weak evidence for a difference in the intercepts, with transmission error remaining higher in the communicative condition.
-This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates more change to the system that was learned.}
+In both conditions, the estimates of the \textit{β} parameters are negative and clearly reject zero, suggesting that the systems do indeed become increasingly learnable over time \secondrevision{as hypothesized.
+Our analysis also suggested that there was little difference between the conditions in terms of how rapidly transmission error decreases over time (i.e., Δ(\textit{β}) is highly compatible with zero), although we did note that transmission error tended to be a little higher in the communicative condition.} This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates greater change to the system taught in training.}
 
@@
 \textbf{D} Transmission error by generation along with regression models fit to the data.
-The panels on the right show the posterior estimates of the intercept ($\alpha$) and slope ($\beta$ and $\gamma$) parameters by condition as well as the posterior differences between conditions.
+The panels on the right show the posterior estimates of the slope ($\beta$ and $\gamma$) parameters by condition as well as the posterior differences between conditions.
 $\beta$ represents the effect of epoch and $\gamma$ represents the effect of generation number within epoch.
@@
 The results, shown in Fig.~\ref{fig11}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition.
-Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this measure.}
+Although there was some weak evidence of a difference in epoch slopes ($\Delta(\beta) = 2.65$; 95\%~HDI: $-0.58$, $5.95$), we could not conclusively reject zero under this first measure of informativeness.}
 
 \newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C.
-Here we find a nonzero effect of both epoch and generation on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
-Like the previous analysis, the $\gamma$ slopes were in close alignment.
+Here we find a nonzero effect of both epoch \secondrevision{($\beta$)} and generation \secondrevision{($\gamma$)} on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
+Like the previous measure, the $\gamma$ slopes were in close alignment\secondrevision{, so $\Delta(\gamma)$ is highly compatible with zero difference}.
 This suggests that the effect of generational turnover is very similar between conditions and that the difference between conditions is mostly driven by the increases in homophony induced in each epoch.
@@
 
-Although the emergent orthographies tended to be slightly more informative under communicative pressure, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
-This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were retained but ineffectually conditioned on shape rather than color.
+Although the emergent orthographies tended to be slightly more informative under communicative pressure, systematic differentiation was rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants.
+This is \textit{not} because participants were unable to learn variant spellings; in many cases variant spellings were retained but ineffectually conditioned on shape.
 Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn and reproduce such systems.
@@
 This experiment was identical to the communicative condition of Experiment~1, except that we removed the spoken forms (thereby replicating previous iterated learning experiments, which are generally only orthographic in nature; e.g., Beckner et al., 2017; Kirby et al., 2008, 2015) and altered the orthographic forms so that no homophony was implied in the spellings.
-This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness, and a commensurate increase in communicative success (a point we return to shortly).
+This experiment is described in Appendix~C of the supplementary material, but in short, we observed much greater levels of innovation and informativeness \secondrevision{in this new experiment, with a clear statistical difference between it and the results of Experiment~1.
+Broadly, the effect of removing the homophony pressure was that the systems remained more informative compared to Experiment~1 (in terms of both the proportion of systems classified as informative and communicative cost).} There was also a commensurate increase in communicative success as a result of the emergence of these more informative systems (a point we return to shortly).
 Importantly, this suggests that the participants in Experiment~1 did indeed understand the communicative imperative, but nevertheless preferred their spellings to encode sound rather than meaning.
@@
 
+\secondrevision{We emphasize, however, that we did not make a-priori predictions about which of the two mechanisms might represent a better theory of the emergence of informative heterography and our experiments were not designed to test the two theories against each other.
+Instead, we draw this conclusion on the basis that it was difficult for informative heterography to get off the ground in Experiment~1 due to the homophony present in the spoken form (as clarified by Experiment 3), while some degree of informative heterography did persist in Experiment 2 in the face of increasing homophony.
+A useful way that future work could explore this hypothesis would be to fit models of differentiation and conservation to historical data on spelling change to see which model offers a better fit to the data.}
+
 \subsection{Limitations}
@@
 This is likely to be related to a combination of two factors: a relatively small effect size combined with a relatively small sample size, with only ten chains---ten independent sampling units---in each condition.
-This is a general issue with iterated learning experiments that results from their resource-intensive nature, and our use of ten chains follows or exceeds norms in the field \parencite[e.g.,][]{Kempe:2015, Kirby:2008, Kirby:2015, Raviv:2018, Roberts:2018, Smith:2010, Tamariz:2015}.
-Nevertheless, our study suggests that future work in this area could benefit from increased power.}
+\secondrevision{Our decision to run ten chains per condition was primarily based on norms in the field \parencite[e.g.,][]{Kempe:2015, Kirby:2008, Kirby:2015, Raviv:2018, Roberts:2018, Smith:2010, Tamariz:2015}, since we had little idea of what effect size we could expect to find when designing the studies.
+Nevertheless, the small effect sizes we observed do suggest some risk of type one error and future work with this paradigm would benefit from increasing the number of chains in light of these relatively small effect sizes}.}
 
-\newmaterial{One important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems.
+\newmaterial{Lastly, one important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems.
 This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage, whether that baggage is for the encoding of sound or the encoding of meaning.
@@
 
-% \section{Acknowledgments}
+\section{Acknowledgments}
 
-% \noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034).
+\noindent This work was funded by a Leverhulme Trust Research Project Grant (grant number: RPG-2020-034) awarded to KR.
 
@@
 \appendix
+
+% \pagestyle{empty}
 
@@
 The quantitative results are plotted in Fig.~\ref{supp3} and are compared against the equivalent results from Experiment~1 (the Transmission~+ Communication condition).
-Overall, the effect of removing the homophony pressure was to prevent the systems from degenerating over time---both measures of informativeness show a clear difference in slopes, and informativeness levels remain fairly constant over time in Experiment~3, mainly in the form of holistic systems.
+\secondrevision{Under both measures of informativeness, Experiment~3 was quite stable over time (purple curves), with little change in how informative the systems are.
+This contrasts with Experiment~1 (orange curves), which saw a more dramatic collapse in informativeness over time.
+The difference between the two experiments clearly rejects zero both in terms of the rate at which the number of informative systems decreases over time ($\Delta(\beta) = 1.69$; 95\%~HDI: 0.4, 3.04) and in terms of the rate at which communicative cost increases over time ($\Delta(\beta) = -0.33$; 95\%~HDI: $-0.54$, $-0.12$).
+These results suggest that the presence of homophony (i.e., in Experiment~1) was a key driver behind the loss of informativeness in that experiment.}
 

### 2024-04-09 manuscript/manuscript.tex
Rename newmaterial to firstrevision

@@
 \usepackage{xcolor}
-\newcommand\newmaterial[1]{\textcolor{black}{#1}}
+\newcommand\firstrevision[1]{\textcolor{black}{#1}}
 \newcommand\secondrevision[1]{\textcolor{black}{#1}}
@@
 Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English affixes \parencite{Berg:2017} and it has been shown that readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.
-\newmaterial{Heterography might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair.
+\firstrevision{Heterography might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair.
 Furthermore, written language has richer vocabulary and more complex syntax than spoken language \parencite{Biber:1988, Korochkina:2024, Nation:2022}, placing different pressures on ambiguity resolution.}
 
-\newmaterial{Heterography is particularly notable in English, but it is also a feature of many other languages and writing systems.
+\firstrevision{Heterography is particularly notable in English, but it is also a feature of many other languages and writing systems.
 French is similar to English in having a large number of heterographic homophones: \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}), for example, are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison).
@@
 
-\newmaterial{Perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese orthography.
+\firstrevision{Perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese orthography.
 The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic writing system---particularly useful \parencite{Frost:2012}.
@@
 In English, the heterographic spellings <meat> and <meet> imply that /iː/ can be spelled <ea> or <ee>.
-Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}.
+Readers are therefore required to learn alternate spellings for a single sound, \firstrevision{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}.
 Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned.
@@
 
-\newmaterial{In this paper, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness.} A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
+\firstrevision{In this paper, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness.} A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology.
 An informative spelling system, on the other hand, would be one that precisely conveys meaning.
@@
 The first model, \textit{the differentiation model} (Fig.~\ref{fig01}A), explains heterography through the creation of new orthographic forms.
-For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\newmaterial{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
+For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (\firstrevision{perhaps influenced by French \textit{chèque}}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}.
 Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by appending <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}.
@@
 All data and code is available from \url{https://osf.io/7auw6/}.
-\newmaterial{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}.
+\firstrevision{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}.
 Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}
@@
 The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus).
-These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent \newmaterial{(e.g., the round torus shape is represented by ``round'' sounds/letters)}.
+These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent \firstrevision{(e.g., the round torus shape is represented by ``round'' sounds/letters)}.
 Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time.
@@
 The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations.
-Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).\footnote{\newmaterial{The 2/3 bottleneck parameter was chosen based on common practices in the field and some piloting.
+Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).\footnote{\firstrevision{The 2/3 bottleneck parameter was chosen based on common practices in the field and some piloting.
 In general, increasing this parameter will lead to a slower evolutionary process, so the value was chosen to allow for sufficient change to occur within nine generations.}} Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items.
@@
 
-All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, lasting around 8~min}.
+All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\firstrevision{, lasting around 8~min}.
 Participants were also told explicitly before starting the training session that the stems looked and sounded like the objects' shapes, allowing participants to focus on learning the suffixes during the training phase.
@@
 To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.
-\newmaterial{The instructions provided to participants are provided in Appendix~A in the supplementary material.}
+\firstrevision{The instructions provided to participants are provided in Appendix~A in the supplementary material.}
 
@@
 The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial.
-Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem correction messages were rarely encountered.
+Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\firstrevision{Although artificial, these stem correction messages were rarely encountered.
 81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it on one trial (out of nine), and 5\% encountered it on two or three trials.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix.
@@
 In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice).
-In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \newmaterial{on the correctness of the signal or object selection}.
+In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \firstrevision{on the correctness of the signal or object selection}.
 Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).
@@
 Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant.
-Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{The seen items were selected independently for each participant.
-We felt this was more ecological than giving both participants identical input.}} The communication game, \newmaterial{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
+Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\firstrevision{The seen items were selected independently for each participant.
+We felt this was more ecological than giving both participants identical input.}} The communication game, \firstrevision{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction.
 On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array).
@@
 The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target.
-\newmaterial{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.}
+\firstrevision{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.}
 
@@
 
-\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate).
+\firstrevision{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate).
 We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain).
@@
 
-\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are.
+\firstrevision{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are.
 In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts\secondrevision{; Table~B1 in the supplementary material}), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0).
@@
 In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.
-\newmaterial{The results are shown in Fig.~\ref{fig07}C.
+\firstrevision{The results are shown in Fig.~\ref{fig07}C.
 We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian).
@@
 
-\newmaterial{Aside from the informativeness of the orthographic systems, we also predicted in our preregistration that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
+\firstrevision{Aside from the informativeness of the orthographic systems, we also predicted in our preregistration that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons.
 In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form.
@@
 The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color.
-\newmaterial{In support of this explanation, we conducted an additional experiment during review, which showed that when homophony is removed, the systems tend to resist degeneration in line with prior work.
+\firstrevision{In support of this explanation, we conducted an additional experiment during review, which showed that when homophony is removed, the systems tend to resist degeneration in line with prior work.
 We discuss this experiment in more detail in the Discussion section and in Appendix~C of the supplementary material.}
@@
 As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose.
-\newmaterial{Similar outcomes have been noted before in the context of artificial language learning experiments \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
-\newmaterial{Interestingly, however, these redundant systems were relatively uncommon under communicative pressure, suggesting that communicating participants recognized the futility of using the suffix to mark shape.}
+\firstrevision{Similar outcomes have been noted before in the context of artificial language learning experiments \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.
+\firstrevision{Interestingly, however, these redundant systems were relatively uncommon under communicative pressure, suggesting that communicating participants recognized the futility of using the suffix to mark shape.}
 
@@
 
-\newmaterial{Our quantitative analysis of Experiment~2 is identical to that of Experiment~1 with a slight change to the statistical model.
+\firstrevision{Our quantitative analysis of Experiment~2 is identical to that of Experiment~1 with a slight change to the statistical model.
 Rather than predict the outcome variables as a function of generation, we now predict the outcome variables as a function of the epoch number (1, 2, or 3) and the generation number within the epoch (1, 2, or 3).\footnote{Model: dependent $\sim$ epoch + (1 + epoch | chain) + generation\_in\_epoch + (1 + generation\_in\_epoch | chain).} This yields two slope parameters: $\beta$, which represents the effect of epoch, and $\gamma$, which represents the additional effect of generational turnover.
@@
 Notably, however, the loss of expressive systems appears to be slower in Transmission~+ Communication, and redundant systems were also less popular under communicative pressure.
-\newmaterial{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time.
+\firstrevision{As in Experiment~1, we further collapsed the typological categories into two broader categories (informative vs.\@ uninformative) to analyze the trends over time.
 The results, shown in Fig.~\ref{fig11}B, show that the probability of a system being informative drops over time in both conditions (primarily as a function of epoch), but does so more slowly in the communicative condition.
@@
 
-\newmaterial{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C.
+\firstrevision{The results in terms of the preregistered measure of informativeness, communicative cost, are presented in Fig.~\ref{fig11}C.
 Here we find a nonzero effect of both epoch \secondrevision{($\beta$)} and generation \secondrevision{($\gamma$)} on cost, as well as a difference between conditions in terms of epoch: $\Delta(\beta) = -0.2$ (95\%~HDI: $-0.4$, $-0.003$).
@@
 
-\newmaterial{For completeness, Fig.~\ref{fig11}D also plots transmission error; however, we did not hypothesize any particular differences in learnability in Experiment~2, either over time or by condition.
+\firstrevision{For completeness, Fig.~\ref{fig11}D also plots transmission error; however, we did not hypothesize any particular differences in learnability in Experiment~2, either over time or by condition.
 The expressive orthographic systems used to initialize the chains start out very easy to learn, and learnability remains fairly consistent throughout the experiment in both conditions, albeit with some constant level of change over time as the systems gradually come into alignment with the spoken forms.}
@@
 When asked to type in a word for a pink pentagon called /buvɪkəʊ/, participants were inclined to type a sequence of graphemes that reflected the sound they heard, without encoding meaning.
-\newmaterial{This behavior might be connected to the concept of ``functional fixedness'' \parencite{German:2000}, which states that learners find it difficult to adduce a new function (e.g., writing meaning) when they are accustomed to another function (e.g., writing sound), which highlights a potentially important role for generational turnover in the development of writing systems, since new learners will be more receptive to new functions.} Second, even when participants \textit{did} appreciate the need to differentiate the written forms to be successful, they often appeared reluctant or unable to do so, perhaps because they viewed the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---was too difficult to overcome \newmaterial{without the ability to coordinate over an extended period of time}.
+\firstrevision{This behavior might be connected to the concept of ``functional fixedness'' \parencite{German:2000}, which states that learners find it difficult to adduce a new function (e.g., writing meaning) when they are accustomed to another function (e.g., writing sound), which highlights a potentially important role for generational turnover in the development of writing systems, since new learners will be more receptive to new functions.} Second, even when participants \textit{did} appreciate the need to differentiate the written forms to be successful, they often appeared reluctant or unable to do so, perhaps because they viewed the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---was too difficult to overcome \firstrevision{without the ability to coordinate over an extended period of time}.
 It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.
 
-\newmaterial{This conclusion is in partial agreement with work by \textcite{Treiman:2015}.
+\firstrevision{This conclusion is in partial agreement with work by \textcite{Treiman:2015}.
 In this study, participants were asked to provide spellings for novel English words (e.g., /hæf/ meaning \textit{alehouse}) that were homophonous with preexisting English words (in this case, \textit{half}).
@@
 
-\newmaterial{During the review process, a concern was raised that our participants might not have fully understood the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling.
+\firstrevision{During the review process, a concern was raised that our participants might not have fully understood the communicative nature of the task, thus explaining why we did not observe the emergence of systematic differentiation in spelling.
 Our position, as outlined above, was that the lack of differentiation was due to the very strong homophony pressure.
@@
 Under this mechanism, expressive forms do not emerge but are simply fossils representing an earlier form of the spoken language that \textit{was} expressive of a particular meaning distinction.
-\newmaterial{Over time, the orthography may experience a ratcheting effect, in which heterographic forms accumulate (due to successive sound changes) but rarely recede (due to the informativeness they provide).
-Over longer periods of time, this mechanism might even shift an orthography from a phonographic principle to a logographic one.} This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into written forms \parencite{Berg:2021}, \newmaterial{and which are sometimes argued to give English a semi-logographic character \parencite{Chomsky:1968, Coulmas:1991, DeFransis:1989, Zachrisson:1931}}.
+\firstrevision{Over time, the orthography may experience a ratcheting effect, in which heterographic forms accumulate (due to successive sound changes) but rarely recede (due to the informativeness they provide).
+Over longer periods of time, this mechanism might even shift an orthography from a phonographic principle to a logographic one.} This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into written forms \parencite{Berg:2021}, \firstrevision{and which are sometimes argued to give English a semi-logographic character \parencite{Chomsky:1968, Coulmas:1991, DeFransis:1989, Zachrisson:1931}}.
 To be clear, this is not to say that such heterography is nonadaptive or an accident of history; rather, such heterographs may have been preserved precisely \textit{because} of the informativeness they inadvertently provide in reading.
-\newmaterial{Thus, just as the spoken language avoids \textit{sound} mergers that increase ambiguity \parencite[e.g.,][]{Wedel:2013}, so the written language might likewise avoid \textit{spelling} mergers that increase ambiguity.} If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
+\firstrevision{Thus, just as the spoken language avoids \textit{sound} mergers that increase ambiguity \parencite[e.g.,][]{Wedel:2013}, so the written language might likewise avoid \textit{spelling} mergers that increase ambiguity.} If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.
 
@@
 In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one \parencite[see also][]{Smith:2002}.
-\newmaterial{This is made particularly salient by Fig.~\ref{fig12}, which compares the experiments in terms of communicative success (the proportion of trials in which the comprehending participant selected the correct target item in response to their partner).
+\firstrevision{This is made particularly salient by Fig.~\ref{fig12}, which compares the experiments in terms of communicative success (the proportion of trials in which the comprehending participant selected the correct target item in response to their partner).
 In Experiment~1, communicative success remains around chance level (a one in three probability of selecting the right color) because the orthographic systems tend to become uninformative, mirroring the spoken form of the language.
@@
 That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography.
-An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers, \newmaterial{especially given that writing systems are often fixed cultural fossils that do not readily adapt to external pressures.} It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many.
+An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers, \firstrevision{especially given that writing systems are often fixed cultural fossils that do not readily adapt to external pressures.} It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many.
 For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space.
@@
 
-\newmaterial{The results of both experiments were relatively weak statistically, with the differences in terms of informativeness between the Transmission-only and Transmission~+ Communication conditions only just (or not quite) meeting the 95\% criterion.
+\firstrevision{The results of both experiments were relatively weak statistically, with the differences in terms of informativeness between the Transmission-only and Transmission~+ Communication conditions only just (or not quite) meeting the 95\% criterion.
 This is likely to be related to a combination of two factors: a relatively small effect size combined with a relatively small sample size, with only ten chains---ten independent sampling units---in each condition.
@@
 
-\newmaterial{Lastly, one important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems.
+\firstrevision{Lastly, one important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be even less forthcoming in other populations used to more transparent writing systems.
 This brings us to a much deeper issue with iterated learning experiments in general: We cannot avoid the fact that our participants come into the lab with prior linguistic baggage, whether that baggage is for the encoding of sound or the encoding of meaning.
@@
 
-\newmaterial{It has long been known that heterography makes reading and learning to read difficult \parencite{Pexman:2001, Seymour:2003}.
+\firstrevision{It has long been known that heterography makes reading and learning to read difficult \parencite{Pexman:2001, Seymour:2003}.
 As a result, heterography has often been derided as a source of unnecessary complexity, and the orthographic reforms implemented in many languages have tended to focus on its elimination.

### 2024-05-22 manuscript/manuscript.tex
Final changes before publication

@@
 
-\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2024).
+\authornote{Carr, J.\@ W., \& Rastle, K.\@ (2024).
 Why do languages tolerate heterography?
 An experimental investigation into the emergence of informative orthography.
-\textit{PsyArXiv}.
-Version~3.
-\url{https://doi.org/10.31234/osf.io/qnz5k}}
+\textit{Cognition}, \textit{249}, Article~105809.
+\url{https://doi.org/10.1016/j.cognition.2024.105809}}
 
@@
 All data and code is available from \url{https://osf.io/7auw6/}.
-\firstrevision{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/YDV_MXV}.
+\firstrevision{To increase the transparency of our work, we created a preregistration at \url{https://aspredicted.org/p8aw9.pdf}.
 Note, however, that due to the more exploratory nature of this project, our preregistration did not specify strong confirmatory hypotheses or precise statistical models, focusing instead on the research question, experimental conditions, general predictions, primary measurement constructs, sample size, and exclusion criteria.}
@@
 Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost.
-For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Roman alphabet, but which trace their origins back to the original seed orthography.
+For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Latin alphabet, but which trace their origins back to the original seed orthography.
 Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without conserving how form patterns with meaning.
@@
 	\caption{Communicative success by generation in the communicative conditions of all three experiments.
-The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).}
+The dotted line shows chance level if the comprehending participant selects from the array of nine items at random (i.e., 1/9), and the dashed line shows chance level if the comprehending participant knows the correct shape but selects color at random (i.e., 1/3).
+Experiment 3 is a replication of Experiment 1 without any auditory component (see Appendix C in the supplementary material).}
 	\label{fig12}
